```{r include=FALSE}
source("R/logistic-models.R")
```

## Predicted probabilities and cutoff rules

This section evaluates predicted probabilities using the models introduced in the previous section. We begin with three baseline models: `logi_age`, `logi_int`, and `logi_multi`. The purpose is to see how a model first produces $\hat p_i$, the estimated probability that applicant $i$ defaults, before any 0/1 decision is made. Later in this section, we bring in `logi_full`, the complete model, and use it for cutoff rules and bank decisions.

We use the first applicant in the test set as a worked example and call that applicant John Doe.

```{r One applicant}
# Define one single observation in test_set.
John_Doe <- as.data.frame(test[1, ])
John_Doe
```

The observed `loan_st` for this test-set applicant is 0. The models cannot use that outcome when making the prediction because they were estimated only on the training set. A good credit risk model should assign this applicant a low predicted probability of default.

The observed values of `loan_st` are either 0 or 1. The logistic models return probability estimates between 0 and 1, as illustrated in @fig-scolcam. For John Doe, the prediction should be closer to 0 than to 1. The cutoff question is how low the probability must be before the bank treats the applicant as an acceptable risk.

This code estimates the default probability for John Doe using the three baseline logistic regression models. Each prediction represents the probability that John Doe defaults on a loan, based on the respective model.

```{r Predictions of one single applicant in the test set}
# List of models
models <- list("logi_age" = logi_age, "logi_int" = logi_int, "logi_multi" = logi_multi)

# Estimate the default probability for John Doe using all models
pred_John <- sapply(models, predict, newdata = John_Doe, type = "response")

# Convert to a table with appropriate column name
pred_John <- as.data.frame(pred_John)
colnames(pred_John) <- "Predicted default probability for John Doe."

# Display the predictions
pred_John
```

In code, `predict(..., type = "response")` returns $\hat p_i$ rather than a final 0/1 label. The values above are close to 0, so the models assign John Doe a low default probability. This single example is only a starting point. We still need a cutoff rule to decide when a predicted probability is low enough to accept an applicant.

The full evaluation must use all `r format(nrow(test), big.mark = ",")` observations in the test set. The code change is small: replace `newdata = John_Doe` with `newdata = test`. Mathematically, this computes one $\hat p_i$ for every applicant $i$ in the test set.

```{r Predictions of the entire test set}
# Estimate default probabilities with the three baseline models.
pred_logi_age <- predict(logi_age, newdata = test, type = "response")
pred_logi_int <- predict(logi_int, newdata = test, type = "response")
pred_logi_multi <- predict(logi_multi, newdata = test,
                           type = "response")

pred_range <- rbind("logi_age" = range(pred_logi_age),
                     "logi_int" = range(pred_logi_int),
                     "logi_multi" = range(pred_logi_multi))
aic <- rbind(logi_age$aic, logi_int$aic, logi_multi$aic)
pred_range <- cbind(pred_range, aic)
colnames(pred_range) <- c("min(predicted PD)", "max(predicted PD)", "AIC")
pred_range
```

At this point, the models have produced predicted probabilities for all `r format(length(pred_logi_age), big.mark = ",")` applicants in the test set. The first column shows the lowest predicted default probability for each model, and the logistic models produce values in the range of 0 to 1. In this case, the predicted ranges are quite narrow.

Narrow ranges, meaning a small difference between the highest and lowest predicted probabilities, can be a warning sign because the model may struggle to discriminate between defaults (predictions closer to 1) and non-defaults (predictions closer to 0), as suggested in @fig-scolcam. The AIC gives a separate in-sample comparison. The higher the AIC, the worse the fit-parsimony trade-off among comparable models. The prediction range only tells us how spread out the predicted probabilities are, so it should be read as an early diagnostic rather than a performance metric. Confusion matrices, bad rates, and ROC/AUC are still needed to evaluate predictive performance on the test set.

The next figure shows all `logi_age` predicted probabilities across the test set.

```{r}
#| label: fig-amp
#| fig-cap: "Predicted PDs from the age-only model across the test set."
# Convert the predictions to a data frame
pred_data <- data.frame(
  Observation = seq_along(pred_logi_age),  # Sequence for observation index
  Predicted_Probabilities = pred_logi_age
)

# Create the plot
ggplot(pred_data, aes(x = Observation, y = Predicted_Probabilities)) +
  geom_point(alpha = 0.6, color = "blue") +  # Points for predictions
  labs(x = "Observation (test set)", 
       y = "Predicted default probability") +
  theme_minimal()
```
The same predictions can also be viewed as a distribution.

```{r}
#| label: fig-amph
#| fig-cap: "Narrow distribution of age-only predicted PDs."
ggplot(data.frame(pred_logi_age), aes(x = pred_logi_age)) + 
  geom_density(fill = "red") +
  labs(y = "Density", x = "Predicted default probability") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

The `logi_age` predictions are concentrated in a very narrow probability range. A useful model need not cover the full 0-to-1 interval, but it does need enough separation to rank applicants by risk. Here the age-only model assigns very similar predicted default probabilities to most applicants, so it has little ability to distinguish likely defaults from likely non-defaults.

The comparison now moves to `logi_int` and `logi_multi`. First, we collect the predictions in a single data frame.

```{r}
pred_logi <- data.frame(cbind(pred_logi_age, pred_logi_int,
                                  pred_logi_multi))
pred_logi <- gather(pred_logi, key = "model",  value = "pred")
```

The density plot shows how widely each model spreads predicted default probabilities.

```{r}
#| label: fig-irammph
#| fig-cap: "Prediction distributions for the interest-rate and multi-factor models."
ggplot(pred_logi[pred_logi$model != "pred_logi_age",], 
       aes(x = pred, fill = model)) + 
  geom_density(alpha = 0.4) +
  labs(y = "Density", x = "Predicted default probability") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Adding the age model makes the contrast clearer.

```{r}
#| label: fig-aairaammpb
#| fig-cap: "Prediction ranges across age, interest-rate, and multi-factor models."
ggplot(pred_logi, aes(x = pred, y = model, fill = model)) + 
  geom_boxplot() +
  labs(y = "Model", x = "Predicted default probability") +
  theme(legend.position = "none", legend.title = element_blank())
```

The previous section estimated `logi_full`, the most complete model in this chapter. Applying that model to the test set creates `pred_logi_full`, an object with one predicted default probability for each applicant in `test`.

```{r logi_full prediction}
# Loan status predictions for all test set elements.
pred_logi_full <- predict(logi_full, newdata = test, type = "response")
# Look at the predictions range.
range(pred_logi_full)
```

```{r Full model prediction range setup, include=FALSE}
full_score_test <- as.numeric(predict(logi_full, newdata = test, type = "link"))
full_score_min_test <- min(full_score_test, na.rm = TRUE)
full_score_max_test <- max(full_score_test, na.rm = TRUE)
full_pd_min_test <- plogis(full_score_min_test)
full_pd_max_test <- plogis(full_score_max_test)
full_pd_range_test <- full_pd_max_test - full_pd_min_test
full_score_max_exp <- exp(full_score_max_test)

full_score_axis_min <- floor(min(full_score_min_test, -8))
full_score_axis_max <- ceiling(max(full_score_max_test, 8))
full_score_curve <- seq(full_score_axis_min, full_score_axis_max, length.out = 1000)
full_score_curve_data <- data.frame(
  score = full_score_curve,
  p = plogis(full_score_curve)
)

full_score_width <- full_score_axis_max - full_score_axis_min
full_score_indicator_data <- data.frame(
  score = c(full_score_min_test, full_score_max_test),
  p = c(full_pd_min_test, full_pd_max_test),
  label_x = c(
    full_score_min_test + 0.14 * full_score_width,
    full_score_max_test + 0.11 * full_score_width
  ),
  label_y = c(
    min(full_pd_min_test + 0.12, 0.95),
    min(full_pd_max_test + 0.08, 0.95)
  ),
  label = c(
    paste0("min score = ", fmt_num(full_score_min_test, 2), "\np = ", fmt_pct(full_pd_min_test, 2)),
    paste0("max score = ", fmt_num(full_score_max_test, 2), "\np = ", fmt_pct(full_pd_max_test, 2))
  )
)
```

Figure @fig-scolcam used age on the horizontal axis because the age-only model has only one predictor. The full model has many predictors, so age alone is no longer the right horizontal axis for the sigmoid. The natural horizontal axis is the full-model score, also called the logit score:

$$
\eta_i = \beta_0 + \beta_1 x_{1i} + \cdots + \beta_k x_{ki}.
$$

This score combines age, interest rate, grade, loan amount, income, employment length, home ownership, sex, and region into one number. The model then converts that score into a predicted probability with the logistic function. In the test set, the full-model score ranges from `r fmt_num(full_score_min_test, 2)` to `r fmt_num(full_score_max_test, 2)`, which maps to predicted default probabilities from `r fmt_pct(full_pd_min_test, 2)` to `r fmt_pct(full_pd_max_test, 2)`.

The direction of the curve is different from @fig-scolcam because the horizontal axis is different. In @fig-scolcam, the horizontal axis is age and the estimated age coefficient is negative, `r fmt_num(beta_age_1, 5)`. As age increases, the age-only score decreases, so the predicted probability decreases. In @fig-full-model-score-sigmoid, the horizontal axis is the score $\eta_i$ itself. The logistic transformation is increasing in $\eta_i$, so moving to the right means a higher model score and therefore a higher predicted probability.

```{r}
#| label: fig-full-model-score-sigmoid
#| fig-cap: "Sigmoid curve or logistic curve: full-model score."
ggplot(data = full_score_curve_data, aes(x = score, y = p)) +
  geom_line(color = "blue", linewidth = 2) +
  geom_segment(
    data = full_score_indicator_data,
    aes(x = score, xend = score, y = 0, yend = p),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    color = "#C0392B"
  ) +
  geom_segment(
    data = full_score_indicator_data,
    aes(x = full_score_axis_min, xend = score, y = p, yend = p),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    color = "#C0392B"
  ) +
  geom_segment(
    data = full_score_indicator_data,
    aes(x = score, xend = label_x, y = p, yend = label_y),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.7,
    color = "#C0392B"
  ) +
  geom_point(
    data = full_score_indicator_data,
    aes(x = score, y = p),
    inherit.aes = FALSE,
    color = "#C0392B",
    size = 2.8
  ) +
  geom_label(
    data = full_score_indicator_data,
    aes(x = label_x, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.2,
    fill = "white",
    color = "#7B241C",
    label.size = 0.2
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +
  labs(x = "Full-model score (logit)", y = "Probability (p)") +
  theme_minimal()
```

This figure should be read together with @fig-scolcam. In the age-only model, the observed age range covers only a small and low-probability part of the sigmoid curve. In the full model, the combined score spans a much wider part of the same logistic transformation. That is why `pred_logi_full` can range from `r fmt_pct(full_pd_min_test, 2)` to `r fmt_pct(full_pd_max_test, 2)`. The wider range comes from the complete score, not from age by itself.

The right dotted point gives the highest predicted PD in the test set. Its score is $\eta = `r fmt_num(full_score_max_test, 2)`$. Substituting that value into the logistic function gives:

$$
p =
\frac{\exp(`r fmt_num(full_score_max_test, 2)`)}
{1+\exp(`r fmt_num(full_score_max_test, 2)`)}
=
\frac{`r fmt_num(full_score_max_exp, 4)`}
{1+`r fmt_num(full_score_max_exp, 4)`}
= `r fmt_num(full_pd_max_test, 4)`
= `r fmt_pct(full_pd_max_test, 2)`.
$$

This calculation is the numerical bridge between the x-axis and the y-axis. The x-axis value is the full-model score. The y-axis value is the probability obtained after applying the logistic transformation.

The `pred_logi_full` prediction range is wider across the test set. Some applicants receive substantially higher predicted default probabilities than others, which can help the model separate likely defaults from likely non-defaults. A wider range is a useful sign, but classification and ranking metrics are needed before concluding that the model is better. The table below compares the prediction ranges across all models.

```{r}
pred_range <- rbind("logi_age" = range(pred_logi_age), 
                     "logi_int" = range(pred_logi_int),
                     "logi_multi" = range(pred_logi_multi),
                     "logi_full" = range(pred_logi_full))
aic <- rbind(logi_age$aic, logi_int$aic, logi_multi$aic, logi_full$aic)
pred_range <- cbind(pred_range, aic)
colnames(pred_range) <- c("min(predicted PD)", "max(predicted PD)", "AIC")
pred_range
```

```{r}
pred_logi <- data.frame(cbind(pred_logi_age, pred_logi_int,
                                  pred_logi_multi, pred_logi_full))
pred_logi <- gather(pred_logi, key = "model",  value = "pred")
```

The next density plot compares the new `pred_logi_full` predictions with `pred_logi_multi`.

```{r}
#| label: fig-mafmph
#| fig-cap: "Prediction distributions for the multi-factor and full models."
ggplot(pred_logi[pred_logi$model != "pred_logi_age" &
                  pred_logi$model != "pred_logi_int" ,], 
       aes(x = pred, fill = model)) + 
  geom_density(alpha = 0.4) +
  labs(y = "Density", x = "Predicted default probability") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

The boxplot adds all four models to the same comparison.

```{r}
#| label: fig-airmafmpb
#| fig-cap: "Prediction ranges across all logistic model candidates."
ggplot(pred_logi, aes(x = pred, y = model, fill = model)) + 
  geom_boxplot() +
  labs(y = "Model", x = "Predicted default probability") +
  theme(legend.position = "none", legend.title = element_blank())
```

The `logi_full` predictions look more useful than the other models because they produce greater separation across applicants. The next question is whether this separation leads to better credit decisions.

To turn predicted probabilities into a lending decision, the bank needs a cutoff. The `logi_full` model produces predicted probabilities of default from `r fmt_num(min(pred_logi_full), 3)` to `r fmt_num(max(pred_logi_full), 3)`. Denote the predicted probability for applicant $i$ by $\hat{p}_i = \widehat{P}(Y_i = 1 \mid x_i)$, where $Y_i$ is the observed value stored in `loan_st` and $x_i$ represents the applicant's predictors. A cutoff $c$ converts those probabilities into an accept/reject rule:

$$
\hat{y}_i(c) =
\begin{cases}
1, & \text{if } \hat{p}_i > c \quad \text{(predicted default, reject)},\\
0, & \text{if } \hat{p}_i \leq c \quad \text{(predicted no default, accept)}.
\end{cases}
$$

In code, the vector $\hat p_i$ is `pred_logi_full`, the cutoff $c$ is `0.15`, and the binary decision $\hat y_i(c)$ is `pred_cutoff_15`. The line `ifelse(pred_logi_full > 0.15, 1, 0)` is the direct code translation of this equation.

For a first pass, set a cutoff of 0.15. This teaching rule makes the conversion from probabilities to binary decisions explicit. Later in the chapter, we compare cutoff rules using confusion matrices, acceptance rates, bad rates, and financial payoff.

This 0.15 rule classifies any predicted probability below or equal to 0.15 as 0 (predicted no default) and any predicted probability above 0.15 as 1 (predicted default). In lending terms, applicants above the cutoff are rejected and applicants below or equal to the cutoff are accepted.

The cutoff is both a policy choice and a statistical choice. A higher cutoff accepts more applicants because more predicted probabilities fall below or equal to the cutoff. This can increase loan volume, but it can also accept more risky borrowers.

It is useful to distinguish two related but different cutoff ideas. A fixed probability cutoff, such as 0.15, represents an absolute risk tolerance: the bank accepts applicants only if their estimated probability of default is at most 15%. An acceptance-rate rule works differently. If the bank wants to accept 80% of applications, the cutoff is not chosen first; it is implied by the 80th percentile of the predicted probabilities. The first rule starts from a risk limit, while the second starts from a portfolio or business target.

Under the 0.15 rule, any predicted probability below or equal to 0.15 is classified as 0 (predicted no default), while any predicted probability above 0.15 is classified as 1 (predicted default). The cutoff is visible in the next scatter plot.

```{r}
#| label: fig-fmpaco
#| fig-cap: "Full-model predicted PDs with a 15% cutoff."
# Convert the predictions to a data frame
cutoff_15 <- 0.15
cutoff_15_below <- mean(pred_logi_full <= cutoff_15)
cutoff_15_above <- mean(pred_logi_full > cutoff_15)

pred_data <- data.frame(
  Observation = seq_along(pred_logi_full),  # Sequence for observation index
  Predicted_Probabilities = pred_logi_full
)

cutoff_15_scatter_labels <- data.frame(
  x = length(pred_logi_full) * 0.72,
  y = c(cutoff_15 * 0.48,
        cutoff_15 + 0.68 * (max(pred_logi_full) - cutoff_15)),
  label = c(
    paste0("Accepted\n", fmt_pct(cutoff_15_below, 1)),
    paste0("Rejected\n", fmt_pct(cutoff_15_above, 1))
  )
)

# Create the plot
ggplot(pred_data, aes(x = Observation, y = Predicted_Probabilities)) +
  geom_point(alpha = 0.6, color = "blue") +  # Points for predictions
  geom_hline(yintercept = cutoff_15, color = "red", size = 1) +  # Horizontal line
  geom_label(
    data = cutoff_15_scatter_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    alpha = 0.92,
    label.size = 0,
    size = 3.5,
    lineheight = 0.95
  ) +
  labs(x = "Observation (test set)", 
       y = "Predicted default probability") +
  theme_minimal()

```
The distribution view gives the same rule from another angle.

```{r}
#| label: fig-fmphaco
#| fig-cap: "Full-model prediction distribution with a 15% cutoff."
full_pred_density <- density(pred_logi_full)
cutoff_15_density_labels <- data.frame(
  x = c(cutoff_15 * 0.55,
        cutoff_15 + 0.48 * (max(pred_logi_full) - cutoff_15)),
  y = max(full_pred_density$y) * 0.82,
  label = c(
    paste0("Accepted\n", fmt_pct(cutoff_15_below, 1)),
    paste0("Rejected\n", fmt_pct(cutoff_15_above, 1))
  )
)

ggplot(pred_logi[pred_logi$model == "pred_logi_full",], 
       aes(x = pred, fill = model)) + 
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = cutoff_15, linetype = "longdash", color = "red", linewidth = 0.8) +
  geom_label(
    data = cutoff_15_density_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    alpha = 0.92,
    label.size = 0,
    size = 3.4,
    lineheight = 0.95
  ) +
  labs(y = "Density", x = "Predicted default probability") +
  theme(legend.position = "none", legend.title = element_blank())
```

Predicted probabilities to the left of the dashed line are classified as no default. Predicted probabilities to the right are classified as default. The labels show the share of test-set applications on each side of the cutoff. The next code block creates that binary variable.

The next rows show the transformation from probability to binary prediction.

```{r Specifying a cutoff of 15}
# Make a binary predictions-vector using a cutoff of 15%
pred_cutoff_15 <- ifelse(pred_logi_full > 0.15, 1, 0)
head(cbind(pred_logi_full, 
           "pred_logi_full_rounded" = round(pred_logi_full, 4), 
           pred_cutoff_15))
```

These are only the first 6 rows in the test set. The rule is doing exactly what the equation says: every estimated default probability below or equal to 0.15 is coded as 0 (predicted no-default), and every estimated default probability above 0.15 is coded as 1 (predicted default). The table above shows the conversion from a probability, `pred_logi_full`, to a binary prediction, `pred_cutoff_15`.

The row numbers in the table above are not necessarily consecutive because the `test` rows were selected randomly out of `dat`. They correspond to the original row positions in `dat`.

To see whether `logi_full` is making useful decisions, we compare predictions with realized outcomes. The first column is the logistic model prediction, the second column is the binary prediction at the 0.15 cutoff, and the third column is the observed outcome. The sample below shows individual correct and incorrect classifications before we move to a full confusion matrix.

```{r}
# Take rows 101 to 110.
(cbind(pred_logi_full, pred_cutoff_15, 
           as.data.frame.numeric(test$loan_st)))[101:110,]
```

## Confusion matrices and bad rates

The previous section converted predicted probabilities into binary decisions through a cutoff. Those decisions can now be evaluated by comparing the predicted loan status with the actual loan status in the test set. The standard summary table is a confusion matrix.

```{r Confusion matrix at cutoff of 15}
cm_cutoff_15 <- table(Actual = test$loan_st, Predicted = pred_cutoff_15)
cm_cutoff_15
```
The `logi_full` model correctly predicts `r cm_cutoff_15["0", "0"]` no-defaults and `r cm_cutoff_15["1", "1"]` defaults. It also misclassifies `r cm_cutoff_15["0", "1"]` no-defaults as defaults and `r cm_cutoff_15["1", "0"]` defaults as no-defaults. In credit terms, the `r cm_cutoff_15["0", "1"]` cases are good applicants that would be incorrectly rejected, while the `r cm_cutoff_15["1", "0"]` cases are bad applicants that would be incorrectly accepted. The sum of these four values equals `r sum(cm_cutoff_15)`, the total number of observations in the test set.

`CrossTable()` reports the same comparison with relative frequencies.

```{r Confusion matrix at cutoff of 15 with CrossTable}
CrossTable(test$loan_st, pred_cutoff_15, 
           prop.r = FALSE, prop.c = FALSE, prop.t = TRUE, 
           prop.chisq = FALSE,
          dnn = c("Actual", "Predicted"))
```

Read the table by actual outcome. The row for actual no-defaults contains two groups: true negatives, which are correctly accepted as no-defaults, and false positives, which are incorrectly rejected as predicted defaults. In this cutoff rule, true negatives represent `r fmt_pct(cm_cutoff_15["0", "0"] / sum(cm_cutoff_15), 1)` of all test-set cases, while false positives represent `r fmt_pct(cm_cutoff_15["0", "1"] / sum(cm_cutoff_15), 1)`.

The row for actual defaults is the credit-risk row. True positives are defaults that the model correctly identifies and rejects; false negatives are defaults that the model misses and accepts. In this cutoff rule, true positives represent `r fmt_pct(cm_cutoff_15["1", "1"] / sum(cm_cutoff_15), 1)` of all test-set cases, while false negatives represent `r fmt_pct(cm_cutoff_15["1", "0"] / sum(cm_cutoff_15), 1)`. The false negatives are especially important for lending because they are accepted loans that later default.

The table also shows class imbalance. Non-defaults represent `r fmt_pct(sum(cm_cutoff_15["0", ]) / sum(cm_cutoff_15), 1)` of the test set, while defaults represent only `r fmt_pct(sum(cm_cutoff_15["1", ]) / sum(cm_cutoff_15), 1)`. This is why accuracy alone can be misleading.

Overall accuracy is defined as:

$$
\text{Accuracy} =
\frac{\text{True negatives} + \text{True positives}}
{\text{True negatives} + \text{False positives} + \text{False negatives} + \text{True positives}}.
$$

In code, `mean(pred_cutoff_15 == actual_default)` computes the same quantity. It checks which binary predictions are equal to the observed outcomes and then averages the TRUE/FALSE values. This also shows why overall accuracy can be misleading in credit risk. Since most applicants in the test set do not default, a very naive rule that predicts "no default" for everyone can obtain a high accuracy simply by following the majority class. That rule would accept every loan that eventually defaults. A credit risk model should therefore be evaluated by the share of correct classifications and by the types of errors it makes.

```{r Accuracy can be misleading}
actual_default <- as.numeric(as.character(test$loan_st))
pred_all_no_default <- rep(0, length(actual_default))

accuracy_summary <- data.frame(
  rule = c("Full model with cutoff 0.15", "Always predict no default"),
  accuracy = c(
    mean(pred_cutoff_15 == actual_default),
    mean(pred_all_no_default == actual_default)
  ),
  defaults_detected = c(
    sum(actual_default == 1 & pred_cutoff_15 == 1),
    sum(actual_default == 1 & pred_all_no_default == 1)
  ),
  accepted_defaults = c(
    sum(actual_default == 1 & pred_cutoff_15 == 0),
    sum(actual_default == 1 & pred_all_no_default == 0)
  )
)

accuracy_summary
```

The second rule has higher accuracy, but it detects no defaults at all. This is unacceptable for credit screening because the bank would accept all defaulting applicants. The full model with a cutoff of 0.15 has lower overall accuracy, but it detects `r accuracy_summary$defaults_detected[accuracy_summary$rule == "Full model with cutoff 0.15"]` defaults and reduces the number of defaulting loans accepted by the bank. In imbalanced classification problems, accuracy alone can reward the wrong behavior.

The 0.15 cutoff was a first teaching rule. A second approach starts from the bank's desired acceptance rate. Suppose the bank wants to reject the 20% riskiest applicants and accept the remaining 80%. In that case, the cutoff should be the 80th percentile of `pred_logi_full`. Applicants above that cutoff are predicted defaults, and applicants below or equal to that cutoff are predicted no-defaults.

```{r cutoff-80-percent-acceptance}
# Cutoff definition.
cutoff <- quantile(pred_logi_full, 0.8)
cutoff
```

The 80% acceptance rule implies a cutoff of `r fmt_num(cutoff, 7)`. The cutoff line is shown below.

```{r}
#| label: fig-fmphncoo0
#| fig-cap: "Full model prediction histogram with acceptance-rate cutoff."
acceptance_cutoff_below <- mean(pred_logi_full <= cutoff)
acceptance_cutoff_above <- mean(pred_logi_full > cutoff)
acceptance_density_labels <- data.frame(
  x = c(
    min(pred_logi_full) + 0.58 * (cutoff - min(pred_logi_full)),
    cutoff + 0.55 * (max(pred_logi_full) - cutoff)
  ),
  y = max(full_pred_density$y) * 0.82,
  label = c(
    paste0("Accepted\n", fmt_pct(acceptance_cutoff_below, 1)),
    paste0("Rejected\n", fmt_pct(acceptance_cutoff_above, 1))
  )
)

ggplot(pred_logi[pred_logi$model == "pred_logi_full",], 
       aes(x = pred, fill = model)) + 
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = cutoff, linetype = "longdash", color = "red", linewidth = 0.8) +
  geom_label(
    data = acceptance_density_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    fill = "white",
    alpha = 0.92,
    label.size = 0,
    size = 3.4,
    lineheight = 0.95
  ) +
  labs(y = "Density", x = "Predicted default probability") +
  theme(legend.position = "none", legend.title = element_blank())
```

The cutoff is now `r fmt_num(cutoff, 7)`. This splits the predicted default probabilities into two parts: values above the cutoff are predicted defaults, and values below or equal to the cutoff are predicted no-defaults. The labels show the actual split in the test set. The next table reports cutoff values at several quantiles.

```{r Cutoff by quantiles}
cutoff_all <- quantile(pred_logi_full, seq(0.1, 1, 0.1))
data.frame("cut_off" = round(cutoff_all,5))
```

The same confusion-matrix logic now applies to the `logi_full` model with the new cutoff of `r fmt_num(cutoff, 7)`.

```{r Confusion matrix pred_full_20}
# Calculate the predictions with the same model and new cutoff.
pred_full_20 <- ifelse(pred_logi_full > cutoff, 1, 0)
cm_full_20 <- table(
  Actual = factor(test$loan_st, levels = c("0", "1")),
  Predicted = factor(pred_full_20, levels = c(0, 1))
)
# Show results in a confusion matrix.
CrossTable(test$loan_st, pred_full_20, prop.r = FALSE,
           prop.c = FALSE, prop.t = TRUE, prop.chisq = FALSE)
```

With a cutoff of `r fmt_num(cutoff, 7)`, we accept `r cm_full_20["0", "0"]` + `r cm_full_20["1", "0"]` = `r sum(cm_full_20[, "0"])` applications as those are the ones that the model predicts a no default. Previously, in the case of a cutoff of 0.15, we accepted `r cm_cutoff_15["0", "0"]` + `r cm_cutoff_15["1", "0"]` = `r sum(cm_cutoff_15[, "0"])`.

The next table compares both confusion matrices.

```{r Cutoff of 15 and 19 comparison}
cat <- c("Correct no-default (true negatives).", 
         "False default (false positives).", 
         "False no-default (false negatives).", 
         "Correct default (true positives).")
cut_15 <- fmt_pct(
  c(cm_cutoff_15["0", "0"], cm_cutoff_15["0", "1"],
    cm_cutoff_15["1", "0"], cm_cutoff_15["1", "1"]) /
    sum(cm_cutoff_15),
  1
)
cut_accept_80 <- fmt_pct(
  c(cm_full_20["0", "0"], cm_full_20["0", "1"],
    cm_full_20["1", "0"], cm_full_20["1", "1"]) /
    sum(cm_full_20),
  1
)
data.frame(
  classification = cat,
  cutoff_0_15 = cut_15,
  acceptance_80 = cut_accept_80
)
```

Type I and Type II errors are key concepts in hypothesis testing and statistical decision-making, particularly relevant in loan default prediction. A Type I error (false positive) occurs when good customers, who would not default, are incorrectly rejected. On the other hand, a Type II error (false negative) happens when bad customers, who will default, are mistakenly accepted.

The new cutoff of `r fmt_num(cutoff, 7)` improves the identification of no-defaults but worsens the identification of defaults. It creates fewer false positives and more false negatives, which is the central classification trade-off in this lending rule.

The next excerpt inspects individual cases under the `r fmt_num(cutoff, 7)` cutoff by comparing actual loan status with the binary prediction created from the estimated default probability.

```{r pred_full_20 for a sample of 10 applicants}
# Comparative table in detail.
real_pred_20 <- cbind.data.frame(test$loan_st, pred_full_20,
                     "Did the model succeed?" = test$loan_st == pred_full_20)
# Show some values.
sample_real_pred_20 <- real_pred_20[131:140,]
sample_correct_rejections <- rownames(sample_real_pred_20)[
  sample_real_pred_20$pred_full_20 == 1 &
    sample_real_pred_20$`test$loan_st` == 1
]
sample_lost_good_customers <- rownames(sample_real_pred_20)[
  sample_real_pred_20$pred_full_20 == 1 &
    sample_real_pred_20$`test$loan_st` == 0
]
sample_model_failures <- sum(!sample_real_pred_20[["Did the model succeed?"]])
sample_real_pred_20
```

In this small excerpt, the model makes `r sample_model_failures` `r ifelse(sample_model_failures == 1, "mistake", "mistakes")` out of `r nrow(sample_real_pred_20)` applications. The acceptance rule is direct: if `pred_full_20 = 0`, the model predicts no default and the application is accepted. According to the excerpt, the model correctly rejects applications `r paste(sample_correct_rejections, collapse = ", ")` because those were indeed defaults. It incorrectly rejects application `r paste(sample_lost_good_customers, collapse = ", ")`, which did not default. The table makes the business meaning of the classification rule visible before we summarize all applications with aggregate metrics.

The next table counts accepted and rejected applications under this rule.

```{r Accepted and rejected applications given the model rule}
# Accepted.
accept_20 <- sum(pred_full_20 == 0)
# Rejected.
reject_20 <- sum(pred_full_20 == 1)
data.frame("total" = length(pred_full_20), accept_20, reject_20)
```

By taking the top 20% of the `pred_logi_full` estimates as defaults and the bottom 80% as non-defaults, we construct a rule that accepts `r accept_20` loan applications (`r fmt_pct(accept_20 / length(pred_full_20), 0)` of the total) and rejects `r reject_20` applications (`r fmt_pct(reject_20 / length(pred_full_20), 0)` of the total). The cutoff criterion determines the number of accepted applications, while the `logi_full` model specifies which applications to accept or reject.

We evaluate the acceptance/rejection rule by comparing it against the actual outcomes recorded in `loan_st`. The next table illustrates the decision process.

```{r Accept and reject decision}
# First 10 accept decisions.
head(data.frame(real_pred_20[,1:2], 
                decision = ifelse(real_pred_20$pred_full_20 == 0, 
                                  "accept", "reject")), 12)
```

An evaluation column labels each decision.

```{r Accept and reject decision plus an evaluation}
# First 10 accept decisions.
head(data.frame(real_pred_20[,1:2], 
                decision = ifelse(real_pred_20$pred_full_20 == 0, 
                                  "accept", "reject"),
                evaluation =
                  ifelse(real_pred_20$pred_full_20 == 0 &
                           real_pred_20$`test$loan_st` == 0,
                         "good decision",
                  ifelse(real_pred_20$pred_full_20 == 0 &
                           real_pred_20$`test$loan_st` == 1,
                         "bad decision",
                  ifelse(real_pred_20$pred_full_20 == 1 &
                           real_pred_20$`test$loan_st` == 1,
                         "correct rejection",
                         "lost good customer")))), 12)
```

The table shows the outcomes of the first 12 loan applications. A "good decision" means the bank accepts a loan that does not default. A "bad decision" means the bank accepts a loan that defaults. A correct rejection avoids a defaulting borrower. A lost good customer is an applicant who would not have defaulted but is rejected by the model rule.

Because this is a historical test set, we can observe the true `loan_st` for every applicant in the sample. That lets us evaluate accepted and rejected cases after the fact. In a live lending process, a rejected applicant usually does not become a loan in the bank's own portfolio, so the bank does not observe that applicant's repayment behavior as one of its borrowers. For operating a credit policy, the most immediate question is focused on the loans the bank actually accepts: among accepted loans, how many default? This is the bad rate.

The `real_pred_20` object contains both the binary model predictions and the actual loan status. In the code below, we first filter for accepted loans, which are cases where the prediction equals 0. If this condition is met, we keep the value from the first column, which is the observed loan status.

As a result, `accepted_loans` becomes a vector of zeros and ones: a value of 0 indicates a loan that did not default, while a value of 1 indicates a loan that defaulted. The length of the `accepted_loans` vector corresponds to the number of loans allocated based on the cutoff value of `r fmt_num(cutoff, 7)` and the predictions from the `logi_full` model.

For a cutoff $c$, the rule above already tells us which loans are accepted. A loan is accepted when $\hat y_i(c)=0$. The bad rate is simply:

$$
\text{bad rate}(c) =
\frac{\text{accepted loans that default}}
{\text{accepted loans}},
$$

where default means $y_i=1$. In the code below, `pred_full_20 == 0` finds the accepted applications, `accepted_loans <- real_pred_20[pred_full_20 == 0, 1]` keeps their observed outcomes, `sum(accepted_loans == 1)` counts the accepted loans that default, and `length(accepted_loans)` counts all accepted loans.


```{r A more efficient way to conduct the model evaluation}
# We accept loans that the model predicts a no-default (0).
# In "accepted_loans" we know whether the accepted loans are in fact
# default or no-default.
accepted_loans <- real_pred_20[pred_full_20 == 0, 1]
length(accepted_loans)
# The code above says: if we accept the application, tell me what happened.
```

```{r}
head(accepted_loans, 10)
```

The small excerpt above shows individual accepted loans and their outcomes. The full bad-rate calculation uses all `r format(length(accepted_loans), big.mark = ",")` accepted loans:

```{r Bad rate calculation}
# bad_rate is the proportion of accepted loans that are in fact default.
bad_rate_80 <- sum(accepted_loans == 1) / length(accepted_loans)
bad_rate_80
```

Following the model-based rule, we accepted `r length(accepted_loans)` loan applications, which represent `r fmt_pct(length(accepted_loans) / length(pred_full_20), 0)` of all applications. Among those accepted applications, `r fmt_pct(bad_rate_80, 2)` were in fact defaults. In counts, we accepted `r sum(accepted_loans == 1)` loans that defaulted, so `r sum(accepted_loans == 1)` / `r length(accepted_loans)` = `r fmt_num(bad_rate_80, 7)`.

The bank can lower the bad rate by becoming more selective. With the same model, that means lowering the acceptance rate and approving fewer applications. This reduces credit risk, but it can also reduce business volume because the bank rejects more potential customers.

Reducing the acceptance rate from 80% to 65% shows how the bad rate changes when the bank becomes more selective.

```{r Bad rate when the acceptance rate is 65}
# New cutoff value.
cutoff <- quantile(pred_logi_full, 0.65)
# Split the pred_logi_full into a binary variable.
pred_full_35 <- ifelse(pred_logi_full > cutoff, 1, 0)
# A data frame with real and predicted loan status.
real_pred_35 <- cbind.data.frame(test$loan_st, pred_full_35)
# Loans that we accept given these new rules.
accepted_loans <- real_pred_35[pred_full_35 == 0, 1]
# Bad rate (accepted loan applications that are defaults).
bad_rate_65 <- sum(accepted_loans == 1)/length(accepted_loans)
# Show the bad rate.
bad_rate_65
```

The bad rate falls from `r fmt_pct(bad_rate_80, 2)` to `r fmt_pct(bad_rate_65, 2)` when the acceptance rate is reduced from 80% to 65%. This illustrates the trade-off. A lower acceptance rate usually means fewer accepted defaults and fewer accepted customers. In the extreme, accepting 0 applications would produce a bad rate of zero and no loan business at all. The next section evaluates this trade-off across many acceptance rates instead of checking only 80% and 65%.

