```{r include=FALSE}
source("R/logistic-models.R")
```

## Acceptance strategy and model comparison

A bank needs to understand how the bad rate changes when it accepts more or fewer applications. The previous section compared two acceptance rates, 80% and 65%. We now evaluate the same idea over a wider grid of acceptance rates.

Once this trade-off is clear, the bank could move beyond a single accept/reject cutoff. For example, it could create risk bands: low-risk applicants are accepted at standard terms, medium-risk applicants are accepted with a higher interest rate or additional conditions, and high-risk applicants are rejected. The simple accept/reject rule below is the first step because it makes the mechanics visible.

```{r}
bank <- function(prob_of_default, actual_default) {
  actual_default <- as.numeric(as.character(actual_default))

  # Pre-define the acceptance rates
  accept_rate <- seq(1, 0, by = -0.05)
  
  # Calculate cutoffs for each acceptance rate
  cutoff <- quantile(prob_of_default, accept_rate)
  
  # Calculate bad rates for each cutoff
  bad_rate <- sapply(cutoff, function(thresh) {
    accepted <- prob_of_default <= thresh
    mean(actual_default[accepted] == 1)
  })
  
  # Create the result table
  table <- cbind(
    accept_rate,
    cutoff = round(cutoff, 4),
    bad_rate = round(bad_rate, 4)
  )
  
  return(list(table = table, bad_rate = bad_rate, accept_rate = accept_rate, cutoff = cutoff))
}
```

The function receives two inputs. The first input is the vector of predicted probabilities, such as `pred_logi_full`. The second input is the observed loan outcome, `test$loan_st`, which is needed to calculate how many accepted loans actually default. Evaluating this function for `logi_full` and `logi_age` lets us compare a stronger model with a weaker benchmark. A stronger model should produce lower bad rates for the same acceptance rate.

The next code applies the function to `pred_logi_full` and `pred_logi_age`.

```{r The bank strategy a good and a bad model}
# Apply the bank function.
bank_logi_full <- bank(pred_logi_full, test$loan_st)
bank_logi_age <- bank(pred_logi_age, test$loan_st)

data.frame(
  `Acceptance rate` = fmt_pct(bank_logi_age$accept_rate, 0),
  `Age-only bad rate` = fmt_pct(bank_logi_age$bad_rate, 2),
  `Full-model bad rate` = fmt_pct(bank_logi_full$bad_rate, 2),
  check.names = FALSE
) |>
  knitr::kable(
    caption = "Bad rate by acceptance rate for the age-only and full logistic models.",
    row.names = FALSE
  )
```
The table suggests that the full model is more useful for decision-making because it achieves a lower bad rate at any given acceptance rate. The next plot highlights the key differences between `logi_age` and `logi_full`.

```{r}
#| label: fig-agaabm
#| fig-cap: "Bad-rate comparison between the age-only model and the full model."
# Combine the two datasets for easier comparison
bank_logi_full$model <- "logi_full"
bank_logi_age$model <- "logi_age"
combined_data <- bind_rows(bank_logi_full, bank_logi_age)

# Add vertical and horizontal lines for visualization
highlight_lines <- tibble(
  model = c("logi_full", "logi_full", "logi_age", "logi_age"),
  accept_rate = c(bank_logi_full[["accept_rate"]][8],
                  bank_logi_full[["accept_rate"]][5],
                  bank_logi_age[["accept_rate"]][8],
                  bank_logi_age[["accept_rate"]][5]),
  bad_rate = c(bank_logi_full[["bad_rate"]][8],
               bank_logi_full[["bad_rate"]][5],
               bank_logi_age[["bad_rate"]][8],
               bank_logi_age[["bad_rate"]][5]),
  color = c("black", "red", "black", "red")
)

# Add a color column to distinguish facets
combined_data <- combined_data %>%
  mutate(facet_color = ifelse(model == "logi_full", "blue", "orange"))

ggplot(combined_data, aes(x = accept_rate, y = bad_rate, color = facet_color)) +
  geom_line(linewidth = 1.7) +  # Use facet-specific colors for the lines
  geom_vline(data = highlight_lines, aes(xintercept = accept_rate, color = color),
             linetype = "dashed", linewidth = 0.65) +
  geom_hline(data = highlight_lines, aes(yintercept = bad_rate, color = color),
             linetype = "dashed", linewidth = 0.65) +
  facet_wrap(~model, nrow = 1, scales = "free_y") +
  labs(x = "Acceptance rate", y = "Bad rate") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  scale_color_identity() +  # Use the colors directly without adding them to the legend
  scale_x_continuous(breaks = c(0, 0.65, 0.8, 1)) +
  scale_y_continuous(
    labels = function(x) fmt_pct(x, 0),
    breaks = scales::breaks_pretty(n = 5)
  )
```


The plot reinforces the same conclusion: at the same acceptance rate, `logi_full` produces a lower bad rate than `logi_age`. The two panels use separate vertical scales so the age-only curve remains readable even though its bad-rate values are very close to each other. The value of a stronger model lies in its capacity to support better business decisions, in this case, more accurate credit evaluations.

The bad rate is useful, but it only describes part of the business decision. A very restrictive strategy can produce a low bad rate by rejecting many applicants, including many applicants who would have repaid their loans. To connect the model with a simple decision criterion, we define a hypothetical *net payoff*. It should be read as a transparent scoring rule, with accepting a good loan treated as a benefit and accepting a bad loan treated as a cost.

Suppose that accepting a loan that does not default gives the bank a payoff of \$1, while accepting a loan that defaults costs \$5. Rejected loans receive a payoff of \$0 because the bank does not issue the loan. Under these assumptions:

$$
\text{net payoff} =
(\text{accepted good loans}) \times \text{\$1} -
(\text{accepted default loans}) \times \text{\$5}.
$$

In the code below, `accepted_good` is the first term, `accepted_defaults` is the second term, `good_loan_payoff = 1` represents \$1, and `bad_loan_cost = 5` represents \$5. The expression `accepted_good * good_loan_payoff - accepted_defaults * bad_loan_cost` is therefore the code version of the equation. This lets us compare acceptance rates using the same model and the same test set.

```{r Net payoff by acceptance rate}
actual_default_strategy <- as.numeric(as.character(test$loan_st))

net_payoff_strategy <- function(prob_of_default, accept_rate,
                                good_loan_payoff = 1,
                                bad_loan_cost = 5) {
  cutoff <- unname(quantile(prob_of_default, accept_rate))
  accept <- prob_of_default <= cutoff
  accepted_good <- sum(actual_default_strategy[accept] == 0)
  accepted_defaults <- sum(actual_default_strategy[accept] == 1)
  accepted <- sum(accept)
  net_payoff <- accepted_good * good_loan_payoff -
    accepted_defaults * bad_loan_cost

  data.frame(
    accept_rate = accept_rate,
    cutoff = cutoff,
    accepted = accepted,
    accepted_good = accepted_good,
    accepted_defaults = accepted_defaults,
    bad_rate = accepted_defaults / accepted,
    net_payoff = net_payoff
  )
}

acceptance_grid <- c(0.50, 0.60, 0.65, 0.70, 0.75, 0.80, 0.85, 0.90)
net_payoff_results <- do.call(
  rbind,
  lapply(acceptance_grid, net_payoff_strategy,
         prob_of_default = pred_logi_full)
)

best_net_payoff_row <- net_payoff_results[
  which.max(net_payoff_results$net_payoff),
]

fmt_dollar <- function(x) {
  paste0(ifelse(x < 0, "-$", "$"), fmt_int(abs(x)))
}

net_payoff_results |>
  mutate(
    cutoff = round(cutoff, 4),
    bad_rate = fmt_pct(bad_rate, 2),
    net_payoff = fmt_dollar(net_payoff)
  ) |>
  knitr::kable(
    caption = "Net payoff across acceptance rates for the full logistic model.",
    row.names = FALSE
  )
```

The table contains the complete payoff calculation, but the best row is worth checking explicitly. At the selected acceptance rate, the rule accepts `r best_net_payoff_row$accepted_good` good loans and `r best_net_payoff_row$accepted_defaults` defaulting loans. Under the assumed payoffs, the calculation is:

$$
`r best_net_payoff_row$accepted_good`(\text{\$1})
-
`r best_net_payoff_row$accepted_defaults`(\text{\$5})
=
\text{\$`r fmt_int(best_net_payoff_row$net_payoff)`}.
$$

```{r Net payoff numerical check}
best_net_payoff_check <- data.frame(
  Quantity = c(
    "Acceptance rate",
    "Accepted good loans",
    "Payoff from good loans ($1 each)",
    "Accepted default loans",
    "Cost from default loans ($5 each)",
    "Net payoff"
  ),
  Value = c(
    fmt_pct(best_net_payoff_row$accept_rate, 0),
    fmt_int(best_net_payoff_row$accepted_good),
    fmt_dollar(best_net_payoff_row$accepted_good * 1),
    fmt_int(best_net_payoff_row$accepted_defaults),
    fmt_dollar(-best_net_payoff_row$accepted_defaults * 5),
    fmt_dollar(best_net_payoff_row$net_payoff)
  )
)

knitr::kable(
  best_net_payoff_check,
  caption = "Numerical check of the best net-payoff row.",
  row.names = FALSE
)
```

In this example, the highest net payoff occurs at an acceptance rate of `r fmt_pct(best_net_payoff_row$accept_rate, 0)`. The result is specific to this payoff assumption: the additional good loans accepted when moving from 65% to 80% more than compensate for the additional defaults. After 80%, accepting more applications still brings in more good loans, but the extra defaults become too costly under this rule.

The preferred acceptance rate depends on how costly defaults are relative to good accepted loans. The calculation can therefore be repeated under different assumptions for the cost of accepting a default.

```{r Net payoff sensitivity}
best_net_payoff <- function(bad_loan_cost) {
  payoff_grid <- do.call(
    rbind,
    lapply(seq(0.50, 0.95, 0.05), net_payoff_strategy,
           prob_of_default = pred_logi_full,
           bad_loan_cost = bad_loan_cost)
  )

  payoff_grid[which.max(payoff_grid$net_payoff), ]
}

net_payoff_sensitivity <- do.call(
  rbind,
  lapply(c(2, 5, 10, 15), best_net_payoff)
)

data.frame(
  bad_loan_cost = fmt_dollar(c(2, 5, 10, 15)),
  best_accept_rate = net_payoff_sensitivity$accept_rate,
  accepted_defaults = net_payoff_sensitivity$accepted_defaults,
  bad_rate = net_payoff_sensitivity$bad_rate,
  net_payoff = fmt_dollar(net_payoff_sensitivity$net_payoff)
) |>
  mutate(
    best_accept_rate = fmt_pct(best_accept_rate, 0),
    accepted_defaults = fmt_int(accepted_defaults),
    bad_rate = fmt_pct(bad_rate, 2)
  ) |>
  knitr::kable(
    caption = "Net-payoff sensitivity to the cost of accepting a default.",
    row.names = FALSE
  )
```

The sensitivity table shows the economic intuition. If accepting a default is relatively less costly, the strategy can tolerate a higher acceptance rate. If accepting a default is much more costly, the strategy becomes more conservative. The assumed payoffs are a transparent cost-benefit exercise for cutoff selection, not real bank profits.

This simple net payoff rule is closely related to the credit-risk idea of expected loss. A standard expected-loss expression is:

$$
\text{Expected loss} = PD \times LGD \times EAD,
$$

where $PD$ is the probability of default, $LGD$ is loss given default, and $EAD$ is exposure at default. In this chapter, the predicted probability $\hat p_i$ plays the role of the model-based $PD$, while \$1 and \$5 are simplified teaching assumptions about benefits and costs. A full profitability model would also include loan amount, interest income, funding costs, recovery rates, and exposure. The simplified net payoff is a first cost-benefit version of the same decision logic before adding those additional banking details.

## ROC, AUC, and calibration

Before introducing ROC curves, we need to fix the classification language. In this chapter, the positive class is default. A predicted positive is therefore a predicted default, which means the lending rule rejects the applicant. A predicted negative is a predicted no-default, which means the lending rule accepts the applicant.

```{r classification-equivalence-table}
classification_equivalence_table <- data.frame(
  `Actual outcome` = c(
    "No default",
    "No default",
    "Default",
    "Default"
  ),
  `Model prediction` = c(
    "No default",
    "Default",
    "No default",
    "Default"
  ),
  `Statistical name` = c(
    "True negative (TN)",
    "False positive (FP)",
    "False negative (FN)",
    "True positive (TP)"
  ),
  `Credit reading` = c(
    "The model expected repayment, and the borrower repaid.",
    "The model expected default, but the borrower would have repaid.",
    "The model expected repayment, but the borrower defaulted.",
    "The model expected default, and the borrower defaulted."
  ),
  `Lending consequence` = c(
    "Good loan accepted",
    "Good applicant rejected",
    "Defaulting borrower accepted",
    "Default avoided"
  ),
  check.names = FALSE
)

knitr::kable(
  classification_equivalence_table,
  caption = "Confusion-matrix language translated into credit decisions.",
  row.names = FALSE
)
```

The table is useful because the words positive and negative can be misleading in credit risk. Positive does not mean good. It means that the event of interest occurred, and here the event of interest is default. A false negative is therefore costly for the lender because the model accepts a borrower who defaults. A false positive is also costly, but in a different way: the bank rejects a borrower who would have repaid.

The same language gives the four rates used in ROC analysis.

```{r classification-metric-equivalence-table}
classification_metric_table <- data.frame(
  Metric = c(
    "Sensitivity",
    "Specificity",
    "False positive rate",
    "False negative rate"
  ),
  Formula = c(
    "$TP/(TP+FN)$",
    "$TN/(TN+FP)$",
    "$FP/(TN+FP)=1-\\text{specificity}$",
    "$FN/(TP+FN)=1-\\text{sensitivity}$"
  ),
  `Credit meaning` = c(
    "Among actual defaults, the share correctly detected as predicted defaults.",
    "Among actual non-defaults, the share correctly kept as predicted no-defaults.",
    "Among good borrowers, the share wrongly rejected.",
    "Among defaulting borrowers, the share wrongly accepted."
  ),
  check.names = FALSE
)

knitr::kable(
  classification_metric_table,
  caption = "Sensitivity, specificity, and error rates in credit language.",
  row.names = FALSE,
  escape = FALSE
)
```

The acceptance-rate analysis compares business rules at selected cutoffs. The next metrics summarize ranking, calibration, and sensitivity/specificity across cutoffs. The ROC (Receiver Operating Characteristic) curve evaluates how well the model ranks defaults above non-defaults as the cutoff changes.

Throughout this book, we draw ROC curves with sensitivity on the vertical axis and the false positive rate on the horizontal axis. The false positive rate is $1 - \text{specificity}$. In credit terms, it is the share of non-default loans that the model incorrectly classifies as predicted defaults at a given cutoff. This convention matches the ROC figures used in the next chapter.

The area under the ROC curve (AUC) quantifies the model's ability to discriminate between the two classes: a higher AUC indicates better ranking performance. For example, a model with an AUC close to 1 tends to assign higher default probabilities to defaulting borrowers than to non-defaulting borrowers, while an AUC of 0.5 indicates no better ranking performance than random guessing. By analyzing the trade-off between sensitivity (correctly identifying defaults) and specificity (correctly identifying non-defaults), the ROC curve helps determine an operational threshold for classification, balancing the cost of misclassifying borrowers.

```{r}
#| label: fig-fmibamio
#| fig-cap: "ROC curves: Full model in blue, age model in orange."

# Calculate ROC curves
ROC_logi_full <- roc(test$loan_st, pred_logi_full, quiet = TRUE)
ROC_logi_age <- roc(test$loan_st, pred_logi_age, quiet = TRUE)
auc_logi_full <- as.numeric(auc(ROC_logi_full))
auc_logi_age <- as.numeric(auc(ROC_logi_age))

roc_to_df_logistic <- function(roc_object, model_name) {
  data.frame(
    false_positive_rate = 1 - roc_object$specificities,
    true_positive_rate = roc_object$sensitivities,
    model = model_name,
    auc = as.numeric(auc(roc_object))
  )
}

roc_logistic_comparison <- bind_rows(
  roc_to_df_logistic(ROC_logi_full, "logi_full"),
  roc_to_df_logistic(ROC_logi_age, "logi_age")
) |>
  mutate(
    model_label = paste0(model, " (AUC = ", fmt_num(auc, 4), ")"),
    model_label = factor(model_label, levels = unique(model_label))
  ) |>
  arrange(model_label, false_positive_rate)

ggplot(
  roc_logistic_comparison,
  aes(x = false_positive_rate, y = true_positive_rate, color = model_label)
) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.7
  ) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    values = c("blue", "orange"),
    name = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  labs(
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

In @fig-fmibamio, the blue curve for `logi_full` dominates the orange curve for `logi_age` over most of the plot. This indicates that `logi_full` gives a better ranking of high-risk and low-risk applicants. The exact AUC values make that comparison explicit:

```{r AUC good and bad model}
# Compute the AUCs:
data.frame(
  model = c("logi_full", "logi_age"),
  auc = round(c(auc_logi_full, auc_logi_age), 4)
)
```

The `logi_age` model has an AUC of `r fmt_num(auc_logi_age, 4)`, which is close to the value produced by random ranking. That means age alone adds very little useful ranking information in this application. The `logi_full` model has an AUC of `r fmt_num(auc_logi_full, 4)`, so it ranks defaulting applicants above non-defaulting applicants much more effectively.

AUC answers a ranking question: do applicants who default tend to receive higher predicted probabilities than applicants who do not default? Calibration answers a probability question. If the model assigns a group of applicants an average predicted probability of default of 20%, do about 20% of those applicants actually default?

This distinction is important because we have been calling the output of the model a probability of default. If the model is well calibrated, the predicted probabilities can be interpreted more literally. If the model is poorly calibrated, the model may still be useful for ranking applicants from low risk to high risk, but the numerical probability values should be interpreted with more caution.

One simple way to inspect calibration is to divide the test set into groups according to the predicted probability of default. Here we use ten groups, or deciles. The first decile contains the applicants with the lowest predicted probabilities, and the tenth decile contains the applicants with the highest predicted probabilities. Then, for each decile and each model, we compare the average predicted probability with the observed default rate.

```{r Calibration table full and age models}
calibration_data <- bind_rows(
  data.frame(
    model = "logi_full",
    observed_default = as.numeric(as.character(test$loan_st)),
    predicted_pd = pred_logi_full
  ),
  data.frame(
    model = "logi_age",
    observed_default = as.numeric(as.character(test$loan_st)),
    predicted_pd = pred_logi_age
  )
)

calibration_models <- calibration_data |>
  group_by(model) |>
  mutate(risk_decile = ntile(predicted_pd, 10)) |>
  group_by(model, risk_decile) |>
  summarize(
    applications = n(),
    average_predicted_pd = mean(predicted_pd),
    observed_default_rate = mean(observed_default),
    .groups = "drop"
  ) |>
  ungroup()

calibration_models |>
  mutate(
    average_predicted_pd = fmt_pct(average_predicted_pd, 2),
    observed_default_rate = fmt_pct(observed_default_rate, 2)
  ) |>
  knitr::kable(
    caption = "Calibration by predicted-risk decile for the full and age-only models.",
    row.names = FALSE
  )
```

The interpretation is direct. If the average predicted probability and the observed default rate are close within each decile, the model is reasonably calibrated for that part of the risk distribution. If the observed default rate is systematically above the predicted probability, the model is underestimating credit risk. If the observed default rate is systematically below the predicted probability, the model is overestimating credit risk.

The same idea appears in the next plot. The dashed line represents perfect calibration. Points close to the dashed line indicate that predicted probabilities and observed default rates are similar. Points far from the dashed line indicate calibration problems. The colors match the ROC figure: blue for the full model and orange for the age model.

```{r}
#| label: fig-calibration-models
#| fig-cap: "Calibration by decile: full model in blue, age model in orange."
ggplot(calibration_models,
       aes(x = average_predicted_pd, y = observed_default_rate,
           color = model)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              color = "gray50") +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = c("logi_full" = "blue", "logi_age" = "orange")) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Average predicted probability of default",
       y = "Observed default rate",
       color = "Model") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The Brier score gives us a compact numerical summary of probability prediction error. It is the average squared distance between the predicted probability and the observed outcome, where the observed outcome is 1 for default and 0 for no default:

$$
\text{Brier score} =
\frac{1}{n}\sum_{i=1}^{n}(\hat{p}_i - y_i)^2.
$$

A lower Brier score indicates better probability predictions. The measure is useful here because it evaluates the probabilities directly, before converting them into a binary accept/reject decision. The function `brier_score(actual, predicted_probability)` implements the formula directly: `actual` is $y_i$, `predicted_probability` is $\hat p_i$, and `mean((predicted_probability - actual)^2)` is the average squared error.

```{r Brier score}
brier_score <- function(actual, predicted_probability) {
  mean((predicted_probability - actual)^2)
}

actual_default <- as.numeric(as.character(test$loan_st))
constant_pd <- rep(mean(actual_default), length(actual_default))

brier_results <- data.frame(
  model = c("logi_full", "logi_age", "constant default rate"),
  brier_score = c(
    brier_score(actual_default, pred_logi_full),
    brier_score(actual_default, pred_logi_age),
    brier_score(actual_default, constant_pd)
  )
)

baseline_brier <- brier_results$brier_score[
  brier_results$model == "constant default rate"
]
brier_results$brier_skill <- 1 - brier_results$brier_score / baseline_brier

brier_results |>
  mutate(
    brier_score = fmt_num(brier_score, 4),
    brier_skill = fmt_num(brier_skill, 4)
  ) |>
  knitr::kable(
    caption = "Brier score and Brier skill against a constant default-rate benchmark.",
    row.names = FALSE
  )
```

The constant default-rate model is a simple benchmark: it assigns the same probability of default to every applicant, equal to the observed default rate in the test set. A useful probability model should improve on that benchmark. The Brier skill score expresses this improvement relative to the benchmark:

$$
\text{Brier skill} =
1 - \frac{\text{Brier score of the model}}
{\text{Brier score of the benchmark}}.
$$

A value close to 0 means the model is barely improving on the constant default-rate benchmark; a positive value means the model reduces probability error relative to that benchmark. In code, `baseline_brier` is the denominator and `1 - brier_results$brier_score / baseline_brier` is the Brier skill calculation. The Brier score combines probability accuracy with the model's ability to separate high-risk and low-risk applicants, so it is a convenient way to summarize whether the predicted probabilities are numerically close to the observed default outcomes.

In practice, discrimination and calibration answer complementary questions. AUC helps us decide whether the model ranks applicants in a useful way. Calibration helps us decide whether the estimated probabilities can be interpreted as realistic probabilities of default. A lending strategy can benefit from both: ranking is useful for deciding whom to accept or reject, while calibration is useful for pricing, provisioning, stress testing, and communicating the meaning of a predicted probability.

We can now return to the business rule. A cutoff converts a predicted probability into an accept/reject decision. Under our convention, a higher sensitivity means fewer defaulting borrowers are mistakenly accepted. A higher specificity means fewer good borrowers are mistakenly rejected.

The trade-off is financial. A very low cutoff rejects many applicants and tends to increase sensitivity, but it can reject many good borrowers. A very high cutoff accepts more applicants and tends to increase specificity, but it can accept more borrowers who default. The right balance depends on risk appetite, profitability, and regulation.

The two extreme policies make the problem clear. Accepting everyone maximizes loan volume and also accepts every default. Rejecting everyone eliminates accepted defaults and issues no loans. Useful credit policy lives between those extremes.

We evaluate two lending policies reviewed before. In the first policy, the bank wants to accept about 80% of applications. In the second policy, the bank wants to accept about 65%. This policy target is chosen before looking at the final classification counts.

The model then converts each policy into a probability cutoff. In R, this is done with `quantile()`. The call `quantile(pred_logi_full, probs = 0.80)` returns the empirical 80th percentile of the full model's predicted default probabilities. We use that value as the cutoff because it places about 80% of the predicted probabilities at or below the cutoff. Those applications are accepted. Applications above the cutoff are rejected because their predicted default probabilities are in the riskiest part of the test set. The same logic gives the 65% policy by using `probs = 0.65`.

The word about is important. The bank chooses the policy target first, such as accepting about 80% of applications. Then `quantile()` finds an empirical cutoff from the ordered predicted probabilities. In R, the default quantile calculation may interpolate between adjacent ordered predictions. It does not smooth the whole distribution; it produces one cutoff value from the sample. After that, the accept/reject rule is applied to the observed predictions. If predicted probabilities are spread smoothly with few ties near the cutoff, the realized acceptance rate will be very close to the policy target. If many predictions are tied or tightly clustered around the cutoff, the realized rate can move a little farther from the target. In our code, applications with predicted default probability at or below the cutoff are accepted, while applications above the cutoff are rejected.


```{r}
# Calculate sensitivity and specificity from a confusion matrix.
classification_metrics <- function(actual, predicted) {
  cm <- table(
    Actual = factor(actual, levels = c("0", "1")),
    Predicted = factor(predicted, levels = c(0, 1))
  )
  tn <- cm["0", "0"]
  fp <- cm["0", "1"]
  fn <- cm["1", "0"]
  tp <- cm["1", "1"]
  c(
    sensitivity = as.numeric(tp / (tp + fn)),
    specificity = as.numeric(tn / (tn + fp))
  )
}

# Cutoffs implied by target acceptance rates.
cutoff_full_80 <- unname(quantile(pred_logi_full, probs = 0.80))
cutoff_age_80 <- unname(quantile(pred_logi_age, probs = 0.80))
cutoff_full_65 <- unname(quantile(pred_logi_full, probs = 0.65))
cutoff_age_65 <- unname(quantile(pred_logi_age, probs = 0.65))

# Binary predictions: 1 = reject as predicted default, 0 = accept.
pred_full_80 <- ifelse(pred_logi_full > cutoff_full_80, 1, 0)
pred_age_80 <- ifelse(pred_logi_age > cutoff_age_80, 1, 0)
pred_full_65 <- ifelse(pred_logi_full > cutoff_full_65, 1, 0)
pred_age_65 <- ifelse(pred_logi_age > cutoff_age_65, 1, 0)

metrics_full_80 <- classification_metrics(test$loan_st, pred_full_80)
metrics_age_80 <- classification_metrics(test$loan_st, pred_age_80)
metrics_full_65 <- classification_metrics(test$loan_st, pred_full_65)
metrics_age_65 <- classification_metrics(test$loan_st, pred_age_65)

# Combine the results.
results_df <- data.frame(
  Model = c("logi_full", "logi_age", "logi_full", "logi_age"),
  Policy = c("Accept about 80%", "Accept about 80%",
             "Accept about 65%", "Accept about 65%"),
  Accepted_in_test_set = c(mean(pred_full_80 == 0),
                           mean(pred_age_80 == 0),
                           mean(pred_full_65 == 0),
                           mean(pred_age_65 == 0)),
  PD_cutoff = c(cutoff_full_80, cutoff_age_80,
                cutoff_full_65, cutoff_age_65),
  Sensitivity = c(metrics_full_80["sensitivity"],
                  metrics_age_80["sensitivity"],
                  metrics_full_65["sensitivity"],
                  metrics_age_65["sensitivity"]),
  Specificity = c(metrics_full_80["specificity"],
                  metrics_age_80["specificity"],
                  metrics_full_65["specificity"],
                  metrics_age_65["specificity"])
)

results_df |>
  mutate(
    Accepted_in_test_set = fmt_pct(Accepted_in_test_set, 1),
    PD_cutoff = fmt_pct(PD_cutoff, 2),
    Sensitivity = fmt_pct(Sensitivity, 1),
    Specificity = fmt_pct(Specificity, 1)
  ) |>
  knitr::kable(
    caption = "Cutoffs and error rates implied by two acceptance policies.",
    col.names = c(
      "Model",
      "Bank policy",
      "Accepted in test set",
      "PD cutoff",
      "Sensitivity",
      "Specificity"
    ),
    row.names = FALSE
  )

```

The summary table should be read as a policy comparison. `Policy` is the acceptance goal chosen by the bank. `PD_cutoff` is the predicted default probability that implements that goal. `Accepted_in_test_set` is the share of applications actually accepted after applying the cutoff. Moving from the 80% policy to the 65% policy makes the rule more conservative because more applicants are rejected. That change usually detects more actual defaults, raising sensitivity. It also rejects more non-defaulting applicants, reducing specificity.

```{r}
#| label: fig-fmibeitar
#| fig-cap: "Full model in blue, evaluated in two acceptance rates: 65% in black, 80% in red."
roc_full_df <- roc_to_df_logistic(ROC_logi_full, "logi_full") |>
  arrange(false_positive_rate)

roc_acceptance_guides <- data.frame(
  target = c("Acceptance rate of 80%", "Acceptance rate of 65%"),
  false_positive_rate = c(
    1 - metrics_full_80["specificity"],
    1 - metrics_full_65["specificity"]
  ),
  true_positive_rate = c(
    metrics_full_80["sensitivity"],
    metrics_full_65["sensitivity"]
  )
)

roc_acceptance_labels <- data.frame(
  target = c("Acceptance rate of 80%", "Acceptance rate of 65%"),
  false_positive_rate = roc_acceptance_guides$false_positive_rate,
  true_positive_rate = roc_acceptance_guides$true_positive_rate,
  label_x = c(0.56, 0.56),
  label_y = c(0.55, 0.84),
  label = c(
    paste0(
      "80% policy\n",
      "Sensitivity = ", fmt_pct(metrics_full_80["sensitivity"], 1), "\n",
      "Specificity = ", fmt_pct(metrics_full_80["specificity"], 1)
    ),
    paste0(
      "65% policy\n",
      "Sensitivity = ", fmt_pct(metrics_full_65["sensitivity"], 1), "\n",
      "Specificity = ", fmt_pct(metrics_full_65["specificity"], 1)
    )
  )
)

ggplot(roc_full_df, aes(x = false_positive_rate, y = true_positive_rate)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.7
  ) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_segment(
    data = roc_acceptance_guides,
    aes(
      x = false_positive_rate,
      xend = false_positive_rate,
      y = 0,
      yend = true_positive_rate,
      color = target
    ),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_segment(
    data = roc_acceptance_guides,
    aes(
      x = 0,
      xend = false_positive_rate,
      y = true_positive_rate,
      yend = true_positive_rate,
      color = target
    ),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 1
  ) +
  geom_point(
    data = roc_acceptance_guides,
    aes(
      x = false_positive_rate,
      y = true_positive_rate,
      color = target
    ),
    inherit.aes = FALSE,
    size = 2.8
  ) +
  geom_segment(
    data = roc_acceptance_labels,
    aes(
      x = false_positive_rate,
      xend = label_x - 0.02,
      y = true_positive_rate,
      yend = label_y,
      color = target
    ),
    inherit.aes = FALSE,
    linewidth = 0.45,
    alpha = 0.7,
    show.legend = FALSE
  ) +
  geom_label(
    data = roc_acceptance_labels,
    aes(
      x = label_x,
      y = label_y,
      label = label,
      color = target
    ),
    inherit.aes = FALSE,
    hjust = 0,
    size = 3.1,
    lineheight = 0.95,
    label.size = 0.25,
    label.padding = grid::unit(0.15, "lines"),
    fill = "white",
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c("Acceptance rate of 80%" = "red",
               "Acceptance rate of 65%" = "black"),
    name = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  labs(
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

@fig-fmibeitar places the two full-model policies on the same ROC curve. The red point corresponds to the 80% acceptance rule, which rejects only the riskiest 20% of applicants. The black point corresponds to the 65% acceptance rule, which rejects the riskiest 35%. The black point is higher because the stricter rule catches more actual defaults. It is also farther to the right because the stricter rule rejects more applicants who would not have defaulted. The figure therefore turns the cutoff choice into a visible risk-management trade-off.

The next two confusion-matrix outputs verify the same logic numerically. Rows are actual outcomes. Columns are lending decisions implied by the model. `Accept` means predicted no-default and `Reject` means predicted default.

For the 80% acceptance target:

```{r Confusion matrix pred_full_80a}
cm_full_80 <- table(
  Actual = factor(
    test$loan_st,
    levels = c("0", "1"),
    labels = c("No default", "Default")
  ),
  Decision = factor(
    pred_full_80,
    levels = c(0, 1),
    labels = c("Accept", "Reject")
  )
)
tn_80 <- cm_full_80["No default", "Accept"]
fp_80 <- cm_full_80["No default", "Reject"]
fn_80 <- cm_full_80["Default", "Accept"]
tp_80 <- cm_full_80["Default", "Reject"]
sens_80 <- as.numeric(tp_80 / (tp_80 + fn_80))
spec_80 <- as.numeric(tn_80 / (tn_80 + fp_80))

cm_full_80
```

For this rule, the model rejects `r tp_80` applicants who actually default and accepts `r fn_80` applicants who later default. Sensitivity is therefore `r round(sens_80, 4)`. The model accepts `r tn_80` applicants who do not default and rejects `r fp_80` applicants who do not default. Specificity is therefore `r round(spec_80, 4)`.

For the 65% acceptance target:

```{r Confusion matrix pred_full_65a}
cm_full_65 <- table(
  Actual = factor(
    test$loan_st,
    levels = c("0", "1"),
    labels = c("No default", "Default")
  ),
  Decision = factor(
    pred_full_65,
    levels = c(0, 1),
    labels = c("Accept", "Reject")
  )
)
tn_65 <- cm_full_65["No default", "Accept"]
fp_65 <- cm_full_65["No default", "Reject"]
fn_65 <- cm_full_65["Default", "Accept"]
tp_65 <- cm_full_65["Default", "Reject"]
sens_65 <- as.numeric(tp_65 / (tp_65 + fn_65))
spec_65 <- as.numeric(tn_65 / (tn_65 + fp_65))

cm_full_65
```

For this rule, the model rejects `r tp_65` applicants who actually default and accepts `r fn_65` applicants who later default. Sensitivity rises to `r round(sens_65, 4)`. The cost is visible among non-defaulting applicants: the model accepts `r tn_65` of them and rejects `r fp_65` of them, so specificity falls to `r round(spec_65, 4)`.

The comparison shows why a cutoff is a business decision. The 65% acceptance rule protects the lender from more defaults, but it also turns away more borrowers who would have repaid. The 80% acceptance rule keeps more business volume and rejects fewer good borrowers, while allowing more defaulting borrowers into the portfolio.

A pure-random score would look like this:

```{r}
#| label: fig-rrm
#| fig-cap: "Random-score ROC benchmark."
set.seed(2020)
pred_rand_model <- runif(length(pred_logi_age))
ROC_rand <- roc(test$loan_st, pred_rand_model, quiet = TRUE)
auc_rand <- as.numeric(auc(ROC_rand))

roc_rand_df <- roc_to_df_logistic(ROC_rand, "Random score") |>
  arrange(false_positive_rate)

ggplot(roc_rand_df, aes(x = false_positive_rate, y = true_positive_rate)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.7
  ) +
  geom_line(color = "red", linewidth = 1.2) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  labs(
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  coord_equal() +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

c(random_score_auc = round(auc_rand, 4))
```

A random score should produce a ROC curve close to the diagonal. The area under that diagonal is 0.5, so an AUC close to 0.5 means the score has almost no ranking power. Because the code uses a finite random sample, the reported AUC can be close to 0.5 without being exactly 0.5.

A perfect score gives the opposite benchmark:

```{r}
#| label: fig-pm
#| fig-cap: "Perfect-score ROC benchmark."
pred_perfect_model <- as.numeric(as.character(test$loan_st))
ROC_perfect <- roc(test$loan_st, pred_perfect_model, quiet = TRUE)
auc_perfect <- as.numeric(auc(ROC_perfect))

roc_perfect_df <- data.frame(
  false_positive_rate = c(0, 0, 1),
  true_positive_rate = c(0, 1, 1)
)

ggplot(roc_perfect_df, aes(x = false_positive_rate, y = true_positive_rate)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "gray60",
    linewidth = 0.7
  ) +
  geom_path(color = "blue", linewidth = 1.2) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0.01, 0.01)
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.02),
    breaks = seq(0, 1, 0.25),
    expand = c(0.01, 0.01)
  ) +
  labs(
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)"
  ) +
  coord_equal() +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

c(perfect_score_auc = round(auc_perfect, 4))
```

In a perfect score, every applicant who defaults receives a higher score than every applicant who does not default. The ROC curve therefore rises vertically from 0% to 100% sensitivity while the false positive rate is still 0%, and then moves horizontally along the top of the graph. The AUC is 1. At the right cutoff, the model identifies all defaults and all non-defaults correctly, giving 100% sensitivity and 100% specificity.

## What do we gain and what do we lose?

Logistic regression gives us a transparent credit-scoring benchmark. The model produces a predicted probability of default, and the coefficients can be interpreted through log-odds and odds ratios. This makes the model easier to audit, explain, and connect to a lending rule.

The cost of that transparency is structure. Unless we manually add nonlinear terms or interactions, the model assumes a relatively simple relationship between the predictors and the log-odds of default. That can be too restrictive if credit risk depends on combinations of variables, such as interest rate and grade, or loan amount and income.

The main gain from this chapter is therefore a disciplined workflow. We started with a probability model, converted probabilities into cutoff decisions, evaluated the resulting accepted portfolio, and connected the cutoff to bad rates, calibration, AUC, Brier score, and net payoff. The next chapter keeps the same credit scoring problem and the same evaluation logic, but changes the model class. That lets us ask a sharper question: can tree-based models capture interactions and nonlinear patterns that the logistic benchmark may miss, and do those extra patterns improve credit decisions out of sample?
