```{r include=FALSE}
source("R/logistic-data.R")
```

## Logistic models

Logistic models allow us to make predictions about loan defaults. Logistic regression is a statistical model that in its basic form uses a logistic function to model a binary dependent variable like <tt>`loan_st`</tt>. In this case, the binary dependent variable is default (1) or no default (0). A good reference for this section is @hull2020machine.

First, load the data and split it into two sets: (1) training and (2) test. The training set is for building and estimating models, and the test set is used to evaluate our model predictions with new data. When estimating models, it is common practice to separate the available data into two parts, *training* and *test* data, where the training data is used to estimate parameters (in-sample) and the test data is used to evaluate its accuracy (out of sample). Because the test data is not used in determining the estimation, it should provide a reliable indication of how well the model is likely to estimate or forecast on new data. In sum, we train the model, we test the model, and once we are OK with the model performance on new data, we are ready to use it in real-life applications. If we ignore this split and use the whole database to estimate our models, we may succeed at explaining defaults in our database but we may fail to explain defaults for new loan applications.

```{r Training and test set construction}
# It is convenient to set the loan status as factor.
dat$loan_st <- as.factor(dat$loan_st)
set.seed(567)
index_train <- cbind(runif(1 : nrow(dat), 0 , 1), c(1 : nrow(dat)))
index_train <- order(index_train[, 1])
index_train <- index_train[1: (2/3 * nrow(dat))]
# Create training set
train <- dat[index_train, ]
# Create test set
test <- dat[-index_train, ]
```

Variables as factors are useful for model estimation and data visualization. Factors are variables in R which take on a limited number of different values; such variables are often referred to as categorical variables.

We have 29,083 observations in <tt>`dat`</tt>. The code above randomly selects $29083 \times (2/3)=19388$ rows to form the <tt>`train`</tt>. The <tt>`test`</tt> are the remaining $29083-19388=9695$ rows. The random selection is highly recommended as the <tt>`dat`</tt> may have some structure or sorting that could bias our model estimation and negatively impact our model test. For example, imagine that for some weird reason the database is sorted in such a way that the first observations are all no default. If that is the case, then the training and the test set would not have portions of default and no default cases and we may distort the whole analysis. The random selection allows us to replicate a real situation in which our database is unsorted, with different characteristics.

Let's see if the recent created <tt>`train`</tt> and <tt>`test`</tt> have the same basic properties than <tt>`dat`</tt>.

```{r}
# Combine data into a list for processing
data_list <- list(dat = dat$loan_st, train = train$loan_st, test = test$loan_st)

# Compute proportions and combine into a data frame
prop <- do.call(rbind, lapply(data_list, function(x) prop.table(table(x))))
colnames(prop) <- c("no defaults", "defaults")
row.names(prop) <- c("dat_prop", "train_prop", "test_prop")
prop
```

The proportions of "defaults" and "no defaults" in the train and test datasets are very similar to those in the original dataset, indicating that both samples are representative. Specifically, the proportion of "defaults" in the train and test datasets closely matches that of the original dataset, as does the proportion of "no defaults." This suggests that the random sampling process has preserved the overall distribution of the target variable, ensuring that the train and test sets reflect the characteristics of the full dataset.

Assume we think that the <tt>`loan_st`</tt> depends on the age of the individual. We can estimate a simple logistic model to learn about the relationship between age and loan status.


$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 \times \text{age}$.

Where:

* $p$ is the probability that `loan_st = 1` (e.g., the probability of default).

* $\beta_0$ is the intercept, and $\beta_1$ is the coefficient for the predictor variable `age`.

The dependent variable in logistic regression is expressed as: $\log\left(\frac{p}{1-p}\right)$. This transformation, known as the *logit function*, maps probabilities $p$ which range between 0 and 1 to the entire real number line $-\infty$ to $+\infty$. This allows the model to establish a linear relationship between the independent variable(s) (e.g., age) and the transformed dependent variable, making estimation feasible. The fraction $\frac{p}{1-p}$, called the *odds*, represents the ratio of the probability of an event happening $p$ to the probability of it not happening $1-p$, making it a natural choice for modeling binary outcomes.

Let's estimate the age model.

```{r logi_age estimation}
# Fitting a simple logistic model.
logi_age <- glm(loan_st ~ age, family = "binomial", data = train)
logi_age
```

The estimated model is: $\log\left(\frac{p}{1-p}\right) = -1.90097 - 0.00623 \times \text{age}$.

The model suggests there is a negative relationship between age and loan status. 

We can solve for $p$ to find the estimate of the probability of default according to the age model: $p = \frac{\exp(-1.90097 - 0.00623 \times \text{age})}{1 + \exp(-1.90097 - 0.00623 \times \text{age})}$. Let's evaluate $p$ for a couple of age values.

Substituting `age = 18`: $p = \frac{\exp(-1.90097 - 0.00623 \times 18)}{1 + \exp(-1.90097 - 0.00623 \times 18)}$,

$p = \frac{\exp(-2.01311)}{1 + \exp(-2.01311)} \approx \frac{0.1333}{1 + 0.1333} \approx 0.1176$.

Substituting `age = 60`: $p = \frac{\exp(-1.90097 - 0.00623 \times 60)}{1 + \exp(-1.90097 - 0.00623 \times 60)}$,

$p = \frac{\exp(-2.27477)}{1 + \exp(-2.27477)} \approx \frac{0.1037}{1 + 0.1037} \approx 0.0939$.

The results of the default probability estimation evaluated at ages 18 and 60 are not very promising because the difference in probabilities is very small. This means that the model does not generate significantly differentiated default probabilities, even when evaluated over a wide age range. This can be problematic because it suggests that this is a poor model.

The following is a graphical representation of the logistic regression model's output.

```{r}
#| label: fig-scolcam
#| fig-cap: "Sigmoid curve or logistic curve: age model."
# Define the function for the logistic regression model
logistic_function <- function(age) {
  beta0 <- -1.90097
  beta1 <- -0.00623
  p <- exp(beta0 + beta1 * age) / (1 + exp(beta0 + beta1 * age))
  return(p)
}

# Generate a sequence of ages
ages <- seq(-1000, 500, by = 1)

# Apply the logistic function to each age
p_values <- sapply(ages, logistic_function)

# Create the plot
library(ggplot2)
ggplot(data = data.frame(age = ages, p = p_values), 
       aes(x = age, y = p)) +
  geom_line(color = "blue", size = 2) +
  labs(x = "Age", y = "Probability (p)") +
  theme_minimal()
```
The logistic curve derived from this model is unreasonable because the probability $p$ changes very slowly with respect to age. This implies that, to observe the full range of probabilities from 0 to 1, we would require age values that are unrealistic or implausibly extreme. Such behavior suggests that the model is poorly specified, as it fails to generate meaningful distinctions in probabilities across a realistic range of ages.

The AIC value of the age model (13,580) is useful when comparing models. The Akaike information criterion (AIC) is a mathematical method for evaluating how well a model fits the data it was generated from. In statistics, AIC is used to compare different possible models and determine which one is the best fit for the data. At the moment we cannot interpret the AIC simply because we only have one model and we cannot compare it with another AIC.

Let's estimate another simple model where the interest rate category is used as a predictor of the <tt>`loan_st`</tt>. Remember we are not conducting any prediction at all at this moment, we are only estimating models using the training set.

```{r logi_int estimation}
# Build a glm model with variable interest rate as a predictor.
logi_int <- glm(formula = loan_st ~ int, family = "binomial", data = train)
# Print the parameter estimates.
logi_int
```

The AIC is lower (13,220 versus 13,580), so we have a better model now.

Using one single predictor as age or interest rate is clearly a limited approach. Let's add some more predictors. Also, let's introduce the <tt>`summary()`</tt> function to extract more information about the model estimation results. The <tt>`logi_multi`</tt> below assumes that the loan status depend on the age, interest rate, grade, loan amount, and annual income.

```{r logi_multi estimation}
# Multiple variables in a logistic regression model.
logi_multi <- glm(loan_st ~ age + int + grade + log(l_amnt) + 
                  log(income) , family = "binomial", data = train)
# Obtain significance levels using summary().
summary(logi_multi)
```

The multi-factor model is a stronger candidate than the previous two. In <tt>`logi_multi`</tt>, the AIC value is the lowest so far (13,050 versus 13,220), so this should be considered the best in-sample model at the moment. The <tt>`summary()`</tt> function shows the significance levels of the estimators, but we are currently more interested in the goodness of fit of the models because we want to conduct predictions about the <tt>`loan_st`</tt>. This is, we are interested to use a model to find out whether new applicants in the test set are expected to default or not, rather than in the applicants' credit risk factors. This is why we are concentrated in AIC now. However, AIC is still an in-sample criterion; the model's practical predictive value must be checked later with the test set.

When a customer fill out a credit application form, we collect information but we do not know for sure whether she or he will eventually default. A credit risk model can help us in this task.

