## Loan analysis

The main objective of this section is to predict whether a person or client applying for a loan will repay it or not. This issue can be considered a classification problem, where based on information about the client and the characteristics of their loan application, the client is classified into one of two categories: whether they will default or not. There are various ways to approach this classification problem; in this section, we will use a simple machine learning model called logistic regression. The model is first estimated or trained and then evaluated using new data to determine how well it performs the classification.

```{r echo=FALSE}
# This removes all items in environment. 
# It is a good practice to start your code this way.
rm(list=ls())
```

Let's load the required R packages first.

```{r Loan analysis}
# Logistic models
library(gmodels) # CrossTable()
library(ggplot2)
library(tidyr) # gather()
library(dplyr)
library(pROC) # roc
library(vembedr)
```

This section relies on the DataCamp course *Credit Risk Modeling in R* by Lore Dirick. However, we incorporate a slightly different database and an extended analysis.

