```{r include=FALSE}
source("R/logistic-data.R")
```

## Logistic regression models

Logistic regression is the first credit-scoring model in the book because it is transparent and directly produces probabilities of default. The dependent variable is `loan_st`, where 1 means default and 0 means no default. A good reference for this section is @hull2020machine.

We split the data into a training set and a test set. The training set is used to estimate the model. The test set is held back and used later to evaluate predictions on observations that were not used for estimation. This separation matters because a model can fit historical defaults well and still perform poorly on new loan applications.

```{r Training and test set construction}
# Set loan status as a factor.
dat$loan_st <- as.factor(dat$loan_st)
set.seed(567)
index_train <- cbind(runif(1 : nrow(dat), 0 , 1), c(1 : nrow(dat)))
index_train <- order(index_train[, 1])
index_train <- index_train[1: (2/3 * nrow(dat))]
# Create training set
train <- dat[index_train, ]
# Create test set
test <- dat[-index_train, ]
```

Factors are useful for categorical variables in R because they tell modeling and plotting functions that the variable takes a limited set of categories.

We have `r format(nrow(dat), big.mark = ",")` observations in `dat`. The code above randomly selects `r format(nrow(train), big.mark = ",")` rows to form `train`. `test` contains the remaining `r format(nrow(test), big.mark = ",")` rows. Random selection helps avoid a split that accidentally reflects some ordering in the data. For example, if the first observations were mostly no-default loans, a non-random split could distort both estimation and evaluation.

The next table checks whether `train` and `test` preserve the basic default proportions in `dat`.

```{r}
# Combine data into a list for processing
data_list <- list(dat = dat$loan_st, train = train$loan_st, test = test$loan_st)

# Compute proportions and combine into a data frame
prop <- do.call(rbind, lapply(data_list, function(x) prop.table(table(x))))
colnames(prop) <- c("no defaults", "defaults")
row.names(prop) <- c("dat_prop", "train_prop", "test_prop")
prop
```

The proportions of "defaults" and "no defaults" in the train and test datasets are very similar to those in the original dataset, indicating that both samples are representative. Specifically, the proportion of "defaults" in the train and test datasets closely matches that of the original dataset, as does the proportion of "no defaults." This suggests that the random sampling process has preserved the overall distribution of the target variable, ensuring that the train and test sets reflect the characteristics of the full dataset.

This split is enough for the teaching objective of the chapter: it lets us estimate models on one sample and evaluate their behavior on observations that were not used for estimation. In a production workflow, however, we would usually add one more layer. Model choices such as variable selection, cutoffs, and tuning decisions should be made with a training/validation structure or cross-validation, while a final holdout test set should be reserved for the last evaluation. In this book we use the test set repeatedly to make the logic visible, so the results should be read as a controlled learning exercise rather than as a complete production validation design.

We begin with a deliberately simple model that uses only age. To keep the mathematical notation separate from the code, let $Y_i$ denote the value stored in `loan_st` for applicant $i$, and let $\text{age}_i$ denote that applicant's age. This first model helps us see the mechanics of logistic regression before adding more predictors.

The model first defines the conditional probability that applicant $i$ defaults:

$$
p_i = P(Y_i = 1 \mid \text{age}_i).
$$

Then logistic regression assumes that the log-odds of that probability are linear in age:

$$
\log\left(\frac{p_i}{1-p_i}\right) = \beta_0 + \beta_1 \times \text{age}_i.
$$

Notice that this equation does not include an additive error term $\varepsilon_i$. In a linear regression, the error term is added to the equation for the observed outcome. In logistic regression, the observed outcome is binary, so the randomness is described by $Y_i \mid \text{age}_i \sim \text{Bernoulli}(p_i)$. The logit equation describes the conditional probability $p_i$, not the realized 0/1 outcome itself.

Where:

* $Y_i$ is the observed outcome in the data: $Y_i=1$ means default and $Y_i=0$ means no default.

* $p_i$ is the probability that applicant $i$ defaults, conditional on age.

* $\beta_0$ is the intercept, and $\beta_1$ is the coefficient for the predictor variable `age`.

The dependent variable in logistic regression remains binary: in the data, `loan_st` is either 0 or 1. What the model transforms is not the observed 0/1 outcome itself, but the conditional probability of default, $p_i$. The transformation $\log\left(\frac{p_i}{1-p_i}\right)$, known as the *logit function*, maps probabilities $p_i$, which range between 0 and 1, to the entire real number line $-\infty$ to $+\infty$. This allows the model to establish a linear relationship between the independent variable(s), such as age, and the transformed probability. The fraction $\frac{p_i}{1-p_i}$, called the *odds*, represents the ratio of the probability of an event happening $p_i$ to the probability of it not happening $1-p_i$, making it a natural choice for modeling binary outcomes.

We estimate the age model with `glm()`.

```{r logi_age estimation}
# Fitting a simple logistic model.
logi_age <- glm(loan_st ~ age, family = "binomial", data = train)
logi_age
```

```{r Age model teaching values, include=FALSE}
age_example_18 <- 18
age_example_60 <- 60
beta_age_0 <- unname(coef(logi_age)[["(Intercept)"]])
beta_age_1 <- unname(coef(logi_age)[["age"]])
age_logit_18 <- beta_age_0 + beta_age_1 * age_example_18
age_exp_18 <- exp(age_logit_18)
age_pd_18 <- age_exp_18 / (1 + age_exp_18)
age_logit_60 <- beta_age_0 + beta_age_1 * age_example_60
age_exp_60 <- exp(age_logit_60)
age_pd_60 <- age_exp_60 / (1 + age_exp_60)
age_min_train <- min(train$age, na.rm = TRUE)
age_max_train <- max(train$age, na.rm = TRUE)
age_logit_min_train <- beta_age_0 + beta_age_1 * age_min_train
age_logit_max_train <- beta_age_0 + beta_age_1 * age_max_train
age_pd_min_train <- exp(age_logit_min_train) / (1 + exp(age_logit_min_train))
age_pd_max_train <- exp(age_logit_max_train) / (1 + exp(age_logit_max_train))
age_pd_observed_low <- min(age_pd_min_train, age_pd_max_train)
age_pd_observed_high <- max(age_pd_min_train, age_pd_max_train)
age_pd_observed_width <- age_pd_observed_high - age_pd_observed_low
aic_age <- AIC(logi_age)
```

The equation and the code correspond directly. In `glm(loan_st ~ age, family = "binomial", data = train)`, the left side of `~` is the observed $Y_i$, the right side is the predictor $\text{age}_i$, and `family = "binomial"` tells R to estimate the logit model. The coefficients $\beta_0$ and $\beta_1$ are the numbers returned by `glm()`, and later `predict(logi_age, newdata = ..., type = "response")` will transform the logit score back into the estimated probability $\hat p_i$.

The estimated model is: $\log\left(\frac{p_i}{1-p_i}\right) = `r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times \text{age}_i$.

The model suggests there is a negative relationship between age and loan status. 

Solving for $p_i$ gives the estimated probability of default according to the age model: $p_i = \frac{\exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times \text{age}_i)}{1 + \exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times \text{age}_i)}$. We now evaluate $p_i$ at two ages.

Substituting $\text{age} = `r age_example_18`$: $p = \frac{\exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times `r age_example_18`)}{1 + \exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times `r age_example_18`)}$,

$p = \frac{\exp(`r fmt_num(age_logit_18, 5)`)}{1 + \exp(`r fmt_num(age_logit_18, 5)`)} = \frac{`r fmt_num(age_exp_18, 4)`}{1 + `r fmt_num(age_exp_18, 4)`} = `r fmt_num(age_pd_18, 4)` = `r fmt_pct(age_pd_18, 2)`$.

Substituting $\text{age} = `r age_example_60`$: $p = \frac{\exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times `r age_example_60`)}{1 + \exp(`r fmt_num(beta_age_0, 5)` `r ifelse(beta_age_1 < 0, "-", "+")` `r fmt_num(abs(beta_age_1), 5)` \times `r age_example_60`)}$,

$p = \frac{\exp(`r fmt_num(age_logit_60, 5)`)}{1 + \exp(`r fmt_num(age_logit_60, 5)`)} = \frac{`r fmt_num(age_exp_60, 4)`}{1 + `r fmt_num(age_exp_60, 4)`} = `r fmt_num(age_pd_60, 4)` = `r fmt_pct(age_pd_60, 2)`$.

The default probability estimates at ages 18 and 60 are very close. The model therefore does not generate much separation in predicted default risk, even when evaluated over a wide age range. This is an early warning that age alone is a weak predictor for this credit scoring problem.

The next figure shows the logistic regression model's output. The training data contain ages from `r fmt_num(age_min_train, 0)` to `r fmt_num(age_max_train, 0)`. The model maps those two endpoints to predicted default probabilities of `r fmt_pct(age_pd_min_train, 2)` and `r fmt_pct(age_pd_max_train, 2)`, respectively. The dotted indicators in the figure mark those two observed age limits and their corresponding probabilities.

```{r}
#| label: fig-scolcam
#| fig-cap: "Sigmoid curve or logistic curve: age model."
# Define the function for the logistic regression model
logistic_function <- function(age) {
  beta0 <- beta_age_0
  beta1 <- beta_age_1
  p <- exp(beta0 + beta1 * age) / (1 + exp(beta0 + beta1 * age))
  return(p)
}

# Generate a sequence of ages
ages <- seq(-1000, 500, by = 1)

# Apply the logistic function to each age
p_values <- sapply(ages, logistic_function)

age_indicator_data <- data.frame(
  age = c(age_min_train, age_max_train),
  p = c(age_pd_min_train, age_pd_max_train),
  label_x = c(age_min_train + 80, age_max_train + 240),
  label_y = c(
    min(age_pd_min_train + 0.08, 0.95),
    min(age_pd_max_train + 0.22, 0.95)
  ),
  label = c(
    paste0("min age = ", fmt_num(age_min_train, 0), "\np = ", fmt_pct(age_pd_min_train, 2)),
    paste0("max age = ", fmt_num(age_max_train, 0), "\np = ", fmt_pct(age_pd_max_train, 2))
  )
)

# Create the plot
library(ggplot2)
ggplot(data = data.frame(age = ages, p = p_values), aes(x = age, y = p)) +
  geom_line(color = "blue", linewidth = 2) +
  geom_segment(
    data = age_indicator_data,
    aes(x = age, xend = age, y = 0, yend = p),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    color = "#C0392B"
  ) +
  geom_segment(
    data = age_indicator_data,
    aes(x = min(ages), xend = age, y = p, yend = p),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.9,
    color = "#C0392B"
  ) +
  geom_segment(
    data = age_indicator_data,
    aes(x = age, xend = label_x, y = p, yend = label_y),
    inherit.aes = FALSE,
    linetype = "dotted",
    linewidth = 0.7,
    color = "#C0392B"
  ) +
  geom_point(
    data = age_indicator_data,
    aes(x = age, y = p),
    inherit.aes = FALSE,
    color = "#C0392B",
    size = 2.8
  ) +
  geom_label(
    data = age_indicator_data,
    aes(x = label_x, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 3.2,
    fill = "white",
    color = "#7B241C",
    label.size = 0.2
  ) +
  labs(x = "Age", y = "Probability (p)") +
  theme_minimal()
```
The logistic curve derived from this model is unreasonable because the probability $p$ changes very slowly with respect to age. The dotted vertical lines show that the realistic age range in the training data is only a small segment of the very wide x-axis needed to display the full sigmoid curve. The dotted horizontal lines show the corresponding predicted probabilities, which move only from `r fmt_pct(age_pd_observed_low, 2)` to `r fmt_pct(age_pd_observed_high, 2)` across the observed age range, a difference of `r fmt_pct(age_pd_observed_width, 2)`. The age-only model therefore does not generate meaningful separation in default probabilities for realistic applicants. To see probabilities close to 0 or 1, the model would require age values that are unrealistic or implausibly extreme.

The AIC value of the age model (`r format(round(aic_age), big.mark = ",")`) is useful when comparing models estimated on the same response variable and sample. The Akaike information criterion (AIC) is a mathematical method for comparing in-sample fit while penalizing model complexity. In statistics, AIC is used to compare different possible models; lower AIC values indicate a better in-sample trade-off between fit and parsimony among the models being compared. At the moment we cannot interpret the AIC very much because we only have one model and we cannot compare it with another AIC.

The next candidate uses the interest rate as the only predictor of `loan_st`. At this stage we are still estimating models on the training set rather than evaluating test-set predictions.

```{r logi_int estimation}
# Build a glm model with variable interest rate as a predictor.
logi_int <- glm(formula = loan_st ~ int, family = "binomial", data = train)
# Print the parameter estimates.
logi_int
```

```{r Interest model teaching values, include=FALSE}
aic_int <- AIC(logi_int)
```

The AIC is lower (`r format(round(aic_int), big.mark = ",")` versus `r format(round(aic_age), big.mark = ",")`), so this model has a better in-sample fit-parsimony trade-off than the age-only model.

Single-predictor models are useful for teaching, but they are too limited for a serious credit score. The `logi_multi` model adds age, interest rate, grade, loan amount, and annual income. We also use `summary()` to inspect more of the estimation output.

```{r logi_multi estimation}
# Multiple variables in a logistic regression model.
logi_multi <- glm(loan_st ~ age + int + grade + log(l_amnt) + 
                  log(income) , family = "binomial", data = train)
# Obtain significance levels using summary().
summary(logi_multi)
```

```{r Multi-factor model teaching values, include=FALSE}
aic_multi <- AIC(logi_multi)
```

The multi-factor model is a stronger candidate than the previous two. In `logi_multi`, the AIC value is the lowest so far (`r format(round(aic_multi), big.mark = ",")` versus `r format(round(aic_int), big.mark = ",")`), so this should be considered the best in-sample model among these candidates at the moment. The `summary()` function shows the significance levels of the estimators, but we are currently more interested in goodness of fit because we want to predict whether new applicants in the test set are expected to default. That is why we focus on AIC at this stage. AIC is still an in-sample criterion, so the model's practical predictive value must be checked later with the test set.

Even if prediction is our main objective, it is useful to understand what the coefficients mean. Logistic regression coefficients are expressed in log-odds units, not directly in probability units. A positive coefficient increases the log-odds of default, while a negative coefficient decreases them. If we exponentiate a coefficient with the `exp()` function, we obtain an odds ratio:

$$
\text{odds ratio}_j = \exp(\beta_j).
$$

An odds ratio above 1 means higher odds of default, and an odds ratio below 1 means lower odds of default. The code below keeps that interpretation but also adds statistical uncertainty. The p-value helps us evaluate whether the coefficient is statistically distinguishable from zero. The 95% interval for the odds ratio helps us see whether the corresponding odds ratio is clearly different from 1.

```{r Odds ratios for the multi-factor model}
multi_summary <- summary(logi_multi)$coefficients

odds_ratios <- data.frame(
  coefficient = multi_summary[, "Estimate"],
  std_error = multi_summary[, "Std. Error"],
  p_value = multi_summary[, "Pr(>|z|)"],
  OR = exp(multi_summary[, "Estimate"]),
  OR_low_95 = exp(multi_summary[, "Estimate"] - 1.96 * multi_summary[, "Std. Error"]),
  OR_high_95 = exp(multi_summary[, "Estimate"] + 1.96 * multi_summary[, "Std. Error"])
)

round(odds_ratios, 4)
```

For practical interpretation, the odds-ratio columns are usually more useful than the raw coefficient column. The coefficient is reported because it is what the logistic model estimates, but it is measured in log-odds units. The OR and its 95% interval translate that estimate into a multiplicative effect on the odds of default.

The interpretation must be careful. For a numerical predictor such as `age`, the odds ratio corresponds to a one-unit increase in age, holding the other variables in the model fixed. For a categorical predictor such as `grade`, each coefficient compares one category with the reference category chosen by R. For transformed variables, the unit is also transformed: a one-unit increase in `log(l_amnt)` is not the same as a one-dollar increase in loan amount. Most importantly, odds ratios are not probability changes. They describe multiplicative changes in the odds of default, while the implied change in probability depends on the starting probability. A final check is uncertainty: if the 95% interval for an odds ratio includes 1, the estimated effect is not statistically clear at the usual 5% level, even if the point estimate is above or below 1.

The next model uses all predictors available in the teaching dataset. We call it `logi_full` because it is the most complete candidate in this chapter. The purpose is to create a richer score before moving to the test-set evaluation, cutoff rules, bad rates, and bank strategy.

```{r logi_full estimation}
# Logistic regression model using all available predictors in the data set.
logi_full <- glm(loan_st ~ age + int + grade + log(l_amnt) +
                   log(income) + emp_len + home + sex +
                   region, family = "binomial", data = train)

aic_full <- AIC(logi_full)
data.frame(
  model = c("logi_age", "logi_int", "logi_multi", "logi_full"),
  AIC = c(aic_age, aic_int, aic_multi, aic_full)
)
```

The AIC table says that `logi_full` has the strongest in-sample fit among the models considered here. That does not settle the problem. It only tells us which candidate fits the training data best after the AIC penalty for complexity. The next section asks a more practical question: when these models face applicants in the test set, do their predicted probabilities support useful lending decisions?

