```{r include=FALSE}
source("R/logistic-packages.R")
```

## Data exploration and cleaning

We begin by loading `loan_data_ARF.rds` and inspecting its structure before any modeling decisions are made. This database is available [here](https://github.com/mlozanoqf/tutorial_arf/blob/main/loan_data_ARF.rds).

```{r Create database, eval=FALSE, include=FALSE}
dat <- readRDS("loan_data_ch1.rds")
# Add sex variable.
# Sex = 1 female; sex = 0 male.
set.seed(1)
sex_1 <- rbinom(n = nrow(dat[dat$loan_st == 1, ]), 
                size = 1, prob = 0.3) # Let's make females default less.
sex_0 <- rbinom(n = nrow(dat[dat$loan_st == 0, ]), 
                size = 1, prob = 0.55) # Let's make females no default more.
sex <- ifelse(dat$loan_st == 1, sex_1, sex_0)
dat$sex <- sex
dat$sex <- as.factor(dat$sex)

# Add region variable.
set.seed(1)
row_def <- nrow(dat[dat$loan_st == 1, ])
row_nodef <- nrow(dat[dat$loan_st == 0, ])

def_region_index <- sample(c("N", "E", "W", "S"), size = row_def, 
                           replace  = TRUE, prob = c(0.1, 0.2, 0.3, 0.4))
nodef_region_index <- sample(c("N", "E", "W", "S"), size = row_nodef, 
                           replace  = TRUE, prob = c(0.4, 0.3, 0.2, 0.1))

region <- ifelse(dat$loan_st == 1, 
                 def_region_index, nodef_region_index)
dat$region <- region
dat$region <- as.factor(dat$region)

# Imputation.
index_NA <- which(is.na(dat$emp_length))
dat$emp_length[index_NA] <- median(dat$emp_length, na.rm = TRUE)

index_NA <- which(is.na(dat$int_rate))
dat$int_rate[index_NA] <- median(dat$int_rate, na.rm = TRUE)

# New database
saveRDS(dat, "loan_data_ARF.rds")
```

```{r Load the modified database}
dat <- readRDS("loan_data_ARF.rds")
initial_n <- nrow(dat)
initial_p <- ncol(dat)
str(dat)
```


This dataset represents the type of loan-level information a financial institution could use before making or reviewing a credit decision. It contains `r format(initial_n, big.mark = ",")` observations across `r initial_p` variables. Each observation corresponds to the personal and loan characteristics of an individual loan. 

The dependent variable is `loan_st`, which indicates loan status. A value of 0 represents no default, while a value of 1 represents default. A default occurs when a borrower fails to make timely payments, misses payments, or ceases payments on the interest or principal owed. The definition of default can vary depending on the goals and interests of the analysis. For the purposes of this study, we classify loans simply as either default or no default. 

The variable `loan_st` is dichotomous, or categorical. Our primary interest lies in predicting whether a new loan application will result in default or not.

`loan_data_ARF.rds` contains historical information because the repayment outcome is already known. Historical data is valuable for understanding how borrower and loan characteristics relate to default. It is also the basis for training quantitative models and evaluating how those models behave on new applicants.

Several variable names can be shortened without losing their meaning. We rename them before continuing.

```{r}
old_names <- colnames(dat)
colnames(dat) <- c("loan_st", "l_amnt", "int", "grade", "emp_len", 
                   "home", "income", "age", "sex", "region")
data.frame(old_names, "new_names" = colnames(dat))
```

The dataset now has shorter variable names. 

The first 10 rows give a quick view of the variables and their coding.

```{r First 10 rows of loan data}
head(dat, 10)
```

Note that `sex` is 1 for female and 0 for male. Instead of looking row by row, we can summarize a categorical variable such as `home` with the `CrossTable()` function.

```{r CrossTable as a way to summarize data}
CrossTable(dat$home)
```

`CrossTable()` can also summarize two variables at once. Adding `loan_st` to the home-ownership table lets us compare default behavior across home-ownership categories.

```{r CrossTable using two dimensions}
CrossTable(dat$home, dat$loan_st, prop.r = TRUE,
           prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE)
```

The table reports defaults by home ownership. For numerical variables, plots are often easier to read than raw tables. The next figure shows the interest-rate distribution.

```{r}
#| label: fig-irh
#| fig-cap: "Interest rate histogram."
ggplot(dat, aes(x = int)) + 
  geom_histogram(aes(y=..density..), binwidth = 0.5, colour = "black", 
                 fill = "white") +
  labs(y = "Density", x = "Interest rate") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

The next dotplot shows annual income.

```{r}
#| label: fig-aib
#| fig-cap: "Annual income dotplot."
ggplot(dat, aes(income)) + 
  geom_dotplot(binwidth = 100000, fill = "blue") +
  labs(y = "Density",
       x = "Annual income") +
  theme(legend.position = "bottom", legend.title = element_blank())
```
The @fig-aib above appears suspicious. The horizontal axis shows a very large annual income value (`r format(max(dat$income), big.mark = ",")`). Additionally, there are only a few individuals with extremely high incomes. We should investigate further to determine whether these observations should remain in the analysis or whether they distort the visual and statistical summary of the rest of the sample.

```{r Extract the row of the high income individuals}
high_income <- dat[(dat$income > 1000000), ]
high_income
```

One individual (ID `r rownames(high_income)[which.max(high_income$age)]`) combines an extremely high income with an age of `r max(high_income$age)`. The age value is not credible, and the income values are extreme relative to the rest of the sample. In this book we continue with an analysis without these extreme observations. This is a discretionary cleaning criterion rather than a universal rule: in an applied project, we would document the decision and usually compare results with and without these observations.

```{r}
high_income_index <- data.frame(value = as.integer(rownames(high_income)))
dat <- dat[-high_income_index$value,]
clean_n <- nrow(dat)
```

The original dataset contains `r format(initial_n, big.mark = ",")` rows. After removing `r nrow(high_income_index)` observations, the cleaned dataset contains `r format(clean_n, big.mark = ",")` rows.

```{r}
#| label: fig-aiwev
#| fig-cap: "Annual income without extreme values."
ggplot(dat, aes(income)) + 
  geom_dotplot(binwidth = 10000, fill = "blue") +
  labs(y = "Density", x = "Annual income") +
  theme(legend.position = "bottom", legend.title = element_blank())
```

The revised plot is easier to read because the extreme observations no longer compress the main income distribution.

We also inspect summary statistics by `sex`.

```{r Gender issues}
summary_table <- dat |>
  group_by(sex) |>
  summarize(
    Mean = mean(income),
    Median = median(income),
    Min = min(income),
    Max = max(income),
    Count = n()
  ) |>
  mutate(
    Gender = case_when(
      sex == 0 ~ "Male",
      sex == 1 ~ "Female"
    )
 )
  
summary_table
```

The table shows that, in this dataset, observations labeled as female have slightly higher average and median incomes compared to observations labeled as male. The income distribution for females also appears more variable, with a wider range of incomes. This summary must be interpreted carefully: the `sex` variable was added for teaching purposes and should not be treated as evidence about gender and income in a real population. In applied credit risk work, variables related to protected characteristics, and variables that may proxy for them, require explicit legal, ethical, and fairness review before they are used in modeling or decision rules.

The exploration above gives us enough context to proceed. Data cleaning can continue indefinitely, so an applied project has to balance inspection with the modeling objective. We have checked the outcome definition, simplified variable names, removed implausible extreme income observations, and reviewed a few basic summaries. The next section uses the cleaned data to estimate logistic credit-scoring models.

