## Risk-neutral simulations and equity payoff

```{r merton-inside-view-setup, include=FALSE}
source("R/merton-common.R")
```

We now return to the main Merton example and simulate the daily evolution of the market value of the firm's assets over one year. The simulation logic follows the geometric Brownian motion framework used in [@Hull], Chapters 14 and 15. Because this section is connected to the valuation formula, the simulations use the risk-neutral drift $rf$. These are valuation scenarios under the same risk-neutral world used to value the equity option. They are useful for visualizing the model mechanics, not for forecasting the firm's real-world asset path.

```{r}
#| label: fig-so10pota
#| fig-cap: "Ten risk-neutral valuation paths for the firm's assets."
# For valuation, Merton assumes V follows a risk-neutral GBM process.
V0.sim <- function(i) {  # the argument i is not used here.
  GBM(N = 365, T = 1, t0 = 0, x0 = V0, theta = rf, sigma = sv) }
set.seed(3) # Reproducibility  
paths <- sapply(1:10, V0.sim) # Create 10 paths for V.
final_row <- nrow(paths)
path_time_10 <- seq(0, final_row - 1)
path_data_10 <- data.frame(
  time_day = rep(path_time_10, times = ncol(paths)),
  path_id = factor(rep(seq_len(ncol(paths)), each = nrow(paths))),
  asset_value = as.vector(paths)
)
terminal_path_data_10 <- data.frame(
  time_day = max(path_time_10),
  asset_value = paths[final_row, ],
  terminal_state = ifelse(paths[final_row, ] < D, "Default", "No default")
)
initial_path_data_10 <- data.frame(
  time_day = 0,
  asset_value = V0
)

ggplot2::ggplot(
  path_data_10,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.45, color = "black") +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = initial_path_data_10,
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::geom_point(
    data = terminal_path_data_10,
    ggplot2::aes(x = time_day, y = asset_value, color = terminal_state),
    inherit.aes = FALSE,
    shape = 1,
    size = 3,
    stroke = 1.1
  ) +
  ggplot2::scale_color_manual(
    values = c("Default" = "red", "No default" = "green"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-so10pota should be read as follows. Each black curve is one risk-neutral path for the firm's asset value, $V_t$. The horizontal red line is the promised debt payment, $D$. The blue point marks the starting value, $V_0$. The open circles at maturity classify the terminal values: green means $V_T>D$ and red means $V_T<D$.

These simulations can be interpreted as 10 risk-neutral valuation scenarios for the firm's assets. They are paths generated under the risk-neutral measure, so the drift is the risk-free rate and not the firm's real-world expected asset return. The formal name is 10 simulated geometric Brownian motion paths under the risk-neutral measure.

Before counting defaults at maturity, let's first count the paths in which $V_t<D$ at least once before or at maturity. Here I use $t$, not $T$, because this first count asks whether a path ever crosses below the promised debt payment during the year. That is different from Merton default, which is determined by whether $V_T<D$ at maturity.

```{r}
# Which path went lower than D?
paths_below_D <- which(colSums(paths < D) > 0)
paths_below_D
```
The value of the assets was lower than the promised debt payment $D$ at some time in risk-neutral scenarios `r paste(paths_below_D, collapse = ", ")`. The next figure isolates those cases. Red paths finish below $D$ and therefore default at maturity. The blue path crosses below $D$ during the year but finishes above $D$.

```{r}
#| label: fig-nttbpfdnd
#| fig-cap: "Paths that cross below D before maturity; only paths ending below D default."
make_path_data <- function(path_matrix, path_ids = seq_len(ncol(path_matrix))) {
  path_matrix <- as.matrix(path_matrix)
  data.frame(
    time_day = rep(seq(0, nrow(path_matrix) - 1), times = ncol(path_matrix)),
    path_id = factor(rep(path_ids, each = nrow(path_matrix)), levels = path_ids),
    asset_value = as.vector(path_matrix)
  )
}

paths_below_D_matrix <- paths[, paths_below_D, drop = FALSE]
paths_below_D_data <- make_path_data(paths_below_D_matrix, paths_below_D)
paths_below_D_data$terminal_state <- rep(
  ifelse(
    paths_below_D_matrix[final_row, ] < D,
    "Default at maturity",
    "Crosses D before maturity, no default"
  ),
  each = nrow(paths_below_D_matrix)
)

ggplot2::ggplot(
  paths_below_D_data,
  ggplot2::aes(
    x = time_day,
    y = asset_value,
    group = path_id,
    color = terminal_state
  )
) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Crosses D before maturity, no default" = "blue",
      "Default at maturity" = "red"
    ),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-nttbpfdnd isolates the paths from the 10-scenario simulation that cross $D$ at least once. The blue path is the key teaching case. It enters the region below the red debt threshold before maturity, then recovers and ends above $D$ at $T$. In the basic Merton model, default is determined by the terminal condition $V_T<D$. In this small risk-neutral simulation, that happens in `r sum(paths[final_row, ] < D)` cases. According to these 10 risk-neutral scenarios, the simulated probability of default is `r fmt_pct(sum(paths[final_row, ] < D) / ncol(paths), 0)`. The code below counts the cases in which $V_T<D$.

```{r}
sim_pd_10 <- sum(paths[final_row, ] < D) / ncol(paths)
knitr::kable(
  data.frame(
    Quantity = c(
      "Simulated default paths",
      "Total simulated paths",
      "Simulated risk-neutral PD",
      "Analytical Merton PD"
    ),
    Value = c(
      fmt_int(sum(paths[final_row, ] < D)),
      fmt_int(ncol(paths)),
      fmt_pct(sim_pd_10, 0),
      fmt_pct(pd_merton, 4)
    )
  ),
  caption = "Default count and simulated risk-neutral PD with 10 paths.",
  row.names = FALSE
)
```

The simulated risk-neutral probability of default of `r fmt_pct(sim_pd_10, 0)` looks high relative to the model value of `r fmt_pct(pd_merton, 4)`. This happens because the number of simulations is small. The following examples show how these values tend to converge as we move from 10 to 100 simulated paths, and then from 100 to 1,000, 10,000, and 100,000 simulated terminal values.

The next simulation uses 100 paths from $V_0$ to $V_T$. The visual coding is the same as before: gray curves are risk-neutral asset paths, the red horizontal line is $D$, and the blue point is the common starting value $V_0$.

```{r}
#| label: fig-so100pota
#| fig-cap: "One hundred risk-neutral valuation paths for the firm's assets."
set.seed(3) # Reproducibility  
paths <- sapply(1:100, V0.sim) # Create 100 paths.
final_row <- nrow(paths)
path_data_100 <- make_path_data(paths)

ggplot2::ggplot(
  path_data_100,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.25, color = "gray35", alpha = 0.55) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

```{r}
sim_pd_100 <- sum(paths[final_row, ] < D) / ncol(paths)
knitr::kable(
  data.frame(
    Quantity = c(
      "Simulated default paths",
      "Total simulated paths",
      "Simulated risk-neutral PD",
      "Analytical Merton PD"
    ),
    Value = c(
      fmt_int(sum(paths[final_row, ] < D)),
      fmt_int(ncol(paths)),
      fmt_pct(sim_pd_100, 0),
      fmt_pct(pd_merton, 4)
    )
  ),
  caption = "Default count and simulated risk-neutral PD with 100 paths.",
  row.names = FALSE
)
```

With 100 paths, individual trajectories are less important than the terminal count. The simulated risk-neutral probability of default is `r fmt_pct(sim_pd_100, 0)`. This is closer to the model value of `r fmt_pct(pd_merton, 4)`.

The next figure keeps only the `r sum(paths[final_row, ] < D)` paths that end in default.

```{r}
#| label: fig-dc
#| fig-cap: "Paths that end below the promised debt payment at maturity."
default_cases_100 <- which(paths[final_row, ] < D)
default_path_data_100 <- make_path_data(
  paths[, default_cases_100, drop = FALSE],
  default_cases_100
)

ggplot2::ggplot(
  default_path_data_100,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.45, color = "gray25", alpha = 0.75) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-dc filters the 100 simulated paths and keeps only the paths with $V_T<D$. These are the default cases under the Merton terminal-default rule. The red line is still $D$, so the important feature is the terminal position of each path relative to that line. Earlier crossings matter for intuition, but the basic Merton default event is decided at maturity.

```{r include=FALSE}
almost_default <- which(colSums(paths < D) > 0)
default <- which(paths[final_row, ] < D)
almost_default_no_default <- setdiff(almost_default, default)
```

The next figure keeps the `r length(almost_default_no_default)` cases in which $V_t$ went below $D$ at some point, but $V_T$ finished above $D$, so the firm did not default at maturity.

```{r}
#| label: fig-atpwl1asp
#| fig-cap: "These paths cross below $D$ before maturity, but finish with $V_T$ above $D$ and therefore do not default."
almost_default_no_default_data <- make_path_data(
  paths[, almost_default_no_default, drop = FALSE],
  almost_default_no_default
)
almost_default_colors <- grDevices::hcl.colors(
  length(almost_default_no_default),
  "Dark 3"
)
names(almost_default_colors) <- levels(almost_default_no_default_data$path_id)

ggplot2::ggplot(
  almost_default_no_default_data,
  ggplot2::aes(
    x = time_day,
    y = asset_value,
    group = path_id,
    color = path_id
  )
) +
  ggplot2::geom_line(linewidth = 0.55, alpha = 0.9) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::scale_color_manual(values = almost_default_colors, guide = "none") +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-atpwl1asp shows the other side of the crossing story. Every colored path falls below $D$ at least once, but each one finishes above $D$ at maturity. The different colors only separate the paths visually. The financial message is that the path history helps us see the simulated movement of assets, while the basic Merton default event is decided at $T$.

We can also simulate terminal asset values directly from the risk-neutral distribution of $V_T$. The resulting distribution is log-normal. This is useful because the default event depends only on whether the terminal value is below $D$.

The next figure shows how the simulated risk-neutral probability of default changes as the number of simulated terminal values increases. The red bars are simulated terminal values with $V_T<D$. The blue bars are simulated terminal values with $V_T\ge D$. The dashed vertical line is the promised debt payment $D$.

```{r}
#| label: fig-vt-convergence-hists
#| fig-cap: "Convergence of the simulated risk-neutral probability of default to the analytical Merton probability."
#| fig-width: 11
#| fig-height: 4.2
simulate_terminal_assets <- function(n) {
  exp(rnorm(n, log(V0) + (rf - (sv^2) / 2) * TT, sv * sqrt(TT)))
}

set.seed(13)
convergence_sizes <- c(100, 1000, 10000)
convergence_labels <- paste0(
  "N = ",
  format(convergence_sizes, big.mark = ",", scientific = FALSE)
)
VT_convergence <- lapply(convergence_sizes, simulate_terminal_assets)
names(VT_convergence) <- convergence_labels

vt_convergence_data <- do.call(
  rbind,
  lapply(names(VT_convergence), function(sample_label) {
    values <- VT_convergence[[sample_label]]
    data.frame(
      sample_label = sample_label,
      terminal_asset_value = values,
      region = ifelse(values < D, "Default region", "No default region")
    )
  })
)
vt_convergence_data$sample_label <- factor(
  vt_convergence_data$sample_label,
  levels = convergence_labels
)

convergence_pd_data <- do.call(
  rbind,
  lapply(names(VT_convergence), function(sample_label) {
    values <- VT_convergence[[sample_label]]
    data.frame(
      sample_label = sample_label,
      x = 17.6,
      y = 380,
      label = paste0(
        "Sim: ", fmt_pct(mean(values < D), 1), "\n",
        "Merton: ", fmt_pct(pd_merton, 1)
      )
    )
  })
)
convergence_pd_data$sample_label <- factor(
  convergence_pd_data$sample_label,
  levels = convergence_labels
)

ggplot2::ggplot(
  vt_convergence_data,
  ggplot2::aes(x = terminal_asset_value, fill = region)
) +
  ggplot2::geom_histogram(
    binwidth = 0.25,
    boundary = D,
    color = "white",
    linewidth = 0.1
  ) +
  ggplot2::geom_vline(xintercept = D, linetype = "dashed", linewidth = 0.65) +
  ggplot2::geom_text(
    data = convergence_pd_data,
    ggplot2::aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 2.9,
    lineheight = 0.95
  ) +
  ggplot2::facet_wrap(~sample_label, ncol = 3) +
  ggplot2::coord_cartesian(xlim = c(4, 30), ylim = c(0, 400), expand = FALSE) +
  ggplot2::scale_fill_manual(
    values = c("Default region" = "red", "No default region" = "blue"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Value of the assets at maturity",
    y = "Frequency"
  ) +
  ggplot2::theme_classic(base_size = 12) +
  ggplot2::theme(
    strip.background = ggplot2::element_blank(),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

Figure @fig-vt-convergence-hists makes the sampling error visible. Each panel answers the same question: what fraction of simulated terminal asset values falls below the promised debt payment $D$? With 100 simulated terminal values, the simulated probability can be noticeably away from the analytical Merton value. With 1,000 and then 10,000 simulated terminal values, the red area becomes a more stable approximation of the analytical risk-neutral probability $N(-d_2)$.

The next figure repeats the same terminal-value simulation with 100,000 draws. It uses the same visual coding as @fig-vt-convergence-hists: red bars for default states, blue bars for non-default states, and the dashed vertical line for $D$. The larger sample makes the simulated red area a close numerical counterpart to the analytical Merton probability.

```{r}
#| label: fig-ho1votaam
#| fig-cap: "Histogram of 100,000 values of the assets at maturity."
set.seed(13)
# 100,000 values of VT at once.
VT <- exp(rnorm(100000, log(V0) + (rf - (sv^2) / 2)*TT, sv*sqrt(TT)))
# Plot results.
h <- hist(VT, 100, plot = FALSE)
hist_data <- data.frame(
  xmin = head(h$breaks, -1),
  xmax = tail(h$breaks, -1),
  count = h$counts
)
hist_data$region <- ifelse(hist_data$xmin < D, "Default region", "No default region")
sim_pd_100k <- mean(VT < D)
hist_label_x <- 18.4
hist_label_y <- max(hist_data$count) * 0.92
hist_label_gap <- max(hist_data$count) * 0.075
hist_merton_label <- paste0(
  "paste('Merton ', N(-d[2]), ' = ", fmt_pct(pd_merton, 4), "')"
)
ggplot2::ggplot(hist_data) +
  ggplot2::geom_rect(
    ggplot2::aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = count,
      fill = region
    ),
    color = "white",
    linewidth = 0.1
  ) +
  ggplot2::geom_vline(xintercept = D, linetype = "dashed", linewidth = 0.7) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y,
    hjust = 0,
    vjust = 1,
    label = paste0("Red area = ", fmt_pct(sim_pd_100k, 3)),
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = "Simulated risk-neutral PD",
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - 2 * hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = hist_merton_label,
    parse = TRUE,
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - 3 * hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = "The two values are close.",
    size = 3.2
  ) +
  ggplot2::coord_cartesian(xlim = c(4, 30), expand = FALSE) +
  ggplot2::scale_fill_manual(
    values = c("Default region" = "red", "No default region" = "blue"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Value of the assets at maturity",
    y = "Frequency"
  ) +
  ggplot2::theme_classic(base_size = 12)
```

The simulated risk-neutral probability of default can be calculated as:

```{r}
knitr::kable(
  data.frame(
    Quantity = c(
      "Simulated terminal asset values",
      "Terminal values below D",
      "Simulated risk-neutral PD",
      "Analytical Merton PD"
    ),
    Value = c(
      fmt_int(length(VT)),
      fmt_int(sum(VT < D)),
      fmt_pct(sim_pd_100k, 4),
      fmt_pct(pd_merton, 4)
    )
  ),
  caption = "Large-sample simulated risk-neutral PD compared with the analytical Merton PD.",
  row.names = FALSE
)
```

Figure @fig-ho1votaam is the large-sample continuation of @fig-vt-convergence-hists. The simulated value `r fmt_num(sim_pd_100k, 5)` is now very close to the model value `r fmt_num(pd_merton, 7)`. The red area in the histogram is therefore the simulation-based approximation of the model-implied risk-neutral probability $N(-d_2)$.

## Equity as a call option payoff

The previous simulation counted default states by comparing the terminal asset value $V_T$ with the promised debt payment $D$. The same terminal comparison also explains why Merton values equity with a call-option formula.

The phrase residual claimant has a concrete meaning here. In the one-period Merton setup, the firm has one promised debt payment due at maturity. At $T$, the firm's assets are valued at their market value $V_T$, and payments follow priority. Debt holders are senior claimants: they are paid first, up to the promised amount $D$. Equity holders are junior claimants: they receive only what remains after debt has been paid. Therefore:

$$
\text{Debt payoff at }T=\min(V_T,D),
$$

and:

$$
E_T=\max(V_T-D,0).
$$

If $V_T>D$, creditors receive $D$ and shareholders receive the surplus $V_T-D$. If $V_T<D$, the firm does not have enough market value to repay the promised debt payment in full. Creditors receive the available asset value, shareholders receive zero, and the firm is in default.

This is a market-value claims view, not an accounting balance sheet. The model is not using book assets from financial statements at maturity. It is describing how the market value of the firm's assets is split between senior debt and junior equity claims.

A European call option has terminal payoff:

$$
c_T=\max(S_T-K,0).
$$

Merton equity has the same payoff shape:

$$
E_T=\max(V_T-D,0).
$$

The mapping is direct. The asset value $V_T$ plays the role of the underlying price $S_T$, and the promised debt payment $D$ plays the role of the strike price $K$.

| European call option | Merton market-value claim |
|---|---|
| Underlying price at maturity, $S_T$ | Market value of assets at maturity, $V_T$ |
| Strike price, $K$ | Promised debt payment due at maturity, $D$ |
| Call is in the money if $S_T>K$ | Equity receives a positive residual if $V_T>D$ |
| Call expires worthless if $S_T<K$ | Equity is zero and the firm defaults if $V_T<D$ |
| Call payoff, $\max(S_T-K,0)$ | Equity payoff, $\max(V_T-D,0)$ |

This mapping is useful because it turns the credit problem into a contingent-claim problem. Default is the out-of-the-money region of equity. Survival is the in-the-money region. The debt payment $D$ becomes the default boundary. The option formula in section 3.2 therefore belongs in the credit-risk argument. It is the pricing formula for the equity claim once equity is understood as a call on the firm's assets. Observed equity value and equity volatility can then be used to infer the unobserved asset value and asset volatility, and those asset quantities determine the model-implied probability of default.

It also clarifies what is being valued. The call formula values the equity claim. The whole-firm market asset value is the broader object. In the Hull/Merton example, section 3.2 inferred a market asset value of $V_0=`r fmt_num(V0, 4)`$. The observed equity value is $E_0=`r fmt_num(E0, 0)`$, so the model-implied market value of debt is $B_0=V_0-E_0=`r fmt_num(market_value_debt, 4)`$. A buyer of the whole firm as a package of market-value assets would look at $V_0$. A buyer of only the shares would be buying the junior residual claim worth $E_0$. The difference between those two values is the debt claim.

This section reads the maturity payoff. The current equity value $E_0$ was already used in section 3.2 to calibrate $V_0$ and $\sigma_V$ through Hull's option-pricing equation. The figure below makes the payoff shape visible.

```{r}
#| label: fig-avotfbssitf
#| fig-cap: "Equity at maturity as a European call payoff on the firm's assets."
ET <- pmax(VT - D, 0) # payoff function of a typical call option.
payoff_data <- data.frame(
  terminal_asset_value = sort(VT),
  terminal_equity_value = sort(ET)
)

ggplot2::ggplot(
  payoff_data,
  ggplot2::aes(x = terminal_asset_value, y = terminal_equity_value)
) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.35, color = "gray78") +
  ggplot2::geom_line(linewidth = 1.25, color = "#0B5CAD") +
  ggplot2::geom_vline(xintercept = D, linetype = "dashed", linewidth = 0.7) +
  ggplot2::annotate(
    "label",
    x = 6.1,
    y = 4.4,
    hjust = 0,
    label = paste0(
      "Equity payoff is zero\n",
      "in ", fmt_pct(sim_pd_100k, 3), " of the\n",
      "100,000 cases."
    ),
    label.size = 0,
    linewidth = 0,
    fill = "white",
    alpha = 0.9,
    size = 3.8
  ) +
  ggplot2::coord_cartesian(xlim = c(5, 20), ylim = c(-0.55, 10), expand = FALSE) +
  ggplot2::labs(
    x = expression(paste("Simulated assets at maturity (", V[T], ")")),
    y = expression(paste("Simulated equity at maturity (", E[T], ")"))
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-avotfbssitf shows the payoff split at maturity. The flat part to the left of $D$ is the limited-liability region: asset value is too low to repay the promised debt payment, so equity is worth zero. To the right of $D$, equity increases one-for-one with assets because every additional unit above the promised debt payment belongs to shareholders.

This payoff relation, $E_T=\max(V_T-D, 0)$, is the key point of the inside view. Equation 24.3 in [@Hull] uses the same call-option logic at time 0. It chooses $V_0$ and $\sigma_V$ so that the model is consistent with the observed equity value $E_0$ and the observed equity volatility $\sigma_E$.

The important distinction is therefore: $E_T$ is the future state-dependent equity payoff, $E_0$ is the observed equity value today, $V_0$ is the model's market value of assets today, and $D$ is the promised debt payment at maturity. The estimate $V_0$ can be compared with the book value of total assets to understand the difference between book values and market values. The market value of debt is another model-implied quantity, and it is different from the promised debt payment $D$.

