## Model-implied PD sensitivity analyses

```{r merton-pd-parameters-setup, include=FALSE}
source("R/merton-pd-function.R")
```

The previous section gives us a convenient function: $pd = f(E_0, \sigma_E, rf, T, D)$. These five observable inputs are used to estimate $V_0$ and $\sigma_V$ before calculating $pd$.

Throughout this section, $pd$ means the Merton risk-neutral/implied probability of default. The exercises below are model-implied sensitivity analyses. We change one or two observable inputs, keep the remaining observable inputs and model assumptions fixed, solve again for the unobservable $V_0$ and $\sigma_V$, and then recompute $N(-d_2)$ under the risk-neutral distribution. The asset quantities $V_0$ and $\sigma_V$ are therefore part of the new Merton solution at each point.

```{r}
# Evaluate the case of the textbook example.
pd1 <- pd(3, 0.8, 0.05, 1, 10)
pd1
```

The code below creates input grids and evaluates the `pd` function across those grids.

```{r}
l = 50 # Vectors of length 50.
# Create vectors of the parameters.
x.E <- seq(from = 1, to = 20, length.out = l)
x.rf <- seq(0, 0.20, length.out = l)
x.D <- seq(1, 20, length.out = l)
x.T <- seq(0.5, 20, length.out = l)
x.se <- seq(0.01, 3, length.out = l)
# Evaluate the pd at different values.
pd.E <- mapply(pd, x.E, se, rf, TT, D)
pd.rf <- mapply(pd, E0, se, x.rf, TT, D)
pd.D <- mapply(pd, E0, se, rf, TT, x.D)
pd.T <- mapply(pd, E0, se, rf, x.T, D)
pd.se <- mapply(pd, E0, x.se, rf, TT, D)

plot_pd_sensitivity <- function(x, y, current_x, x_label,
                                y_limits = NULL, x_as_percent = FALSE,
                                y_digits = 0) {
  plot_data <- data.frame(
    input_value = x,
    risk_neutral_pd = y
  )
  current_point <- data.frame(
    input_value = current_x,
    risk_neutral_pd = pd1
  )
  if (is.null(y_limits)) {
    y_limits <- c(0, max(c(y, pd1), na.rm = TRUE) * 1.06)
  }

  p <- ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = input_value, y = risk_neutral_pd)
  ) +
    ggplot2::geom_hline(
      yintercept = 0,
      linewidth = 0.35,
      color = "gray82"
    ) +
    ggplot2::geom_line(linewidth = 1.1, color = "#0b5cad") +
    ggplot2::geom_vline(
      xintercept = current_x,
      linetype = "dashed",
      linewidth = 0.55,
      color = "gray35"
    ) +
    ggplot2::geom_hline(
      yintercept = pd1,
      linetype = "dashed",
      linewidth = 0.55,
      color = "gray35"
    ) +
    ggplot2::geom_point(
      data = current_point,
      ggplot2::aes(x = input_value, y = risk_neutral_pd),
      shape = 21,
      size = 3.6,
      stroke = 1.1,
      color = "#b84300",
      fill = "white"
    ) +
    ggplot2::coord_cartesian(ylim = y_limits) +
    ggplot2::scale_y_continuous(labels = function(x) fmt_pct(x, y_digits)) +
    ggplot2::labs(x = x_label, y = "Risk-neutral PD") +
    ggplot2::theme_classic(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_line(
        color = "gray90",
        linewidth = 0.25
      ),
      panel.grid.minor = ggplot2::element_blank()
    )

  if (x_as_percent) {
    p <- p + ggplot2::scale_x_continuous(labels = function(x) fmt_pct(x, 0))
  }

  p
}
```

The risk-free-rate grid ranges from 0% to 20%. This keeps the plot focused on a moderate range around the textbook example, rather than on extreme interest-rate environments.

::: {.merton-subhead}
One-input sensitivity curves
:::

The first curve asks how the risk-neutral PD changes as the observed equity value $E_0$ changes. The code evaluates the probability-of-default function 50 times, once for each value of $E_0$ from 1 to 20. At each value of $E_0$, the code re-estimates $V_0$ and $\sigma_V$ and then computes a new risk-neutral probability of default.

```{r}
#| label: fig-tpodaafoteaatz
#| fig-cap: "The risk-neutral probability of default as a function of the equity at time zero."
plot_pd_sensitivity(
  x = x.E,
  y = pd.E,
  current_x = E0,
  x_label = "Equity at time zero",
  y_limits = c(0, 0.15)
)
```

These relationships between the observable inputs and the risk-neutral probability of default are nonlinear. The current level of each input matters because the same absolute change can have a high or low impact depending on where the firm starts.

```{r}
#| label: fig-tpodaafotrfr
#| fig-cap: "The risk-neutral probability of default as a function of the risk-free rate."
plot_pd_sensitivity(
  x = x.rf,
  y = pd.rf,
  current_x = rf,
  x_label = "Risk-free rate",
  y_limits = range(pd.rf) + c(-0.0006, 0.0006),
  x_as_percent = TRUE,
  y_digits = 1
)
```

The next curve changes the promised debt payment due at maturity, $D$. Increasing or decreasing $D$ by 1 unit can have a different impact on the model-implied probability of default because the curve is nonlinear.

```{r}
#| label: fig-tpodaafotd
#| fig-cap: "The risk-neutral probability of default as a function of the promised debt payment due at maturity."
plot_pd_sensitivity(
  x = x.D,
  y = pd.D,
  current_x = D,
  x_label = "Promised debt payment at maturity",
  y_limits = c(0, 0.15)
)
```

The maturity curve requires a careful reading. In this model-implied sensitivity exercise, changing maturity also triggers a new solution for $V_0$ and $\sigma_V$. For each maturity value on the horizontal axis, the remaining observable inputs are kept fixed, the unobserved asset quantities are re-estimated, and the risk-neutral PD is recomputed. This is a model comparison, not a standalone recommendation about debt negotiation. A real negotiation would also depend on cash flows, covenants, recovery, refinancing risk, and the bank's required spread.

```{r}
#| label: fig-tpodaafotdm
#| fig-cap: "The risk-neutral probability of default as a function of debt maturity."
plot_pd_sensitivity(
  x = x.T,
  y = pd.T,
  current_x = TT,
  x_label = "Maturity"
)
```

The last one-input curve changes the observed equity volatility $\sigma_E$.

```{r}
#| label: fig-tpodaafote
#| fig-cap: "The risk-neutral probability of default as a function of the volatility of the equity."
plot_pd_sensitivity(
  x = x.se,
  y = pd.se,
  current_x = se,
  x_label = "Equity volatility",
  x_as_percent = TRUE
)
```


::: {.merton-subhead}
Two-input views
:::

The previous plots changed one observable input at a time: $E_0$, $\sigma_E$, $rf$, $T$ or $D$. We were able to do that because we constructed a vector of 50 different values for these inputs. Again, $V_0$ and $\sigma_V$ do not remain fixed because they are inferred from $E_0$, $\sigma_E$, $rf$, $T$ and $D$ each time the function is evaluated.

The two-input view lets two market inputs move at the same time: $E_0$ and $\sigma_E$. This is financially useful because equity value measures how much market cushion the firm has above the promised debt payment, while equity volatility measures how unstable that cushion is. A firm with low equity value and high equity volatility should imply a higher probability of default than a firm with high equity value and low equity volatility.

To make the comparison complete, we evaluate all combinations of 50 values of $E_0$ and 50 values of $\sigma_E$. For this two-input view, we cap $E_0$ at 10 and $\sigma_E$ at 150% so the relevant region around the original example remains readable. The result is a $50 \times 50$ grid. Each cell answers a concrete question: if the observed equity value were this level and the observed equity volatility were that level, what Merton risk-neutral PD would the model imply?

```{r}
x.E_two_input <- seq(from = 1, to = 10, length.out = l)
x.se_two_input <- seq(from = 0.01, to = 1.5, length.out = l)
# Create the empty matrix.
p_E_se <- matrix(0, nrow = l, ncol = l)
# Fill the empty matrix with probability of default values.
for(i in 1 : l){ # Is there an easier way to do this?
  for(j in 1 : l){
    p_E_se[i,j] <- mapply(pd, x.E_two_input[i], x.se_two_input[j], rf, TT, D) } }
```

The grid is easiest to read as a contour plot. Each contour line connects combinations of $E_0$ and $\sigma_E$ that produce the same risk-neutral PD after the model solves again for $V_0$ and $\sigma_V$.

```{r}
#| label: fig-tpodaafoteatzatvoeacv
#| fig-cap: "A contour view of the risk-neutral probability of default as a function of equity value and equity volatility."
pd_grid <- expand.grid(
  equity_at_time_zero = x.E_two_input,
  equity_volatility = x.se_two_input
)
pd_grid$risk_neutral_pd <- as.vector(p_E_se)
current_pd_point <- data.frame(
  equity_at_time_zero = E0,
  equity_volatility = se,
  risk_neutral_pd = pd1
)
pd_contour_breaks <- seq(0.05, 0.95, by = 0.05)
pd_contour_label_levels <- c(0.05, 0.10, 0.15, 0.20, 0.40)
pd_contour_label_lines <- contourLines(
  x = x.E_two_input,
  y = x.se_two_input,
  z = p_E_se,
  levels = pd_contour_label_levels
)
pd_contour_label_list <- lapply(pd_contour_label_levels, function(level_value) {
  level_lines <- pd_contour_label_lines[
    vapply(
      pd_contour_label_lines,
      function(line) isTRUE(all.equal(line$level, level_value)),
      logical(1)
    )
  ]
  if (length(level_lines) == 0) {
    return(NULL)
  }
  longest_line <- level_lines[[which.max(vapply(level_lines, function(line) {
    length(line$x)
  }, numeric(1)))]]
  label_index <- max(1, round(length(longest_line$x) * 0.68))
  data.frame(
    equity_at_time_zero = longest_line$x[label_index],
    equity_volatility = longest_line$y[label_index],
    risk_neutral_pd = level_value
  )
})
pd_contour_label_data <- do.call(rbind, pd_contour_label_list)
if (is.null(pd_contour_label_data)) {
  pd_contour_label_data <- data.frame(
    equity_at_time_zero = numeric(),
    equity_volatility = numeric(),
    risk_neutral_pd = numeric()
  )
}
pd_contour_label_data$label <- fmt_pct(
  pd_contour_label_data$risk_neutral_pd,
  0
)

ggplot2::ggplot(
  pd_grid,
  ggplot2::aes(
    x = equity_at_time_zero,
    y = equity_volatility,
    z = risk_neutral_pd
  )
) +
  ggplot2::geom_contour(
    ggplot2::aes(color = ggplot2::after_stat(level)),
    breaks = pd_contour_breaks,
    linewidth = 0.85
  ) +
  ggplot2::geom_label(
    data = pd_contour_label_data,
    ggplot2::aes(
      x = equity_at_time_zero,
      y = equity_volatility,
      label = label
    ),
    inherit.aes = FALSE,
    size = 4,
    label.size = 0,
    fill = "white",
    alpha = 0.86
  ) +
  ggplot2::geom_vline(xintercept = E0, linetype = "dashed", linewidth = 0.55) +
  ggplot2::geom_hline(yintercept = se, linetype = "dashed", linewidth = 0.55) +
  ggplot2::geom_point(
    data = current_pd_point,
    ggplot2::aes(x = equity_at_time_zero, y = equity_volatility),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) fmt_pct(x, 0),
    limits = c(0, 1.5),
    breaks = seq(0, 1.5, by = 0.5),
    expand = ggplot2::expansion(mult = c(0, 0))
  ) +
  ggplot2::scale_x_continuous(
    limits = c(1, 10),
    breaks = seq(2, 10, by = 2),
    expand = ggplot2::expansion(mult = c(0, 0))
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    end = 0.9,
    guide = "none"
  ) +
  ggplot2::labs(
    x = expression(E[0]),
    y = expression(sigma[E])
  ) +
  ggplot2::theme_classic(base_size = 12)
```      

The red point is the original case, where $E_0=`r fmt_num(E0, 0)`$, $\sigma_E=`r fmt_pct(se, 0)`$ and the model-implied PD is `r fmt_pct(pd1, 2)`. The point lies between the 10% and 15% contour lines, which is consistent with the numerical value. The plot also shows the main financial pattern: lower equity value and higher equity volatility move the firm toward higher default probabilities.

```{r}
l = 40 # Vectors of length 40.
# Create vectors of the parameters.
x.E <- seq(from = 2, to = 4, length.out = l)
x.se <- seq(0.2, 1.5, length.out = l)
# Create the empty matrix.
p_E_se <- matrix(0, nrow = l, ncol = l)
# Fill the empty matrix with probability of default values.
for(i in 1 : l){ # Is there an easier way to do this?
  for(j in 1 : l){
    p_E_se[i,j] <- mapply(pd, x.E[i], x.se[j], rf, TT, D) } }
```

```{r}
#| label: fig-tpodaafoteatzatvoeaiv
#| fig-cap: "The risk-neutral probability of default as a function of the equity at time zero and the volatility of equity: An interactive view."
plot_ly(type = "surface" , x = x.se, y = x.E , z = p_E_se ) %>%
layout(scene = list(xaxis = list(title = "Volatility of equity"), 
                    yaxis = list(title = "Market value of equity"), 
                    zaxis = list(title = "Risk-neutral PD")))  %>%
  hide_colorbar()
```

