## Capital-structure scenarios

```{r merton-pd-setup, include=FALSE}
source("R/merton-pd-function.R")
```

The problem. Daenerys Targaryen owns a big manufacturing firm that produces fire extinguishers. The firm's inputs are exactly the same as in Hull's example 24.3 [@Hull]. She is planning to ask the Iron Bank of Braavos for a loan, but she would like to reduce the firm's model-implied probability of default before negotiating credit conditions. Basic finance gives her two possible levers: reduce the promised debt payment $D$, or increase the market value of equity $E_0$ through a capital injection. Doing both at the same time would be more expensive, so she wants to compare the alternatives one at a time.

In this section, probability of default means the Merton risk-neutral/implied probability of default. The scenarios are useful for understanding how the model reacts to changes in capital structure; they are not historical forecasts of how often firms like this default.

These are model-implied sensitivity scenarios. In each scenario, we change one observed input while keeping the remaining observed inputs and model assumptions fixed. The unobserved asset value $V_0$ and asset volatility $\sigma_V$ are then re-estimated by the Merton equations, so they are not held fixed. The resulting probability of default is the new risk-neutral PD implied by the full Merton solution.

::: {.merton-subhead}
Scenario 1: reduce debt or inject equity
:::

Which action produces the lower model-implied probability of default: reduce the promised debt payment $D$ by 2, or increase the equity value $E_0$ by 2 through a capital injection?

```{r}
D.seq <- seq(from = 0.1, to = 13, by = 0.1)
pd_initial <- pd1
pd_scenario_2_initial <- pd(2, se, rf, TT, 2)
pd_scenario_3_initial <- pd(E0, se, rf, 1.5, 6)
scenario1_reduce_D <- pd(E0, se, rf, TT, D - 2)
scenario1_raise_E <- pd(E0 + 2, se, rf, TT, D)
scenario2_E0 <- 2
scenario2_D <- 2
scenario2_reduce_D <- pd(scenario2_E0, se, rf, TT, scenario2_D - 1)
scenario2_raise_E <- pd(scenario2_E0 + 1, se, rf, TT, scenario2_D)
scenario3_T <- 1.5
scenario3_D <- 6
scenario3_more_debt <- pd(E0, se, rf, scenario3_T, scenario3_D + 2)
scenario3_more_time <- pd(E0, se, rf, scenario3_T + 0.5, scenario3_D)
# Evaluate pd function: E0 changes from 1 to 5; D goes from 0.1 to 13.
E1 <- mapply(pd, 1, se, rf, TT, D.seq)
E2 <- mapply(pd, 2, se, rf, TT, D.seq)
E3 <- mapply(pd, 3, se, rf, TT, D.seq)
E4 <- mapply(pd, 4, se, rf, TT, D.seq)
E5 <- mapply(pd, 5, se, rf, TT, D.seq)
```

Plot the results.

```{r}
#| label: fig-gppodacs
#| fig-cap: "GoT problem. Risk-neutral probability of default and capital structure."
# We use these vectors for colors and legends in the following plots.
colors <- c("green", "purple", "black", "blue", "red")
leg <- c("E0=1", "E0=2", "E0=3 (initial value)", "E0=4", "E0=5")
# Plot.
plot(D.seq, E1, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_initial, 2)))
lines(D.seq, E2, col = "purple", lwd = 3)
lines(D.seq, E3, col = "black", lwd = 3)
lines(D.seq, E4, col = "blue", lwd = 3)
lines(D.seq, E5, col = "red", lwd = 3)
abline(v = D, lty = 2)
abline(h = pd_initial, lty = 2)
legend("bottomright", legend = leg, lwd = rep(3, 5), col = colors, 
       bg = "white")
points(D, pd_initial, pch = 19, cex = 2)
```

Let's illustrate the alternatives as clearly as possible.

```{r}
#| label: fig-gspodacs
#| fig-cap: "GoT solution. Risk-neutral probability of default and capital structure."
plot(D.seq, E1, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_initial, 2)))
lines(D.seq, E2, col = "purple", lwd = 3)
lines(D.seq, E3, col = "black", lwd = 3)
lines(D.seq, E4, col = "blue", lwd = 3)
lines(D.seq, E5, col = "red", lwd = 3)
abline(v = D, lty = 2)
abline(v = D - 2, lty = 2)
abline(h = pd_initial, lty = 2)
abline(h = scenario1_reduce_D, lty = 2)
abline(h = scenario1_raise_E, lty = 2)
legend("bottomright", legend = leg, lwd = rep(3, 5), col = colors, 
       bg = "white")
points(D, pd_initial, pch = 19, cex = 2)
points(D, scenario1_raise_E, pch = 19, cex = 2, col = "red")
points(D - 2, scenario1_reduce_D, pch = 19, cex = 2, col = "yellow")
```

In the first scenario, reducing the promised debt payment from $D=`r fmt_num(D, 0)`$ to $D=`r fmt_num(D - 2, 0)`$ gives a risk-neutral PD of `r fmt_pct(scenario1_reduce_D, 2)`. Increasing the equity value from $E_0=`r fmt_num(E0, 0)`$ to $E_0=`r fmt_num(E0 + 2, 0)`$ gives a risk-neutral PD of `r fmt_pct(scenario1_raise_E, 2)`. Under this model-implied comparison, the lower PD comes from `r ifelse(scenario1_reduce_D < scenario1_raise_E, "reducing the promised debt payment", "increasing the equity value")`. This statement is conditional on the Merton setup: the remaining observed inputs stay fixed, while $V_0$ and $\sigma_V$ are re-estimated in each alternative.

::: {.merton-subhead}
Scenario 2: a different starting balance sheet
:::

Consider a different initial situation. Now, $E_0=2$ and $D=2$ while the remaining observed inputs are unchanged. In this case, the risk-neutral probability of default is `r fmt_pct(pd_scenario_2_initial, 2)`. Which action produces the lower model-implied probability of default: reduce the promised debt payment $D$ by 1, or increase the equity value $E_0$ by 1?

```{r}
#| label: fig-gp2podacs
#| fig-cap: "GoT problem 2. Risk-neutral probability of default and capital structure."
leg2 <- c("E0=1", "E0=2 (initial value)", "E0=3", "E0=4", "E0=5")

plot(D.seq, E1, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(scenario2_D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_scenario_2_initial, 2)), 
     xlim = c(0, 2.5), ylim = c(0, 0.08))
lines(D.seq, E2, col = "purple", lwd = 3)
lines(D.seq, E3, col = "black", lwd = 3)
lines(D.seq, E4, col = "blue", lwd = 3)
lines(D.seq, E5, col = "red", lwd = 3)
abline(v = scenario2_D, lty = 2)
abline(h = pd_scenario_2_initial, lty = 2)
legend("bottomright", legend = leg2, lwd = rep(3, 5), col = colors, 
       bg = "white", cex = 0.6)
points(scenario2_D, pd_scenario_2_initial, pch = 19, cex = 2, col = "purple")
```

The solution can be illustrated as follows:

```{r}
#| label: fig-gs2podacs
#| fig-cap: "GoT solution 2. Risk-neutral probability of default and capital structure."
plot(D.seq, E1, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(scenario2_D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_scenario_2_initial, 2)), 
     xlim = c(0, 2.5), ylim = c(0, 0.08))
lines(D.seq, E2, col = "purple", lwd = 3)
lines(D.seq, E3, col = "black", lwd = 3)
lines(D.seq, E4, col = "blue", lwd = 3)
lines(D.seq, E5, col = "red", lwd = 3)
abline(v = scenario2_D, lty = 2)
abline(v = scenario2_D - 1, lty = 2)
abline(h = pd_scenario_2_initial, lty = 2)
abline(h = scenario2_reduce_D, lty = 2)
abline(h = scenario2_raise_E, lty = 2)
legend("bottomright", legend = leg2, lwd = rep(3, 5), col = colors, 
       bg = "white", cex = 0.6)
points(scenario2_D, pd_scenario_2_initial, pch = 19, cex = 2, col = "purple")
points(scenario2_D - 1, scenario2_reduce_D, pch = 19, cex = 2, col = "yellow")
points(scenario2_D, scenario2_raise_E, pch = 19, cex = 2)
```

In the second scenario, reducing the promised debt payment from $D=`r fmt_num(scenario2_D, 0)`$ to $D=`r fmt_num(scenario2_D - 1, 0)`$ gives a risk-neutral PD of `r fmt_pct(scenario2_reduce_D, 2)`. Increasing the equity value from $E_0=`r fmt_num(scenario2_E0, 0)`$ to $E_0=`r fmt_num(scenario2_E0 + 1, 0)`$ gives a risk-neutral PD of `r fmt_pct(scenario2_raise_E, 2)`. The lower PD comes from `r ifelse(scenario2_reduce_D < scenario2_raise_E, "reducing the promised debt payment", "increasing the equity value")`. The point is not that this action is universally better, but that the model's nonlinear response depends on the starting balance-sheet position.

::: {.merton-subhead}
Scenario 3: more debt or more time
:::

Consider a third scenario. Daenerys Targaryen's firm has some short-term liquidity troubles. She needs either more cash now or more time to pay the promised debt payment. In the model, more borrowing is represented as a higher promised payment $D$ at maturity. She would like to know which alternative leads to the lowest increase in the risk-neutral probability of default: increase $D$ by 2, or ask for a half year more time to pay the promised debt payment?

```{r}
T1 <- mapply(pd, E0, se, rf, 0.5, D.seq)
T2 <- mapply(pd, E0, se, rf, 1, D.seq)
T3 <- mapply(pd, E0, se, rf, 1.5, D.seq)
T4 <- mapply(pd, E0, se, rf, 2, D.seq)
T5 <- mapply(pd, E0, se, rf, 2.5, D.seq)
```

```{r}
#| label: fig-gp3mdolm
#| fig-cap: "GoT problem 3. Higher promised debt payment or longer maturity?"
leg3 <- c("T=2.5", "T=2", "T=1.5", "T=1", "T=0.5")

plot(D.seq, T5, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(scenario3_D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_scenario_3_initial, 2)))
lines(D.seq, T4, col = "purple", lwd = 3)
lines(D.seq, T3, col = "black", lwd = 3)
lines(D.seq, T2, col = "blue", lwd = 3)
lines(D.seq, T1, col = "red", lwd = 3)
abline(v = scenario3_D, lty = 2)
abline(h = pd_scenario_3_initial, lty = 2)
legend("bottomright", legend = leg3, lwd = rep(3, 5), col = colors, 
       bg = "white")
points(scenario3_D, pd_scenario_3_initial, pch = 19, cex = 2)
```

Let's visualize the result.

```{r}
#| label: fig-ga3mdolm
#| fig-cap: "GoT answer 3. Higher promised debt payment or longer maturity?"
plot(D.seq, T5, type = "l", col = "green", lwd = 3,
     xlab = paste0("Promised debt payment. Initially, D=", fmt_num(scenario3_D, 0)),
     ylab = paste0("Risk-neutral PD. Initially PD=", fmt_pct(pd_scenario_3_initial, 2)))
lines(D.seq, T4, col = "purple", lwd = 3)
lines(D.seq, T3, col = "black", lwd = 3)
lines(D.seq, T2, col = "blue", lwd = 3)
lines(D.seq, T1, col = "red", lwd = 3)
abline(v = scenario3_D, lty = 2)
abline(v = scenario3_D + 2, lty = 2)
abline(h = pd_scenario_3_initial, lty = 2)
abline(h = scenario3_more_time, lty = 2)
abline(h = scenario3_more_debt, lty = 2)
legend("bottomright", legend = leg3, lwd = rep(3, 5), col = colors, 
       bg = "white")
points(scenario3_D, pd_scenario_3_initial, pch = 19, cex = 2)
points(scenario3_D + 2, scenario3_more_debt, pch = 19, cex = 2, col = "yellow")
points(scenario3_D, scenario3_more_time, pch = 19, cex = 2, col = "purple")
```

In the third scenario, increasing the promised debt payment from $D=`r fmt_num(scenario3_D, 0)`$ to $D=`r fmt_num(scenario3_D + 2, 0)`$ while keeping maturity at $T=`r fmt_num(scenario3_T, 1)`$ gives a risk-neutral PD of `r fmt_pct(scenario3_more_debt, 2)`. Extending maturity from $T=`r fmt_num(scenario3_T, 1)`$ to $T=`r fmt_num(scenario3_T + 0.5, 1)`$ while keeping $D=`r fmt_num(scenario3_D, 0)`$ gives a risk-neutral PD of `r fmt_pct(scenario3_more_time, 2)`. Both alternatives increase the PD relative to the initial `r fmt_pct(pd_scenario_3_initial, 2)`. The smaller increase comes from `r ifelse(scenario3_more_debt < scenario3_more_time, "increasing the promised debt payment", "extending maturity")`. This does not mean that higher debt improves credit risk; it means that, in this particular Merton configuration, increasing the promised payment is less adverse than extending maturity. This is a structural-model result, not a complete financing recommendation; a real decision would also require cash-flow projections, funding spreads, covenants, and business constraints.
