## Risk-neutral simulations and equity payoff

```{r merton-inside-view-setup, include=FALSE}
source("R/merton-common.R")
```

Let's go back to our main example and simulate the daily evolution of the market value of the firm's assets over a year. Here are 10 simulated paths of $V_t$ from $t=0,\ldots,T$. The reference of this section is [@Hull] Chapter 14 and 15. Because this section is connected to the valuation formula, the simulations use the risk-neutral drift $rf$. These are valuation scenarios under the same risk-neutral world used to value the equity option. They are useful for visualizing the model mechanics, not for forecasting the firm's real-world asset path.

```{r}
#| label: fig-so10pota
#| fig-cap: "Ten risk-neutral valuation paths for the firm's assets."
# For valuation, Merton assumes V follows a risk-neutral GBM process.
V0.sim <- function(i) {  # the argument i is not used here.
  GBM(N = 365, T = 1, t0 = 0, x0 = V0, theta = rf, sigma = sv) }
set.seed(3) # Reproducibility  
paths <- sapply(1:10, V0.sim) # Create 10 paths for V.
final_row <- nrow(paths)
path_time_10 <- seq(0, final_row - 1)
path_data_10 <- data.frame(
  time_day = rep(path_time_10, times = ncol(paths)),
  path_id = factor(rep(seq_len(ncol(paths)), each = nrow(paths))),
  asset_value = as.vector(paths)
)
terminal_path_data_10 <- data.frame(
  time_day = max(path_time_10),
  asset_value = paths[final_row, ],
  terminal_state = ifelse(paths[final_row, ] < D, "Default", "No default")
)
initial_path_data_10 <- data.frame(
  time_day = 0,
  asset_value = V0
)

ggplot2::ggplot(
  path_data_10,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.45, color = "black") +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = initial_path_data_10,
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::geom_point(
    data = terminal_path_data_10,
    ggplot2::aes(x = time_day, y = asset_value, color = terminal_state),
    inherit.aes = FALSE,
    shape = 1,
    size = 3,
    stroke = 1.1
  ) +
  ggplot2::scale_color_manual(
    values = c("Default" = "red", "No default" = "green"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-so10pota should be read as follows. Each black curve is one risk-neutral path for the firm's asset value, $V_t$. The horizontal red line is the promised debt payment, $D$. The blue point marks the starting value, $V_0$. The open circles at maturity classify the terminal values: green means $V_T>D$ and red means $V_T<D$.

These simulations can be interpreted as 10 risk-neutral valuation scenarios for the firm's assets. They are paths generated under the risk-neutral measure, so the drift is the risk-free rate rather than the firm's real-world expected asset return. In the past, some students have called this figure *la grafica de los pelitos*, which is fine as long as the interpretation is clear. The formal name is 10 simulated geometric Brownian motion processes under the risk-neutral measure.

Before counting defaults at maturity, let's first count the paths in which $V_t<D$ at least once before or at maturity. Here I use $t$, not $T$, because this first count asks whether a path ever crosses below the promised debt payment during the year. That is different from Merton default, which is determined by whether $V_T<D$ at maturity.

```{r}
# Which path went lower than D?
paths_below_D <- which(colSums(paths < D) > 0)
paths_below_D
```
The value of the assets was lower than the promised debt payment $D$ at some time in risk-neutral scenarios `r paste(paths_below_D, collapse = ", ")`. Let's plot these cases. In the next figure, red paths finish below $D$ and therefore default at maturity. The blue path crosses below $D$ during the year but finishes above $D$.

```{r}
#| label: fig-nttbpfdnd
#| fig-cap: "Paths that cross below D before maturity; only paths ending below D default."
make_path_data <- function(path_matrix, path_ids = seq_len(ncol(path_matrix))) {
  path_matrix <- as.matrix(path_matrix)
  data.frame(
    time_day = rep(seq(0, nrow(path_matrix) - 1), times = ncol(path_matrix)),
    path_id = factor(rep(path_ids, each = nrow(path_matrix)), levels = path_ids),
    asset_value = as.vector(path_matrix)
  )
}

paths_below_D_matrix <- paths[, paths_below_D, drop = FALSE]
paths_below_D_data <- make_path_data(paths_below_D_matrix, paths_below_D)
paths_below_D_data$terminal_state <- rep(
  ifelse(
    paths_below_D_matrix[final_row, ] < D,
    "Default at maturity",
    "Crosses D before maturity, no default"
  ),
  each = nrow(paths_below_D_matrix)
)

ggplot2::ggplot(
  paths_below_D_data,
  ggplot2::aes(
    x = time_day,
    y = asset_value,
    group = path_id,
    color = terminal_state
  )
) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::scale_color_manual(
    values = c(
      "Crosses D before maturity, no default" = "blue",
      "Default at maturity" = "red"
    ),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-nttbpfdnd isolates the paths from the 10-scenario simulation that cross $D$ at least once. The blue path is the key teaching case. It enters the region below the red debt threshold before maturity, then recovers and ends above $D$ at $T$. In the basic Merton model, default is determined by the terminal condition $V_T<D$. In this small risk-neutral simulation, that happens in `r sum(paths[final_row, ] < D)` cases. Thus, according to these 10 risk-neutral scenarios, the simulated probability of default is `r fmt_pct(sum(paths[final_row, ] < D) / ncol(paths), 0)`. Here is how we calculate the cases in which $V_T<D$.

```{r}
sim_pd_10 <- sum(paths[final_row, ] < D) / ncol(paths)
sim_pd_10
```

The simulated risk-neutral probability of default of `r fmt_pct(sim_pd_10, 0)` looks high relative to the model value of `r fmt_pct(pd_merton, 4)`. This happens because the number of simulations is small. The following examples show how these values tend to converge as we increase the number of simulations from 10 to 100 and to 100,000.

Here are 100 simulated paths from $V_0$ to $V_T$. The visual coding is the same as before: gray curves are risk-neutral asset paths, the red horizontal line is $D$, and the blue point is the common starting value $V_0$.

```{r}
#| label: fig-so100pota
#| fig-cap: "One hundred risk-neutral valuation paths for the firm's assets."
set.seed(3) # Reproducibility  
paths <- sapply(1:100, V0.sim) # Create 100 paths.
final_row <- nrow(paths)
path_data_100 <- make_path_data(paths)

ggplot2::ggplot(
  path_data_100,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.25, color = "gray35", alpha = 0.55) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

```{r}
sim_pd_100 <- sum(paths[final_row, ] < D) / ncol(paths)
sim_pd_100
```

With 100 paths, individual trajectories are less important than the terminal count. The simulated risk-neutral probability of default is `r fmt_pct(sim_pd_100, 0)`. This is closer to the model value of `r fmt_pct(pd_merton, 4)`.

Let's see all the `r sum(paths[final_row, ] < D)` cases that end in default.

```{r}
#| label: fig-dc
#| fig-cap: "Default cases."
default_cases_100 <- which(paths[final_row, ] < D)
default_path_data_100 <- make_path_data(
  paths[, default_cases_100, drop = FALSE],
  default_cases_100
)

ggplot2::ggplot(
  default_path_data_100,
  ggplot2::aes(x = time_day, y = asset_value, group = path_id)
) +
  ggplot2::geom_line(linewidth = 0.45, color = "gray25", alpha = 0.75) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-dc filters the 100 simulated paths and keeps only the paths with $V_T<D$. These are the default cases under the Merton terminal-default rule. The red line is still $D$, so the important feature is where each path ends relative to that line at maturity.

```{r include=FALSE}
almost_default <- which(colSums(paths < D) > 0)
default <- which(paths[final_row, ] < D)
almost_default_no_default <- setdiff(almost_default, default)
```

Now, let's see the `r length(almost_default_no_default)` cases in which $V_t$ went below $D$ at some point, but $V_T$ finished above $D$, so the firm did not default at maturity.

```{r}
#| label: fig-atpwl1asp
#| fig-cap: "These paths cross below D before maturity, but finish with V_T above D and therefore do not default."
almost_default_no_default_data <- make_path_data(
  paths[, almost_default_no_default, drop = FALSE],
  almost_default_no_default
)
almost_default_colors <- grDevices::hcl.colors(
  length(almost_default_no_default),
  "Dark 3"
)
names(almost_default_colors) <- levels(almost_default_no_default_data$path_id)

ggplot2::ggplot(
  almost_default_no_default_data,
  ggplot2::aes(
    x = time_day,
    y = asset_value,
    group = path_id,
    color = path_id
  )
) +
  ggplot2::geom_line(linewidth = 0.55, alpha = 0.9) +
  ggplot2::geom_hline(yintercept = D, linewidth = 0.9, color = "red") +
  ggplot2::geom_point(
    data = data.frame(time_day = 0, asset_value = V0),
    ggplot2::aes(x = time_day, y = asset_value),
    inherit.aes = FALSE,
    size = 3,
    color = "blue"
  ) +
  ggplot2::scale_color_manual(values = almost_default_colors, guide = "none") +
  ggplot2::labs(
    x = "Time in days (from 0 to T)",
    y = expression(V[t])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Figure @fig-atpwl1asp shows the other side of the crossing story. Every colored path falls below $D$ at least once, but each one finishes above $D$ at maturity. The different colors only separate the paths visually. The financial message is that the path history helps us see the simulated movement of assets, while the basic Merton default event is decided at $T$.

Here are 100,000 simulated values of $V_T$ shown in a histogram. The reference for this section is section 15.1 in @Hull. Note that instead of drawing the complete paths, we can directly get the risk-neutral distribution of $V_T$. The resulting distribution is log-normal.

```{r}
#| label: fig-ho1votaam
#| fig-cap: "Histogram of 100,000 values of the assets at maturity."
set.seed(13)
# 100,000 values of VT at once.
VT <- exp(rnorm(100000, log(V0) + (rf - (sv^2) / 2)*TT, sv*sqrt(TT)))
# Plot results.
h <- hist(VT, 100, plot = FALSE)
hist_data <- data.frame(
  xmin = head(h$breaks, -1),
  xmax = tail(h$breaks, -1),
  count = h$counts
)
hist_data$region <- ifelse(hist_data$xmin < D, "Default region", "No default region")
sim_pd_100k <- mean(VT < D)
hist_label_x <- 18.4
hist_label_y <- max(hist_data$count) * 0.92
hist_label_gap <- max(hist_data$count) * 0.075
hist_merton_label <- paste0(
  "paste('Merton ', N(-d[2]), ' = ", fmt_pct(pd_merton, 4), "')"
)
ggplot2::ggplot(hist_data) +
  ggplot2::geom_rect(
    ggplot2::aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = count,
      fill = region
    ),
    color = "white",
    linewidth = 0.1
  ) +
  ggplot2::geom_vline(xintercept = D, linetype = "dashed", linewidth = 0.7) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y,
    hjust = 0,
    vjust = 1,
    label = paste0("Red area = ", fmt_pct(sim_pd_100k, 3)),
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = "Simulated risk-neutral PD",
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - 2 * hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = hist_merton_label,
    parse = TRUE,
    size = 3.2
  ) +
  ggplot2::annotate(
    "text",
    x = hist_label_x,
    y = hist_label_y - 3 * hist_label_gap,
    hjust = 0,
    vjust = 1,
    label = "The two values are close.",
    size = 3.2
  ) +
  ggplot2::coord_cartesian(xlim = c(4, 30), expand = FALSE) +
  ggplot2::scale_fill_manual(
    values = c("Default region" = "red", "No default region" = "blue"),
    guide = "none"
  ) +
  ggplot2::labs(
    x = "Value of the assets at maturity",
    y = "Frequency"
  ) +
  ggplot2::theme_classic(base_size = 12)
```

The simulated risk-neutral probability of default can be calculated as:

```{r}
sim_pd_100k
```

Note that `r fmt_num(sim_pd_100k, 5)` is now much closer to the model value `r fmt_num(pd_merton, 7)`. The larger simulation is doing what we expect: it is approximating the model-implied risk-neutral probability $N(-d_2)$.

This firm can be represented as a European call option payoff. The payoff of a typical stock option is $c=max(S_T-K, 0)$ or equivalently in terms of the Merton model: $E_T=max(V_T-D, 0)$.

```{r}
#| label: fig-avotfbssitf
#| fig-cap: "A view of the firm's balance sheet in the future. Looks like a typical European call option payoff."
ET <- pmax(VT - D, 0) # payoff function of a typical call option.
plot(sort(VT), sort(ET), type = "l", lwd = 4, xlim = c(5, 20), 
     ylim = c(0, 10), xlab = "Simulated assets at maturity (V_T)",
     ylab = "Simulated equity at maturity (E_T)")
abline(v = 10, lty = 2)
et_legend_text <- paste0(
  "In the simulation,\n",
  "E_T=0 happens\n",
  "in ", fmt_pct(sim_pd_100k, 3), " of the\n",
  "100,000 cases."
)
legend("left", legend = et_legend_text,
bg = "white", bty = "n", cex = 0.9)
```

This diagram shows the firm's market values at maturity. The equity value at maturity, $E_T$, is a function of $V_T$ and $D$. If $V_T<D$, then $E_T=0$. If $V_T>D$, then $E_T=V_T-D$, just as the balance-sheet logic suggests in future-value terms. At time 0, $E_T$ is a future random payoff. The Merton model values equity today and assigns a risk-neutral probability to the default case, $V_T<D$, which is the same case in which $E_T=0$.

This payoff relation, $E_T=\max(V_T-D, 0)$, is the key point of the inside view. Equation 24.3 in [@Hull] was already used in the estimation step to choose $V_0$ and $\sigma_V$ so that the model is consistent with the observed equity value $E_0$. Recomputing $E_0$ here would only repeat that calibration check, so we do not treat it as a new valuation exercise.

The important distinction is therefore: $E_T$ is the future state-dependent equity payoff, $E_0$ is the observed equity value today, $V_0$ is the model's market value of assets today, and $D$ is the promised debt payment at maturity. The estimate $V_0$ can be compared with the book value of total assets to understand the difference between book values and market values. We can also analyze the market value of debt, which is different from the promised debt payment $D$.

