## Estimating asset value and risk-neutral PD

```{r merton-estimation-setup, include=FALSE}
source("R/merton-packages.R")
source("R/format-helpers.R")
```

::: {.merton-subhead}
Inputs and Hull equations
:::

We use Example 24.3 in [@Hull]. The observed inputs are $E_0=3$, $\sigma_E=0.8$, $rf=0.05$, $T=1$, and $D=10$. Here, $D$ is the promised debt payment due at maturity. The current market value of debt is a different quantity. Using these inputs, we estimate the value of the assets today, $V_0$, and the volatility of the assets, $\sigma_V$. Those estimated asset quantities are then used to calculate the Merton risk-neutral probability of default $pd$, among other results.

The risk-free rate appears here because equation 24.3 is a valuation equation. We are not assuming that the firm's assets truly earn the risk-free rate in the real world. We are valuing equity under the risk-neutral measure, which is the same logic used to price a European call option. In that artificial valuation world, default is still the event $V_T < D$. After estimating $V_0$ and $\sigma_V$, we will write the risk-neutral probability of that event as $N(-d_2)$.

In this section, probability of default means the model-implied risk-neutral default probability unless stated otherwise.

These are the known parameters.

```{r}
# List of known parameters.
E0 <- 3 # Value of equity today; this is market capitalization.
se <- 0.8 # Stock return volatility.
rf <- 0.05 # Risk-free rate.
TT <- 1 # Maturity.
D <- 10 # Promised debt payment due at maturity. The Bloomberg function DDIS is useful.
```

The reason $V_0$ and $\sigma_V$ are difficult is that neither is directly observed. Hull's example uses two observable equity quantities, $E_0$ and $\sigma_E$, to infer them. We therefore replicate equations 24.3 and 24.4 from [@Hull] before translating them into R.

The first equation comes from the payoff of equity at maturity. If the firm's assets are worth more than the promised debt payment, equity receives the residual value. If the firm's assets are worth less than the promised debt payment, equity receives zero:

$$
E_T=\max(V_T-D,0).
$$

Equation 24.3 is the closed-form value of that payoff. It is the risk-neutral expected equity payoff, discounted at the risk-free rate. In other words, the observed market value of equity today must equal the model value of the equity claim on the firm's assets.

$$
E_0 = V_0N(d_1)-De^{-rT}N(d_2).
\tag{24.3}
$$

The terms $d_1$ and $d_2$ are:

$$
d_1 =
\frac{\ln(V_0/D)+\left(r+\sigma_V^2/2\right)T}
{\sigma_V\sqrt{T}},
\qquad
d_2=d_1-\sigma_V\sqrt{T}.
$$

Here $N(\cdot)$ is the cumulative standard normal distribution. In the code, Hull's $r$ is our `rf`. This equation says that, given an asset value $V_0$ and asset volatility $\sigma_V$, the option-pricing formula must reproduce the observed equity value $E_0$.

The two normal terms do not play the same role. In this chapter, $N(d_2)$ will become the survival probability under the risk-neutral distribution, so $N(-d_2)$ will be the risk-neutral default probability. The term $N(d_1)$ enters through the sensitivity of equity value to the firm's asset value.

Hull obtains equation 24.4 from It&ocirc;'s lemma. It links the observed equity volatility to the unobserved asset volatility:

$$
\sigma_E E_0
=
\frac{\partial E}{\partial V}\sigma_V V_0
=
N(d_1)\sigma_V V_0.
\tag{24.4}
$$

This equation is needed because equity is levered. Equity volatility is usually higher than asset volatility, and the scaling depends on the option delta $N(d_1)$ and on the ratio between asset value and equity value. Together, equations 24.3 and 24.4 give two conditions for the two unknowns, $V_0$ and $\sigma_V$. The code below writes each condition as an error. The numerical solution is the pair $(V_0,\sigma_V)$ that makes both errors as close to zero as possible.

To make the link with the code explicit, define the error from equation 24.3 as

$$
F(V_0,\sigma_V)
=
V_0N(d_1)-De^{-rT}N(d_2)-E_0,
$$

and the error from equation 24.4 as

$$
G(V_0,\sigma_V)
=
N(d_1)\sigma_V V_0-\sigma_EE_0.
$$

The target is not to make one equation fit while ignoring the other. We want both errors to be close to zero at the same time. Hull's footnote 10 implements that idea by minimizing the sum of squared errors:

$$
\min_{V_0,\sigma_V}
\left[
F(V_0,\sigma_V)^2+G(V_0,\sigma_V)^2
\right].
$$

This is exactly what the R function `min.footnote10()` computes below.

The numerical search also needs starting values. These starting values give the algorithm an economically sensible place to begin. For $V_0$, we start with equity value plus the risk-free present value of the promised debt payment, $E_0+De^{-rfT}$. This is a starting scale for the unobserved asset value, not an assumption that debt is risk-free. For $\sigma_V$, we use a simple deleveraging guess, $\sigma_E E_0/(E_0+De^{-rfT})$, because asset volatility is usually lower than equity volatility when equity is levered.

::: {.merton-subhead}
Solving the two Merton equations
:::

The optimizer is then run with `L-BFGS-B` and positive lower bounds. This keeps both $V_0$ and $\sigma_V$ in economically meaningful territory during the search. Because $V_0$ is measured in units of value while $\sigma_V$ is a decimal volatility, the code also gives `optim()` scale information for the numerical search.

```{r}
start_V0 <- E0 + D * exp(-rf * TT)
start_sv <- se * E0 / start_V0

d1_merton <- function(V0, sv) {
  (log(V0 / D) + (rf + sv^2 / 2) * TT) / (sv * sqrt(TT))
}

d2_merton <- function(V0, sv) {
  d1_merton(V0, sv) - sv * sqrt(TT)
}

eq24.3 <- function(V0, sv) { 
  V0 * pnorm(d1_merton(V0, sv)) -
    D * exp(-rf * TT) * pnorm(d2_merton(V0, sv)) -
    E0
}

eq24.4 <- function(V0, sv) { 
  pnorm(d1_merton(V0, sv)) * sv * V0 - se * E0
}

# Footnote 10 indicates that we should minimize F(x,y)^2+G(x,y)^2
min.footnote10 <- function(x) 
  (eq24.3(x[1], x[2]))^2 + (eq24.4(x[1], x[2]))^2
# The minimization leads to the values of V0 and sv.
V0_sv <- optim(
  par = c(V0 = start_V0, sv = start_sv),
  fn = min.footnote10,
  method = "L-BFGS-B",
  lower = c(V0 = 1e-6, sv = 1e-6),
  control = list(
    parscale = c(V0 = 10, sv = 0.2),
    ndeps = c(V0 = 1e-6, sv = 1e-8)
  )
)
# Define the values as parameters.
V0 <- unname(V0_sv$par["V0"])
sv <- unname(V0_sv$par["sv"])
```

The optimizer therefore starts at

$$
V_0^{start}
=E_0+De^{-rT}
=`r fmt_num(start_V0, 4)`,
$$

and

$$
\sigma_V^{start}
=
\frac{\sigma_EE_0}{E_0+De^{-rT}}
=`r fmt_num(start_sv, 4)`.
$$

These are only starting values. The final estimates are allowed to move away from them when the optimizer minimizes $F(V_0,\sigma_V)^2+G(V_0,\sigma_V)^2$.

The optimizer returns the following Merton asset estimates:

$$
\widehat{V}_0=`r fmt_num(V0, 4)`,
\qquad
\widehat{\sigma}_V=`r fmt_num(sv, 7)`
=`r fmt_pct(sv, 4)`.
$$

These two estimates are the bridge to the rest of the section. First, $V_0$ and $\sigma_V$ determine the risk-neutral default weight $N(-d_2)$. Then they imply the market value of debt, which lets us infer the Merton-implied bond yield spread.

::: {.merton-subhead}
From $d_2$ to the risk-neutral PD
:::

Calculate the Merton risk-neutral probability of default as a function of $V_0$ and $\sigma_V$. In Hull's notation, this probability is $N(-d_2)$. The next paragraphs unpack what that notation means.

```{r}
PD <- function(V0, sv) { 
  pnorm(-(((log(V0 / D) + (rf + sv^2 / 2) * TT) / 
             (sv * sqrt(TT))) - sv * sqrt(TT))) }
# Calculate the risk-neutral probability of default given the previous parameters.
pd <- PD(V0, sv)
knitr::kable(
  data.frame(
    Quantity = c(
      "Estimated asset value",
      "Estimated asset volatility",
      "Merton risk-neutral PD"
    ),
    Value = c(
      fmt_num(V0, 4),
      fmt_pct(sv, 4),
      fmt_pct(pd, 4)
    )
  ),
  caption = "Estimated Merton asset value, asset volatility, and risk-neutral default probability.",
  row.names = FALSE
)
```
The risk-neutral probability of default is `r fmt_num(pd, 7)`, or `r fmt_pct(pd, 4)`. This number is market-implied by the option-pricing structure of the model. It answers a pricing question. Under the risk-neutral distribution used to value equity, how much weight is assigned to default states? A direct historical or real-world default probability would require a separate calibration from risk-neutral probabilities to real-world probabilities.

The next table isolates $d_1$, $d_2$, $N(d_1)$ and $N(d_2)$ so we can connect the formula to the default region.

```{r}
# Inspect the role of d and N(d).
d1 <- d1_merton(V0, sv)
d2 <- d2_merton(V0, sv)
Nd1 <- pnorm(d1)
Nd2 <- pnorm(d2)
merton_d_table <- data.frame(
  Quantity = c("$d_1$", "$N(d_1)$", "$d_2$", "$N(d_2)$", "$N(-d_2)$"),
  Value = c(
    fmt_num(d1, 7),
    fmt_num(Nd1, 7),
    fmt_num(d2, 7),
    fmt_num(Nd2, 7),
    fmt_num(pd, 7)
  )
)

knitr::kable(
  merton_d_table,
  caption = "The d-values and normal probabilities behind the Merton PD.",
  escape = FALSE,
  row.names = FALSE
)
```

Before connecting this result with the figure, unpack the notation. The symbol $N(\cdot)$ means the cumulative standard normal distribution. In plain words, $N(x)$ is the area to the left of $x$ under the standard normal curve. The value $d_2$ comes from the Black-Scholes-Merton formula. In this chapter, $d_2$ has a concrete credit-risk role. After the model writes future asset values in standard normal units, $d_2$ determines the cutoff between survival and default.

The default condition remains the original Merton condition:

$$
V_T<D.
$$

The formula converts that asset-value condition into standard normal units:

$$
Z<-d_2.
$$

This is only a change of scale. We read the same future-asset event in standard normal units. Survival corresponds to the right-tail region, and default corresponds to the left-tail region. The survival probability under the risk-neutral distribution is

$$
P^Q(V_T>D)=N(d_2),
$$

and the default probability is

$$
P^Q(V_T<D)=N(-d_2).
$$

The value $N(d_1)$ has a different role. It should be read as the option delta in the Merton equity formula. It measures how sensitive the market value of equity is to a change in the market value of the firm's assets. The clean credit-risk probability is $N(-d_2)$.

This is why the `r fmt_pct(pd, 4)` probability should be read as a risk-neutral default probability implied by the Merton pricing structure. A direct empirical default frequency would require a separate real-world calibration.

The next figure illustrates how $N(-d_2)$ appears as the left-tail default area under the standard normal distribution.

Figure @fig-trartpod shows exactly this area in red. Every point in the red tail represents a standardized future asset value low enough to make $V_T<D$ at maturity.

```{r}
#| label: fig-trartpod
#| fig-cap: "The red area represents the risk-neutral probability of default."
xseq <- seq(-4, 4, 0.01) 
densities <- dnorm(xseq, 0, 1) # Standard normal distribution.
legend_nd2_expression <- paste0("N(d[2])==", round(Nd2, 7))
legend_pd_expression <- paste0("N(-d[2])==", round(pd, 7), "~'in red.'")
normal_density_data <- data.frame(
  x = xseq,
  density = densities
)
left_tail_data <- normal_density_data[normal_density_data$x < -d2, ]
right_tail_data <- normal_density_data[normal_density_data$x >= -d2, ]

ggplot2::ggplot(normal_density_data, ggplot2::aes(x = x, y = density)) +
  ggplot2::geom_area(
    data = right_tail_data,
    fill = "green",
    color = NA
  ) +
  ggplot2::geom_area(
    data = left_tail_data,
    fill = "red",
    color = NA
  ) +
  ggplot2::geom_line(linewidth = 0.9, color = "black") +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, color = "black") +
  ggplot2::geom_vline(
    xintercept = -d2,
    linetype = "dotted",
    linewidth = 1.2,
    color = "black"
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.377,
    hjust = 0,
    label = legend_nd2_expression,
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.352,
    hjust = 0,
    label = "'in green and'",
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.327,
    hjust = 0,
    label = legend_pd_expression,
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.305,
    hjust = 0,
    label = "N(d[2])+N(-d[2])==1",
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.42), expand = FALSE) +
  ggplot2::labs(
    x = expression(paste("d values (", -d[2], " is the dotted line)")),
    y = "Density"
  ) +
  ggplot2::theme_classic(base_size = 12)
```

::: {.merton-subhead}
Checking the numerical solution
:::

The next diagnostic checks the minimization that leads to $V_0$ and $\sigma_V$. Equations 24.3 and 24.4 in [@Hull] are solved numerically in this implementation, so $V_0$ and $\sigma_V$ come from minimizing the squared pricing errors. To check the solution, we retrieve the value of $F(V_0,\sigma_V)^2+G(V_0,\sigma_V)^2$ evaluated at the estimated pair $(\widehat{V}_0,\widehat{\sigma}_V)$. This objective function is positive because both errors are squared, so a good solution should produce a value close to zero.

```{r}
merton_objective_at_solution <- min.footnote10(x = c(V0, sv))

knitr::kable(
  data.frame(
    Quantity = c(
      "Objective function value",
      "Rounded objective function value"
    ),
    Value = c(
      fmt_num(merton_objective_at_solution, 10),
      fmt_num(round(merton_objective_at_solution, 6), 6)
    )
  ),
  caption = "Objective function value at the estimated Merton solution.",
  row.names = FALSE
)
```

The constrained optimization above worked well. The objective function is zero in practical terms.

The minimization can also be checked visually. To show the objective function on two axes, we first hold $\sigma_V$ fixed at its estimated value and vary $V_0$. Then we hold $V_0$ fixed and vary $\sigma_V$. These are one-dimensional slices of the same two-parameter objective function.

```{r}
# Fix sv and evaluate different values of V0.
V0.seq <- seq(from = 4, to = 20, length.out = 100)
sv.rep <- rep(sv, 100)
# Function to be evaluated.
FG <- function(V0, sv) { (eq24.3(V0, sv))^2 + (eq24.4(V0, sv))^2 }
# Apply the function with fixed sv and different V0 values.
FG.V0 <- mapply(FG, V0.seq, sv.rep)
```

The first slice is the objective function as a function of $V_0$ while $\sigma_V$ is held at its estimated value.

```{r}
#| label: fig-hwftvotaactvoaatz
#| fig-cap: "Objective-function slice when asset volatility is held fixed."
objective_at_solution <- FG(V0, sv)
slice_v0_data <- data.frame(
  asset_value = V0.seq,
  objective = FG.V0
)
solution_v0_data <- data.frame(
  asset_value = V0,
  objective = objective_at_solution
)

ggplot2::ggplot(slice_v0_data, ggplot2::aes(x = asset_value, y = objective)) +
  ggplot2::geom_line(linewidth = 1.1, color = "black") +
  ggplot2::geom_vline(xintercept = V0, linetype = "dashed", linewidth = 0.7) +
  ggplot2::geom_hline(
    yintercept = objective_at_solution,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    data = solution_v0_data,
    ggplot2::aes(x = asset_value, y = objective),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::labs(
    x = expression(V[0]),
    y = expression(F(x, y)^2 + G(x, y)^2)
  ) +
  ggplot2::theme_classic(base_size = 12)
```

@fig-hwftvotaactvoaatz is a one-dimensional slice of the minimization problem. In this slice, $\sigma_V$ is held fixed at the estimated value and only $V_0$ changes. The vertical dashed line marks the estimated asset value, and the red circle marks the value of the objective function at that estimate. The curve reaches its lowest point near the estimated $V_0$, so nearby alternatives for the asset value produce larger squared errors in equations 24.3 and 24.4. Economically, moving away from this $V_0$ makes it harder for the model to match both the observed equity value and the observed equity volatility.

This figure is a diagnostic check separate from the estimation method. It shows that, conditional on the estimated $\sigma_V$, the optimizer has found a sensible asset value. For completeness, we now make the symmetric check by fixing $V_0$ and varying $\sigma_V$.

```{r}
#| label: fig-hwftvotaatzactvota
#| fig-cap: "Objective-function slice when asset value is held fixed."
# Now fix V0, and evaluate different sv values.
sv.seq <- seq(0.001, 0.6, length.out = 100)
V0.rep <- rep(V0, 100)
# Evaluate the function with these parameters.
FG.sv <- mapply(FG, V0.rep, sv.seq)
slice_sv_data <- data.frame(
  asset_volatility = sv.seq,
  objective = FG.sv
)
solution_sv_data <- data.frame(
  asset_volatility = sv,
  objective = objective_at_solution
)

ggplot2::ggplot(
  slice_sv_data,
  ggplot2::aes(x = asset_volatility, y = objective)
) +
  ggplot2::geom_line(linewidth = 1.1, color = "black") +
  ggplot2::geom_vline(xintercept = sv, linetype = "dashed", linewidth = 0.7) +
  ggplot2::geom_hline(
    yintercept = objective_at_solution,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    data = solution_sv_data,
    ggplot2::aes(x = asset_volatility, y = objective),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::labs(
    x = expression(sigma[V]),
    y = expression(F(x, y)^2 + G(x, y)^2)
  ) +
  ggplot2::theme_classic(base_size = 12)
```

@fig-hwftvotaatzactvota gives the second slice. Now $V_0$ is fixed at the estimated value and $\sigma_V$ changes. The estimated asset volatility again sits near the bottom of the curve. If $\sigma_V$ is too low, the model understates how strongly equity reacts to asset risk. If $\sigma_V$ is too high, the model overstates that risk transmission. Both mistakes increase the objective function because the two Hull equations are no longer matched as closely.

Taken together, @fig-hwftvotaactvoaatz and @fig-hwftvotaatzactvota show that the numerical solution is locally coherent in each parameter direction. The next diagnostic evaluates the objective function over a grid where $V_0$ and $\sigma_V$ move at the same time.

```{r}
# outer evaluates the function for each combination of the vectors.
z <- outer(V0.seq, sv.seq, FG) # z is now 100x100 matrix.
```
Contour plots, also called level plots, show a three-dimensional objective function on a two-dimensional plane.

```{r}
#| label: fig-motfacv
#| fig-cap: "Minimization of the function: a contour view."
contour_data <- expand.grid(
  asset_value = V0.seq,
  asset_volatility = sv.seq
)
contour_data$objective <- as.vector(z)
solution_surface_data <- data.frame(
  asset_value = V0,
  asset_volatility = sv
)
contour_breaks <- c(
  0.001, 0.01, 0.05, 0.1, 0.25, 0.5,
  1, 2, 5, 10, 20, 50, 100
)
contour_label_levels <- c(1, 5, 10, 50, 100)
contour_label_lines <- contourLines(
  x = V0.seq,
  y = sv.seq,
  z = z,
  levels = contour_label_levels
)
contour_label_data <- do.call(
  rbind,
  lapply(contour_label_levels, function(level_value) {
    level_lines <- contour_label_lines[
      vapply(
        contour_label_lines,
        function(line) isTRUE(all.equal(line$level, level_value)),
        logical(1)
      )
    ]
    longest_line <- level_lines[[which.max(vapply(level_lines, function(line) {
      length(line$x)
    }, numeric(1)))]]
    label_index <- round(length(longest_line$x) * 0.72)
    data.frame(
      asset_value = longest_line$x[label_index],
      asset_volatility = longest_line$y[label_index],
      objective = level_value
    )
  })
)
contour_label_data$label <- sub(
  "\\.0$",
  "",
  format(contour_label_data$objective, trim = TRUE)
)

ggplot2::ggplot(
  contour_data,
  ggplot2::aes(x = asset_value, y = asset_volatility, z = objective)
) +
  ggplot2::geom_contour(
    ggplot2::aes(color = ggplot2::after_stat(level)),
    breaks = contour_breaks,
    linewidth = 0.75
  ) +
  ggplot2::geom_label(
    data = contour_label_data,
    ggplot2::aes(
      x = asset_value,
      y = asset_volatility,
      label = label
    ),
    inherit.aes = FALSE,
    size = 4,
    label.size = 0,
    fill = "white",
    alpha = 0.85
  ) +
  ggplot2::geom_vline(xintercept = V0, linetype = "dashed", linewidth = 0.6) +
  ggplot2::geom_hline(yintercept = sv, linetype = "dashed", linewidth = 0.6) +
  ggplot2::geom_point(
    data = solution_surface_data,
    ggplot2::aes(x = asset_value, y = asset_volatility),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    end = 0.9,
    trans = "log10",
    guide = "none"
  ) +
  ggplot2::labs(
    x = expression(V[0]),
    y = expression(sigma[V])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Each contour line connects combinations of $V_0$ and $\sigma_V$ that produce the same value of $F(V_0,\sigma_V)^2+G(V_0,\sigma_V)^2$. The numbers printed on selected lines give direct reference values for the objective function. The red point marks the estimated pair. The dashed lines make its two coordinates easier to read. A good numerical solution should sit near the innermost low-objective region because that is where the two Hull equations are matched most closely.

The interactive surface shows the same objective function in three dimensions. In HTML, it can be rotated to inspect how the low-objective region sits around the estimated pair.


```{r}
#| label: fig-motfaiv
#| fig-cap: "Minimization of the function: an interactive view."
plot_ly(type = "surface" , x = sv.seq, y = V0.seq , z = z ) %>%
layout(scene = list(xaxis = list(title = "Assets volatility"), 
                    yaxis = list(title = "Assets", 
                    zaxis = list(title = "f=F^2+G^2"))))  %>%
hide_colorbar()
```

The diagnostics can be summarized as follows.

```{r merton-optimization-diagnostic-summary}
merton_optimization_diagnostic <- data.frame(
  Diagnostic = c(
    "Objective value at solution",
    "Slice in $V_0$",
    "Slice in $\\sigma_V$",
    "Contour plot",
    "Interactive surface"
  ),
  `What is checked` = c(
    "Whether both Hull equation errors are essentially zero at the estimated pair.",
    "Whether moving only $V_0$ away from the solution increases the objective.",
    "Whether moving only $\\sigma_V$ away from the solution increases the objective.",
    "Whether $(\\widehat{V}_0,\\widehat{\\sigma}_V)$ sits in the low-objective region when both variables move.",
    "Whether the two-dimensional contour view matches the three-dimensional objective surface."
  ),
  `Reading in this example` = c(
    paste0("The rounded objective is ", fmt_num(round(merton_objective_at_solution, 6), 6), "."),
    "The red point lies near the lowest part of the $V_0$ slice.",
    "The red point lies near the lowest part of the $\\sigma_V$ slice.",
    "The estimated pair is inside the innermost low-error region.",
    "The valley of the objective surface is centered near the estimated pair."
  ),
  check.names = FALSE
)

knitr::kable(
  merton_optimization_diagnostic,
  caption = "How to read the Merton numerical-solution diagnostics.",
  escape = FALSE,
  row.names = FALSE
)
```

::: {.merton-subhead}
Model-implied debt value and spread
:::

The fitted Merton model also implies debt-related measures. These are not new observed inputs. They are computed from the estimated asset value, the observed equity value, and the promised debt payment due at maturity.

```{r}
# Model-implied debt quantities.
market_value_debt <- V0 - E0
pv_promised_payment_debt <- D * exp(-rf * TT)
credit_value_gap <- pv_promised_payment_debt - market_value_debt
implied_loss_rate <- credit_value_gap /
  pv_promised_payment_debt
implied_lgd <- implied_loss_rate / pd
implied_recovery_rate <- 1 - implied_lgd
reconstructed_debt_value <- pv_promised_payment_debt *
  (1 - pd * implied_lgd)
debt_summary <- data.frame(
  Measure = c(
    "Model-implied market value of debt",
    "Risk-free present value of the promised payment",
    "Credit-risk value gap",
    "Credit-risk value discount",
    "Risk-neutral probability of default",
    "Constant-equivalent implied LGD",
    "Constant-equivalent implied recovery rate",
    "Debt value reconstructed from PD and implied recovery"
  ),
  Value = c(
    fmt_num(market_value_debt, 4),
    fmt_num(pv_promised_payment_debt, 4),
    fmt_num(credit_value_gap, 4),
    fmt_pct(implied_loss_rate, 4),
    fmt_pct(pd, 4),
    fmt_pct(implied_lgd, 4),
    fmt_pct(implied_recovery_rate, 4),
    fmt_num(reconstructed_debt_value, 4)
  )
)
knitr::kable(
  debt_summary,
  caption = "Debt measures implied by the fitted Merton model.",
  row.names = FALSE
)
```

The first row is the model-implied market value of debt, denoted $B_0$. There is one debt value here, but two useful ways to read it. The payoff view starts at maturity. In Merton, debt holders receive the promised payment $D$ only if the firm has enough asset value, $V_T>D$. If the firm defaults, $V_T<D$, debt holders receive the available asset value $V_T$ instead. Therefore, the debt payoff is

$$
B_T=\min(V_T,D).
$$

The same payoff can be written as a risk-free promise minus a default shortfall:

$$
B_T
=
D-\max(D-V_T,0).
$$

This expression shows where default enters the debt value. The risk-free promise is $D$. The term $\max(D-V_T,0)$ is the shortfall absorbed by debt holders when the firm ends below the promised debt payment. If $V_T\geq D$, the shortfall is zero. If $V_T<D$, the shortfall is positive.

The payoff view values that default-contingent debt payoff under the risk-neutral distribution:

$$
B_0
=
e^{-rfT}E^Q[\min(V_T,D)].
$$

The balance-sheet view gets the same debt value from the fact that firm assets are split between equity and debt claims:

$$
B_0=V_0-E_0.
$$

In this implementation we use the balance-sheet expression because $V_0$ has already been estimated and $E_0$ is observed. Numerically,

$$
B_0
=
V_0-E_0
=
`r fmt_num(V0, 4)`-`r fmt_num(E0, 4)`
=
`r fmt_num(market_value_debt, 4)`.
$$

This is not a second definition of debt value. It is the same $B_0$ read from the other side of the Merton balance sheet. The equity value $E_0$ has already been calibrated with a payoff that depends on whether $V_T$ is above or below $D$, so the residual debt value also carries the effect of default.

The second row of the table is the present value that the promised debt payment would have if it were risk-free, $De^{-rfT}$. The difference between $De^{-rfT}$ and $B_0$ is the credit-risk value gap. It measures how much present value is lost because debt holders may receive less than $D$ at maturity.

We calculate this gap because it is the bridge between the Merton balance sheet and credit-market language. Once we know how much value is lost because of default risk, we can express that loss as a percentage, connect it to $PD^Q\times LGD$, infer an implied recovery rate, and later translate the same debt value into a yield spread.

$$
De^{-rfT}-B_0
=
`r fmt_num(pv_promised_payment_debt, 4)`-
`r fmt_num(market_value_debt, 4)`
=
`r fmt_num(credit_value_gap, 4)`.
$$

At this point, the gap is still measured in value units. In the example, Merton prices the debt at `r fmt_num(market_value_debt, 4)` instead of the risk-free present value `r fmt_num(pv_promised_payment_debt, 4)`. The difference, `r fmt_num(credit_value_gap, 4)`, is the amount of present value removed by default risk.

To connect this value gap with default probabilities and recovery rates, we need to put it in rate form. A probability is a rate, and loss given default is also a rate. Therefore, divide the value gap by the risk-free present value of the promised payment:

$$
\text{credit-risk value discount}
=
\frac{De^{-rfT}-B_0}{De^{-rfT}}.
$$

Substituting the numbers gives:

$$
\text{credit-risk value discount}
=
\frac{`r fmt_num(credit_value_gap, 4)`}
{`r fmt_num(pv_promised_payment_debt, 4)`}
=
`r fmt_num(implied_loss_rate, 6)`
=
`r fmt_num(100 * implied_loss_rate, 4)`\%.
$$

This `r fmt_num(100 * implied_loss_rate, 4)`% is not a default probability. It says that default risk removes `r fmt_num(100 * implied_loss_rate, 4)`% of the risk-free present value of the promised payment. It combines two forces, the probability of default and the severity of the loss if default occurs.

Credit-risk models often write expected loss as default probability times loss given default. Here the probability is the Merton risk-neutral PD, so the value discount can be read as a risk-neutral expected loss rate:

$$
\text{credit-risk value discount}
=
PD^Q\times LGD
=
PD^Q(1-R).
$$

The Merton model has already produced the risk-neutral default probability, $PD^Q=`r fmt_num(100 * pd, 4)`\%$. The model has not separately given us a fixed recovery rate. In the structural model, recovery in default depends on the terminal asset value $V_T$. The calculation below therefore backs out a constant-equivalent $LGD$. This is the single loss-given-default rate that reproduces the Merton debt value when paired with $PD^Q$.

Solve the expected-loss relation for $LGD$:

$$
LGD^{impl}
=
\frac{\text{credit-risk value discount}}{PD^Q}
=
\frac{`r fmt_num(100 * implied_loss_rate, 4)`\%}
{`r fmt_num(100 * pd, 4)`\%}
=
`r fmt_num(100 * implied_lgd, 4)`\%.
$$

The constant-equivalent implied recovery rate is the remaining part of the promised value after this implied loss:

$$
R^{impl}
=
1-LGD^{impl}
=
1-`r fmt_num(implied_lgd, 6)`
=
`r fmt_num(100 * implied_recovery_rate, 4)`\%.
$$

This means the Merton debt value behaves, in this simplified expected-loss reading, as if default occurs with risk-neutral probability `r fmt_num(100 * pd, 4)`% and the constant-equivalent recovery rate needed to reproduce the Merton debt value is `r fmt_num(100 * implied_recovery_rate, 4)`% of the promised value. The final row checks the calculation by reconstructing the debt value. Start with the risk-free present value of the promise and subtract the risk-neutral expected loss rate:

$$
De^{-rfT}\left[1-PD^Q(1-R^{impl})\right]
=
`r fmt_num(pv_promised_payment_debt, 4)`
\left[
1-
`r fmt_num(pd, 7)`
\left(1-`r fmt_num(implied_recovery_rate, 6)`\right)
\right]
=
`r fmt_num(reconstructed_debt_value, 4)`,
$$

which matches the model-implied market value of debt, $B_0=`r fmt_num(market_value_debt, 4)`$.

Could we figure out the bond yield spread implied by the Merton model? Yes. The cleanest way to do it here is to use the already computed model-implied market value of debt, $B_0=`r fmt_num(market_value_debt, 4)`$.

In this simplified Merton setup, debt is treated like a zero-coupon promise to pay $D$ at maturity $T$. The continuously compounded yield $y$ on that debt is the value that satisfies:

$$
B_0=De^{-yT}.
$$

Solving for $y$ gives:

$$
y=\frac{1}{T}\log\left(\frac{D}{B_0}\right).
$$

The Merton-implied credit spread is then the difference between this implied debt yield and the risk-free rate:

$$
s=y-rf.
$$

The code below calculates these quantities.

```{r}
merton_implied_debt_yield <- (1 / TT) * log(D / market_value_debt)
merton_implied_spread <- merton_implied_debt_yield - rf
spread_summary <- data.frame(
  Measure = c(
    "Model-implied market value of debt",
    "Promised debt payment at maturity",
    "Risk-free rate",
    "Implied debt yield",
    "Merton-implied credit spread",
    "Merton-implied credit spread in basis points"
  ),
  Value = c(
    fmt_num(market_value_debt, 4),
    fmt_num(D, 4),
    fmt_pct(rf, 4),
    fmt_pct(merton_implied_debt_yield, 4),
    fmt_pct(merton_implied_spread, 4),
    fmt_num(10000 * merton_implied_spread, 1)
  )
)
knitr::kable(
  spread_summary,
  caption = "Merton-implied bond yield and credit spread.",
  row.names = FALSE
)
```

Numerically, the model-implied market value of debt is `r fmt_num(market_value_debt, 4)`. Since the promised debt payment is `r fmt_num(D, 0)` and maturity is `r fmt_num(TT, 0)` in this example, the implied debt yield is `r fmt_pct(merton_implied_debt_yield, 4)`. With a risk-free rate of `r fmt_pct(rf, 2)`, the Merton-implied spread is `r fmt_pct(merton_implied_spread, 4)`, or about `r fmt_num(10000 * merton_implied_spread, 1)` basis points.

This is a model-implied benchmark. It is the spread that makes the model-implied value of debt consistent with the promised debt payment at maturity. In practice, an observed corporate bond spread can differ from this benchmark because of liquidity, taxes, seniority, covenants, multiple debt maturities, and broader risk-premium effects. The comparison is still useful. If the market spread is much higher than the Merton-implied spread, the bond is paying more compensation than this structural default model suggests. If it is much lower, the bond is paying less.

Later sections develop that comparison directly. They take default probabilities, recovery assumptions, corporate bond prices, and CDS spreads, then ask how default risk becomes market compensation.

