## Model-implied PD sensitivity analyses

```{r merton-pd-parameters-setup, include=FALSE}
source("R/merton-pd-function.R")
```

The previous section gives us a convenient function, $pd = f(E_0, \sigma_E, rf, T, D)$. These five observable inputs are used to estimate $V_0$ and $\sigma_V$ before calculating $pd$.

Throughout this section, $pd$ means the Merton risk-neutral/implied probability of default. The exercises below are model-implied sensitivity analyses. We change one or two observable inputs, keep the remaining observable inputs and model assumptions fixed, solve again for the unobservable $V_0$ and $\sigma_V$, and then recompute $N(-d_2)$ under the risk-neutral distribution. The asset quantities $V_0$ and $\sigma_V$ are therefore part of the new Merton solution at each point.

The function below makes that workflow explicit. Every time we call `pd(E0, se, rf, TT, D)`, the code rebuilds equations 24.3 and 24.4, solves again for $V_0$ and $\sigma_V$ with `optim()`, and then returns $N(-d_2)$. This is why the sensitivity curves are model-implied recalculations rather than simple one-variable plots that hold the hidden asset quantities fixed.

```{r merton-pd-function-visible}
pd <- function(E0, se, rf, TT, D) {
  d1_merton <- function(V0, sv) {
    (log(V0 / D) + (rf + sv^2 / 2) * TT) / (sv * sqrt(TT))
  }

  d2_merton <- function(V0, sv) {
    d1_merton(V0, sv) - sv * sqrt(TT)
  }

  eq24.3 <- function(V0, sv) {
    V0 * pnorm(d1_merton(V0, sv)) -
      D * exp(-rf * TT) * pnorm(d2_merton(V0, sv)) -
      E0
  }

  eq24.4 <- function(V0, sv) {
    pnorm(d1_merton(V0, sv)) * sv * V0 - se * E0
  }

  objective <- function(x) {
    eq24.3(x[1], x[2])^2 + eq24.4(x[1], x[2])^2
  }

  start_V0 <- E0 + D * exp(-rf * TT)
  start_sv <- se * E0 / start_V0

  solution <- optim(
    par = c(V0 = start_V0, sv = start_sv),
    fn = objective,
    method = "L-BFGS-B",
    lower = c(V0 = 1e-6, sv = 1e-6),
    control = list(
      parscale = c(V0 = 10, sv = 0.2),
      ndeps = c(V0 = 1e-6, sv = 1e-8)
    )
  )

  V0_hat <- unname(solution$par["V0"])
  sv_hat <- unname(solution$par["sv"])

  pnorm(-d2_merton(V0_hat, sv_hat))
}
```

```{r}
# Evaluate the case of the textbook example.
pd1 <- pd(3, 0.8, 0.05, 1, 10)
knitr::kable(
  data.frame(
    Quantity = c(
      "Observed equity value",
      "Observed equity volatility",
      "Risk-free rate",
      "Debt maturity",
      "Promised debt payment",
      "Merton risk-neutral PD"
    ),
    Value = c(
      fmt_num(E0, 0),
      fmt_pct(se, 0),
      fmt_pct(rf, 0),
      fmt_num(TT, 0),
      fmt_num(D, 0),
      fmt_pct(pd1, 4)
    )
  ),
  caption = "Original Hull/Merton input position for the sensitivity analysis.",
  row.names = FALSE
)
```

The code below creates input grids and evaluates the `pd` function across those grids.

```{r}
l = 50 # Vectors of length 50.
# Create vectors of the parameters.
x.E <- seq(from = 1, to = 20, length.out = l)
x.rf <- seq(0, 0.20, length.out = l)
x.D <- seq(1, 20, length.out = l)
x.T <- seq(0.5, 20, length.out = l)
x.se <- seq(0.01, 3, length.out = l)
# Evaluate the pd at different values.
pd.E <- mapply(pd, x.E, se, rf, TT, D)
pd.rf <- mapply(pd, E0, se, x.rf, TT, D)
pd.D <- mapply(pd, E0, se, rf, TT, x.D)
pd.T <- mapply(pd, E0, se, rf, x.T, D)
pd.se <- mapply(pd, E0, x.se, rf, TT, D)

plot_pd_sensitivity <- function(x, y, current_x, x_label,
                                y_limits = NULL, x_as_percent = FALSE,
                                y_digits = 0) {
  plot_data <- data.frame(
    input_value = x,
    risk_neutral_pd = y
  )
  current_point <- data.frame(
    input_value = current_x,
    risk_neutral_pd = pd1
  )
  if (is.null(y_limits)) {
    y_limits <- c(0, max(c(y, pd1), na.rm = TRUE) * 1.06)
  }

  p <- ggplot2::ggplot(
    plot_data,
    ggplot2::aes(x = input_value, y = risk_neutral_pd)
  ) +
    ggplot2::geom_hline(
      yintercept = 0,
      linewidth = 0.35,
      color = "gray82"
    ) +
    ggplot2::geom_line(linewidth = 1.45, color = "#0b5cad") +
    ggplot2::geom_vline(
      xintercept = current_x,
      linetype = "dashed",
      linewidth = 0.55,
      color = "gray35"
    ) +
    ggplot2::geom_hline(
      yintercept = pd1,
      linetype = "dashed",
      linewidth = 0.55,
      color = "gray35"
    ) +
    ggplot2::geom_point(
      data = current_point,
      ggplot2::aes(x = input_value, y = risk_neutral_pd),
      shape = 21,
      size = 3.6,
      stroke = 1.1,
      color = "#b84300",
      fill = "white"
    ) +
    ggplot2::coord_cartesian(ylim = y_limits) +
    ggplot2::scale_y_continuous(labels = function(x) fmt_pct(x, y_digits)) +
    ggplot2::labs(x = x_label, y = "Risk-neutral PD") +
    ggplot2::theme_classic(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.y = ggplot2::element_line(
        color = "gray90",
        linewidth = 0.25
      ),
      panel.grid.minor = ggplot2::element_blank()
    )

  if (x_as_percent) {
    p <- p + ggplot2::scale_x_continuous(labels = function(x) fmt_pct(x, 0))
  }

  p
}
```

The risk-free-rate grid ranges from 0% to 20%. This keeps the plot focused on a moderate range around the textbook example and away from extreme interest-rate environments.

::: {.merton-subhead}
One-input sensitivity curves
:::

The first curve asks how the risk-neutral PD changes as the observed equity value $E_0$ changes. The code evaluates the probability-of-default function 50 times, once for each value of $E_0$ from 1 to 20. At each value of $E_0$, the code re-estimates $V_0$ and $\sigma_V$ and then computes a new risk-neutral probability of default.

```{r}
#| label: fig-tpodaafoteaatz
#| fig-cap: "The risk-neutral probability of default as a function of the equity at time zero."
plot_pd_sensitivity(
  x = x.E,
  y = pd.E,
  current_x = E0,
  x_label = "Equity at time zero",
  y_limits = c(0, 0.15)
)
```

These relationships between the observable inputs and the risk-neutral probability of default are nonlinear. The current level of each input shapes the result because the same absolute change can have a high or low impact depending on where the firm starts.

```{r}
#| label: fig-tpodaafotrfr
#| fig-cap: "The risk-neutral probability of default as a function of the risk-free rate."
plot_pd_sensitivity(
  x = x.rf,
  y = pd.rf,
  current_x = rf,
  x_label = "Risk-free rate",
  y_limits = range(pd.rf) + c(-0.0006, 0.0006),
  x_as_percent = TRUE,
  y_digits = 1
)
```

The next curve changes the promised debt payment due at maturity, $D$. Increasing or decreasing $D$ by 1 unit can have a different impact on the model-implied probability of default because the curve is nonlinear.

```{r}
#| label: fig-tpodaafotd
#| fig-cap: "The risk-neutral probability of default as a function of the promised debt payment due at maturity."
plot_pd_sensitivity(
  x = x.D,
  y = pd.D,
  current_x = D,
  x_label = "Promised debt payment at maturity",
  y_limits = c(0, 0.15)
)
```

The maturity curve requires a careful reading. In this model-implied sensitivity exercise, changing maturity also triggers a new solution for $V_0$ and $\sigma_V$. For each maturity value on the horizontal axis, the remaining observable inputs are kept fixed, the unobserved asset quantities are re-estimated, and the risk-neutral PD is recomputed. This is a model comparison. A real debt negotiation would also depend on cash flows, covenants, recovery, refinancing risk, and the bank's required spread.

```{r}
#| label: fig-tpodaafotdm
#| fig-cap: "The risk-neutral probability of default as a function of debt maturity."
plot_pd_sensitivity(
  x = x.T,
  y = pd.T,
  current_x = TT,
  x_label = "Maturity"
)
```

The last one-input curve changes the observed equity volatility $\sigma_E$.

```{r}
#| label: fig-tpodaafote
#| fig-cap: "The risk-neutral probability of default as a function of the volatility of the equity."
plot_pd_sensitivity(
  x = x.se,
  y = pd.se,
  current_x = se,
  x_label = "Equity volatility",
  x_as_percent = TRUE
)
```

The five one-input figures are easier to use when their messages are read together. The next table summarizes the grid calculation behind the plots. Each row changes one observed input, solves the Merton equations again, and reports the lowest and highest risk-neutral PD on that grid.

```{r}
sensitivity_summary_table <- data.frame(
  Input = c(
    "$E_0$",
    "$r_f$",
    "$D$",
    "$T$",
    "$\\sigma_E$"
  ),
  `Grid range` = c(
    paste0(fmt_num(min(x.E), 0), " to ", fmt_num(max(x.E), 0)),
    paste0(fmt_pct(min(x.rf), 0), " to ", fmt_pct(max(x.rf), 0)),
    paste0(fmt_num(min(x.D), 0), " to ", fmt_num(max(x.D), 0)),
    paste0(fmt_num(min(x.T), 1), " to ", fmt_num(max(x.T), 0), " years"),
    paste0(fmt_pct(min(x.se), 0), " to ", fmt_pct(max(x.se), 0))
  ),
  `PD range on grid` = c(
    paste0(fmt_pct(min(pd.E, na.rm = TRUE), 2), " to ", fmt_pct(max(pd.E, na.rm = TRUE), 2)),
    paste0(fmt_pct(min(pd.rf, na.rm = TRUE), 2), " to ", fmt_pct(max(pd.rf, na.rm = TRUE), 2)),
    paste0(fmt_pct(min(pd.D, na.rm = TRUE), 2), " to ", fmt_pct(max(pd.D, na.rm = TRUE), 2)),
    paste0(fmt_pct(min(pd.T, na.rm = TRUE), 2), " to ", fmt_pct(max(pd.T, na.rm = TRUE), 2)),
    paste0(fmt_pct(min(pd.se, na.rm = TRUE), 2), " to ", fmt_pct(max(pd.se, na.rm = TRUE), 2))
  ),
  `Financial reading` = c(
    "More equity market value gives the firm a larger cushion above the promised debt payment.",
    "The effect is small in this example because the option valuation equation and asset re-estimation both move.",
    "A higher promised debt payment raises the default boundary.",
    "More time gives the asset value more room to move under the risk-neutral distribution.",
    "Higher observed equity volatility implies a more unstable residual claim and usually higher asset risk."
  )
)

knitr::kable(
  sensitivity_summary_table,
  caption = "Reading the one-input Merton sensitivity curves.",
  escape = FALSE,
  row.names = FALSE
)
```


::: {.merton-subhead}
Two-input views
:::

The previous plots changed one observable input at a time. The inputs were $E_0$, $\sigma_E$, $rf$, $T$ or $D$. We were able to do that because we constructed a vector of 50 different values for these inputs. Again, $V_0$ and $\sigma_V$ are re-estimated each time because they are inferred from $E_0$, $\sigma_E$, $rf$, $T$ and $D$ each time the function is evaluated.

The two-input view lets two market inputs move at the same time, $E_0$ and $\sigma_E$. This is financially useful because equity value measures how much market cushion the firm has above the promised debt payment, while equity volatility measures how unstable that cushion is. A firm with low equity value and high equity volatility should imply a higher probability of default than a firm with high equity value and low equity volatility.

To make the comparison complete, we evaluate all combinations of 50 values of $E_0$ and 50 values of $\sigma_E$. For this two-input view, we cap $E_0$ at 10 and $\sigma_E$ at 150% so the relevant region around the original example remains readable. The result is a $50 \times 50$ grid. Each cell answers a concrete question. If the observed equity value were this level and the observed equity volatility were that level, what Merton risk-neutral PD would the model imply?

```{r}
x.E_two_input <- seq(from = 1, to = 10, length.out = l)
x.se_two_input <- seq(from = 0.01, to = 1.5, length.out = l)
# Create the empty matrix.
p_E_se <- matrix(0, nrow = l, ncol = l)
# Fill the empty matrix with probability of default values.
for(i in 1 : l){ # Is there an easier way to do this?
  for(j in 1 : l){
    p_E_se[i,j] <- mapply(pd, x.E_two_input[i], x.se_two_input[j], rf, TT, D) } }
```

The grid is easiest to read as a contour plot. Each contour line connects combinations of $E_0$ and $\sigma_E$ that produce the same risk-neutral PD after the model solves again for $V_0$ and $\sigma_V$.

```{r}
#| label: fig-tpodaafoteatzatvoeacv
#| fig-cap: "A contour view of the risk-neutral probability of default as a function of equity value and equity volatility."
pd_grid <- expand.grid(
  equity_at_time_zero = x.E_two_input,
  equity_volatility = x.se_two_input
)
pd_grid$risk_neutral_pd <- as.vector(p_E_se)
current_pd_point <- data.frame(
  equity_at_time_zero = E0,
  equity_volatility = se,
  risk_neutral_pd = pd1
)
pd_contour_breaks <- seq(0.05, 0.95, by = 0.05)
pd_contour_label_levels <- c(0.05, 0.10, 0.15, 0.20, 0.40)
pd_contour_label_lines <- contourLines(
  x = x.E_two_input,
  y = x.se_two_input,
  z = p_E_se,
  levels = pd_contour_label_levels
)
pd_contour_label_list <- lapply(pd_contour_label_levels, function(level_value) {
  level_lines <- pd_contour_label_lines[
    vapply(
      pd_contour_label_lines,
      function(line) isTRUE(all.equal(line$level, level_value)),
      logical(1)
    )
  ]
  if (length(level_lines) == 0) {
    return(NULL)
  }
  longest_line <- level_lines[[which.max(vapply(level_lines, function(line) {
    length(line$x)
  }, numeric(1)))]]
  label_index <- max(1, round(length(longest_line$x) * 0.68))
  data.frame(
    equity_at_time_zero = longest_line$x[label_index],
    equity_volatility = longest_line$y[label_index],
    risk_neutral_pd = level_value
  )
})
pd_contour_label_data <- do.call(rbind, pd_contour_label_list)
if (is.null(pd_contour_label_data)) {
  pd_contour_label_data <- data.frame(
    equity_at_time_zero = numeric(),
    equity_volatility = numeric(),
    risk_neutral_pd = numeric()
  )
}
pd_contour_label_data$label <- fmt_pct(
  pd_contour_label_data$risk_neutral_pd,
  0
)

ggplot2::ggplot(
  pd_grid,
  ggplot2::aes(
    x = equity_at_time_zero,
    y = equity_volatility,
    z = risk_neutral_pd
  )
) +
  ggplot2::geom_contour(
    ggplot2::aes(color = ggplot2::after_stat(level)),
    breaks = pd_contour_breaks,
    linewidth = 0.85
  ) +
  ggplot2::geom_label(
    data = pd_contour_label_data,
    ggplot2::aes(
      x = equity_at_time_zero,
      y = equity_volatility,
      label = label
    ),
    inherit.aes = FALSE,
    size = 4,
    label.size = 0,
    fill = "white",
    alpha = 0.86
  ) +
  ggplot2::geom_vline(xintercept = E0, linetype = "dashed", linewidth = 0.55) +
  ggplot2::geom_hline(yintercept = se, linetype = "dashed", linewidth = 0.55) +
  ggplot2::geom_point(
    data = current_pd_point,
    ggplot2::aes(x = equity_at_time_zero, y = equity_volatility),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::scale_y_continuous(
    labels = function(x) fmt_pct(x, 0),
    limits = c(0.4, 1.2),
    breaks = seq(0.4, 1.2, by = 0.2),
    expand = ggplot2::expansion(mult = c(0, 0))
  ) +
  ggplot2::scale_x_continuous(
    limits = c(1, 10),
    breaks = seq(2, 10, by = 2),
    expand = ggplot2::expansion(mult = c(0, 0))
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    end = 0.9,
    guide = "none"
  ) +
  ggplot2::labs(
    x = expression(E[0]),
    y = expression(sigma[E])
  ) +
  ggplot2::theme_classic(base_size = 12)
```      

The red point is the original case, where $E_0=`r fmt_num(E0, 0)`$, $\sigma_E=`r fmt_pct(se, 0)`$ and the model-implied PD is `r fmt_pct(pd1, 4)`. The point lies between the 10% and 15% contour lines, which is consistent with the numerical value. The plot also shows the main financial pattern. Lower equity value and higher equity volatility move the firm toward higher default probabilities.

```{r}
l = 40 # Vectors of length 40.
# Create vectors of the parameters.
x.E <- seq(from = 2, to = 4, length.out = l)
x.se <- seq(0.2, 1.5, length.out = l)
# Create the empty matrix.
p_E_se <- matrix(0, nrow = l, ncol = l)
# Fill the empty matrix with probability of default values.
for(i in 1 : l){ # Is there an easier way to do this?
  for(j in 1 : l){
    p_E_se[i,j] <- mapply(pd, x.E[i], x.se[j], rf, TT, D) } }
```

```{r}
#| label: fig-tpodaafoteatzatvoeaiv
#| fig-cap: "The risk-neutral probability of default as a function of the equity at time zero and the volatility of equity: An interactive view."
plot_ly(type = "surface" , x = x.se, y = x.E , z = p_E_se ) %>%
layout(scene = list(xaxis = list(title = "Volatility of equity"), 
                    yaxis = list(title = "Market value of equity"), 
                    zaxis = list(title = "Risk-neutral PD")))  %>%
  hide_colorbar()
```

