## Minimize a function in R

```{r merton-minimize-setup, include=FALSE}
source("R/merton-packages.R")
source("R/format-helpers.R")
```

The Merton estimation problem is numerical. We observe equity value and equity volatility, but the asset value and asset volatility that appear inside the model are unobserved. The practical task is to choose the unobserved quantities so that the model-implied equity value and equity volatility match the observed ones as closely as possible.

Before solving the Merton equations, it helps to see the same numerical idea in a smaller example. Suppose we want to fit a straight line,

$$
\hat y_i=a+bx_i,
$$

to six observations. For any candidate pair $(a,b)$, the residual for observation $i$ is $a+bx_i-y_i$. The objective function adds the squared residuals:

$$
f(a,b)=\sum_i(a+bx_i-y_i)^2.
$$

The numerical question is direct: which values of $a$ and $b$ make this objective function as small as possible? The data are:

```{r}
# Data.
dat <- data.frame(x = c(1, 2, 3, 4, 5, 6), y = c(1, 3, 5, 6, 8, 12))
kable(dat, caption = "Observations.", row.names = TRUE) # The table.
```

The same six observations are shown in the next figure.

```{r}
#| label: fig-od
#| fig-cap: "Six observations used in a simple minimization example."
ggplot2::ggplot(dat, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_point(size = 3, color = "#0B5CAD") +
  ggplot2::scale_x_continuous(breaks = dat$x) +
  ggplot2::labs(x = "x", y = "y") +
  ggplot2::theme_classic(base_size = 12)
```

The code below defines the residual-sum-of-squares objective and asks `optim()` to search for the values of $a$ and $b$ that minimize it.

```{r}
# Function that calculates the residual sum of squares.
min.RSS <- function(data, par) {
  with(data, sum((par[1] + par[2] * x - y)^2)) }
result <- optim(par = c(0, 1), min.RSS, data = dat)
a <- result$par[1]
b <- result$par[2]
f <- sum((a + b * dat$x - dat$y)^2)
ans <- data.frame(a, b, f)
kable(ans, caption = "Minimization results f(a,b).")
```

The minimization result is a line with intercept $`r fmt_num(a, 6)`$ and slope $`r fmt_num(b, 5)`$. The fitted value for observation $i$ is therefore $\hat y_i=`r fmt_num(a, 6)`+`r fmt_num(b, 5)`x_i$.

The next figure shows the fitted line and the residuals. Each gray segment is the vertical distance between an observed value and the fitted line.

```{r}
#| label: fig-ltmtrsos
#| fig-cap: "Fitted line that minimizes the residual sum of squares."
dat$y_hat <- a + b * dat$x

ggplot2::ggplot(dat, ggplot2::aes(x = x, y = y)) +
  ggplot2::geom_segment(
    ggplot2::aes(xend = x, yend = y_hat),
    color = "gray65",
    linewidth = 0.8
  ) +
  ggplot2::geom_abline(
    intercept = a,
    slope = b,
    color = "#C0392B",
    linewidth = 1.1
  ) +
  ggplot2::geom_point(size = 3, color = "#0B5CAD") +
  ggplot2::scale_x_continuous(breaks = dat$x) +
  ggplot2::labs(x = "x", y = "y") +
  ggplot2::theme_classic(base_size = 12)
```

This objective function is the ordinary least squares criterion. The regression model gives the same answer because it minimizes the same sum of squared residuals.

```{r}
reg <- lm(y ~ x, data = dat)
summary(reg)
```

The Merton objective function is more complex, but the numerical structure is the same. We choose candidate parameters, compute model errors, square those errors, and let an optimizer search for the parameter values that make the objective function small. In the next section, the two parameters are $V_0$ and $\sigma_V$, and the two errors come from Hull's equity-value and equity-volatility equations.

