## Minimize a function in R

```{r merton-minimize-setup, include=FALSE}
source("R/merton-packages.R")
```

The estimation of the credit risk Merton's model requires a minimization of a function. In this section, we show a simple minimization example. Here are some useful online resources for further references.

* [FRM: How d2 in Black-Scholes becomes PD in Merton model.](https://youtu.be/lV0ytJYbVzc)
* [Normal Probability Distribution Graph. Interactive.](https://www.intmath.com/counting-probability/normal-distribution-graph-interactive.php)
* [How to use optim in R.](https://www.r-bloggers.com/how-to-use-optim-in-r/)

The objective function to be minimized is $f(a,b)=\sum(a + bx_i-y_i)^2$. We have six values for $x$ and $y$, and we are expected to minimize the function $f$ by finding the parameter values $a$ and $b$. The $x_i$ and $y_i$ data looks like this:

```{r}
# Data.
dat <- data.frame(x = c(1, 2, 3, 4, 5, 6), y = c(1, 3, 5, 6, 8, 12))
kable(dat, caption = "Observations.", row.names = TRUE) # The table.
```

Graphically:

```{r}
#| label: fig-od
#| fig-cap: "Original data."
# A scatter plot.
plot(y ~ x, data = dat, pch = 19, cex = 2)
```

Let's implement the $f(a,b)=\sum(a + bx_i-y_i)^2$ minimization.

```{r}
# Function that calculates the residual sum of squares.
min.RSS <- function(data, par) {
  with(data, sum((par[1] + par[2] * x - y)^2)) }
# Function minimization.
result <- optim(par = c(0, 1), min.RSS, data = dat)
a <- result$par[1]
b <- result$par[2]
f <- sum((a + b * dat$x - dat$y)^2)
# Minimization output.
ans <- data.frame(a, b, f)
kable(ans, caption = "Minimization results f(a,b).")
```

The minimization result is represented by a line, and the line is characterized by an intercept equal to $-1.266846$ and a slope equal to $2.02862$. 

We can plot it and verify that this is indeed the right answer.

```{r}
#| label: fig-ltmtrsos
#| fig-cap: "Line that minimize the residual sum of squares."
# View the results.
plot(y ~ x, data = dat, pch = 19, ylim = c(-2, 12), xlim = c(0, 6), cex = 2)
abline(a, b, col = "red", lwd = 3)
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
```

The function that we minimize here is basically the OLS criterion. This is why the regression model leads to the same results.

```{r}
# The previous example is equivalent as a linear regression model.
reg <- lm(y ~ x, data = dat)
# See the results.
summary(reg)
```

The function that we minimize in the estimation of the credit risk Merton model is more elaborated, but we follow the same principle as above.

