## Estimating asset value and risk-neutral PD

```{r merton-estimation-setup, include=FALSE}
source("R/merton-packages.R")
source("R/format-helpers.R")
```

::: {.merton-subhead}
Inputs and Hull equations
:::

Let's follow the example 24.3 in [@Hull]. We observe five inputs: $E_0=3$, $\sigma_E=0.8$, $rf=0.05$, $T=1$, and $D=10$. Here, $D$ is the promised debt payment due at maturity, not the current market value of debt. Using these inputs, we estimate the value of the assets today, $V_0$, and the volatility of the assets, $\sigma_V$. We then use those estimated asset quantities to calculate the Merton risk-neutral probability of default $pd$, among other results.

The risk-free rate appears here because equation 24.3 is a valuation equation. We are not assuming that the firm's assets truly earn the risk-free rate in the real world. We are valuing equity under the risk-neutral measure, which is the same logic used to price a European call option. In that artificial valuation world, default is still the event $V_T < D$. After estimating $V_0$ and $\sigma_V$, we will write the risk-neutral probability of that event as $N(-d_2)$.

Therefore, whenever this section says probability of default, it means the model-implied risk-neutral default probability unless we explicitly say otherwise.

These are the known parameters.

```{r}
# List of known parameters.
E0 <- 3 # Value of equity today; this is market capitalization.
se <- 0.8 # Stock return volatility.
rf <- 0.05 # Risk-free rate.
TT <- 1 # Maturity.
D <- 10 # Promised debt payment due at maturity. The Bloomberg function DDIS is useful.
```

The reason $V_0$ and $\sigma_V$ are difficult is that neither is directly observed. Hull's example uses two observable equity quantities, $E_0$ and $\sigma_E$, to infer them. We therefore replicate equations 24.3 and 24.4 from [@Hull] before translating them into R.

Equation 24.3 values equity as a European call option on the firm's assets:

$$
E_0 = V_0N(d_1)-De^{-rT}N(d_2).
\tag{24.3}
$$

The terms $d_1$ and $d_2$ are:

$$
d_1 =
\frac{\ln(V_0/D)+\left(r+\sigma_V^2/2\right)T}
{\sigma_V\sqrt{T}},
\qquad
d_2=d_1-\sigma_V\sqrt{T}.
$$

Here $N(\cdot)$ is the cumulative standard normal distribution. In the code, Hull's $r$ is our `rf`. This equation says: given an asset value $V_0$ and asset volatility $\sigma_V$, the option-pricing formula must reproduce the observed equity value $E_0$.

Hull obtains equation 24.4 from Ito's lemma. It links the observed equity volatility to the unobserved asset volatility:

$$
\sigma_E E_0
=
\frac{\partial E}{\partial V}\sigma_V V_0
=
N(d_1)\sigma_V V_0.
\tag{24.4}
$$

This equation matters because equity is levered. Equity volatility is usually higher than asset volatility, and the scaling depends on the option delta $N(d_1)$ and on the ratio between asset value and equity value. Together, equations 24.3 and 24.4 give two conditions for the two unknowns, $V_0$ and $\sigma_V$. The code below writes each condition as an error. The numerical solution is the pair $(V_0,\sigma_V)$ that makes both errors as close to zero as possible.

The numerical search also needs starting values. These starting values do not impose the answer; they only give the algorithm an economically sensible place to begin. For $V_0$, we start with equity value plus the risk-free present value of the promised debt payment, $E_0+De^{-rfT}$. This does not assume that debt is risk-free. It only gives a reasonable starting scale for the unobserved asset value. For $\sigma_V$, we use a simple deleveraging guess, $\sigma_E E_0/(E_0+De^{-rfT})$, because asset volatility is usually lower than equity volatility when equity is levered.

::: {.merton-subhead}
Solving the two Merton equations
:::

The optimizer is then run with `L-BFGS-B` and positive lower bounds. This keeps both $V_0$ and $\sigma_V$ in economically meaningful territory during the search. Because $V_0$ is measured in units of value while $\sigma_V$ is a decimal volatility, the code also gives `optim()` scale information for the numerical search.

```{r}
start_V0 <- E0 + D * exp(-rf * TT)
start_sv <- se * E0 / start_V0

d1_merton <- function(V0, sv) {
  (log(V0 / D) + (rf + sv^2 / 2) * TT) / (sv * sqrt(TT))
}

d2_merton <- function(V0, sv) {
  d1_merton(V0, sv) - sv * sqrt(TT)
}

eq24.3 <- function(V0, sv) { 
  V0 * pnorm(d1_merton(V0, sv)) -
    D * exp(-rf * TT) * pnorm(d2_merton(V0, sv)) -
    E0
}

eq24.4 <- function(V0, sv) { 
  pnorm(d1_merton(V0, sv)) * sv * V0 - se * E0
}

# Footnote 10 indicates that we should minimize F(x,y)^2+G(x,y)^2
min.footnote10 <- function(x) 
  (eq24.3(x[1], x[2]))^2 + (eq24.4(x[1], x[2]))^2
# The minimization leads to the values of V0 and sv.
V0_sv <- optim(
  par = c(V0 = start_V0, sv = start_sv),
  fn = min.footnote10,
  method = "L-BFGS-B",
  lower = c(V0 = 1e-6, sv = 1e-6),
  control = list(
    parscale = c(V0 = 10, sv = 0.2),
    ndeps = c(V0 = 1e-6, sv = 1e-8)
  )
)
# Define the values as parameters.
V0 <- unname(V0_sv$par["V0"])
sv <- unname(V0_sv$par["sv"])
kable(data.frame("Market value" = V0, 
                 "Volatility" = sv), 
      caption = "Assets.")
```

In this example, those formulas give starting values of about `r fmt_num(start_V0, 4)` for $V_0$ and `r fmt_num(start_sv, 4)` for $\sigma_V$. The final estimates are allowed to move away from those starting values.

These two estimates are the bridge to the rest of the section. First, $V_0$ and $\sigma_V$ determine the risk-neutral default weight $N(-d_2)$. Then they imply the market value of debt, which lets us infer the Merton-implied bond yield spread.

::: {.merton-subhead}
From d2 to the risk-neutral PD
:::

Calculate the Merton risk-neutral probability of default as a function of $V_0$ and $\sigma_V$. In Hull's notation, this probability is $N(-d_2)$. The next paragraphs unpack what that notation means.

```{r}
PD <- function(V0, sv) { 
  pnorm(-(((log(V0 / D) + (rf + sv^2 / 2) * TT) / 
             (sv * sqrt(TT))) - sv * sqrt(TT))) }
# Calculate the risk-neutral probability of default given the previous parameters.
pd <- PD(V0, sv)
pd
```
The risk-neutral probability of default is `r fmt_num(pd, 7)`, or `r fmt_pct(pd, 4)`. This number is market-implied by the option-pricing structure of the model. It answers a pricing question: under the risk-neutral distribution used to value equity, how much weight is assigned to default states? A direct historical or real-world default probability would require a separate calibration from risk-neutral probabilities to real-world probabilities.

Let's analyze the role of the values of $d_1$, $d_2$, $N(d_1)$ and $N(d_2)$. First, we isolate these values.

```{r}
# Inspect the role of d and N(d).
d1 <- d1_merton(V0, sv)
d2 <- d2_merton(V0, sv)
Nd1 <- pnorm(d1)
Nd2 <- pnorm(d2)
ans <- data.frame(d1, Nd1, d2, Nd2)
ans
```

Before connecting this result with the figure, unpack the notation. The symbol $N(\cdot)$ means the cumulative standard normal distribution. In plain words, $N(x)$ is the area to the left of $x$ under the standard normal curve. The value $d_2$ comes from the Black-Scholes-Merton formula. In this chapter, $d_2$ has a concrete role: it is the cutoff that separates survival from default after the model writes future asset values in standard normal units.

The red default area will be $N(-d_2)$, the area to the left of $-d_2$. The green survival area is the rest of the curve. Because the standard normal distribution is symmetric, that survival area is $1-N(-d_2)=N(d_2)$.

Now we can bring back the generic risk-neutral example. In the simple one-period example, the artificial probability $q=0.80$ was the pricing weight of the survival branch, and $1-q=0.20$ was the pricing weight of the default branch. In the Merton model, a continuous distribution of possible future asset values $V_T$ replaces the two branches. The numbers are different because this is now the Hull/Merton example, but the interpretation is the same:

| One-period example | Merton model |
|---|---|
| Survival branch | Region where $V_T>D$ |
| Default branch | Region where $V_T<D$ |
| Survival pricing weight: $q=0.80$ | Survival pricing weight: $N(d_2)$ = `r fmt_num(Nd2, 7)` |
| Default pricing weight: $1-q=0.20$ | Default pricing weight: $N(-d_2)$ = `r fmt_num(pd, 7)` |

This is why the `r fmt_pct(pd, 4)` probability should be read as a risk-neutral default weight, not as an empirical default frequency. It is the continuous-model version of the bad branch in the generic example.

Now, let's illustrate how $N(-d_2)$ leads to the risk-neutral probability of default given the properties of a standard normal distribution.

This is the continuous version of the one-period example. In the one-period example, there was one default branch, so the risk-neutral probability of default was simply $1-q$. In Merton, there is a full range of possible values for $V_T$, and every value below $D$ is a default state.

The model converts the default condition into standard normal units. In asset-value terms, default is:

$$
V_T<D.
$$

In standardized normal terms, the same event is:

$$
Z<-d_2.
$$

This is only a change of scale. Instead of looking at future assets in units of money, we look at the same event in standard normal units. In our example, $d_2$ is positive, so default is the left-tail event $Z<-d_2$. Therefore:

$$
P^Q(V_T<D)=P^Q(Z<-d_2)=N(-d_2).
$$

Figure @fig-trartpod shows exactly this area in red. The red area is the continuous-model equivalent of the bad branch in the one-period example.

```{r}
#| label: fig-trartpod
#| fig-cap: "The red area represents the risk-neutral probability of default."
xseq <- seq(-4, 4, 0.01) 
densities <- dnorm(xseq, 0, 1) # Standard normal distribution.
legend_nd2_expression <- paste0("N(d[2])==", round(Nd2, 7))
legend_pd_expression <- paste0("N(-d[2])==", round(pd, 7), "~'in red.'")
normal_density_data <- data.frame(
  x = xseq,
  density = densities
)
left_tail_data <- normal_density_data[normal_density_data$x < -d2, ]
right_tail_data <- normal_density_data[normal_density_data$x >= -d2, ]

ggplot2::ggplot(normal_density_data, ggplot2::aes(x = x, y = density)) +
  ggplot2::geom_area(
    data = right_tail_data,
    fill = "green",
    color = NA
  ) +
  ggplot2::geom_area(
    data = left_tail_data,
    fill = "red",
    color = NA
  ) +
  ggplot2::geom_line(linewidth = 0.9, color = "black") +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, color = "black") +
  ggplot2::geom_vline(
    xintercept = -d2,
    linetype = "dotted",
    linewidth = 1.2,
    color = "black"
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.377,
    hjust = 0,
    label = legend_nd2_expression,
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.352,
    hjust = 0,
    label = "'in green and'",
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.327,
    hjust = 0,
    label = legend_pd_expression,
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::annotate(
    "text",
    x = -3.80,
    y = 0.305,
    hjust = 0,
    label = "N(d[2])+N(-d[2])==1",
    parse = TRUE,
    size = 3.4
  ) +
  ggplot2::coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.42), expand = FALSE) +
  ggplot2::labs(
    x = expression(paste("d values (", -d[2], " is the dotted line)")),
    y = "Density"
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Here you can find an alternative explanation about $N(-d_2)$:

```{r}
embed_url("https://youtu.be/lV0ytJYbVzc")
```

::: {.merton-subhead}
Checking the numerical solution
:::

Now, let's analyze the minimization that leads to $V_0$ and $\sigma_V$. Equations 24.3 and 24.4 in [@Hull] are solved numerically in this implementation, so $V_0$ and $\sigma_V$ come from minimizing the squared pricing errors. To see how good this approximation is, we can retrieve the value of $[F(x,y)]^2+[G(x,y)]^2$ evaluated at $x=V_0$ and $y=\sigma_V$. This objective function is positive because both $F(x,y)$ and $G(x,y)$ are squared, so a good solution should produce a value close to zero.

```{r}
# Evaluate how good is the minimization.
min.footnote10(x = c(V0, sv))
round(min.footnote10(x = c(V0, sv)), 6)
```

The constrained optimization above worked well: the objective function is zero in practical terms.

There is another way to see how the minimization works. We can propose a graphical analysis. In particular, we can verify whether the estimated values of $V_0$ and $\sigma_V$ minimize the objective function. To illustrate this on two axes, we need to fix either $V_0$ or $\sigma_V$, and evaluate $[F(x,y)]^2+[G(x,y)]^2$ with different values of $V_0$ or $\sigma_V$. Let's do that.

```{r}
# Fix sv and evaluate different values of V0.
V0.seq <- seq(from = 4, to = 20, length.out = 100)
sv.rep <- rep(sv, 100)
# Function to be evaluated.
FG <- function(V0, sv) { (eq24.3(V0, sv))^2 + (eq24.4(V0, sv))^2 }
# Apply the function with fixed sv and different V0 values.
FG.V0 <- mapply(FG, V0.seq, sv.rep)
```

We are ready to plot.

```{r}
#| label: fig-hwftvotaactvoaatz
#| fig-cap: "Here we fix the volatility of the assets and change the value of the assets at time zero. The minimization looks fine."
objective_at_solution <- FG(V0, sv)
slice_v0_data <- data.frame(
  asset_value = V0.seq,
  objective = FG.V0
)
solution_v0_data <- data.frame(
  asset_value = V0,
  objective = objective_at_solution
)

ggplot2::ggplot(slice_v0_data, ggplot2::aes(x = asset_value, y = objective)) +
  ggplot2::geom_line(linewidth = 1.1, color = "black") +
  ggplot2::geom_vline(xintercept = V0, linetype = "dashed", linewidth = 0.7) +
  ggplot2::geom_hline(
    yintercept = objective_at_solution,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    data = solution_v0_data,
    ggplot2::aes(x = asset_value, y = objective),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::labs(
    x = expression(V[0]),
    y = expression(F(x, y)^2 + G(x, y)^2)
  ) +
  ggplot2::theme_classic(base_size = 12)
```

@fig-hwftvotaactvoaatz is a one-dimensional slice of the minimization problem. In this slice, $\sigma_V$ is held fixed at the estimated value and only $V_0$ changes. The vertical dashed line marks the estimated asset value, and the red circle marks the value of the objective function at that estimate. The curve reaches its lowest point near the estimated $V_0$, so nearby alternatives for the asset value produce larger squared errors in equations 24.3 and 24.4. Economically, moving away from this $V_0$ makes it harder for the model to match both the observed equity value and the observed equity volatility.

This figure is a diagnostic check rather than a separate estimation method. It shows that, conditional on the estimated $\sigma_V$, the optimizer has found a sensible asset value. For completeness, we now make the symmetric check by fixing $V_0$ and varying $\sigma_V$.

```{r}
#| label: fig-hwftvotaatzactvota
#| fig-cap: "Here we fix the value of the assets at time zero and change the volatility of the assets. The minimization looks fine."
# Now fix V0, and evaluate different sv values.
sv.seq <- seq(0.001, 0.6, length.out = 100)
V0.rep <- rep(V0, 100)
# Evaluate the function with these parameters.
FG.sv <- mapply(FG, V0.rep, sv.seq)
slice_sv_data <- data.frame(
  asset_volatility = sv.seq,
  objective = FG.sv
)
solution_sv_data <- data.frame(
  asset_volatility = sv,
  objective = objective_at_solution
)

ggplot2::ggplot(
  slice_sv_data,
  ggplot2::aes(x = asset_volatility, y = objective)
) +
  ggplot2::geom_line(linewidth = 1.1, color = "black") +
  ggplot2::geom_vline(xintercept = sv, linetype = "dashed", linewidth = 0.7) +
  ggplot2::geom_hline(
    yintercept = objective_at_solution,
    linetype = "dashed",
    linewidth = 0.7
  ) +
  ggplot2::geom_point(
    data = solution_sv_data,
    ggplot2::aes(x = asset_volatility, y = objective),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::labs(
    x = expression(sigma[V]),
    y = expression(F(x, y)^2 + G(x, y)^2)
  ) +
  ggplot2::theme_classic(base_size = 12)
```

@fig-hwftvotaatzactvota gives the second slice. Now $V_0$ is fixed at the estimated value and $\sigma_V$ changes. The estimated asset volatility again sits near the bottom of the curve. If $\sigma_V$ is too low, the model understates how strongly equity reacts to asset risk. If $\sigma_V$ is too high, the model overstates that risk transmission. Both mistakes increase the objective function because the two Hull equations are no longer matched as closely.

Taken together, @fig-hwftvotaactvoaatz and @fig-hwftvotaatzactvota show that the numerical solution is locally coherent in each parameter direction. We can also evaluate the objective function over a grid where $V_0$ and $\sigma_V$ move at the same time.

```{r}
# outer evaluates the function for each combination of the vectors.
z <- outer(V0.seq, sv.seq, FG) # z is now 100x100 matrix.
```
Contour plots or level plots are a way to show a three-dimensional surface on a two-dimensional plane.

```{r}
#| label: fig-motfacv
#| fig-cap: "Minimization of the function: a contour view."
contour_data <- expand.grid(
  asset_value = V0.seq,
  asset_volatility = sv.seq
)
contour_data$objective <- as.vector(z)
solution_surface_data <- data.frame(
  asset_value = V0,
  asset_volatility = sv
)
contour_breaks <- c(
  0.001, 0.01, 0.05, 0.1, 0.25, 0.5,
  1, 2, 5, 10, 20, 50, 100
)
contour_label_levels <- c(1, 5, 10, 50, 100)
contour_label_lines <- contourLines(
  x = V0.seq,
  y = sv.seq,
  z = z,
  levels = contour_label_levels
)
contour_label_data <- do.call(
  rbind,
  lapply(contour_label_levels, function(level_value) {
    level_lines <- contour_label_lines[
      vapply(
        contour_label_lines,
        function(line) isTRUE(all.equal(line$level, level_value)),
        logical(1)
      )
    ]
    longest_line <- level_lines[[which.max(vapply(level_lines, function(line) {
      length(line$x)
    }, numeric(1)))]]
    label_index <- round(length(longest_line$x) * 0.72)
    data.frame(
      asset_value = longest_line$x[label_index],
      asset_volatility = longest_line$y[label_index],
      objective = level_value
    )
  })
)
contour_label_data$label <- sub(
  "\\.0$",
  "",
  format(contour_label_data$objective, trim = TRUE)
)

ggplot2::ggplot(
  contour_data,
  ggplot2::aes(x = asset_value, y = asset_volatility, z = objective)
) +
  ggplot2::geom_contour(
    ggplot2::aes(color = ggplot2::after_stat(level)),
    breaks = contour_breaks,
    linewidth = 0.75
  ) +
  ggplot2::geom_label(
    data = contour_label_data,
    ggplot2::aes(
      x = asset_value,
      y = asset_volatility,
      label = label
    ),
    inherit.aes = FALSE,
    size = 4,
    label.size = 0,
    fill = "white",
    alpha = 0.85
  ) +
  ggplot2::geom_vline(xintercept = V0, linetype = "dashed", linewidth = 0.6) +
  ggplot2::geom_hline(yintercept = sv, linetype = "dashed", linewidth = 0.6) +
  ggplot2::geom_point(
    data = solution_surface_data,
    ggplot2::aes(x = asset_value, y = asset_volatility),
    inherit.aes = FALSE,
    shape = 21,
    size = 4.2,
    stroke = 1.2,
    color = "red",
    fill = "white"
  ) +
  ggplot2::scale_color_viridis_c(
    option = "C",
    end = 0.9,
    trans = "log10",
    guide = "none"
  ) +
  ggplot2::labs(
    x = expression(V[0]),
    y = expression(sigma[V])
  ) +
  ggplot2::theme_classic(base_size = 12)
```

Each contour line connects combinations of $V_0$ and $\sigma_V$ that produce the same value of $[F(x,y)]^2+[G(x,y)]^2$. The numbers printed on selected lines give direct reference values for the objective function. The red point marks the estimated pair. The dashed lines make its two coordinates easier to read. A good numerical solution should sit near the innermost low-objective region because that is where the two Hull equations are matched most closely.

The following plot is useful in HTML because it can be shown as an interactive surface that can be rotated.


```{r}
#| label: fig-motfaiv
#| fig-cap: "Minimization of the function: an interactive view."
plot_ly(type = "surface" , x = sv.seq, y = V0.seq , z = z ) %>%
layout(scene = list(xaxis = list(title = "Assets volatility"), 
                    yaxis = list(title = "Assets", 
                    zaxis = list(title = "f=F^2+G^2"))))  %>%
hide_colorbar()
```

::: {.merton-subhead}
Model-implied debt value and spread
:::

The fitted Merton model also implies additional debt-related quantities. These are not new observed inputs. They are computed from the estimated asset value, the observed equity value, and the promised debt payment due at maturity.

```{r}
# Model-implied debt quantities.
market_value_debt <- V0 - E0
pv_promised_payment_debt <- D * exp(-rf * TT)
implied_loss_rate <- (pv_promised_payment_debt - market_value_debt) /
  pv_promised_payment_debt
implied_recovery_rate <- 1 - (implied_loss_rate / pd)
ans <- data.frame(
  model_implied_debt_value = market_value_debt,
  risk_free_debt_value = pv_promised_payment_debt,
  implied_loss_rate = implied_loss_rate,
  implied_recovery_rate = implied_recovery_rate,
  risk_neutral_pd = pd
)
kable(t(ans), caption = "Model-implied additional quantities.")
```

The risk-free debt value is the value that the promised payment $D$ would have if it were certain. The lower model-implied debt value reflects the credit-risk discount inside the Merton balance sheet.

Could we figure out the bond yield spread implied by the Merton model? Yes. The cleanest way to do it here is to use the model-implied market value of debt directly.

The Merton model gives the asset value $V_0$ and we observe the equity value $E_0$. Since the firm's assets are split between equity and debt claims, the model-implied market value of debt is:

$$
B_0=V_0-E_0.
$$

In this simplified Merton setup, debt is treated like a zero-coupon promise to pay $D$ at maturity $T$. The continuously compounded yield $y$ on that debt is the value that satisfies:

$$
B_0=De^{-yT}.
$$

Solving for $y$ gives:

$$
y=\frac{1}{T}\log\left(\frac{D}{B_0}\right).
$$

The Merton-implied credit spread is then the difference between this implied debt yield and the risk-free rate:

$$
s=y-rf.
$$

Let's calculate these quantities.

```{r}
merton_implied_debt_yield <- (1 / TT) * log(D / market_value_debt)
merton_implied_spread <- merton_implied_debt_yield - rf
ans <- data.frame(
  market_value_debt = market_value_debt,
  promised_debt_payment = D,
  risk_free_rate = rf,
  implied_debt_yield = merton_implied_debt_yield,
  merton_implied_spread = merton_implied_spread,
  spread_bps = 10000 * merton_implied_spread
)
kable(t(ans), caption = "Merton-implied bond yield spread.")
```

Numerically, the model-implied market value of debt is `r fmt_num(market_value_debt, 4)`. Since the promised debt payment is `r fmt_num(D, 0)` and maturity is `r fmt_num(TT, 0)` in this example, the implied debt yield is `r fmt_pct(merton_implied_debt_yield, 4)`. With a risk-free rate of `r fmt_pct(rf, 2)`, the Merton-implied spread is `r fmt_pct(merton_implied_spread, 4)`, or about `r fmt_num(10000 * merton_implied_spread, 1)` basis points.

This is a model-implied benchmark, not a market quote and not a complete corporate bond pricing model. It is the spread that makes the model-implied value of debt consistent with the promised debt payment at maturity. In practice, an observed corporate bond spread can differ from this benchmark because of liquidity, taxes, seniority, covenants, multiple debt maturities, and broader risk-premium effects. Still, the comparison is useful: if the market spread is much higher than the Merton-implied spread, the bond is paying more compensation than this structural default model suggests; if it is much lower, the bond is paying less.

