## Estimating the final XGBoost model

The previous sections prepared the XGBoost matrices and inspected how boosting rounds change predictions for selected applicants. We now estimate the final XGBoost model used in the benchmark. The objects `xgb_train`, `xgb_test`, `xgb_train_label`, and `xgb_test_label` already contain the encoded training and test data. The settings `xgb_params` and `xgb_nrounds` are the same settings used in the checkpoint exercise.

The estimation happens in the call to `xgboost::xgb.train()`. The inputs to that call identify what is being learned, from which data, and for how many boosting rounds. The printed training log is suppressed with `verbose = 0` so the book output stays compact, but the model still stores the final fitted trees and the evaluation history.

```{r}
xgb_estimation_inputs <- data.frame(
  Argument = c(
    "params = xgb_params",
    "data = xgb_train",
    "nrounds = xgb_nrounds",
    "watchlist = list(train = xgb_train, test = xgb_test)",
    "verbose = 0"
  ),
  Role = c(
    "Objective, evaluation metrics, learning rate, depth, regularization, and sampling choices.",
    "Encoded training matrix with the observed default labels.",
    "Number of boosted trees added to the score.",
    "Training and test matrices used to store diagnostic metrics during estimation.",
    "Suppresses console printing; it does not change the estimated model."
  ),
  check.names = FALSE
)

knitr::kable(
  xgb_estimation_inputs,
  caption = "Inputs used by the XGBoost estimation call.",
  row.names = FALSE
)
```

The object `xgb_model` stores the fitted boosted trees. The object `pred_xgb` stores one predicted default probability for each applicant in the test set.

```{r XGBoost model estimation}
set.seed(567)
xgb_model <- xgboost::xgb.train(
  params = xgb_params,
  data = xgb_train,
  nrounds = xgb_nrounds,
  watchlist = list(train = xgb_train, test = xgb_test),
  verbose = 0
)

pred_xgb <- predict(xgb_model, newdata = xgb_test)
ROC_xgb <- pROC::roc(test$loan_st, pred_xgb, quiet = TRUE)
```

After this chunk, the model has been estimated. The next table makes the fitted object visible. The most important out-of-sample quantities are the test-set predictions, test AUC, and test Brier score. The training metrics are useful as diagnostics, but they are not the final criterion for credit-scoring performance.

```{r}
pred_xgb_train <- predict(xgb_model, newdata = xgb_train)
ROC_xgb_train <- pROC::roc(train$loan_st, pred_xgb_train, quiet = TRUE)

xgb_estimation_log_loss <- function(actual, predicted_pd) {
  predicted_pd <- pmin(pmax(predicted_pd, 1e-8), 1 - 1e-8)
  mean(-(actual * log(predicted_pd) +
           (1 - actual) * log(1 - predicted_pd)))
}

xgb_train_auc_estimated <- as.numeric(auc(ROC_xgb_train))
xgb_test_auc_estimated <- as.numeric(auc(ROC_xgb))
xgb_test_logloss_estimated <- xgb_estimation_log_loss(
  actual_default_tree,
  pred_xgb
)

xgb_fitted_summary <- data.frame(
  Quantity = c(
    "Fitted object",
    "Training applicants",
    "Test applicants",
    "Encoded predictors",
    "Boosting rounds",
    "Learning rate",
    "Maximum tree depth",
    "Predicted test-set PDs",
    "Final train AUC",
    "Final test AUC",
    "Final test log-loss",
    "Test Brier score"
  ),
  Value = c(
    "xgb_model",
    fmt_int(nrow(xgb_train_matrix)),
    fmt_int(nrow(xgb_test_matrix)),
    fmt_int(ncol(xgb_train_matrix)),
    fmt_int(xgb_nrounds),
    fmt_num(xgb_params$eta, 2),
    fmt_int(xgb_params$max_depth),
    fmt_int(length(pred_xgb)),
    fmt_num(xgb_train_auc_estimated, 4),
    fmt_num(xgb_test_auc_estimated, 4),
    fmt_num(xgb_test_logloss_estimated, 4),
    fmt_num(brier_score(actual_default_tree, pred_xgb), 4)
  )
)

knitr::kable(
  xgb_fitted_summary,
  caption = "Estimated XGBoost model summary.",
  row.names = FALSE
)
```

The fitted object is valuable because it produces applicant-level PDs. The next two figures make those final predictions visible. The first figure compares the final PD distributions from the full logistic model and XGBoost. The second figure compares each XGBoost PD with the full logistic-model PD from Chapter 1 for the same test-set applicant, with separate panels for observed defaults and non-defaults.

```{r}
#| label: fig-xgb-final-pd-distribution
#| fig-cap: "Final predicted PD distributions for the full logistic model and XGBoost."
xgb_final_pd_plot_data <- data.frame(
  predicted_pd = c(pred_logi_full, pred_xgb),
  model = factor(
    rep(c("Full logistic model", "XGBoost"), each = length(pred_xgb)),
    levels = c("Full logistic model", "XGBoost")
  )
)
xgb_logistic_pd_quantiles <- quantile(
  pred_logi_full,
  probs = c(0.50, 0.90, 0.95, 0.99)
)
xgb_final_pd_quantiles <- quantile(
  pred_xgb,
  probs = c(0.50, 0.90, 0.95, 0.99)
)

ggplot(
  xgb_final_pd_plot_data,
  aes(x = predicted_pd, color = model, fill = model)
) +
  geom_density(
    alpha = 0.20,
    linewidth = 1.05
  ) +
  scale_x_continuous(labels = fmt_pct) +
  scale_color_manual(
    values = c("Full logistic model" = "#1F618D", "XGBoost" = "#C0392B")
  ) +
  scale_fill_manual(
    values = c("Full logistic model" = "#1F618D", "XGBoost" = "#C0392B")
  ) +
  labs(
    x = "Predicted PD",
    y = "Density",
    color = "Model",
    fill = "Model"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

Both models assign low PDs to many applicants, and the right tail is where the comparison becomes informative. The median PD is `r fmt_pct(xgb_logistic_pd_quantiles[["50%"]], 2)` for `logi_full` and `r fmt_pct(xgb_final_pd_quantiles[["50%"]], 2)` for XGBoost. At the 90th percentile, the corresponding values are `r fmt_pct(xgb_logistic_pd_quantiles[["90%"]], 2)` and `r fmt_pct(xgb_final_pd_quantiles[["90%"]], 2)`. At the 99th percentile, they are `r fmt_pct(xgb_logistic_pd_quantiles[["99%"]], 2)` and `r fmt_pct(xgb_final_pd_quantiles[["99%"]], 2)`. The plot therefore asks a credit-scoring question. Does the challenger model merely reproduce the logistic score, or does it redistribute applicants across the risk scale in a way that may change approvals, rejections, and expected losses?

```{r}
#| label: fig-xgb-logistic-pd-comparison
#| fig-cap: "Full logistic-model PDs compared with final XGBoost PDs by observed outcome."
xgb_logistic_pd_comparison <- data.frame(
  logistic_pd = pred_logi_full,
  xgboost_pd = pred_xgb,
  observed_outcome = factor(
    actual_default_tree,
    levels = c(0, 1),
    labels = c("No default", "Default")
  )
)
xgb_pd_comparison_limit <- max(
  xgb_logistic_pd_comparison$logistic_pd,
  xgb_logistic_pd_comparison$xgboost_pd
) * 1.04
xgb_above_logistic_share <- mean(
  xgb_logistic_pd_comparison$xgboost_pd >
    xgb_logistic_pd_comparison$logistic_pd
)
xgb_below_logistic_share <- mean(
  xgb_logistic_pd_comparison$xgboost_pd <
    xgb_logistic_pd_comparison$logistic_pd
)
xgb_default_rows <- xgb_logistic_pd_comparison$observed_outcome == "Default"
xgb_nondefault_rows <- xgb_logistic_pd_comparison$observed_outcome == "No default"
xgb_above_logistic_default_share <- mean(
  xgb_logistic_pd_comparison$xgboost_pd[xgb_default_rows] >
    xgb_logistic_pd_comparison$logistic_pd[xgb_default_rows]
)
xgb_below_logistic_nondefault_share <- mean(
  xgb_logistic_pd_comparison$xgboost_pd[xgb_nondefault_rows] <
    xgb_logistic_pd_comparison$logistic_pd[xgb_nondefault_rows]
)
xgb_default_median_logistic_pd <- median(
  xgb_logistic_pd_comparison$logistic_pd[xgb_default_rows]
)
xgb_default_median_xgboost_pd <- median(
  xgb_logistic_pd_comparison$xgboost_pd[xgb_default_rows]
)
xgb_nondefault_median_logistic_pd <- median(
  xgb_logistic_pd_comparison$logistic_pd[xgb_nondefault_rows]
)
xgb_nondefault_median_xgboost_pd <- median(
  xgb_logistic_pd_comparison$xgboost_pd[xgb_nondefault_rows]
)

ggplot(
  xgb_logistic_pd_comparison,
  aes(x = logistic_pd, y = xgboost_pd, color = observed_outcome)
) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    linewidth = 0.65,
    color = "gray35"
  ) +
  geom_point(alpha = 0.22, size = 1.45) +
  facet_wrap(~ observed_outcome, nrow = 1) +
  coord_equal(
    xlim = c(0, xgb_pd_comparison_limit),
    ylim = c(0, xgb_pd_comparison_limit)
  ) +
  scale_x_continuous(labels = fmt_pct) +
  scale_y_continuous(labels = fmt_pct) +
  scale_color_manual(
    values = c("No default" = "gray55", "Default" = "#C0392B"),
    guide = "none"
  ) +
  labs(
    x = "Full logistic-model predicted PD",
    y = "Final XGBoost predicted PD"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

The dashed line is the equal-PD line. Points above the line are applicants for whom XGBoost assigns a higher PD than the logistic benchmark; points below the line receive a lower XGBoost PD. Separating the panels by observed outcome makes the comparison easier to read. Among applicants who defaulted, XGBoost assigns a higher PD than `logi_full` to `r fmt_pct(xgb_above_logistic_default_share, 2)` of cases. Their median PD moves from `r fmt_pct(xgb_default_median_logistic_pd, 2)` under `logi_full` to `r fmt_pct(xgb_default_median_xgboost_pd, 2)` under XGBoost. Among applicants who repaid, XGBoost assigns a lower PD than `logi_full` to `r fmt_pct(xgb_below_logistic_nondefault_share, 2)` of cases. Their median PD moves from `r fmt_pct(xgb_nondefault_median_logistic_pd, 2)` to `r fmt_pct(xgb_nondefault_median_xgboost_pd, 2)`. The visual conclusion is mixed. XGBoost changes the risk ranking locally, while the panels do not show a uniform shift in the desired direction for all realized defaults and non-defaults. This prepares the reader for the broader evidence below. AUC, calibration, bad rates, and payoff must decide whether those local changes improve the credit policy.

The model was estimated with shallow trees and a moderate number of boosting rounds. The checkpoint exercise followed individual applicants; the learning curve below evaluates aggregate performance across the training and test sets. We re-estimate the model at selected boosting rounds and calculate AUC on both samples. The next code builds the data used in the learning-curve figure.

```{r XGBoost learning curve data}
xgb_round_grid <- seq(20, xgb_nrounds, 20)
xgb_learning_curve <- do.call(
  rbind,
  lapply(xgb_round_grid, function(rounds) {
    set.seed(567)
    model_i <- xgboost::xgb.train(
      params = xgb_params,
      data = xgb_train,
      nrounds = rounds,
      verbose = 0
    )

    pred_train_i <- predict(model_i, newdata = xgb_train)
    pred_test_i <- predict(model_i, newdata = xgb_test)

    data.frame(
      rounds = rounds,
      series = c("Train AUC", "Test AUC"),
      auc = c(
        as.numeric(pROC::auc(train$loan_st, pred_train_i, quiet = TRUE)),
        as.numeric(pROC::auc(test$loan_st, pred_test_i, quiet = TRUE))
      )
    )
  })
)
```

```{r include=FALSE}
xgb_final_round <- max(xgb_learning_curve$rounds)
xgb_final_train_auc <- xgb_learning_curve$auc[
  xgb_learning_curve$rounds == xgb_final_round &
    xgb_learning_curve$series == "Train AUC"
]
xgb_final_test_auc <- xgb_learning_curve$auc[
  xgb_learning_curve$rounds == xgb_final_round &
    xgb_learning_curve$series == "Test AUC"
]
xgb_final_auc_gap <- xgb_final_train_auc - xgb_final_test_auc
xgb_test_curve <- xgb_learning_curve[xgb_learning_curve$series == "Test AUC", ]
xgb_train_curve <- xgb_learning_curve[xgb_learning_curve$series == "Train AUC", ]
xgb_best_test_row <- xgb_test_curve[which.max(xgb_test_curve$auc), ]
xgb_previous_test_auc <- xgb_test_curve$auc[nrow(xgb_test_curve) - 1]
xgb_test_auc_last_change <- xgb_final_test_auc - xgb_previous_test_auc
xgb_test_auc_declines_after_best <- xgb_best_test_row$rounds < xgb_final_round &&
  xgb_final_test_auc < xgb_best_test_row$auc
```

```{r}
#| label: fig-xgboost-learning
#| fig-cap: "XGBoost learning curve: test AUC with train AUC as a diagnostic reference."
ggplot(xgb_learning_curve, aes(x = rounds, y = auc, color = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Train AUC" = "steelblue",
                                "Test AUC" = "firebrick")) +
  labs(x = "Boosting round", y = "AUC", color = "Series") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The red test-AUC line is the main out-of-sample quantity in @fig-xgboost-learning. The blue train-AUC line is included as a diagnostic reference. A direct overfitting check has two parts. First, inspect whether test AUC deteriorates as more boosting rounds are added. Second, compare how far train AUC sits above test AUC at the final model.

```{r}
xgb_overfitting_check <- data.frame(
  Check = c(
    "Best test AUC in the grid",
    "Final test AUC",
    "Change in test AUC from previous checkpoint",
    "Final train AUC",
    "Final train-test AUC gap",
    "Test AUC declines after its best checkpoint?"
  ),
  Value = c(
    paste0(
      fmt_num(xgb_best_test_row$auc, 4),
      " at ",
      fmt_int(xgb_best_test_row$rounds),
      " rounds"
    ),
    fmt_num(xgb_final_test_auc, 4),
    fmt_num(xgb_test_auc_last_change, 4),
    fmt_num(xgb_final_train_auc, 4),
    fmt_num(xgb_final_auc_gap, 4),
    ifelse(xgb_test_auc_declines_after_best, "Yes", "No")
  ),
  Reading = c(
    "Highest out-of-sample ranking result among the inspected checkpoints.",
    "Ranking result for the model used in the chapter.",
    "Positive values mean the last added rounds still improved test ranking.",
    "In-sample ranking result.",
    "Larger gaps call for validation, calibration, and monitoring.",
    "A decline would be direct evidence that additional rounds are hurting test ranking."
  ),
  check.names = FALSE
)

knitr::kable(
  xgb_overfitting_check,
  caption = "Direct overfitting diagnostic from the XGBoost learning curve.",
  row.names = FALSE
)
```

In this run, the best inspected test AUC occurs at `r fmt_int(xgb_best_test_row$rounds)` boosting rounds, and the final test AUC is `r fmt_num(xgb_final_test_auc, 4)`. The last inspected movement in test AUC is `r fmt_num(xgb_test_auc_last_change, 4)`. The training AUC is `r fmt_num(xgb_final_train_auc, 4)`, so the final train-test gap is `r fmt_num(xgb_final_auc_gap, 4)`. The evidence therefore supports a cautious reading. The selected grid does not show test-AUC deterioration after the best checkpoint, while the train-test gap still deserves validation and monitoring.

The curve is a diagnostic. A careful optimization workflow would use cross-validation or a separate validation set to select the number of rounds and other hyperparameters, and only then evaluate the final model on the test set.

The distinction between the teaching workflow and a production workflow is important. This chapter uses the same test set repeatedly because the objective is to make the comparison visible. In production, the analyst should separate three tasks. Fit the model, choose tuning parameters, and report final performance.

```{r XGBoost validation workflow}
xgb_validation_workflow <- data.frame(
  Stage = c(
    "Training",
    "Validation or cross-validation",
    "Final holdout test",
    "Monitoring after deployment"
  ),
  `What it answers` = c(
    "Which patterns can the model learn from historical applications?",
    "Which hyperparameters give stable out-of-sample behavior?",
    "How well does the chosen model perform on untouched borrowers?",
    "Do ranking, calibration, bad rates, and approval patterns remain stable over time?"
  ),
  `Typical XGBoost checks` = c(
    "Fit shallow boosted trees with regularization and subsampling.",
    "Tune rounds, depth, learning rate, minimum child weight, subsampling, and calibration.",
    "Report AUC, Brier score, calibration, bad rates, and payoff once.",
    "Track drift, overrides, calibration decay, fairness checks, and reject-inference issues."
  ),
  check.names = FALSE
)

knitr::kable(
  xgb_validation_workflow,
  caption = "Teaching workflow versus production validation for XGBoost credit scoring.",
  row.names = FALSE
)
```

The learning curve should therefore be read as a diagnostic rather than a full tuning exercise. If test AUC rose and then deteriorated, the analyst would stop adding trees. If validation calibration worsened while AUC improved, the analyst might keep the ranking model and recalibrate the probabilities before using them for pricing or provisioning.

```{r XGBoost prediction range}
xgb_prediction_range_table <- data.frame(
  Quantity = c(
    "Minimum predicted PD",
    "Median predicted PD",
    "Maximum predicted PD",
    "AUC",
    "Brier score"
  ),
  Value = c(
    fmt_pct(min(pred_xgb), 2),
    fmt_pct(median(pred_xgb), 2),
    fmt_pct(max(pred_xgb), 2),
    fmt_num(as.numeric(auc(ROC_xgb)), 4),
    fmt_num(brier_score(actual_default_tree, pred_xgb), 4)
  )
)

knitr::kable(
  xgb_prediction_range_table,
  caption = "XGBoost prediction range and test-set performance summary.",
  row.names = FALSE
)
```

The XGBoost model returns predicted probabilities of default, just like the logistic model. In code, those probabilities are stored in `pred_xgb`; mathematically, they are the $\hat p_i$ values used by the same evaluation equations from Chapter 1. Therefore, all the credit-risk tools developed in Chapter 1 still apply. These include cutoffs, acceptance rates, bad rates, calibration, Brier score, and net payoff.

The prediction range table should be read as a risk-segmentation summary. The minimum, median, and maximum predicted PD tell us how widely the model spreads applicants across the risk scale. A model that assigns nearly the same PD to everyone would have little value for screening, even if that average PD were reasonable. A useful scoring model must separate applicants enough to support different decisions.
