## Preparing and estimating the XGBoost model

The previous section inspected how boosting rounds change the prediction for selected applicants. We now show the full estimation workflow behind those objects. The chapter setup already created `xgb_train`, `xgb_test`, and the related labels so the checkpoint exercise could use them. This section prints that construction explicitly and then estimates the XGBoost model used for comparison.

XGBoost requires a numerical matrix. This is why we use `model.matrix()`: it converts categorical variables such as `grade`, `home`, `sex`, and `region` into indicator variables, while keeping numerical variables such as `age`, `int`, `l_amnt`, and `income` in numerical form. This encoding step is important because it makes explicit what the algorithm receives as input. The mathematical object $x_i$ is represented in code by one row of `xgb_train_matrix` or `xgb_test_matrix`; the observed outcome $y_i$ is represented by `xgb_train_label` or `xgb_test_label`.

One row of the matrix is one applicant. Each column is one numerical signal that the algorithm can use when it searches for splits. After this transformation, XGBoost no longer sees a variable called `grade` as a word or category. It sees numerical columns indicating the applicant's grade level. This is the bridge between the credit application data and the algorithm.

The translation from the financial object to the R object is:

```{r XGBoost object bridge}
knitr::kable(
  data.frame(
    step = c(
      "Borrower and loan information",
      "Observed default outcome",
      "Encoded numerical matrix",
      "XGBoost score",
      "Predicted probability of default",
      "Credit decision"
    ),
    R_object = c(
      "`train` and `test`",
      "`loan_st`",
      "`xgb_train_matrix` and `xgb_test_matrix`",
      "`predict(..., outputmargin = TRUE)` conceptually",
      "`pred_xgb`",
      "cutoff or acceptance-rate rule"
    ),
    interpretation = c(
      "Applicant characteristics available before the lending decision",
      "Historical repayment result used for training and evaluation",
      "One row per applicant and one numerical column per model signal",
      "Log-odds scale value produced by the boosted trees",
      "Default-risk estimate after the logistic transformation",
      "Rule that converts the PD into accept or reject"
    )
  ),
  caption = "Bridge between credit-scoring objects and XGBoost objects.",
  escape = FALSE
)
```

```{r XGBoost design matrices}
xgb_formula <- ~ age + int + grade + log(l_amnt) + log(income) +
  emp_len + home + sex + region

xgb_train_matrix <- model.matrix(xgb_formula, data = train)[, -1, drop = FALSE]
xgb_test_matrix <- model.matrix(xgb_formula, data = test)[, -1, drop = FALSE]
xgb_train_label <- as.numeric(as.character(train$loan_st))
xgb_test_label <- actual_default_tree

xgb_train <- xgboost::xgb.DMatrix(
  data = xgb_train_matrix,
  label = xgb_train_label
)
xgb_test <- xgboost::xgb.DMatrix(
  data = xgb_test_matrix,
  label = xgb_test_label
)

xgb_matrix_dimensions <- data.frame(
  Object = c("Training matrix", "Test matrix"),
  Rows = fmt_int(c(nrow(xgb_train_matrix), nrow(xgb_test_matrix))),
  Columns = fmt_int(c(ncol(xgb_train_matrix), ncol(xgb_test_matrix)))
)

knitr::kable(
  xgb_matrix_dimensions,
  caption = "Dimensions of the encoded XGBoost design matrices.",
  row.names = FALSE
)
```

The estimation happens in the call to `xgboost::xgb.train()`. The inputs to that call identify what is being learned, from which data, and for how many boosting rounds. The printed training log is suppressed with `verbose = 0` so the book output stays compact, but the model still stores the final fitted trees and the evaluation history.

```{r}
xgb_estimation_inputs <- data.frame(
  Argument = c(
    "params = xgb_params",
    "data = xgb_train",
    "nrounds = xgb_nrounds",
    "watchlist = list(train = xgb_train, test = xgb_test)",
    "verbose = 0"
  ),
  Role = c(
    "Objective, evaluation metrics, learning rate, depth, regularization, and sampling choices.",
    "Encoded training matrix with the observed default labels.",
    "Number of boosted trees added to the score.",
    "Training and test matrices used to store diagnostic metrics during estimation.",
    "Suppresses console printing; it does not change the estimated model."
  ),
  check.names = FALSE
)

knitr::kable(
  xgb_estimation_inputs,
  caption = "Inputs used by the XGBoost estimation call.",
  row.names = FALSE
)
```

The object `xgb_model` stores the fitted boosted trees. The object `pred_xgb` stores one predicted default probability for each applicant in the test set.

```{r XGBoost model estimation}
set.seed(567)
xgb_model <- xgboost::xgb.train(
  params = xgb_params,
  data = xgb_train,
  nrounds = xgb_nrounds,
  watchlist = list(train = xgb_train, test = xgb_test),
  verbose = 0
)

pred_xgb <- predict(xgb_model, newdata = xgb_test)
ROC_xgb <- pROC::roc(test$loan_st, pred_xgb, quiet = TRUE)
```

After this chunk, the model has been estimated. The next table makes the fitted object visible. The most important out-of-sample quantities are the test-set predictions, test AUC, and test Brier score. The training metrics are useful as diagnostics, but they are not the final criterion for credit-scoring performance.

```{r}
pred_xgb_train <- predict(xgb_model, newdata = xgb_train)
ROC_xgb_train <- pROC::roc(train$loan_st, pred_xgb_train, quiet = TRUE)

xgb_estimation_log_loss <- function(actual, predicted_pd) {
  predicted_pd <- pmin(pmax(predicted_pd, 1e-8), 1 - 1e-8)
  mean(-(actual * log(predicted_pd) +
           (1 - actual) * log(1 - predicted_pd)))
}

xgb_train_auc_estimated <- as.numeric(auc(ROC_xgb_train))
xgb_test_auc_estimated <- as.numeric(auc(ROC_xgb))
xgb_test_logloss_estimated <- xgb_estimation_log_loss(
  actual_default_tree,
  pred_xgb
)

xgb_fitted_summary <- data.frame(
  Quantity = c(
    "Fitted object",
    "Training applicants",
    "Test applicants",
    "Encoded predictors",
    "Boosting rounds",
    "Learning rate",
    "Maximum tree depth",
    "Predicted test-set PDs",
    "Final train AUC",
    "Final test AUC",
    "Final test log-loss",
    "Test Brier score"
  ),
  Value = c(
    "xgb_model",
    fmt_int(nrow(xgb_train_matrix)),
    fmt_int(nrow(xgb_test_matrix)),
    fmt_int(ncol(xgb_train_matrix)),
    fmt_int(xgb_nrounds),
    fmt_num(xgb_params$eta, 2),
    fmt_int(xgb_params$max_depth),
    fmt_int(length(pred_xgb)),
    fmt_num(xgb_train_auc_estimated, 4),
    fmt_num(xgb_test_auc_estimated, 4),
    fmt_num(xgb_test_logloss_estimated, 4),
    fmt_num(brier_score(actual_default_tree, pred_xgb), 4)
  )
)

knitr::kable(
  xgb_fitted_summary,
  caption = "Estimated XGBoost model summary.",
  row.names = FALSE
)
```

The model was estimated with shallow trees and a moderate number of boosting rounds. The checkpoint exercise followed individual applicants; the learning curve below evaluates aggregate performance across the training and test sets. We re-estimate the model at selected boosting rounds and calculate AUC on both samples. The next code builds the data used in the learning-curve figure.

```{r XGBoost learning curve data}
xgb_round_grid <- seq(20, xgb_nrounds, 20)
xgb_learning_curve <- do.call(
  rbind,
  lapply(xgb_round_grid, function(rounds) {
    set.seed(567)
    model_i <- xgboost::xgb.train(
      params = xgb_params,
      data = xgb_train,
      nrounds = rounds,
      verbose = 0
    )

    pred_train_i <- predict(model_i, newdata = xgb_train)
    pred_test_i <- predict(model_i, newdata = xgb_test)

    data.frame(
      rounds = rounds,
      series = c("Train AUC", "Test AUC"),
      auc = c(
        as.numeric(pROC::auc(train$loan_st, pred_train_i, quiet = TRUE)),
        as.numeric(pROC::auc(test$loan_st, pred_test_i, quiet = TRUE))
      )
    )
  })
)
```

```{r include=FALSE}
xgb_final_round <- max(xgb_learning_curve$rounds)
xgb_final_train_auc <- xgb_learning_curve$auc[
  xgb_learning_curve$rounds == xgb_final_round &
    xgb_learning_curve$series == "Train AUC"
]
xgb_final_test_auc <- xgb_learning_curve$auc[
  xgb_learning_curve$rounds == xgb_final_round &
    xgb_learning_curve$series == "Test AUC"
]
xgb_final_auc_gap <- xgb_final_train_auc - xgb_final_test_auc
```

```{r}
#| label: fig-xgboost-learning
#| fig-cap: "XGBoost learning curve: test AUC with train AUC as a diagnostic reference."
ggplot(xgb_learning_curve, aes(x = rounds, y = auc, color = series)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Train AUC" = "steelblue",
                                "Test AUC" = "firebrick")) +
  labs(x = "Boosting round", y = "AUC", color = "Series") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The red test-AUC line is the main out-of-sample quantity in @fig-xgboost-learning. The blue train-AUC line is included as a diagnostic reference. A gap between train and test AUC does not prove overfitting by itself. It tells us to inspect out-of-sample performance, calibration, Brier score, bad rates, and economic payoff before trusting the model. In this specification, we keep the model intentionally restrained: shallow trees, a low learning rate, and subsampling.

At `r xgb_final_round` boosting rounds, the training AUC is `r round(xgb_final_train_auc, 4)` and the test AUC is `r round(xgb_final_test_auc, 4)`. The train-test gap is `r round(xgb_final_auc_gap, 4)`. This is a warning signal to monitor, not a final diagnosis. The test AUC remains the relevant line for out-of-sample ranking performance.

The curve is a diagnostic. A careful optimization workflow would use cross-validation or a separate validation set to select the number of rounds and other hyperparameters, and only then evaluate the final model on the test set.

The distinction between the teaching workflow and a production workflow is important. This chapter uses the same test set repeatedly because the objective is to make the comparison visible. In production, the analyst should separate three tasks: fit the model, choose tuning parameters, and report final performance.

```{r XGBoost validation workflow}
xgb_validation_workflow <- data.frame(
  Stage = c(
    "Training",
    "Validation or cross-validation",
    "Final holdout test",
    "Monitoring after deployment"
  ),
  `What it answers` = c(
    "Which patterns can the model learn from historical applications?",
    "Which hyperparameters give stable out-of-sample behavior?",
    "How well does the chosen model perform on untouched borrowers?",
    "Do ranking, calibration, bad rates, and approval patterns remain stable over time?"
  ),
  `Typical XGBoost checks` = c(
    "Fit shallow boosted trees with regularization and subsampling.",
    "Tune rounds, depth, learning rate, minimum child weight, subsampling, and calibration.",
    "Report AUC, Brier score, calibration, bad rates, and payoff once.",
    "Track drift, overrides, calibration decay, fairness checks, and reject-inference issues."
  ),
  check.names = FALSE
)

knitr::kable(
  xgb_validation_workflow,
  caption = "Teaching workflow versus production validation for XGBoost credit scoring.",
  row.names = FALSE
)
```

The learning curve should therefore be read as a diagnostic, not as a full tuning exercise. If test AUC rose and then deteriorated, the analyst would not keep adding trees. If validation calibration worsened while AUC improved, the analyst might keep the ranking model but recalibrate the probabilities before using them for pricing or provisioning.

```{r XGBoost prediction range}
xgb_prediction_range_table <- data.frame(
  Quantity = c(
    "Minimum predicted PD",
    "Median predicted PD",
    "Maximum predicted PD",
    "AUC",
    "Brier score"
  ),
  Value = c(
    fmt_pct(min(pred_xgb), 2),
    fmt_pct(median(pred_xgb), 2),
    fmt_pct(max(pred_xgb), 2),
    fmt_num(as.numeric(auc(ROC_xgb)), 4),
    fmt_num(brier_score(actual_default_tree, pred_xgb), 4)
  )
)

knitr::kable(
  xgb_prediction_range_table,
  caption = "XGBoost prediction range and test-set performance summary.",
  row.names = FALSE
)
```

The XGBoost model returns predicted probabilities of default, just like the logistic model. In code, those probabilities are stored in `pred_xgb`; mathematically, they are the $\hat p_i$ values used by the same evaluation equations from Chapter 1. Therefore, all the credit-risk tools developed in Chapter 1 still apply: cutoffs, acceptance rates, bad rates, calibration, Brier score, and net payoff.

The prediction range table should be read as a risk-segmentation summary. The minimum, median, and maximum predicted PD tell us how widely the model spreads applicants across the risk scale. A model that assigns nearly the same PD to everyone would have little value for screening, even if that average PD were reasonable. A useful scoring model must separate applicants enough to support different decisions.
