## A single decision tree

A decision tree is the simplest tree-based model. It divides the data by asking a sequence of questions. Each question creates a split. At the end of a sequence of splits, each terminal node contains a group of observations with a default rate. That default rate becomes the predicted probability of default for applicants who fall into that node.

If a terminal node, or leaf, contains $n_L$ loans and $d_L$ of them defaulted, the leaf prediction is:

$$
\hat{p}_L = \frac{d_L}{n_L}.
$$

Any new applicant that reaches that leaf receives the same predicted probability $\hat{p}_L$. This makes the model easy to inspect: each prediction is tied to a path through the tree.

This is very different from logistic regression. Logistic regression creates one smooth scoring equation. A tree creates local rules. The model may say, in effect: applicants in this region of the data have a default rate of 3%, while applicants in another region have a default rate of 20%. This rule-based structure is why a single tree is useful pedagogically.

```{r Single decision tree}
simple_tree
```

The printed object describes the tree in text form. Each split sends observations to different branches. The variables that appear near the top are the first questions asked by the model, which means they are especially important for this particular tree.

```{r}
#| label: fig-single-decision-tree
#| fig-cap: "A single decision tree for credit scoring."
#| fig-width: 10
#| fig-height: 7
plot(simple_tree, uniform = TRUE, margin = 0.10)
text(simple_tree, use.n = TRUE, all = TRUE, cex = 0.70)
```

The figure is useful because it makes the logic visible. The model is not estimating a coefficient for each variable. It is building a set of conditional rules. The first split partitions the data into two groups. Then each group can be split again. The final leaves summarize the predicted class and the number of observations reaching each leaf.

However, a single tree has an important weakness: it can be unstable. A small change in the training data can produce a different set of splits. A tree can also be too simple if it is shallow, or too noisy if it is allowed to grow too deep. This creates a natural motivation for ensemble methods, where many trees are combined instead of relying on one tree.

```{r Single tree predictions}
data.frame(
  min_predicted_pd = min(pred_tree_simple),
  max_predicted_pd = max(pred_tree_simple),
  auc = as.numeric(auc(ROC_tree_simple)),
  brier_score = brier_score(actual_default_tree, pred_tree_simple)
)
```

The single tree already produces default probabilities and can be evaluated like the logistic model. In code, `simple_tree <- rpart(...)` learns the leaves, and `predict(simple_tree, newdata = test, type = "prob")[, "1"]` returns the leaf default rate assigned to each test applicant. This is the code counterpart of $\hat p_L = d_L/n_L$. The main role of the single tree in this chapter is conceptual: it shows the basic building block used by more powerful tree-based methods.
