## Two firms and joint defaults

```{r gaussian-two-firms-setup, include=FALSE}
source("R/gaussian-common.R")

two_firm_sims <- 100000
two_firm_threshold <- default_threshold[5]
two_firm_pd <- pd_cum[5]
two_firm_rhos <- c(0, 0.2, 0.5)

simulate_two_firm_summary <- function(rho) {
  latent <- simulate_latent_credit(
    n_sims = two_firm_sims,
    n_firms = 2,
    rho = rho,
    seed = 130575
  )
  defaults <- make_default_matrix(latent, two_firm_threshold)
  default_a <- defaults[, 1]
  default_b <- defaults[, 2]
  data.frame(
    rho = rho,
    pd_firm_a = mean(default_a),
    pd_firm_b = mean(default_b),
    both_default = mean(default_a & default_b),
    at_least_one = mean(default_a | default_b),
    default_indicator_corr = cor(as.numeric(default_a), as.numeric(default_b))
  )
}

two_firm_numeric <- do.call(rbind, lapply(two_firm_rhos, simulate_two_firm_summary))
two_firm_independent_both <- two_firm_pd * two_firm_pd
two_firm_independent_any <- 1 - (1 - two_firm_pd)^2
two_firm_plot_rhos <- c(-0.5, -0.2, 0.2, 0.5)
two_firm_plot_layout_rhos <- c(-0.5, -0.2, 0.2, 0.5)

two_firm_table <- data.frame(
  Copula_correlation = fmt_num(two_firm_numeric$rho, 1),
  Simulated_PD_A = fmt_pct(two_firm_numeric$pd_firm_a, 2),
  Simulated_PD_B = fmt_pct(two_firm_numeric$pd_firm_b, 2),
  P_both_default = fmt_pct(two_firm_numeric$both_default, 2),
  P_at_least_one = fmt_pct(two_firm_numeric$at_least_one, 2),
  Default_variable_corr = fmt_num(two_firm_numeric$default_indicator_corr, 4)
)
names(two_firm_table) <- c(
  "rho",
  "PD A",
  "PD B",
  "P(A and B)",
  "P(A or B)",
  "Default corr."
)

two_firm_plot_data <- lapply(two_firm_plot_rhos, function(rho) {
  latent <- simulate_latent_credit(
    n_sims = 6000,
    n_firms = 2,
    rho = rho,
    seed = 130575
  )
  data.frame(firm_a = latent[, 1], firm_b = latent[, 2], rho = rho)
})
```

The two-firm case shows what the copula adds. Suppose both firms have the same 5-year cumulative PD, `r fmt_pct(two_firm_pd, 2)`. If the two default events were independent, the probability that both default by year 5 would be

$$
P(A\text{ defaults and }B\text{ defaults})
=`r fmt_pct(two_firm_pd, 2)`\times `r fmt_pct(two_firm_pd, 2)`
=`r fmt_pct(two_firm_independent_both, 2)`.
$$

The probability that at least one defaults would be

$$
P(A\text{ or }B\text{ defaults})
=1-(1-`r fmt_pct(two_firm_pd, 2)`)^2
=`r fmt_pct(two_firm_independent_any, 2)`.
$$

Those formulas rely on independence. Under independence, firm A's credit state and firm B's credit state are unrelated. The Gaussian copula adds shared shocks, so low values of $X_A$ and $X_B$ can occur together more often while each firm keeps a marginal PD near `r fmt_pct(two_firm_pd, 2)`.

The covariance matrix tells R how strongly the two credit-state scores share shocks.

$$
\Sigma=
\begin{bmatrix}
1 & \rho\\
\rho & 1
\end{bmatrix}.
$$

```{r}
rho <- 0.2
sigma <- matrix(c(1, rho, rho, 1), nrow = 2)

set.seed(130575)
credit_state <- MASS::mvrnorm(
  n = 100000,
  mu = c(0, 0),
  Sigma = sigma
)

threshold <- qnorm(0.15)
default_a <- credit_state[, 1] <= threshold
default_b <- credit_state[, 2] <= threshold

mean(default_a)
mean(default_b)
mean(default_a & default_b)
```

The first two probabilities check the marginal PDs. The third probability is the joint default probability. That is where the shared-shock assumption becomes visible.

```{r}
kable(
  two_firm_table,
  caption = "Two-firm default probabilities under different Gaussian copula correlations.",
  escape = FALSE
)
```

The table demonstrates the main point. The marginal PDs stay close to `r fmt_pct(two_firm_pd, 2)`, while the probability of both firms defaulting rises as $\rho$ rises. The default-variable correlation is lower than the correlation between credit-state scores because default is a tail event created after applying the threshold.

The union probability also helps explain what correlation changes. For two firms,

$$
P(A \text{ or } B)=P(A)+P(B)-P(A \text{ and } B).
$$

The marginal probabilities $P(A)$ and $P(B)$ remain close to `r fmt_pct(two_firm_pd, 2)`. As $\rho$ moves from 0 to 0.5, the simulated joint-default probability rises from `r fmt_pct(two_firm_numeric$both_default[1], 2)` to `r fmt_pct(two_firm_numeric$both_default[3], 2)`. Because the joint-default term is subtracted in the union formula, the probability that at least one firm defaults falls from `r fmt_pct(two_firm_numeric$at_least_one[1], 2)` to `r fmt_pct(two_firm_numeric$at_least_one[3], 2)`. Higher correlation moves defaults into the same scenarios.

The scatter plots show the same mechanism and widen the comparison to negative and positive correlations. Each dot is one simulated pair $(X_A,X_B)$, meaning one credit scenario for the two firms. A firm defaults when its credit-state score is below the threshold. The lower-left region is the joint-default region. Negative correlation pushes the firms' credit states in opposite directions more often, while positive correlation makes weak credit states arrive together more often.

```{r}
#| label: fig-two-firm-copula-scatter
#| fig-cap: "Two-firm Gaussian copula simulations across negative and positive correlations."
#| fig-width: 7.5
#| fig-height: 7.5
two_firm_scatter_data <- do.call(rbind, lapply(two_firm_plot_data, function(plot_data) {
  rho <- unique(plot_data$rho)
  rho_label <- paste0("rho = ", fmt_num(rho, 1))
  data.frame(
    firm_a = plot_data$firm_a,
    firm_b = plot_data$firm_b,
    rho = rho,
    rho_label = rho_label,
    joint_default = plot_data$firm_a <= two_firm_threshold &
      plot_data$firm_b <= two_firm_threshold
  )
}))
two_firm_scatter_data$rho_label <- factor(
  two_firm_scatter_data$rho_label,
  levels = paste0("rho = ", fmt_num(two_firm_plot_layout_rhos, 1))
)
two_firm_panel_data <- unique(
  two_firm_scatter_data[, c("rho", "rho_label")]
)
two_firm_joint_counts <- aggregate(
  joint_default ~ rho_label,
  data = two_firm_scatter_data,
  FUN = sum
)
two_firm_joint_counts$label <- paste0("Joint defaults: ", fmt_int(two_firm_joint_counts$joint_default))

ggplot(two_firm_scatter_data, aes(x = firm_a, y = firm_b)) +
  geom_rect(
    data = two_firm_panel_data,
    aes(xmin = -4, xmax = two_firm_threshold, ymin = -4, ymax = two_firm_threshold),
    inherit.aes = FALSE,
    fill = "#E74C3C",
    alpha = 0.10
  ) +
  geom_point(aes(color = joint_default), size = 0.45, alpha = 0.55) +
  geom_vline(xintercept = two_firm_threshold, linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = two_firm_threshold, linetype = "dashed", linewidth = 0.8) +
  geom_label(
    data = two_firm_joint_counts,
    aes(x = 1.0, y = -3.55, label = label),
    inherit.aes = FALSE,
    size = 2.8,
    fill = "white",
    label.size = 0,
    linewidth = 0
  ) +
  facet_wrap(~rho_label, ncol = 2) +
  scale_color_manual(
    values = c("FALSE" = adjustcolor("#566573", alpha.f = 0.55), "TRUE" = "#E74C3C"),
    labels = c("Other scenarios", "Joint defaults"),
    name = NULL
  ) +
  coord_equal(xlim = c(-4, 4), ylim = c(-4, 4)) +
  labs(
    x = "Firm A credit-state score",
    y = "Firm B credit-state score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "white", color = NA),
    panel.grid.minor = element_blank()
  )
```

This first credit-risk application of the copula separates the individual default probability of each firm from the co-default pattern between firms. The PD of each firm stays fixed, while the probability of clustered defaults changes.
