## Two firms

```{r gaussian-two-firms-setup, include=FALSE}
source("R/gaussian-common.R")
```

Consider the case in which we have two firms instead of one. The main difference now is that instead of simulating one firm we need two. Moreover, each firm follows a standard normal distribution function and both of them are correlated by a given correlation value so the firms are not independent. If they are not independent, then what happens to one firm has some impact on what happens to the other. 

In a different context, here we have an old video in which we drop many blue balls. Balls are dropped randomly given a multi-variate normal distribution. Note how the 3-D bell-shape distribution is formed as we increase the number of balls. 

```{r}
embed_url("https://youtu.be/5iv5Zksf_zo")
```

The case of two firms is not the one presented in @Hull example but it can help us to visualize how the Gaussian copula approach works in a two-dimension plot where the default correlation is 0.2.

The simulation of both firms' performance measures is <tt>`x2`</tt>.

```{r}
m <- 2 # number of firms
n <- 10000 # number of simulations
rho_pos02 <- 0.2 # correlation
corr_pos02 <- matrix(rep(rho_pos02, m * m), m, m) # correlation matrix
diag(corr_pos02) <- 1
set.seed(130575)
x2 <- mvrnorm(n, mu = rep(0, m), Sigma = corr_pos02)
x2 <- data.frame(x2)
colnames(x2) <- c("f1", "f2")
```

The matrix <tt>`x2`</tt> length is 10,000 for each firm. In other words, we have 10,000 observations of the performance measure or $z$-score for two firms that are related. This matrix is big, but we can visualize the header (the first six observations).

```{r}
kable(head(x2), caption = "Firms' performance x2.", row.names = TRUE)
```

Remember the cumulative probabilities of default are 1%, 3%, 6%, 10% and 15% for the maturities of 1, 2, 3, 4 and 5 years respectively. How do we extract those cases in which both firms will default in 5 years? In rows 15, 53, 61 and so on both firms default at the same time. Note that in all cases the $x$ values are indeed below -1.036433.

```{r}
# These names are going to be useful later.
n.year <- c("year 1", "year 2", "year 3", "year 4", "year 5")
n.pd <- c("pd.y1", "pd.y2", "pd.y3", "pd.y4", "pd.y5")
n.f <- c("f1", "f2", "f1 default?", "f2 default?")

# Function to calculate cases in which firms default and probabilities.
fun.X <- function(x) {
  both <- x2[x2$f2 < x & x2$f1 < x, ] # both default.
  any <- x2[x2$f2 < x | x2$f1 < x, ] # at least one firm default.
  onlyf1 <- x2[x2$f2 > x & x2$f1 < x, ] # only firm 1 default.
  onlyf2 <- x2[x2$f1 > x & x2$f2 < x, ] # only firm 2 default.
  one <- x2[(x2$f2 < x & x2$f1 > x | # only one firm default.
                   x2$f2 > x & x2$f1 < x),]
  none <- x2[x2$f2 > x & x2$f1 > x, ] # no firm default.
# Gather results and probabilities in a list.
ans <- list(both = both, any = any, onlyf1 = onlyf1,
  onlyf2 = onlyf2, one = one, none = none,
  both.pd = (nrow(both) / n), any.pd = (nrow(any) / n),
  onlyf1.pd = (nrow(onlyf1) / n), onlyf2.pd = (nrow(onlyf2)/ n),
  one.pd = (nrow(one) / n), none.pd = (nrow(none) / n))
}
# X has all the relevant results for x2.
X <- mapply(fun.X, x.y)
```

See the cases in which both firms default in 5 years.

```{r}
# Extract "both" cases, year 5.
both <- data.frame(X[["both", 5]], X[["both", 5]] < x.y[5])
colnames(both) <- n.f
kable(head(both), caption = "Cases in which both firms default in 5 years.", 
      row.names = TRUE)
```
In total, we have 343 cases in which both firms default at the same time. The first case is number 15, the second 53, the third 61 and so on. It is easy to know the total cases if we count the number of rows.

How do we extract those cases in which any firm will default in 5 years? This is, only firm 1, only firm 2, or both at the same time. This is a less strict condition so we would expect to have more cases to match this new criteria compared with <tt>`both`</tt>. Note that in all cases at least one one firm is indeed below -1.036433. In row 1, 2 and 10 firm 2 defaults. In row 15 both firms default. In row 18 and 20 firm 1 and firm 2 default respectively.


```{r}
any <- data.frame(X[["any", 5]], X[["any", 5]] < x.y[5])
colnames(any) <- n.f
kable(head(any), caption = "Cases in which at least one firm default.", 
      row.names = TRUE)
```

Now we have 2,660 cases. Considerably more as the <tt>`|`</tt> restriction is less strict than the <tt>`&`</tt>. Let's see the cases in which only firm 1 defaults.

```{r}
onlyf1 <- data.frame(X[["onlyf1", 5]], X[["onlyf1", 5]] < x.y[5])
colnames(onlyf1) <- n.f
kable(head(onlyf1), caption = "Cases in which only firm 1 default.")
```

Only firm 2 defaults. 

```{r}
onlyf2 <- data.frame(X[["onlyf2", 5]], X[["onlyf2", 5]] < x.y[5])
colnames(onlyf2) <- n.f
kable(head(onlyf2), caption = "Cases in which only firm 2 default.")
```

Only one firm default at year 5.

```{r}
one <- data.frame(X[["one", 5]], X[["one", 5]] < x.y[5])
colnames(one) <- n.f
kable(head(one), caption = "Cases in which only one firm default.")
```

Lastly, when no firm defaults.

```{r}
none <- data.frame(X[["none", 5]], X[["none", 5]] < x.y[5])
colnames(none) <- n.f
kable(head(none), caption = "Cases in which no firm default.")
```

Finally, probabilities.

```{r}
both.pd <- t(data.frame(X["both.pd",]))
any.pd <- t(data.frame(X["any.pd",]))
onlyf1.pd <- t(data.frame(X["onlyf1.pd",]))
onlyf2.pd <- t(data.frame(X["onlyf2.pd",]))
one.pd <- t(data.frame(X["one.pd",]))
none.pd <- t(data.frame(X["none.pd",]))
ans <- data.frame(both.pd, any.pd, onlyf1.pd, onlyf2.pd, 
                 one.pd, none.pd)
rownames(ans) <- n.year
kable(ans, caption = "Probabilities of default.")
```

So interesting. 

Note that 15% is the theoretical probability that one firm will default in 5 years. Here, this 15% is 12.1% for firm 1 and 11.07% for firm 2 when the data is simulated. 

We can perform a simple test to verify that everything is alright. For example, this equation must hold: <tt>`onlyf1+onlyf2=one`</tt>. Substituting for the year 5: $0.121+0.1107=0.2317$. As you can see, everything is alright. This equation must hold as well: <tt>`any-both=one`</tt>. Substituting for the year 5: $0.266-0.0343=0.2317$.

Let's visualize all 10,000 cases. Each dot represents a couple of Firm 1 and Firm 2 $x$ values and the dotted lines the threshold that represents the probability of default in 5 years.

```{r}
#| label: fig-a1sc
#| fig-cap: "All 10,000 simulated cases."
par(pty = "s") # Figures are shown in a perfect square (not a rectangle).
plot(x2, pch = ".", cex = 1) 
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("bottomright", legend = c(paste(nrow(x2))), bty = "n")
```

These 10,000 observations are highly concentrated around the mean which is very close to zero. This can be also easily seen in the following density plot.


```{r}
#| label: fig-a1scadv
#| fig-cap: "All 10,000 simulated cases: A density view."
ggplot(x2, aes(x = f1, y = f2) ) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", 
                  colour = "white") +
    coord_fixed()
```


```{r eval=FALSE, include=FALSE}
densray <- ggplot(x2, aes(x = f1, y = f2) ) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", 
                  colour = "white") +
    coord_fixed() +
  ggtitle("All simulated cases: A rayshader view.")

plot_gg(densray, width = 4, height = 4, scale = 400, zoom = 0.7,
        multicore = TRUE)
# Movie and picture.
render_movie("dens.mp4", frames = 1840) #460)
render_snapshot("dens.png", clear = TRUE)

```

This is the 3-D <tt>`rayshader`</tt> video version of the plot above.

```{r}
embed_url("https://youtu.be/3p-B19R8RwA")
```

And a third interactive version.

```{r}
#| label: fig-aiv
#| fig-cap: "An interactive view."
f <- function(x, y) dmnorm(cbind(x, y), c(0, 0), corr_pos02)
z <- outer(sort(x2[1:100,1]), sort(x2[1:100,2]), f)

#create contour plot
#contour(x2[1:100,1], x2[1:100,2], z)
plot_ly(type = "surface" , x = sort(x2[1:100,2]), 
        y = sort(x2[1:100,1]) , z = z ) %>%
layout(#title = "Surface",
       scene = list(xaxis = list(title = "f1", range = c(-2,2)),
                    yaxis = list(title = "f2", range = c(-2,2)), 
                    zaxis = list(title = "Density"))) %>%
hide_colorbar()
```


```{r eval=FALSE, include=FALSE}
df <- tibble(x2)
par(pty = "s")
rayplot <- ggplot(df, aes(x = x2$f1, y = x2$f2)) +
  stat_density2d(aes(fill = ..density..), contour = F, 
                 geom = 'tile') +
  scale_fill_viridis()+
  #scale_fill_viridis_c(option = "A")+
  coord_fixed() +
  theme(legend.position = "none")
plot_gg(rayplot, width = 4, height = 4, scale = 400, zoom = 0.7,
        multicore = TRUE)
# Movie and picture.
#render_movie("p25.mp4", frames = 460)
#render_snapshot("p25.png", clear = TRUE)

```

```{r eval=FALSE, include=FALSE}
df <- tibble(x2)
par(pty = "s")
rayplot <- ggplot(df, aes(x = x2$f1, y = x2$f2)) +
  stat_density2d(aes(fill = ..density..), contour = F, geom = 'tile') +
    geom_hex(bins = 35, size = 0.5, color = "black") +
  #scale_fill_viridis()+
  #scale_fill_viridis_c(option = "A")+
    scale_fill_viridis_c(option = "C")+
  coord_fixed() +
  theme(legend.position = "none")

plot_gg(rayplot, width = 4, height = 4, scale = 400, zoom = 0.7,
        multicore = TRUE)

```

We can visualize the default cases. First, the case when both firms default at year 5.

```{r}
#| label: fig-bfday5
#| fig-cap: "Both firms default at year 5."
par(pty = "s")
plot(X[["both", 5]], xlim  = L, ylim = L, pch = ".", cex = 1) 
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(both.pd[5]*n)), bty = "n")
```

The values within the plot represent the number of cases. Here, we have 343 times (out of 10,000) in which both firms default at the same time in 5 years. Note that this is a cumulative probability of default.

Now, the case in which any firm defaults in 5 years.

```{r}
#| label: fig-afdi5y
#| fig-cap: "Any firm defaults in 5 years."
par(pty = "s")
plot(X[["any", 5]], xlim = L, ylim = L, pch = ".", cex = 1)
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(any.pd[5]*n)), bty = "n")
```

Only firm 1 defaults in 5 years.

```{r}
#| label: fig-of1di5y
#| fig-cap: "Only firm 1 defaults in 5 years."
par(pty = "s")
plot(X[["onlyf1", 5]], xlim = L, ylim = L, pch = ".", cex = 1)
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(onlyf1.pd[5]*n)), bty = "n")
```

Only firm 2 defaults in 5 years.

```{r}
#| label: fig-of2di5y
#| fig-cap: "Only firm 2 defaults in 5 years."
par(pty = "s")
plot(X[["onlyf2", 5]], xlim = L, ylim = L, pch = ".", cex = 1)
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(onlyf2.pd[5]*n)), bty = "n")
```

This is the case in which only one firm defaults.

```{r}
#| label: fig-oofdi5y
#| fig-cap: "Only one firm defaults in 5 years."
par(pty = "s")
plot(X[["one", 5]], xlim = L, ylim = L, pch = ".", cex = 1)
abline(h = x.y[5], lty =2)
abline(v = x.y[5], lty =2)
legend("topright", legend = c(paste(one.pd[5]*n)), bty = "n")
```

This is the case in which no one firm defaults.

```{r}
#| label: fig-nfdi5y
#| fig-cap: "No firm defaults in 5 years."
par(pty = "s")
plot(X[["none", 5]], xlim = L, ylim = L, pch = ".", cex = 1)
abline(h = x.y[5], lty =2)
abline(v = x.y[5], lty =2)
legend("topright", legend = c(paste(none.pd[5]*n)), bty = "n")
```

It is a good idea to summarize all previous plots in one.

```{r}
#| label: fig-wfday5
#| fig-cap: "Which firm defaults at year 5?"
# none
par(mfrow = c(2, 3), oma = c(0, 0, 2, 0))
par(pty = "s")
plot(X[["none", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "None.") 
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("bottomright", legend = c(paste(none.pd[5]*n)), bty = "n")
# both
par(pty = "s")
plot(X[["both", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "Both.") 
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(both.pd[5]*n)), bty = "n")
# any
par(pty = "s")
plot(X[["any", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "Any.")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(any.pd[5]*n)), bty = "n")
# onlyfirm1
par(pty = "s")
plot(X[["onlyf1", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "Only f1.")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(onlyf1.pd[5]*n)), bty = "n")
# onlyfirm2
par(pty = "s")
plot(X[["onlyf2", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "Only f2.")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("topright", legend = c(paste(onlyf2.pd[5]*n)), bty = "n")
# one
par(pty = "s")
plot(X[["one", 5]], xlim = L, ylim = L, pch = ".", cex = 1, 
     main = "One.")
abline(h = x.y[5], lty = 2)
abline(v = x.y[5], lty = 2)
legend("topright", legend = c(paste(one.pd[5]*n)), bty = "n")
```

It is interesting to compare two different correlation values. Here, we compare 0 versus 0.2.

```{r}
m <- 2 # number of firms
n <- 10000 # number of simulations
rho_pos02 <- 0 # correlation
corr_pos02 <- matrix(rep(rho_pos02, m * m), m, m) # correlation matrix
diag(corr_pos02) <- 1
set.seed(130575)
X0 <- mvrnorm(n, mu = rep(0, m), Sigma = corr_pos02)
X0 <- data.frame(X0)
colnames(X0) <- c("f1", "f2")
```

Just as before, we create a function, evaluate it, and store results in <tt>`X0`</tt>.

```{r}
fun.X0 <- function(x) {
  both <- X0[X0$f2 < x & X0$f1 < x, ]
  any <- X0[X0$f2 < x | X0$f1 < x, ]
  onlyf1 <- X0[X0$f2 > x & X0$f1 < x, ]
  onlyf2 <- X0[X0$f1 > x & X0$f2 < x, ]
  one <- X0[(X0$f2 < x & X0$f1 > x |
                   X0$f2 > x & X0$f1 < x),]
  none <- X0[X0$f2 > x & X0$f1 > x, ]

ans <- list(both = both, any = any, onlyf1 = onlyf1,
  onlyf2 = onlyf2, one = one, none = none,
  both.pd = (nrow(both) / n), any.pd = (nrow(any) / n),
  onlyf1.pd = (nrow(onlyf1) / n), onlyf2.pd = (nrow(onlyf2) /n),
  one.pd = (nrow(one) / n), none.pd = (nrow(none) / n)) }

X0 <- mapply(fun.X0, x.y)

none.pd0 <- data.frame(X0["none.pd",]) * n
both.pd0 <- data.frame(X0["both.pd",]) * n

onlyf1.pd0 <- data.frame(X0["onlyf1.pd",]) * n
onlyf2.pd0 <- data.frame(X0["onlyf2.pd",]) * n
```


A graphical analysis shows that in the case of 0.2 it is more likely that both firms default at the same time, and it is less likely that any firm default at the same time.


```{r}
#| label: fig-cpqdlcty5
#| fig-cap: "Cases per quadrant. Dotted lines corresponds to year 5."
par(mfrow=c(1, 2), oma = c(0, 0, 2, 0))
par(pty = "s")
plot(X[["none", 5]], pch = ".", xlim = L, ylim = L, 
     cex = 1, main = "Correlation = 0.2") 
points(X[["both", 5]], pch = ".", col = "red")
points(X[["onlyf1", 5]], pch = ".", col = "purple")
points(X[["onlyf2", 5]], pch = ".", col = "blue")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("bottomleft", legend = c(paste(both.pd[5]*n)), bty = "n")
legend("topright", legend = c(paste(none.pd[5]*n)), bty = "n")
legend("topleft", legend = c(paste(onlyf1.pd[5]*n)), bty = "n")
legend("bottomright", legend = c(paste(onlyf2.pd[5]*n)), bty = "n")
par(pty = "s")
plot(X0[["none", 5]], pch = ".", xlim = L, ylim = L,
     cex = 1, main = "Correlation = 0.") 
points(X0[["both", 5]], pch = ".", col = "red")
points(X0[["onlyf1", 5]], pch = ".", col = "purple")
points(X0[["onlyf2", 5]], pch = ".", col = "blue")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("bottomleft", legend = c(paste(both.pd0[5])), bty = "n")
legend("topright", legend = c(paste(none.pd0[5])), bty = "n")
legend("topleft", legend = c(paste(onlyf1.pd0[5])), bty = "n")
legend("bottomright", legend = c(paste(onlyf2.pd0[5])), bty = "n")
par(pty = "s")
```
Very interesting indeed. 

The Gaussian copula model is not the only copula model. See how other distributions work.

t-coupla:

```{r}
embed_url("https://youtu.be/XZPVnanmEtw")
```

Gumbel coupla:

```{r}
embed_url("https://youtu.be/Oq27fu86ahU")
```

Frank coupla:

```{r}
embed_url("https://youtu.be/4OOF_FTrZ3w")
```

Clayton coupla:

```{r}
embed_url("https://youtu.be/vtzkgnWjFQ4")
```

```{r eval=FALSE, include=FALSE}
m <- 2 # number of firms
n <- 10000 # number of simulations
rho_pos00 <- 0 # correlation
rho_pos02 <- 0.2
corr_pos00 <- matrix(rep(rho_pos00, m * m), m, m) # correlation matrix
corr_pos02 <- matrix(rep(rho_pos02, m * m), m, m) # correlation matrix
diag(corr_pos00) <- 1
diag(corr_pos02) <- 1
set.seed(130575)
X00 <- mvrnorm(n, mu = rep(0, m), Sigma = corr_pos00)
set.seed(130575)
X02 <- mvrnorm(n, mu = rep(0, m), Sigma = corr_pos02)
X00.02 <- data.frame(rbind(X00, X02))
colnames(X00.02) <- c("f1", "f2")

n=20000
fun.X0 <- function(x) {
  both <- X00.02[X00.02$f2 < x & X00.02$f1 < x, ]
  any <- X00.02[X00.02$f2 < x | X00.02$f1 < x, ]
  onlyf1 <- X00.02[X00.02$f2 > x & X00.02$f1 < x, ]
  onlyf2 <- X00.02[X00.02$f1 > x & X00.02$f2 < x, ]
  one <- X00.02[(X00.02$f2 < x & X00.02$f1 > x | 
                   X00.02$f2 > x & X00.02$f1 < x),]
  none <- X00.02[X00.02$f2 > x & X00.02$f1 > x, ]
  
ans <- list(both = both, any = any, onlyf1 = onlyf1,
  onlyf2 = onlyf2, one = one, none = none,
  both.pd = (nrow(both) / n), any.pd = (nrow(any) / n),
  onlyf1.pd = (nrow(onlyf1) / n), onlyf2.pd = (nrow(onlyf2) /n),
  one.pd = (nrow(one) / n), none.pd = (nrow(none) / n)) }

X00.02 <- mapply(fun.X0, x.y)

none.pd0002 <- data.frame(X00.02["none.pd",]) * n
both.pd0002 <- data.frame(X00.02["both.pd",]) * n

onlyf1.pd0002 <- data.frame(X00.02["onlyf1.pd",]) * n
onlyf2.pd0002 <- data.frame(X00.02["onlyf2.pd",]) * n

```


```{r eval=FALSE, include=FALSE}
par(pty = "s")
plot(X00.02[["none", 5]], pch = ".", xlim = L, ylim = L, 
     cex = 0.8, main = "Correlation=0.2") 
points(X00.02[["both", 5]], pch = ".", col = "red")
points(X00.02[["onlyf1", 5]], pch = ".", col = "purple")
points(X00.02[["onlyf2", 5]], pch = ".", col = "blue")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
legend("bottomleft", legend = c(paste(both.pd0002[5]/n*100)), bty = "n")
legend("topright", legend = c(paste(none.pd0002[5]/n*100)), bty = "n")
legend("topleft", legend = c(paste(onlyf1.pd0002[5]/n*100)), bty = "n")
legend("bottomright", legend = c(paste(onlyf2.pd0002[5]/n*100)), bty = "n")

```
```{r eval=FALSE, include=FALSE}
par(pty = "s")
plot(X00.02[["none", 5]], pch = ".", xlim = L, ylim = L, 
     cex = 0.8, main = "Correlation=0.2") 
points(X00.02[["both", 5]], pch = ".", col = "black")
points(X00.02[["onlyf1", 5]], pch = ".", col = "red")
points(X00.02[["onlyf2", 5]], pch = ".", col = "red")
points(X00.02[["both", 4]], pch = ".", col = "black")
points(X00.02[["onlyf1", 4]], pch = ".", col = "black")
points(X00.02[["onlyf2", 4]], pch = ".", col = "black")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
abline(v = x.y[4], lty = 2)
abline(h = x.y[4], lty = 2)
x.y[5]
x.y[4]
```

```{r eval=FALSE, include=FALSE}
par(pty = "s")
plot(X00.02[["none", 5]], pch = ".", xlim = L, ylim = L, 
     cex = 0.8, main = "Correlation=0.2") 
points(X00.02[["both", 5]], pch = ".", col = "black")
points(X00.02[["onlyf1", 5]], pch = ".", col = "red")
points(X00.02[["onlyf2", 5]], pch = ".", col = "red")
points(X00.02[["both", 5]], pch = ".", col = "red")
points(X00.02[["onlyf1", 4]], pch = ".", col = "black")
points(X00.02[["onlyf2", 4]], pch = ".", col = "black")
points(X00.02[["both", 5]], pch = ".", col = "red")
points(X00.02[["both", 4]], pch = ".", col = "black")
abline(v = x.y[5], lty = 2)
abline(h = x.y[5], lty = 2)
abline(v = x.y[4], lty = 2)
abline(h = x.y[4], lty = 2)
x.y[5]
x.y[4]
```

