## Tail dependence with a t-copula

```{r gaussian-t-copula-setup, include=FALSE}
source("R/gaussian-common.R")

tcopula_sims <- 100000
tcopula_n_firms <- 10
tcopula_rho <- hull_example_247_rho
tcopula_pd <- pd_cum[5]
tcopula_total_ead <- 100e6
tcopula_ead <- rep(tcopula_total_ead / tcopula_n_firms, tcopula_n_firms)
tcopula_recovery <- rep(0.40, tcopula_n_firms)

tcopula_design <- data.frame(
  copula = c("Gaussian", "t-copula", "t-copula"),
  df = c(NA, 10, 4)
)

simulate_tail_dependence_case <- function(copula, df) {
  if (copula == "Gaussian") {
    latent <- simulate_latent_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      seed = 130575
    )
    threshold <- qnorm(tcopula_pd)
  } else {
    latent <- simulate_t_copula_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      df = df,
      seed = 130575
    )
    threshold <- qt(tcopula_pd, df = df)
  }

  defaults <- make_default_matrix(latent, threshold)
  losses <- portfolio_losses(defaults, tcopula_ead, tcopula_recovery)

  list(
    defaults = defaults,
    default_count = rowSums(defaults),
    losses = losses,
    events = summarize_default_events(defaults),
    loss_summary = loss_risk_summary(losses),
    mean_pd = mean(defaults)
  )
}

tcopula_results <- Map(
  simulate_tail_dependence_case,
  tcopula_design$copula,
  tcopula_design$df
)
names(tcopula_results) <- c("Gaussian", "t df = 10", "t df = 4")

tcopula_events <- do.call(rbind, lapply(tcopula_results, function(x) x$events))
tcopula_losses <- do.call(rbind, lapply(tcopula_results, function(x) x$loss_summary))

tcopula_df_label <- ifelse(is.na(tcopula_design$df), "-", fmt_int(tcopula_design$df))

tcopula_default_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  Mean_PD = fmt_pct(sapply(tcopula_results, function(x) x$mean_pd), 2),
  Expected_defaults = fmt_num(tcopula_events[, "expected_defaults"], 3),
  P_three_or_more = fmt_pct(tcopula_events[, "three_or_more"], 2),
  P_five_or_more = fmt_pct(tcopula_events[, "five_or_more"], 2),
  P_all_ten = fmt_pct(tcopula_events[, "all_default"], 3)
)
names(tcopula_default_table) <- c(
  "Copula",
  "df",
  "Mean PD",
  "E[defaults]",
  "P(>=3)",
  "P(>=5)",
  "P(10)"
)

tcopula_loss_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  EL = fmt_musd(tcopula_losses[, "expected_loss"]),
  VaR_99 = fmt_musd(tcopula_losses[, "var_99"]),
  VaR_99_9 = fmt_musd(tcopula_losses[, "var_999"]),
  ES_99_9 = fmt_musd(tcopula_losses[, "es_999"]),
  Capital_99_9 = fmt_musd(tcopula_losses[, "economic_capital_999"])
)
names(tcopula_loss_table) <- c(
  "Copula",
  "df",
  "EL",
  "VaR 99%",
  "VaR 99.9%",
  "ES 99.9%",
  "Capital 99.9%"
)

tcopula_count_matrix <- sapply(tcopula_results, function(x) {
  as.numeric(tabulate(x$default_count + 1, nbins = tcopula_n_firms + 1)) /
    tcopula_sims
})
colnames(tcopula_count_matrix) <- names(tcopula_results)
```

The Gaussian copula uses normal latent variables. Correlation creates joint default clustering. The normal distribution controls the shape of the tails. A t-copula keeps the same threshold idea and changes the tail shape.

For the Gaussian copula, one convenient one-factor representation is

$$
X_i^{G}=\sqrt{\rho}\,Z_0+\sqrt{1-\rho}\,Z_i.
$$

For the t-copula, use the same normal factor structure and divide by a random scale shock:

$$
X_i^{t}=
\frac{\sqrt{\rho}\,Z_0+\sqrt{1-\rho}\,Z_i}
{\sqrt{S/\nu}},
\qquad S\sim\chi^2_{\nu}.
$$

The parameter $\nu$ is the number of degrees of freedom. Lower $\nu$ means heavier tails. The default threshold changes from $N^{-1}(PD)$ to $t_{\nu}^{-1}(PD)$, so each firm keeps the same marginal PD.

The comparison below keeps the 5-year PD at `r fmt_pct(tcopula_pd, 2)`, the copula correlation at `r fmt_num(tcopula_rho, 1)`, the same 10 firms, the same total exposure, and the same recovery rate. The only change is the copula tail shape.

```{r}
kable(
  tcopula_default_table,
  caption = "Default clustering under Gaussian and t-copulas.",
  row.names = FALSE,
  escape = FALSE
)
```

The average PD stays near `r fmt_pct(tcopula_pd, 2)`. The tail probabilities move because the t-copula gives more weight to joint stress states. This is the credit-risk reason to care about tail dependence.

The same default simulations can be translated into losses using the portfolio loss formula from the previous section.

```{r}
kable(
  tcopula_loss_table,
  caption = "Portfolio loss measures under Gaussian and t-copulas.",
  row.names = FALSE,
  escape = FALSE
)
```

Expected loss remains anchored by PD, EAD, and LGD. Tail loss responds to the dependence model. This ten-firm portfolio has coarse $6 million loss steps, so VaR can land on the same step for two copulas. Expected shortfall is a useful companion because it averages the losses beyond the percentile.

```{r}
#| label: fig-gaussian-vs-t-default-counts
#| fig-cap: "Default-count distribution under Gaussian and t-copulas."
default_counts <- 0:tcopula_n_firms
tcopula_count_colors <- adjustcolor(
  c("#2C7FB8", "#F39C12", "#C0392B"),
  alpha.f = 0.78
)
matplot(
  default_counts,
  tcopula_count_matrix,
  type = "b",
  pch = 19,
  cex = 1.45,
  lwd = 2.4,
  lty = 1,
  col = tcopula_count_colors,
  xlab = "Number of defaults by year 5",
  ylab = "Simulated probability"
)
legend(
  "topright",
  legend = colnames(tcopula_count_matrix),
  col = c("#2C7FB8", "#F39C12", "#C0392B"),
  lty = 1,
  pch = 19,
  bty = "n"
)
```

This section gives the key model-risk message. A copula is more than a correlation input. It also chooses how much probability is assigned to joint tail events.
