## Hull Example 24.7 from cumulative PDs to thresholds

```{r gaussian-example-247-setup, include=FALSE}
source("R/gaussian-common.R")

hull_threshold_table <- data.frame(
  Horizon = paste(credit_horizons, ifelse(credit_horizons == 1, "year", "years")),
  Cumulative_PD = fmt_pct(pd_cum, 2),
  Incremental_PD = fmt_pct(pd_incremental, 2),
  Default_threshold = fmt_num(default_threshold, 6),
  Check_pnorm_threshold = fmt_pct(threshold_check, 2)
)
names(hull_threshold_table) <- c(
  "Horizon",
  "Cumulative PD",
  "Incremental PD",
  "Default threshold",
  "PD recovered from threshold"
)

hull_threshold_numeric <- data.frame(
  horizon = credit_horizons,
  pd_cum = pd_cum,
  pd_incremental = pd_incremental,
  threshold = default_threshold,
  threshold_check = threshold_check
)
```

The example that belongs to this chapter is Hull's Example 24.7. The example immediately before it, Example 24.6, is about CVA for a gold forward contract. The Gaussian-copula credit portfolio example starts after Hull motivates default correlation in section 24.8.

Hull's Example 24.7 considers ten firms. Each firm has the same cumulative probabilities of default over five horizons.

In this example, the cumulative PDs are inputs. In practice, they would have to be estimated before using the copula. They could come from an internal credit-scoring model like the models in Chapters 1 and 2, from a structural model such as Merton for public firms, from historical default rates for comparable rating or industry groups, or from market-implied credit information. The copula starts after these marginal PDs are available. Its job is to combine the individual default probabilities with a dependence assumption so we can simulate which firms default together.

$$
Q(1)=1\%,\quad Q(2)=3\%,\quad Q(3)=6\%,\quad Q(4)=10\%,\quad Q(5)=15\%.
$$

The word cumulative is important. A cumulative 5-year PD of 15% means default occurs sometime between today and year 5. A separate 15% default probability inside year 5 would describe a different object. The incremental probability for a specific year is obtained by differencing the cumulative curve. For example, the probability of default between year 4 and year 5 is

$$
Q(5)-Q(4)=15\%-10\%=5\%.
$$

The table below keeps both objects visible because they answer different questions. The cumulative PD is used to define the default threshold by horizon. The incremental PD tells us how much new default probability is added between two adjacent horizons. The last column is a check. It applies $N(\cdot)$ back to the threshold, so it should reproduce the cumulative PD exactly apart from rounding. In other words, the cumulative PD is the input, the default threshold is the normal-scale version of that input, and the recovered PD confirms that the transformation preserves the marginal default probability.

```{r}
kable(
  hull_threshold_table,
  caption = "Hull Example 24.7 inputs and the corresponding Gaussian default thresholds.",
  escape = FALSE
)
```

The code below shows the transformation and its check.

```{r}
pd_cum <- c(0.01, 0.03, 0.06, 0.10, 0.15)
default_threshold <- qnorm(pd_cum)
pnorm(default_threshold)
```

The financial meaning is longer than the code. `pd_cum` contains marginal cumulative default probabilities. The command `qnorm(pd_cum)` converts those probabilities into normal-scale default thresholds. The command `pnorm(default_threshold)` walks back from the threshold to the probability scale. It gives the same numerical values as `pd_cum` by construction, and that equality is useful because it proves that the threshold rule preserves the individual default probabilities before we link firms through shared shocks.

For the 5-year horizon, the threshold is `r fmt_num(default_threshold[5], 6)`. A simulated firm defaults by year 5 if its credit-state score is below that value.

$$
Y_i(5)=
\begin{cases}
1, & \text{if } X_i \leq `r fmt_num(default_threshold[5], 6)`,\\
0, & \text{if } X_i > `r fmt_num(default_threshold[5], 6)`.
\end{cases}
$$

$Y_i(5)$ is therefore a binary default variable. It equals 1 when firm $i$ defaults by year 5 and equals 0 when the firm survives through year 5. The cutoff rule is simple: values of $X_i$ below the threshold are counted as defaults, and values above the threshold are counted as survival.

Because $X_i$ is standard normal, this rule gives

$$
P(Y_i(5)=1)=P(X_i \leq `r fmt_num(default_threshold[5], 6)`)=N(`r fmt_num(default_threshold[5], 6)`) = `r fmt_pct(pd_cum[5], 2)`.
$$

A single simulated scenario can be read directly from the same rule. The next table uses the 5-year threshold and a few illustrative credit-state scores. A score below the threshold becomes $Y_i(5)=1$; a score above the threshold becomes $Y_i(5)=0$.

```{r Gaussian threshold scenario example}
threshold_scenario_example <- data.frame(
  Firm = c("Firm A", "Firm B", "Firm C", "Firm D"),
  Credit_state_score = c(-1.40, -1.05, -0.80, 0.25),
  Default_threshold = rep(default_threshold[5], 4)
)
threshold_scenario_example$`Y_i(5)` <- ifelse(
  threshold_scenario_example$Credit_state_score <=
    threshold_scenario_example$Default_threshold,
  1,
  0
)
threshold_scenario_example$Interpretation <- ifelse(
  threshold_scenario_example$`Y_i(5)` == 1,
  "Default by year 5",
  "Survival through year 5"
)

kable(
  transform(
    threshold_scenario_example,
    Credit_state_score = fmt_num(Credit_state_score, 2),
    Default_threshold = fmt_num(Default_threshold, 6)
  ),
  caption = "Illustrative threshold rule for one simulated credit scenario.",
  row.names = FALSE
)
```

The figure below shows the same logic graphically for year 1 and year 5. The shaded area is the cumulative probability of default by that horizon.

```{r}
#| label: fig-hull-247-threshold-areas
#| fig-cap: "Default thresholds implied by Hull Example 24.7 cumulative PDs."
#| fig-width: 8
#| fig-height: 4
threshold_area_horizons <- c(1, 5)
threshold_area_labels <- paste0(
  "Year ", threshold_area_horizons,
  " PD = ", fmt_pct(pd_cum[threshold_area_horizons], 2)
)
threshold_area_colors <- c("#5DADE2", "#E74C3C")
names(threshold_area_colors) <- threshold_area_labels

threshold_area_data <- do.call(rbind, lapply(seq_along(threshold_area_horizons), function(j) {
  horizon_index <- threshold_area_horizons[j]
  data.frame(
    Horizon = threshold_area_labels[j],
    x = x,
    Density = D,
    Threshold = default_threshold[horizon_index],
    Shade = x <= default_threshold[horizon_index]
  )
}))
threshold_area_data$Horizon <- factor(threshold_area_data$Horizon, levels = threshold_area_labels)

threshold_label_data <- data.frame(
  Horizon = factor(threshold_area_labels, levels = threshold_area_labels),
  x = -3.75,
  y = 0.375,
  label = paste0("z = ", fmt_num(default_threshold[threshold_area_horizons], 3))
)

ggplot(threshold_area_data, aes(x = x, y = Density)) +
  geom_area(
    data = subset(threshold_area_data, Shade),
    aes(fill = Horizon),
    alpha = 0.88
  ) +
  geom_line(linewidth = 1.1, color = "#2C3E50") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#777777") +
  geom_vline(
    aes(xintercept = Threshold, color = Horizon),
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_label(
    data = threshold_label_data,
    aes(x = x, y = y, label = label, color = Horizon),
    inherit.aes = FALSE,
    hjust = 0,
    size = 3.1,
    label.size = 0.2,
    fill = "white"
  ) +
  facet_wrap(~Horizon, nrow = 1) +
  scale_fill_manual(values = threshold_area_colors, guide = "none") +
  scale_color_manual(values = threshold_area_colors, guide = "none") +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.42)) +
  labs(
    x = "Credit-state score X",
    y = "Standard normal density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    aspect.ratio = 1,
    strip.text = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

A second view uses the cumulative normal distribution directly. In this graph, the vertical axis already is the cumulative probability. The points show that each default threshold maps back to the PD input that produced it.

```{r}
#| label: fig-hull-247-threshold-check
#| fig-cap: "Checking that each Gaussian threshold reproduces the original cumulative PD."
threshold_curve_data <- data.frame(x = x, Probability = P)
threshold_points <- data.frame(
  Threshold = default_threshold,
  Cumulative_PD = pd_cum,
  Horizon = paste0("Y", credit_horizons)
)

ggplot(threshold_curve_data, aes(x = x, y = Probability)) +
  geom_line(linewidth = 1.1, color = "#2C3E50") +
  geom_hline(data = threshold_points, aes(yintercept = Cumulative_PD), linetype = "dotted", color = "#BBBBBB") +
  geom_vline(data = threshold_points, aes(xintercept = Threshold), linetype = "dotted", color = "#BBBBBB") +
  geom_point(data = threshold_points, aes(x = Threshold, y = Cumulative_PD, color = Horizon), size = 3) +
  geom_text(data = threshold_points, aes(x = Threshold, y = Cumulative_PD, label = Horizon), nudge_x = 0.12, size = 3.4) +
  scale_color_manual(values = colors2, guide = "none") +
  scale_y_continuous(labels = function(x) fmt_pct(x, 0)) +
  labs(
    x = "Credit-state score X",
    y = "Cumulative probability N(X)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

A zoomed version makes the relevant threshold region easier to inspect.

```{r}
#| label: fig-hull-247-threshold-check-zoom
#| fig-cap: "Zoom: Gaussian thresholds and cumulative PDs between -2.5 and -0.5."
ggplot(threshold_curve_data, aes(x = x, y = Probability)) +
  geom_line(linewidth = 1.1, color = "#2C3E50") +
  geom_hline(data = threshold_points, aes(yintercept = Cumulative_PD), linetype = "dotted", color = "#BBBBBB") +
  geom_vline(data = threshold_points, aes(xintercept = Threshold), linetype = "dotted", color = "#BBBBBB") +
  geom_point(data = threshold_points, aes(x = Threshold, y = Cumulative_PD, color = Horizon), size = 3.4) +
  geom_text(data = threshold_points, aes(x = Threshold, y = Cumulative_PD, label = Horizon), nudge_x = 0.04, size = 3.5) +
  scale_color_manual(values = colors2, guide = "none") +
  scale_y_continuous(labels = function(x) fmt_pct(x, 0)) +
  coord_cartesian(xlim = c(-2.5, -0.5), ylim = c(0, 0.2)) +
  labs(
    x = "Credit-state score X",
    y = "Cumulative probability N(X)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

At this stage, the chapter has built marginal default rules for each firm. The next step draws several credit-state scores in the same scenario. Independent draws represent borrowers whose credit states move separately. Correlated draws represent borrowers exposed to shared shocks. With a common copula correlation of `r fmt_num(hull_example_247_rho, 1)`, the probability of joint defaults changes while the individual PDs shown in this section are preserved.
