## One firm and the marginal default rule

```{r gaussian-one-firm-setup, include=FALSE}
source("R/gaussian-common.R")

one_firm_sims <- 100000
one_firm_horizon <- 5
one_firm_threshold <- default_threshold[one_firm_horizon]
one_firm_pd <- pd_cum[one_firm_horizon]

set.seed(130575)
one_firm_x <- rnorm(one_firm_sims)
one_firm_default <- one_firm_x <= one_firm_threshold
one_firm_sim_pd <- mean(one_firm_default)

one_firm_check <- data.frame(
  Horizon = paste(one_firm_horizon, "years"),
  Input_PD = fmt_pct(one_firm_pd, 2),
  Default_threshold = fmt_num(one_firm_threshold, 6),
  Simulated_PD = fmt_pct(one_firm_sim_pd, 2),
  Difference = fmt_pct(one_firm_sim_pd - one_firm_pd, 2)
)
names(one_firm_check) <- c(
  "Horizon",
  "Input PD",
  "Default threshold",
  "Simulated PD",
  "Difference"
)
```

Before linking firms through common shocks, it is useful to check the one-firm rule. For the 5-year horizon in Hull's Example 24.7, the cumulative PD is `r fmt_pct(one_firm_pd, 2)` and the default threshold is `r fmt_num(one_firm_threshold, 6)`.

The rule can be written as a two-case definition:

$$
D_i(5)=
\begin{cases}
1, & \text{if } X_i \leq N^{-1}(0.15),\\
0, & \text{if } X_i > N^{-1}(0.15).
\end{cases}
$$

The R code applies the same rule.

```{r}
one_firm_sims <- 100000
one_firm_threshold <- qnorm(0.15)

set.seed(130575)
one_firm_x <- rnorm(one_firm_sims)
one_firm_default <- one_firm_x <= one_firm_threshold
mean(one_firm_default)
```

The object `one_firm_default` is a TRUE/FALSE vector. R treats TRUE as 1 and FALSE as 0 when it computes a mean, so `mean(one_firm_default)` is the fraction of simulations in which the firm defaults.

With a finite number of random draws, the simulated PD lands close to `r fmt_pct(one_firm_pd, 2)`. That is the first sanity check. The simulation rule preserves the marginal PD before we link this firm to other firms.

```{r}
kable(
  one_firm_check,
  caption = "One-firm simulation check for the 5-year marginal PD.",
  escape = FALSE
)
```

Graphically, each simulated value of $X_i$ is a possible credit state for the firm. Lower values are weaker credit states. The firm defaults by year 5 when the simulated value falls to the left of the threshold.

```{r}
#| label: fig-one-firm-marginal-default-check
#| fig-cap: "One-firm simulation check with threshold comparison."
one_firm_hist <- hist(
  one_firm_x,
  breaks = 80,
  probability = TRUE,
  plot = FALSE
)
one_firm_hist_data <- data.frame(
  x_mid = one_firm_hist$mids,
  density = one_firm_hist$density,
  default_region = one_firm_hist$mids <= one_firm_threshold
)

ggplot(one_firm_hist_data, aes(x = x_mid, y = density, fill = default_region)) +
  geom_col(width = diff(one_firm_hist$breaks)[1], color = "white", linewidth = 0.1) +
  geom_vline(xintercept = one_firm_threshold, linewidth = 1, linetype = "dashed", color = "#E74C3C") +
  annotate(
    "label",
    x = 1.05,
    y = 0.37,
    hjust = 0,
    label = paste0(
      "Threshold = ", fmt_num(one_firm_threshold, 3),
      "\nSimulated PD = ", fmt_pct(one_firm_sim_pd, 2)
    ),
    size = 3.5,
    fill = "white",
    label.size = 0,
    linewidth = 0
  ) +
  scale_fill_manual(
    values = c("FALSE" = "#D6EAF8", "TRUE" = "#E74C3C"),
    labels = c("Survival region", "Default region"),
    name = NULL
  ) +
  coord_cartesian(xlim = c(-4, 4), ylim = c(0, 0.43)) +
  labs(
    x = "Simulated credit-state score X",
    y = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

This one-firm check verifies the marginal default mechanism. The copula model begins when we draw several firms' credit-state scores in the same scenario and let shared shocks make those scores correlated.
