## Default correlation and the portfolio problem

```{r gaussian-package-setup, include=FALSE}
source("R/gaussian-common.R")

copula_input_table <- data.frame(
  Object = c(
    "$Q_i(T)$",
    "$X_i$",
    "$z_i(T)=N^{-1}(Q_i(T))$",
    "$\\rho$",
    "$EAD_i$",
    "$R_i$",
    "$LGD_i=1-R_i$"
  ),
  Code = c(
    "`pd_cum`",
    "simulated normal variable",
    "`default_threshold <- qnorm(pd_cum)`",
    "`rho`",
    "`ead`",
    "`recovery`",
    "`lgd <- 1 - recovery`"
  ),
  Financial_role = c(
    "Cumulative default probability of firm $i$ by horizon $T$.",
    "Latent credit-quality state used by the copula.",
    "Default boundary on the latent normal scale.",
    "Copula correlation controlling how strongly firms move together.",
    "Exposure at default measuring the amount at risk.",
    "Fraction recovered after default.",
    "Fraction lost after default."
  )
)
names(copula_input_table) <- c("Object", "Code", "Financial role")
```

Hull introduces default correlation before introducing the Gaussian copula model. That order matters. The financial problem is portfolio co-default, where several borrowers can fail in the same bad state of the world.

Defaults cluster for economic reasons. Firms in the same industry can be exposed to the same demand shock. Firms in the same country can be affected by the same interest-rate, exchange-rate, or funding shock. A large default can also weaken suppliers, customers, banks, or other connected firms. This dependence leaves diversification incomplete even when a portfolio contains many borrowers.

The Gaussian copula model gives us a practical simulation machine for this problem. It takes individual default probabilities as inputs, adds a dependence structure, and produces simulated portfolio default events. Once we know which firms default in each simulation, we can add exposures and recoveries to obtain a loss distribution.

```{r}
#| label: fig-copula-credit-risk-pipeline
#| fig-cap: "Gaussian copula workflow for portfolio credit risk."
#| fig-width: 9
#| fig-height: 1.35
old_par <- par(mar = c(0.1, 0.2, 0.1, 0.2))
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0.35, 0.65))

box_x <- c(0.14, 0.38, 0.62, 0.86)
box_labels <- c(
  "Individual\nPDs",
  "Copula\ncorrelation",
  "Simulated\ndefaults",
  "Portfolio\nloss distribution"
)
box_width <- 0.18
box_height <- 0.22

for (i in seq_along(box_x)) {
  rect(
    box_x[i] - box_width / 2, 0.5 - box_height / 2,
    box_x[i] + box_width / 2, 0.5 + box_height / 2,
    col = "#F4F8FB", border = "#2C7FB8", lwd = 2
  )
  text(box_x[i], 0.5, box_labels[i], cex = 0.95)
}

arrows(
  box_x[-length(box_x)] + box_width / 2 + 0.015, 0.5,
  box_x[-1] - box_width / 2 - 0.015, 0.5,
  length = 0.08, lwd = 2, col = "#555555"
)
par(old_par)
```

The workflow separates three objects. Marginal default risk gives each firm its own cumulative default probability, denoted $Q_i(T)$. Dependence enters through the copula correlation $\rho$, which controls how much the latent credit-quality variables move together. Loss severity comes from exposures and recoveries, which convert default events into money losses.

This separation is why copulas are useful in credit risk. We can hold each firm's default probability fixed and change only the dependence assumption. When the tail of the loss distribution changes, joint default behavior drives that change.

```{r}
kable(
  copula_input_table,
  caption = "Objects used in the Gaussian copula credit-risk workflow.",
  escape = FALSE
)
```

One detail is worth making explicit before we code. In the Gaussian copula simulation, $\rho$ is the correlation between the latent normal variables $X_i$ and $X_j$. The sample correlation between the default indicators $D_i$ and $D_j$ is a derived quantity, where $D_i$ is 1 after default and 0 otherwise. The indicator correlation depends on both $\rho$ and the default threshold. Later in the chapter we compute joint default probabilities directly from the simulated defaults.

We now need one bridge between a default probability and a simulation rule. The bridge is a default threshold. Suppose firm $i$ has cumulative probability of default $Q_i(T)$ by horizon $T$. The Gaussian copula works with a latent standard normal variable $X_i$. The default threshold is

$$
z_i(T)=N^{-1}(Q_i(T)).
$$

The simulation rule is then

$$
\text{firm } i \text{ defaults by } T
\quad \Longleftrightarrow \quad
X_i \leq z_i(T).
$$

This rule is constructed so that the marginal default probability is preserved.

$$
P(X_i \leq z_i(T))=P(X_i \leq N^{-1}(Q_i(T)))=Q_i(T).
$$

In R, the equation is exactly the pair `qnorm()` and `pnorm()`. The function `qnorm()` translates a given cumulative PD into a normal threshold. The function `pnorm()` checks that the threshold reproduces the same cumulative PD.
