## Default correlation and the portfolio problem

```{r gaussian-package-setup, include=FALSE}
source("R/gaussian-common.R")

copula_input_table <- data.frame(
  Object = c(
    "$Q_i(T)$",
    "$X_i$",
    "$z_i(T)=N^{-1}(Q_i(T))$",
    "$\\rho$",
    "$EAD_i$",
    "$R_i$",
    "$LGD_i=1-R_i$"
  ),
  Code = c(
    "`pd_cum`",
    "simulated standard normal credit-state score",
    "`default_threshold <- qnorm(pd_cum)`",
    "`rho`",
    "`ead`",
    "`recovery`",
    "`lgd <- 1 - recovery`"
  ),
  Financial_role = c(
    "Cumulative default probability of firm $i$ by horizon $T$.",
    "Model score summarizing firm $i$'s credit condition in a scenario; lower values mean weaker credit.",
    "Default threshold; values below this boundary are defaults.",
    "Correlation between credit-state scores before converting them into 0/1 defaults.",
    "Exposure at default measuring the amount at risk.",
    "Fraction recovered after default.",
    "Fraction lost after default."
  )
)
names(copula_input_table) <- c("Object", "Code", "Financial role")
```

Imagine a bank with loans to ten manufacturing firms. The credit team may know the 5-year cumulative PD of each firm. That information is useful, but it answers a borrower-level question. It says how often each name defaults over five years. The portfolio manager also needs a portfolio-level answer. Do defaults tend to arrive one by one, or can several borrowers fail in the same bad credit environment?

That second question is the co-default problem. Co-default means that several borrowers default within the same horizon. A lender can absorb isolated defaults more easily than a cluster of defaults arriving together, so co-default directly affects portfolio loss risk.

Defaults cluster for economic reasons. Firms in the same industry can face the same fall in demand. Firms in the same country can face the same interest-rate, exchange-rate, or funding shock. A supplier can be hurt when a large customer defaults. A borrower that depends on refinancing can become fragile when market liquidity dries up. These shared forces leave diversification incomplete even when a portfolio contains many borrowers.

The Gaussian copula turns this idea into a disciplined scenario model. In each scenario, firm $i$ receives a credit-state score, denoted $X_i$. Think of $X_i$ as a compact summary of the credit environment facing that firm in that scenario. It stands in for the combined effect of demand, margins, funding access, refinancing conditions, collateral values, sector stress, and other credit-relevant forces. A high value of $X_i$ means a strong credit state. A low value means a weak credit state.

The modeling target is deliberately narrow: one standardized credit-state score per firm and per scenario. The default threshold then gives that score financial meaning. If a firm's 5-year cumulative PD is 15%, the threshold is chosen so that the worst 15% of simulated credit states are counted as defaults by year 5.

Simulation is useful here because a portfolio has many possible default combinations. With ten firms there are already $2^{10}=1{,}024$ possible default sets by a given horizon. The simulation creates many credit scenarios, applies the same default rule in each scenario, and counts the resulting portfolio defaults. The first check is mechanical: simulated default frequencies should reproduce the input PDs. After that check passes, the financial question is which value of $\rho$ is a reasonable description of common credit stress.

Dependence has a concrete credit-risk meaning here. The firms' credit-state scores tend to be weak together more often, or less often, than they would under independence. The copula correlation $\rho$ controls that shared movement. If $\rho$ is high, a bad scenario for one borrower is more likely to be a bad scenario for other borrowers too. The Gaussian copula procedure starts from individual cumulative PDs, converts them into default thresholds, draws correlated credit-state scores, and records which firms default in each scenario. Once we know which firms default, exposures and recoveries convert those default events into portfolio losses.

For the equal-correlation examples in this chapter, a simple one-factor version makes the shared shock visible:

$$
X_i=\sqrt{\rho}\,Z_0+\sqrt{1-\rho}\,Z_i.
$$

The score $X_i$ is the simulated credit state of firm $i$. The variable $Z_0$ is the common credit shock in the scenario, such as a recession, a funding squeeze, or a sector-wide stress event. The variable $Z_i$ is the firm-specific shock. If $Z_0$ is very low, many firms receive weak credit-state scores in the same simulation. The parameter $\rho$ controls how much each firm's score depends on that common shock. This is the mechanism that creates co-default risk.

The main simulation functions are shown below. They are reused throughout the chapter. The first function creates an equal-correlation matrix. The second function draws correlated credit-state scores. The third function converts those scores into default indicators by comparing each score with its default threshold. The remaining functions summarize default counts, convert defaults into portfolio losses, and report tail-risk measures.

```{r gaussian-core-simulation-functions}
make_correlation_matrix <- function(n_firms, rho) {
  sigma <- matrix(rho, nrow = n_firms, ncol = n_firms)
  diag(sigma) <- 1
  sigma
}

simulate_latent_credit <- function(n_sims, n_firms, rho, seed = 130575) {
  set.seed(seed)
  MASS::mvrnorm(
    n = n_sims,
    mu = rep(0, n_firms),
    Sigma = make_correlation_matrix(n_firms, rho)
  )
}

make_default_matrix <- function(latent_credit, threshold) {
  latent_credit <= threshold
}

summarize_default_counts <- function(default_matrix) {
  default_count <- rowSums(default_matrix)
  count_table <- as.data.frame(table(default_count))
  names(count_table) <- c("defaults", "frequency")
  count_table$defaults <- as.integer(as.character(count_table$defaults))
  count_table$probability <- count_table$frequency / nrow(default_matrix)
  count_table
}

portfolio_losses <- function(default_matrix, ead, recovery) {
  lgd_amount <- ead * (1 - recovery)
  as.numeric(default_matrix %*% lgd_amount)
}

loss_risk_summary <- function(losses) {
  var_95 <- unname(quantile(losses, 0.95, type = 1))
  var_99 <- unname(quantile(losses, 0.99, type = 1))
  var_999 <- unname(quantile(losses, 0.999, type = 1))

  c(
    expected_loss = mean(losses),
    median_loss = unname(quantile(losses, 0.50, type = 1)),
    var_95 = var_95,
    var_99 = var_99,
    var_999 = var_999,
    es_99 = mean(losses[losses >= var_99]),
    es_999 = mean(losses[losses >= var_999]),
    economic_capital_999 = var_999 - mean(losses)
  )
}
```

In an applied model, $\rho$ must be defended. It is a portfolio dependence assumption, so it should come from evidence, policy, or a transparent sensitivity exercise. The chapter treats $\rho$ as a controlled input because the pedagogical goal is to isolate the effect of co-default behavior on losses.

```{r rho-selection-table}
rho_selection_table <- data.frame(
  Source = c(
    "Historical defaults or migrations",
    "Equity or asset-return proxies",
    "CDS or bond-spread co-movement",
    "Regulatory or internal-model benchmark",
    "Stress and sensitivity analysis"
  ),
  `How it helps` = c(
    "Uses observed credit events by sector, rating, or region.",
    "Uses market data for public firms when default histories are sparse.",
    "Uses credit-market prices that react quickly to common stress.",
    "Uses a documented benchmark when a portfolio policy or capital framework requires one.",
    "Reports low, base, and high dependence cases when direct calibration is uncertain."
  ),
  `Main caution` = c(
    "Defaults are rare, so estimates can be noisy.",
    "Equity and asset returns are proxies for credit states, not defaults themselves.",
    "Spread co-movement includes liquidity, risk premia, and technical market effects.",
    "Benchmarks can be too coarse for a specific portfolio.",
    "Scenario choices need a clear reason; otherwise the parameter becomes arbitrary."
  ),
  check.names = FALSE
)

kable(
  rho_selection_table,
  caption = "Practical ways to justify the copula correlation input.",
  row.names = FALSE
)
```

A good credit report would state the base value of $\rho$, explain why it is reasonable for the portfolio, and show how the loss tail changes when $\rho$ moves. That is why the examples below repeatedly change $\rho$ while holding the marginal PDs fixed. The exercise is not cosmetic. It reveals how much of the portfolio loss comes from default clustering rather than from the individual default probabilities alone.

```{r}
#| label: fig-copula-credit-risk-pipeline
#| fig-cap: "Gaussian copula workflow for portfolio credit risk."
#| fig-width: 9
#| fig-height: 1.35
box_x <- c(0.14, 0.38, 0.62, 0.86)
box_labels <- c(
  "Individual\nPDs",
  "Shared\ncredit shocks",
  "Simulated\ndefaults",
  "Portfolio\nloss distribution"
)
box_width <- 0.18
box_height <- 0.22

pipeline_boxes <- data.frame(
  x = box_x,
  xmin = box_x - box_width / 2,
  xmax = box_x + box_width / 2,
  ymin = 0.5 - box_height / 2,
  ymax = 0.5 + box_height / 2,
  label = box_labels
)
pipeline_arrows <- data.frame(
  x = box_x[-length(box_x)] + box_width / 2 + 0.015,
  xend = box_x[-1] - box_width / 2 - 0.015,
  y = 0.5,
  yend = 0.5
)

ggplot() +
  geom_segment(
    data = pipeline_arrows,
    aes(x = x, xend = xend, y = y, yend = yend),
    linewidth = 0.8,
    color = "#555555",
    arrow = arrow(length = grid::unit(0.16, "cm"), type = "closed")
  ) +
  geom_rect(
    data = pipeline_boxes,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "#F4F8FB",
    color = "#2C7FB8",
    linewidth = 0.8
  ) +
  geom_text(
    data = pipeline_boxes,
    aes(x = x, y = 0.5, label = label),
    lineheight = 0.95,
    size = 3.5
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0.35, 0.65), clip = "off") +
  theme_void()
```

The workflow separates marginal default risk, co-default behavior, and loss severity. Marginal default risk describes each firm one at a time, before adding common shocks across firms. For firm $i$, the cumulative default probability $Q_i(T)$ is the probability that the firm has defaulted at some point between today and horizon $T$. It accumulates default risk over the interval from 0 to $T$, so a 5-year cumulative PD includes defaults that occur in years 1, 2, 3, 4, or 5.

Co-default behavior enters through the copula correlation $\rho$. The parameter $\rho$ controls how strongly the firms' credit-state scores share common shocks. When $\rho$ is higher, weak credit states tend to arrive for several firms in the same scenario. Lower values of the credit-state score mean weaker credit conditions, and default occurs when the score falls below its threshold. Loss severity comes from exposures and recoveries. After defaults are simulated, exposures and recoveries convert each default event into a money loss.

This separation is why copulas are useful in credit risk. We can hold each firm's default probability fixed and change the common-shock correlation. When the tail of the loss distribution changes, joint default behavior drives that change.

```{r}
kable(
  copula_input_table,
  caption = "Objects used in the Gaussian copula credit-risk workflow.",
  escape = FALSE
)
```

One detail is worth making explicit before we code. In the Gaussian copula simulation, $\rho$ is the correlation between the standardized normal credit-state scores $X_i$ and $X_j$. These are the continuous credit states we simulate first.

The binary default variable $Y_i(T)$ is created after the threshold rule is applied. It equals 1 when firm $i$ defaults by horizon $T$ and 0 otherwise. We use $Y_i(T)$ here to avoid confusing the default indicator with the promised debt payment $D$ from the Merton chapter. The sample correlation between $Y_i(T)$ and $Y_j(T)$ is therefore a result of the simulation, and it depends on both $\rho$ and the default thresholds. Later in the chapter we compute joint default probabilities directly from the simulated defaults.

We now need one bridge between a default probability and a simulation rule. The bridge is a default threshold. Suppose firm $i$ has cumulative probability of default $Q_i(T)$ by horizon $T$. The Gaussian copula represents the firm's credit state with a standard normal score $X_i$. The threshold is the point on the normal scale that leaves probability $Q_i(T)$ to its left:

$$
z_i(T)=N^{-1}(Q_i(T)).
$$

The simulation rule is then

$$
\text{firm } i \text{ defaults by } T
\quad \Longleftrightarrow \quad
X_i \leq z_i(T).
$$

The rule preserves the marginal default probability by construction. The inverse normal function $N^{-1}(\cdot)$ finds the value whose left-tail area is exactly $Q_i(T)$, so the probability of falling below the threshold is the original cumulative PD.

$$
P(X_i \leq z_i(T))=P(X_i \leq N^{-1}(Q_i(T)))=Q_i(T).
$$

For the 5-year PD used below, `pd_cum[5]` is `r fmt_pct(pd_cum[5], 1)`. The threshold is `r fmt_num(default_threshold[5], 4)`, obtained with `qnorm(pd_cum[5])`. A simulated value of $X_i$ below `r fmt_num(default_threshold[5], 4)` is counted as a default by year 5. Since `pnorm(default_threshold[5])` returns `r fmt_pct(threshold_check[5], 1)`, the threshold check recovers the same cumulative PD.

This is why the code uses the pair `qnorm()` and `pnorm()`. The function `qnorm()` translates a cumulative PD into a normal threshold. The function `pnorm()` is the check that the threshold reproduces the same cumulative PD.
