## Ten firms

```{r gaussian-ten-firms-setup, include=FALSE}
source("R/gaussian-common.R")
```

The original @Hull example proposes a 10 firm case and here we implement this example following a simulation approach. First, we need a $10\times10$ correlation matrix to produce the new $x$ values using a random multi-variate distribution algorithm. According to @Hull example the copula default correlations between each pair of companies is 0.2. The code below has the option to vary the default correlation given a uniform random distribution.

The new $10\times10$ correlation matrix is then:

```{r}
# Create the correlation matrix.
m <- 10 # number of firms.
n <- 1000000 # number of simulations per firm.
x <- matrix(rep(0.2, m * m), m, m) 
ind <- lower.tri(x) 
x[ind] <- t(x)[ind] 
diag(x) = 1
kable(x, caption = "Correlation matrix 0.2.")
```
Now, we can simulate the multivariate normal distribution. The variable <tt>`X10`</tt> length is 1,000,000 for each firm. We choose a higher number of simulations because otherwise the most restrictive constraints would not lead to any observations.

```{r}
# Create the simulated cases.
set.seed(130575) # Reproducibility.
X10 <- mvrnorm(n, mu = rep(0, m), Sigma = x)
X10 <- data.frame(X10) # 10,000,000 observations.
colnames(X10) <- c("f1", "f2", "f3", "f4", "f5", "f6", "f7", "f8", "f9", "f10")
kable(head(X10), caption = "10 firms' performance, 1,000,000 simulations.",
      digits = 3)
```

How do we extract those cases in which all 10 firms will default in 5 years (at the same time)? Here are the first 6 of those cases. Note that in all cases the <tt>`X10`</tt> values are lower than $-1.036433$.

```{r}
# Given that we have 10 firms, it is easier to use filter_all function.
# Although probably this could be simplified even further.
y5.all.2 <- filter_all(X10, all_vars(. < x.y[5]))
y4.all.2 <- filter_all(X10, all_vars(. < x.y[4])) 
y3.all.2 <- filter_all(X10, all_vars(. < x.y[3])) 
y2.all.2 <- filter_all(X10, all_vars(. < x.y[2])) 
y1.all.2 <- filter_all(X10, all_vars(. < x.y[1])) 
```

Let's analyze the case of 5 years.

```{r}
kable(head(y5.all.2), 
      caption = "Cases in which all 10 firms default in five years.", 
      digits = 3)
```

```{r}
kable((head(y5.all.2) < x.y[5]), caption = "Check if all of them default.")
```

How many cases are there?

```{r}
nrow(y5.all.2)
```
Which are those 72 cases?

```{r}
# Here, I compare only firm 1 as if firm 1 defaults, then the rest default.
kable(matrix(which(X10[,1] %in% y5.all.2[,1]), 6, 9), 
      caption = "Which of the 1,000,000 cases represent a default of all 
      firms in five years?")
```

In total, we only have 72 cases. This is, the 10 firms will default at the same time in 5 years in 72 out of 1,000,000 total cases. For the rest of the years, the cases are less frequent, in fact we have zero cases for year 1, 2 and 3.

How do we extract those cases in which at least one of the 10 firms will default? 

```{r}
y5.any.2 <- filter_all(X10, any_vars(. < x.y[5]))
y4.any.2 <- filter_all(X10, any_vars(. < x.y[4]))
y3.any.2 <- filter_all(X10, any_vars(. < x.y[3]))
y2.any.2 <- filter_all(X10, any_vars(. < x.y[2]))
y1.any.2 <- filter_all(X10, any_vars(. < x.y[1]))
```

Here are the first 6 cases for the 5-year default.

```{r}
kable(head(y5.any.2), caption = "At least one firm default in five years.",
      digits = 3)
```

```{r}
kable((head(y5.any.2) < x.y[5]), caption = "Check which one(s) default.")
```

How many cases are these?

```{r}
nrow(y5.any.2)
```

In total, we have 682,148 cases. This is, at least one of 10 firms will default in 5 years in 682,148 of 1,000,000 cases. 

And how to convert them into probabilities?

```{r}
atleastone02 <- t(data.frame(nrow(y1.any.2), nrow(y2.any.2),
                nrow(y3.any.2), nrow(y4.any.2), nrow(y5.any.2)))
all02 <- t(data.frame(nrow(y1.all.2), nrow(y2.all.2),
                nrow(y3.all.2), nrow(y4.all.2), nrow(y5.all.2)))
res02 <- data.frame(all02 / n, atleastone02 / n)
rownames(res02) <- n.year
colnames(res02) <- c("All firms", "At least one")
kable(res02, caption = "Probabilities of default (10 firms, corr=0.2).")
```

Let's see the difference when we assume a different correlation matrix. This case, the correlation vary randomly between 0.45 and 0.65.

```{r}
m <- 10 # number of firms
n <- 1000000 # number of simulations
set.seed(130575)
x <- matrix(runif(m * m, 0.45, 0.65), m, m) 
ind <- lower.tri(x) 
x[ind] <- t(x)[ind] 
diag(x) = 1
kable(x, caption = "Correlation between 0.45 and 0.65.", digits = 3)
```

The resulting values of <tt>`Xr`</tt> are:

```{r}
set.seed(130575)
Xr <- mvrnorm(n, mu = rep(0, m), Sigma = x)
Xr <- data.frame(Xr)
kable(head(Xr), caption = "Firms' performance.", digits = 3)
```

Extract the cases in which all ten firms default at the same time in 5 years and the cases in which any firm default at the same time in 5 years.

```{r}
Xr <- as_tibble(Xr)
# All firms.
Xr.y1.all <- filter_all(Xr, all_vars(. < x.y[1])) 
Xr.y2.all <- filter_all(Xr, all_vars(. < x.y[2])) 
Xr.y3.all <- filter_all(Xr, all_vars(. < x.y[3])) 
Xr.y4.all <- filter_all(Xr, all_vars(. < x.y[4]))
Xr.y5.all <- filter_all(Xr, all_vars(. < x.y[5])) 
# At least one firm.
Xr.y1.any <- filter_all(Xr, any_vars(. < x.y[1]))
Xr.y2.any <- filter_all(Xr, any_vars(. < x.y[2]))
Xr.y3.any <- filter_all(Xr, any_vars(. < x.y[3]))
Xr.y4.any <- filter_all(Xr, any_vars(. < x.y[4]))
Xr.y5.any <- filter_all(Xr, any_vars(. < x.y[5]))
```

All firms default in 5,935 cases out of 1,000,000.

```{r}
nrow(Xr.y5.all)
```
Let's see the first 6 cases:

```{r}
kable(head(Xr.y5.all), 
      caption = "Cases in which all 10 firms default in five years.", 
      digits = 3)
```

Verify that those cases default.

```{r}
kable((head(Xr.y5.all) < x.y[5]), caption = "Check if all of them default.")
```

At least one firm default in 497,987 out of 1,000,000.

```{r}
nrow(Xr.y5.any)
```

```{r}
kable(head(Xr.y5.any), caption = "At least one firm default in five years.",
      digits = 3)
```

```{r}
kable((head(Xr.y5.any) < x.y[5]), caption = "Check which one(s) default.")
```

And the corresponding probabilities:

```{r}
allR <- t(data.frame(nrow(Xr.y1.all), nrow(Xr.y2.all),
        nrow(Xr.y3.all), nrow(Xr.y4.all), nrow(Xr.y5.all)))
atleastoneR <- t(data.frame(nrow(Xr.y1.any), nrow(Xr.y2.any),
        nrow(Xr.y3.any), nrow(Xr.y4.any), nrow(Xr.y5.any)))
resR <- data.frame(allR / n, atleastoneR / n)
ans <- data.frame(res02, resR)
rownames(ans) <- n.year
colnames(ans) <- c("All (corr=0.2)", "At least one (corr=0.2)",
                    "All (corr=rand)", "At least one (corr=rand)")
kable(ans, caption = "Probability of default.", 
      row.names = TRUE)
```
