## One firm and the marginal default rule

```{r gaussian-one-firm-setup, include=FALSE}
source("R/gaussian-common.R")

one_firm_sims <- 100000
one_firm_horizon <- 5
one_firm_threshold <- default_threshold[one_firm_horizon]
one_firm_pd <- pd_cum[one_firm_horizon]

set.seed(130575)
one_firm_x <- rnorm(one_firm_sims)
one_firm_default <- one_firm_x <= one_firm_threshold
one_firm_sim_pd <- mean(one_firm_default)

one_firm_check <- data.frame(
  Horizon = paste(one_firm_horizon, "years"),
  Input_PD = fmt_pct(one_firm_pd, 2),
  Default_threshold = fmt_num(one_firm_threshold, 6),
  Simulated_PD = fmt_pct(one_firm_sim_pd, 2),
  Difference = fmt_pct(one_firm_sim_pd - one_firm_pd, 2)
)
names(one_firm_check) <- c(
  "Horizon",
  "Input PD",
  "Default threshold",
  "Simulated PD",
  "Difference"
)
```

Before adding dependence across firms, it is useful to check the one-firm rule. For the 5-year horizon in Hull's Example 24.7, the cumulative PD is `r fmt_pct(one_firm_pd, 2)` and the default threshold is `r fmt_num(one_firm_threshold, 6)`.

The equation is

$$
D_i(5)=1\{X_i \leq N^{-1}(0.15)\}.
$$

The R code applies the same rule.

```{r}
one_firm_sims <- 100000
one_firm_threshold <- qnorm(0.15)

set.seed(130575)
one_firm_x <- rnorm(one_firm_sims)
one_firm_default <- one_firm_x <= one_firm_threshold
mean(one_firm_default)
```

With a finite number of random draws, the simulated PD lands close to `r fmt_pct(one_firm_pd, 2)`. That is the first sanity check. The simulation rule preserves the marginal PD before we ask it to model dependence.

```{r}
kable(
  one_firm_check,
  caption = "One-firm simulation check for the 5-year marginal PD.",
  escape = FALSE
)
```

Graphically, each simulated value of $X_i$ is a possible latent credit-quality state. The firm defaults by year 5 when the simulated value falls to the left of the threshold.

```{r}
#| label: fig-one-firm-marginal-default-check
#| fig-cap: "One-firm simulation check with threshold comparison."
one_firm_hist <- hist(
  one_firm_x,
  breaks = 80,
  probability = TRUE,
  plot = FALSE
)
one_firm_bar_color <- ifelse(
  one_firm_hist$mids <= one_firm_threshold,
  adjustcolor("#E74C3C", alpha.f = 0.85),
  "#D6EAF8"
)
plot(
  one_firm_hist,
  freq = FALSE,
  col = one_firm_bar_color,
  border = "white",
  main = NULL,
  ylab = "Density",
  xlab = "Simulated latent credit-quality variable X"
)
lines(x, D, lwd = 3, col = "#2C3E50")
abline(v = one_firm_threshold, lwd = 3, lty = 2, col = "#E74C3C")
legend(
  "topright",
  legend = c(
    "Default region",
    paste0("Threshold = ", fmt_num(one_firm_threshold, 3)),
    paste0("Simulated PD = ", fmt_pct(one_firm_sim_pd, 2))
  ),
  fill = c(adjustcolor("#E74C3C", alpha.f = 0.85), NA, NA),
  border = c("white", NA, NA),
  lty = c(NA, 2, NA),
  lwd = c(NA, 3, NA),
  col = c(NA, "#E74C3C", NA),
  bty = "n",
  text.col = c("#E74C3C", "#E74C3C", "#2C3E50")
)
```

This one-firm check verifies the marginal default mechanism. The copula model begins when we simulate several latent variables together and allow them to be correlated.
