## Two firms and joint defaults

```{r gaussian-two-firms-setup, include=FALSE}
source("R/gaussian-common.R")

two_firm_sims <- 100000
two_firm_threshold <- default_threshold[5]
two_firm_pd <- pd_cum[5]
two_firm_rhos <- c(0, 0.2, 0.5)

simulate_two_firm_summary <- function(rho) {
  latent <- simulate_latent_credit(
    n_sims = two_firm_sims,
    n_firms = 2,
    rho = rho,
    seed = 130575
  )
  defaults <- make_default_matrix(latent, two_firm_threshold)
  default_a <- defaults[, 1]
  default_b <- defaults[, 2]
  data.frame(
    rho = rho,
    pd_firm_a = mean(default_a),
    pd_firm_b = mean(default_b),
    both_default = mean(default_a & default_b),
    at_least_one = mean(default_a | default_b),
    default_indicator_corr = cor(as.numeric(default_a), as.numeric(default_b))
  )
}

two_firm_numeric <- do.call(rbind, lapply(two_firm_rhos, simulate_two_firm_summary))
two_firm_independent_both <- two_firm_pd * two_firm_pd
two_firm_independent_any <- 1 - (1 - two_firm_pd)^2

two_firm_table <- data.frame(
  Copula_correlation = fmt_num(two_firm_numeric$rho, 1),
  Simulated_PD_A = fmt_pct(two_firm_numeric$pd_firm_a, 2),
  Simulated_PD_B = fmt_pct(two_firm_numeric$pd_firm_b, 2),
  P_both_default = fmt_pct(two_firm_numeric$both_default, 2),
  P_at_least_one = fmt_pct(two_firm_numeric$at_least_one, 2),
  Default_indicator_corr = fmt_num(two_firm_numeric$default_indicator_corr, 4)
)
names(two_firm_table) <- c(
  "rho",
  "PD A",
  "PD B",
  "P(A and B)",
  "P(A or B)",
  "Default corr."
)

two_firm_plot_data <- lapply(two_firm_rhos, function(rho) {
  latent <- simulate_latent_credit(
    n_sims = 6000,
    n_firms = 2,
    rho = rho,
    seed = 130575
  )
  data.frame(firm_a = latent[, 1], firm_b = latent[, 2], rho = rho)
})
```

Now we can see what the copula adds. Suppose both firms have the same 5-year cumulative PD, `r fmt_pct(two_firm_pd, 2)`. If the two default events were independent, the probability that both default by year 5 would be

$$
P(A\text{ defaults and }B\text{ defaults})
=`r fmt_pct(two_firm_pd, 2)`\times `r fmt_pct(two_firm_pd, 2)`
=`r fmt_pct(two_firm_independent_both, 2)`.
$$

The probability that at least one defaults would be

$$
P(A\text{ or }B\text{ defaults})
=1-(1-`r fmt_pct(two_firm_pd, 2)`)^2
=`r fmt_pct(two_firm_independent_any, 2)`.
$$

Those formulas rely on independence. The Gaussian copula lets us keep each marginal PD near `r fmt_pct(two_firm_pd, 2)` while changing how often the firms default together.

The covariance matrix carries the dependence.

$$
\Sigma=
\begin{bmatrix}
1 & \rho\\
\rho & 1
\end{bmatrix}.
$$

```{r}
rho <- 0.2
sigma <- matrix(c(1, rho, rho, 1), nrow = 2)

set.seed(130575)
latent <- MASS::mvrnorm(
  n = 100000,
  mu = c(0, 0),
  Sigma = sigma
)

threshold <- qnorm(0.15)
default_a <- latent[, 1] <= threshold
default_b <- latent[, 2] <= threshold

mean(default_a)
mean(default_b)
mean(default_a & default_b)
```

The first two probabilities check the marginal PDs. The third probability is the joint default probability. That is where dependence becomes visible.

```{r}
kable(
  two_firm_table,
  caption = "Two-firm default probabilities under different Gaussian copula correlations.",
  escape = FALSE
)
```

The table demonstrates the main point. The marginal PDs stay close to `r fmt_pct(two_firm_pd, 2)`, while the probability of both firms defaulting rises as $\rho$ rises. The default-indicator correlation is lower than the latent copula correlation because default is a tail event created after applying the threshold.

The scatter plots show the same mechanism. Each dot is one simulated pair $(X_A,X_B)$. A firm defaults when its latent variable is below the threshold. The lower-left region is the joint-default region.

```{r}
#| label: fig-two-firm-copula-scatter
#| fig-cap: "Two-firm Gaussian copula simulations with more joint tail outcomes."
#| fig-width: 9.2
#| fig-height: 3.4
old_par <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1), pty = "s")

for (plot_data in two_firm_plot_data) {
  rho_value <- unique(plot_data$rho)
  both_default <- plot_data$firm_a <= two_firm_threshold &
    plot_data$firm_b <= two_firm_threshold
  point_color <- ifelse(both_default, "#E74C3C", "#56657355")
  plot(
    plot_data$firm_a,
    plot_data$firm_b,
    pch = 19,
    cex = 0.35,
    col = point_color,
    xlim = c(-4, 4),
    ylim = c(-4, 4),
    asp = 1,
    xlab = "Firm A latent variable",
    ylab = "Firm B latent variable",
    main = paste0("rho = ", fmt_num(rho_value, 1))
  )
  abline(v = two_firm_threshold, h = two_firm_threshold, lty = 2, lwd = 2)
  legend(
    "bottomright",
    legend = paste0("Joint defaults ", fmt_int(sum(both_default))),
    bty = "n"
  )
}

par(old_par)
```

This first credit-risk application of the copula separates the individual credit quality of each firm from the dependence between firms. The PD of each firm stays fixed, while the probability of clustered defaults changes.
