## Tail dependence with a t-copula

```{r gaussian-t-copula-setup, include=FALSE}
source("R/gaussian-common.R")

tcopula_sims <- 100000
tcopula_n_firms <- 10
tcopula_rho <- hull_example_247_rho
tcopula_pd <- pd_cum[5]
tcopula_total_ead <- 100e6
tcopula_ead <- rep(tcopula_total_ead / tcopula_n_firms, tcopula_n_firms)
tcopula_recovery <- rep(0.40, tcopula_n_firms)

tcopula_design <- data.frame(
  copula = c("Gaussian", "t-copula", "t-copula"),
  df = c(NA, 10, 4)
)

simulate_tail_dependence_case <- function(copula, df) {
  if (copula == "Gaussian") {
    latent <- simulate_latent_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      seed = 130575
    )
    threshold <- qnorm(tcopula_pd)
  } else {
    latent <- simulate_t_copula_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      df = df,
      seed = 130575
    )
    threshold <- qt(tcopula_pd, df = df)
  }

  defaults <- make_default_matrix(latent, threshold)
  losses <- portfolio_losses(defaults, tcopula_ead, tcopula_recovery)

  list(
    defaults = defaults,
    default_count = rowSums(defaults),
    losses = losses,
    events = summarize_default_events(defaults),
    loss_summary = loss_risk_summary(losses),
    mean_pd = mean(defaults)
  )
}

tcopula_results <- Map(
  simulate_tail_dependence_case,
  tcopula_design$copula,
  tcopula_design$df
)
names(tcopula_results) <- c("Gaussian", "t df = 10", "t df = 4")

tcopula_events <- do.call(rbind, lapply(tcopula_results, function(x) x$events))
tcopula_losses <- do.call(rbind, lapply(tcopula_results, function(x) x$loss_summary))

tcopula_df_label <- ifelse(is.na(tcopula_design$df), "-", fmt_int(tcopula_design$df))

tcopula_default_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  Mean_PD = fmt_pct(sapply(tcopula_results, function(x) x$mean_pd), 2),
  Expected_defaults = fmt_num(tcopula_events[, "expected_defaults"], 3),
  P_three_or_more = fmt_pct(tcopula_events[, "three_or_more"], 2),
  P_five_or_more = fmt_pct(tcopula_events[, "five_or_more"], 2),
  P_all_ten = fmt_pct(tcopula_events[, "all_default"], 3)
)
names(tcopula_default_table) <- c(
  "Copula",
  "df",
  "Mean PD",
  "E[defaults]",
  "P(>=3)",
  "P(>=5)",
  "P(10)"
)

tcopula_loss_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  EL = fmt_musd(tcopula_losses[, "expected_loss"]),
  VaR_99 = fmt_musd(tcopula_losses[, "var_99"]),
  VaR_99_9 = fmt_musd(tcopula_losses[, "var_999"]),
  ES_99_9 = fmt_musd(tcopula_losses[, "es_999"]),
  Capital_99_9 = fmt_musd(tcopula_losses[, "economic_capital_999"])
)
names(tcopula_loss_table) <- c(
  "Copula",
  "df",
  "EL",
  "VaR 99%",
  "VaR 99.9%",
  "ES 99.9%",
  "Capital 99.9%"
)

tcopula_count_matrix <- sapply(tcopula_results, function(x) {
  as.numeric(tabulate(x$default_count + 1, nbins = tcopula_n_firms + 1)) /
    tcopula_sims
})
colnames(tcopula_count_matrix) <- names(tcopula_results)
```

The Gaussian copula uses normal credit-state scores. The correlation parameter makes weak credit states arrive together more often. The normal distribution controls how much probability sits in extreme credit states. A t-copula keeps the same threshold idea and changes the tail shape.

For the Gaussian copula, one convenient one-factor representation is

$$
X_i^{G}=\sqrt{\rho}\,Z_0+\sqrt{1-\rho}\,Z_i.
$$

For the t-copula, use the same normal factor structure and divide by a random scale shock:

$$
X_i^{t}=
\frac{\sqrt{\rho}\,Z_0+\sqrt{1-\rho}\,Z_i}
{\sqrt{S/\nu}},
\qquad S\sim\chi^2_{\nu}.
$$

The parameter $\nu$ is the number of degrees of freedom. Lower $\nu$ means heavier tails, so extreme credit states occur more often. The default threshold changes from $N^{-1}(PD)$ to $t_{\nu}^{-1}(PD)$, so each firm keeps the same marginal PD. In code, this is the move from `qnorm(tcopula_pd)` to `qt(tcopula_pd, df = df)`.

The scale shock is shared by all firms in a scenario. When $S/\nu$ is small, the denominator is small and the scenario becomes more extreme for the whole portfolio. Negative credit shocks are pushed further into the left tail, which creates more scenarios where several firms cross their default thresholds together. The model therefore changes the clustering of defaults without changing the average PD assigned to each firm.

The code below builds the comparison. It keeps the same number of firms, PD, exposure, recovery, and correlation. Then it changes only the copula family and, for the t-copula, the degrees of freedom.

```{r t-copula-simulation-workflow}
simulate_t_copula_credit <- function(n_sims, n_firms, rho, df, seed = 130575) {
  set.seed(seed)
  gaussian_latent <- MASS::mvrnorm(
    n = n_sims,
    mu = rep(0, n_firms),
    Sigma = make_correlation_matrix(n_firms, rho)
  )
  scale_shock <- sqrt(rchisq(n_sims, df = df) / df)
  gaussian_latent / scale_shock
}

tcopula_sims <- 100000
tcopula_n_firms <- 10
tcopula_rho <- hull_example_247_rho
tcopula_pd <- pd_cum[5]
tcopula_total_ead <- 100e6
tcopula_ead <- rep(tcopula_total_ead / tcopula_n_firms, tcopula_n_firms)
tcopula_recovery <- rep(0.40, tcopula_n_firms)

tcopula_design <- data.frame(
  copula = c("Gaussian", "t-copula", "t-copula"),
  df = c(NA, 10, 4)
)

simulate_tail_dependence_case <- function(copula, df) {
  if (copula == "Gaussian") {
    latent <- simulate_latent_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      seed = 130575
    )
    threshold <- qnorm(tcopula_pd)
  } else {
    latent <- simulate_t_copula_credit(
      n_sims = tcopula_sims,
      n_firms = tcopula_n_firms,
      rho = tcopula_rho,
      df = df,
      seed = 130575
    )
    threshold <- qt(tcopula_pd, df = df)
  }

  defaults <- make_default_matrix(latent, threshold)
  losses <- portfolio_losses(defaults, tcopula_ead, tcopula_recovery)

  list(
    defaults = defaults,
    default_count = rowSums(defaults),
    losses = losses,
    events = summarize_default_events(defaults),
    loss_summary = loss_risk_summary(losses),
    mean_pd = mean(defaults)
  )
}

tcopula_results <- Map(
  simulate_tail_dependence_case,
  tcopula_design$copula,
  tcopula_design$df
)
names(tcopula_results) <- c("Gaussian", "t df = 10", "t df = 4")

tcopula_events <- do.call(rbind, lapply(tcopula_results, function(x) x$events))
tcopula_losses <- do.call(rbind, lapply(tcopula_results, function(x) x$loss_summary))

tcopula_df_label <- ifelse(is.na(tcopula_design$df), "-", fmt_int(tcopula_design$df))

tcopula_default_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  Mean_PD = fmt_pct(sapply(tcopula_results, function(x) x$mean_pd), 2),
  Expected_defaults = fmt_num(tcopula_events[, "expected_defaults"], 3),
  P_three_or_more = fmt_pct(tcopula_events[, "three_or_more"], 2),
  P_five_or_more = fmt_pct(tcopula_events[, "five_or_more"], 2),
  P_all_ten = fmt_pct(tcopula_events[, "all_default"], 3)
)
names(tcopula_default_table) <- c(
  "Copula",
  "df",
  "Mean PD",
  "E[defaults]",
  "P(>=3)",
  "P(>=5)",
  "P(10)"
)

tcopula_loss_table <- data.frame(
  Copula = tcopula_design$copula,
  df = tcopula_df_label,
  EL = fmt_musd(tcopula_losses[, "expected_loss"]),
  VaR_99 = fmt_musd(tcopula_losses[, "var_99"]),
  VaR_99_9 = fmt_musd(tcopula_losses[, "var_999"]),
  ES_99_9 = fmt_musd(tcopula_losses[, "es_999"]),
  Capital_99_9 = fmt_musd(tcopula_losses[, "economic_capital_999"])
)
names(tcopula_loss_table) <- c(
  "Copula",
  "df",
  "EL",
  "VaR 99%",
  "VaR 99.9%",
  "ES 99.9%",
  "Capital 99.9%"
)

tcopula_count_matrix <- sapply(tcopula_results, function(x) {
  as.numeric(tabulate(x$default_count + 1, nbins = tcopula_n_firms + 1)) /
    tcopula_sims
})
colnames(tcopula_count_matrix) <- names(tcopula_results)
```

The comparison below keeps the 5-year PD at `r fmt_pct(tcopula_pd, 2)`, the copula correlation at `r fmt_num(tcopula_rho, 1)`, the same 10 firms, the same total exposure, and the same recovery rate. The only change is the copula tail shape.

```{r}
kable(
  tcopula_default_table,
  caption = "Default clustering under Gaussian and t-copulas.",
  row.names = FALSE,
  escape = FALSE
)
```

The average PD stays near `r fmt_pct(tcopula_pd, 2)`. The tail probabilities move because the t-copula gives more weight to scenarios where many firms experience weak credit states together. This is the credit-risk reason to care about tail dependence.

For example, the probability of five or more defaults rises from `r fmt_pct(tcopula_events["Gaussian", "five_or_more"], 2)` under the Gaussian copula to `r fmt_pct(tcopula_events["t df = 4", "five_or_more"], 2)` under the heavier-tailed t-copula. The probability that all ten firms default rises from `r fmt_pct(tcopula_events["Gaussian", "all_default"], 3)` to `r fmt_pct(tcopula_events["t df = 4", "all_default"], 3)`. The marginal PD remains anchored near 15%, so the difference is coming from joint-tail behavior.

The same default simulations can be translated into losses using the portfolio loss formula from the previous section.

```{r}
kable(
  tcopula_loss_table,
  caption = "Portfolio loss measures under Gaussian and t-copulas.",
  row.names = FALSE,
  escape = FALSE
)
```

Expected loss remains anchored by PD, EAD, and LGD. Tail loss responds to the co-default model. This ten-firm portfolio has coarse $6 million loss steps, so VaR can land on the same step for two copulas. Expected shortfall is a useful companion because it averages the losses beyond the percentile.

```{r}
#| label: fig-gaussian-vs-t-default-counts
#| fig-cap: "Default-count distribution under Gaussian and t-copulas."
default_counts <- 0:tcopula_n_firms
tcopula_count_colors <- c("#2C7FB8", "#F39C12", "#C0392B")
tcopula_count_plot_data <- data.frame(
  Defaults = rep(default_counts, times = ncol(tcopula_count_matrix)),
  Probability = as.vector(tcopula_count_matrix),
  Copula = factor(
    rep(colnames(tcopula_count_matrix), each = length(default_counts)),
    levels = colnames(tcopula_count_matrix)
  )
)

ggplot(
  tcopula_count_plot_data,
  aes(x = Defaults, y = Probability, color = Copula, group = Copula)
) +
  geom_line(linewidth = 1.1, alpha = 0.85) +
  geom_point(size = 3.2, alpha = 0.82) +
  scale_color_manual(values = tcopula_count_colors) +
  scale_x_continuous(breaks = default_counts) +
  scale_y_continuous(
    labels = function(x) fmt_pct(x, 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    x = "Number of defaults by year 5",
    y = "Simulated probability",
    color = "Copula"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 12, 5.5, 5.5)
  )
```

This section gives the key model-risk message. A copula is more than a correlation input. It also chooses how much probability is assigned to joint tail events.
