## Credit VaR under copula model risk

```{r gaussian-credit-var-bridge-setup, include=FALSE}
source("R/gaussian-common.R")

credit_var_inputs <- list(
  ead = 100e6,
  pd = 0.02,
  recovery = 0.60,
  rho = 0.10,
  confidence = c(0.95, 0.99, 0.999)
)

vasicek_default_rate <- function(pd, rho, confidence) {
  pnorm(
    (qnorm(pd) + sqrt(rho) * qnorm(confidence)) /
      sqrt(1 - rho)
  )
}

vasicek_credit_var <- function(ead, pd, recovery, rho, confidence) {
  ead * (1 - recovery) * vasicek_default_rate(pd, rho, confidence)
}

simulate_granular_gaussian_loss <- function(n_sims, n_names, ead, pd, recovery, rho, seed) {
  set.seed(seed)
  systematic_factor <- rnorm(n_sims)
  conditional_pd <- pnorm(
    (qnorm(pd) - sqrt(rho) * systematic_factor) / sqrt(1 - rho)
  )
  default_count <- rbinom(n_sims, n_names, conditional_pd)
  default_count * (ead / n_names) * (1 - recovery)
}

simulate_granular_t_loss <- function(n_sims, n_names, ead, pd, recovery, rho, df, seed) {
  set.seed(seed)
  systematic_factor <- rnorm(n_sims)
  scale_shock <- sqrt(rchisq(n_sims, df = df) / df)
  conditional_pd <- pnorm(
    (qt(pd, df = df) * scale_shock - sqrt(rho) * systematic_factor) /
      sqrt(1 - rho)
  )
  default_count <- rbinom(n_sims, n_names, conditional_pd)
  default_count * (ead / n_names) * (1 - recovery)
}

credit_var_rate <- vasicek_default_rate(
  credit_var_inputs$pd,
  credit_var_inputs$rho,
  credit_var_inputs$confidence
)

credit_var_loss <- vasicek_credit_var(
  credit_var_inputs$ead,
  credit_var_inputs$pd,
  credit_var_inputs$recovery,
  credit_var_inputs$rho,
  credit_var_inputs$confidence
)

credit_expected_loss <- credit_var_inputs$ead *
  (1 - credit_var_inputs$recovery) *
  credit_var_inputs$pd

credit_var_focus <- length(credit_var_inputs$confidence)
credit_var_focus_alpha <- credit_var_inputs$confidence[credit_var_focus]
credit_var_q_pd <- qnorm(credit_var_inputs$pd)
credit_var_q_conf <- qnorm(credit_var_inputs$confidence[credit_var_focus])
credit_var_numerator <- credit_var_q_pd +
  sqrt(credit_var_inputs$rho) * credit_var_q_conf
credit_var_denominator <- sqrt(1 - credit_var_inputs$rho)
credit_var_standardized <- credit_var_numerator / credit_var_denominator

set.seed(130575)
systematic_factor <- rnorm(200000)
conditional_default_rate <- pnorm(
  (qnorm(credit_var_inputs$pd) -
     sqrt(credit_var_inputs$rho) * systematic_factor) /
    sqrt(1 - credit_var_inputs$rho)
)
vasicek_loss <- credit_var_inputs$ead *
  (1 - credit_var_inputs$recovery) *
  conditional_default_rate

simulated_var_loss <- as.numeric(
  quantile(vasicek_loss, credit_var_inputs$confidence, type = 1)
)
simulated_es_loss <- sapply(simulated_var_loss, function(var_value) {
  mean(vasicek_loss[vasicek_loss >= var_value])
})

credit_var_input_table <- data.frame(
  Quantity = c(
    "Total exposure",
    "One-year PD",
    "Recovery rate",
    "LGD",
    "Copula correlation"
  ),
  Value = c(
    fmt_musd(credit_var_inputs$ead),
    fmt_pct(credit_var_inputs$pd, 2),
    fmt_pct(credit_var_inputs$recovery, 2),
    fmt_pct(1 - credit_var_inputs$recovery, 2),
    fmt_num(credit_var_inputs$rho, 2)
  )
)

credit_var_code_bridge_table <- data.frame(
  `Financial object` = c(
    "$Q(T)$",
    "$\\rho$",
    "$\\alpha$",
    "$EAD$",
    "$LGD$",
    "$V(\\alpha,T)$",
    "Credit VaR"
  ),
  `Code object` = c(
    "`pd`",
    "`rho`",
    "`confidence`",
    "`ead`",
    "`1 - recovery`",
    "`vasicek_default_rate()`",
    "`vasicek_credit_var()`"
  ),
  `Role in the calculation` = c(
    "Average cumulative default probability for each name.",
    "Common-factor correlation that controls co-default risk.",
    "Tail probability level used to select a bad common-factor state.",
    "Portfolio exposure at default.",
    "Loss severity after recovery.",
    "Tail default rate implied by the common-factor state.",
    "Dollar loss percentile after multiplying the tail default rate by EAD and LGD."
  )
)

credit_var_summary_table <- data.frame(
  Confidence = fmt_pct(credit_var_inputs$confidence, 1),
  Tail_default_rate = fmt_pct(credit_var_rate, 2),
  Credit_VaR = fmt_musd(credit_var_loss),
  EL = fmt_musd(rep(credit_expected_loss, length(credit_var_loss))),
  Capital = fmt_musd(credit_var_loss - credit_expected_loss),
  ES = fmt_musd(simulated_es_loss)
)
names(credit_var_summary_table) <- c(
  "Confidence",
  "Tail default rate",
  "Credit VaR",
  "EL",
  "Capital",
  "ES"
)

credit_var_step_table <- data.frame(
  Step = c(
    "Default threshold",
    "Tail quantile",
    "Factor contribution",
    "Numerator",
    "Denominator",
    "Standardized value",
    "Tail default rate",
    "Credit VaR"
  ),
  Calculation = c(
    "$N^{-1}(Q(T))$",
    "$N^{-1}(\\alpha)$",
    "$\\sqrt{\\rho}\\,N^{-1}(\\alpha)$",
    "$N^{-1}(Q(T))+\\sqrt{\\rho}\\,N^{-1}(\\alpha)$",
    "$\\sqrt{1-\\rho}$",
    "$\\frac{N^{-1}(Q(T))+\\sqrt{\\rho}\\,N^{-1}(\\alpha)}{\\sqrt{1-\\rho}}$",
    "$N(\\cdot)$",
    "$EAD\\times LGD\\times V(\\alpha,T)$"
  ),
  Value = c(
    fmt_num(credit_var_q_pd, 4),
    fmt_num(credit_var_q_conf, 4),
    fmt_num(sqrt(credit_var_inputs$rho) * credit_var_q_conf, 4),
    fmt_num(credit_var_numerator, 4),
    fmt_num(credit_var_denominator, 4),
    fmt_num(credit_var_standardized, 4),
    fmt_pct(credit_var_rate[credit_var_focus], 2),
    fmt_musd(credit_var_loss[credit_var_focus])
  )
)

credit_var_999_table <- data.frame(
  Quantity = c(
    "Tail default rate at 99.9%",
    "Credit VaR at 99.9%",
    "Expected loss",
    "Economic capital at 99.9%",
    "Expected shortfall at 99.9%"
  ),
  Value = c(
    fmt_pct(credit_var_rate[3], 2),
    fmt_musd(credit_var_loss[3]),
    fmt_musd(credit_expected_loss),
    fmt_musd(credit_var_loss[3] - credit_expected_loss),
    fmt_musd(simulated_es_loss[3])
  )
)

credit_model_sims <- 200000
credit_model_n_names <- 1000
credit_gaussian_sim_loss <- simulate_granular_gaussian_loss(
  n_sims = credit_model_sims,
  n_names = credit_model_n_names,
  ead = credit_var_inputs$ead,
  pd = credit_var_inputs$pd,
  recovery = credit_var_inputs$recovery,
  rho = credit_var_inputs$rho,
  seed = 130575
)
credit_t10_sim_loss <- simulate_granular_t_loss(
  n_sims = credit_model_sims,
  n_names = credit_model_n_names,
  ead = credit_var_inputs$ead,
  pd = credit_var_inputs$pd,
  recovery = credit_var_inputs$recovery,
  rho = credit_var_inputs$rho,
  df = 10,
  seed = 130575
)
credit_t4_sim_loss <- simulate_granular_t_loss(
  n_sims = credit_model_sims,
  n_names = credit_model_n_names,
  ead = credit_var_inputs$ead,
  pd = credit_var_inputs$pd,
  recovery = credit_var_inputs$recovery,
  rho = credit_var_inputs$rho,
  df = 4,
  seed = 130575
)

credit_model_loss_list <- list(
  "Gaussian sim" = credit_gaussian_sim_loss,
  "t df = 10" = credit_t10_sim_loss,
  "t df = 4" = credit_t4_sim_loss
)
credit_model_summary <- do.call(
  rbind,
  lapply(credit_model_loss_list, loss_risk_summary)
)

credit_model_risk_table <- rbind(
  data.frame(
    Model = "Vasicek formula",
    df = "-",
    EL = fmt_musd(credit_expected_loss),
    VaR_99 = fmt_musd(credit_var_loss[2]),
    VaR_99_9 = fmt_musd(credit_var_loss[3]),
    ES_99_9 = fmt_musd(simulated_es_loss[3]),
    Capital_99_9 = fmt_musd(credit_var_loss[3] - credit_expected_loss)
  ),
  data.frame(
    Model = c("Gaussian sim", "t-copula sim", "t-copula sim"),
    df = c("-", "10", "4"),
    EL = fmt_musd(credit_model_summary[, "expected_loss"]),
    VaR_99 = fmt_musd(credit_model_summary[, "var_99"]),
    VaR_99_9 = fmt_musd(credit_model_summary[, "var_999"]),
    ES_99_9 = fmt_musd(credit_model_summary[, "es_999"]),
    Capital_99_9 = fmt_musd(credit_model_summary[, "economic_capital_999"])
  )
)
names(credit_model_risk_table) <- c(
  "Model",
  "df",
  "EL",
  "VaR 99%",
  "VaR 99.9%",
  "ES 99.9%",
  "Capital 99.9%"
)

credit_tail_grid <- seq(
  0,
  max(sapply(credit_model_loss_list, quantile, probs = 0.9995, type = 1)),
  length.out = 160
)
credit_tail_survival <- sapply(credit_model_loss_list, function(losses) {
  sapply(credit_tail_grid, function(loss_cutoff) mean(losses >= loss_cutoff))
})
credit_tail_survival[credit_tail_survival == 0] <- NA

credit_gaussian_capital <- credit_model_summary["Gaussian sim", "economic_capital_999"]
credit_t4_capital <- credit_model_summary["t df = 4", "economic_capital_999"]
```

The chapter now has all the pieces needed to make Credit VaR concrete. We have individual PDs, a co-default model, exposures, recoveries, a loss distribution, and tail-risk measures. Hull's Credit VaR section, section 24.9 of [@Hull], gives a compact large-portfolio approximation.

The numerical setting changes in this section. Earlier we used Hull's ten-firm, 5-year default example to learn how the copula creates co-default scenarios. Hull's Credit VaR example uses a large homogeneous portfolio, a one-year PD of `r fmt_pct(credit_var_inputs$pd, 2)`, total exposure of `r fmt_musd(credit_var_inputs$ead)`, recovery of `r fmt_pct(credit_var_inputs$recovery, 2)`, and copula correlation of `r fmt_num(credit_var_inputs$rho, 2)`. The financial question also changes. We now ask how much capital is needed to absorb rare losses caused by a bad common credit factor.

The approximation applies to a large homogeneous portfolio. Each loan has the same one-year PD, the same recovery rate, and the same pairwise copula correlation. In a highly granular portfolio, borrower-specific noise largely diversifies away. Conditional on the common credit factor, the realized default fraction is close to the conditional default probability.

In the ten-firm simulations, the outcome was a default count from 0 to 10. In the large-portfolio approximation, the outcome is a default rate. Multiplying that rate by EAD and LGD turns the rate into a dollar loss.

The large-portfolio formula compresses the same one-factor logic into a percentile calculation. We use $\alpha$ for the confidence level, leaving $X_i$ for the credit-state score of firm $i$. The question is practical. When the common credit factor reaches the bad tail state associated with confidence level $\alpha$, what default rate would a highly diversified portfolio experience? That default rate is then multiplied by exposure and loss given default.

Under those assumptions, the tail default rate at confidence level $\alpha$ is

$$
V(\alpha,T)=
N\left(
\frac{N^{-1}(Q(T))+\sqrt{\rho}\,N^{-1}(\alpha)}
{\sqrt{1-\rho}}
\right).
$$

The formula uses the same objects as the copula simulation. $Q(T)$ is the marginal cumulative PD. The parameter $\rho$ controls the strength of the common credit factor. The confidence level $\alpha$ selects how far into the bad tail we look. The output $V(\alpha,T)$ is Hull's tail default-rate function for a large homogeneous portfolio. This is local Credit VaR notation; the asset-value notation $V_T$ in the Merton chapter referred to a different financial object.

The plus sign in the numerator comes from the bad state of the common credit factor. In the one-factor representation, low values of the common factor raise conditional default probabilities. The loss percentile at confidence level $\alpha$ corresponds to the factor quantile $N^{-1}(1-\alpha)$. Substituting that bad factor into the conditional PD formula gives the same expression as above because $-N^{-1}(1-\alpha)=N^{-1}(\alpha)$.

The value $\rho=`r fmt_num(credit_var_inputs$rho, 2)`$ below is part of Hull's numerical example. In practice, changing this input is one of the most important sensitivity checks in Credit VaR. A small increase in common-factor dependence can have little effect on expected loss and a large effect on economic capital, because the capital number is about clustered losses in bad states.

At $\alpha=`r fmt_pct(credit_var_focus_alpha, 1)`$, the bad common-factor quantile is $N^{-1}(1-\alpha)=`r fmt_num(qnorm(1 - credit_var_focus_alpha), 4)`$. Its opposite is `r fmt_num(credit_var_q_conf, 4)`, which is the tail quantile used in the table below.

Hull's Example 24.8 uses the following inputs.

```{r}
kable(
  credit_var_input_table,
  caption = "Inputs for the Vasicek-style Credit VaR approximation.",
  row.names = FALSE,
  escape = FALSE
)
```

Before writing the function, it helps to align the financial notation with the code objects. The first function produces the tail default rate. The second function turns that rate into a dollar loss.

```{r}
kable(
  credit_var_code_bridge_table,
  caption = "Mapping the Credit VaR notation to the R implementation.",
  row.names = FALSE,
  escape = FALSE
)
```

The R function below is the equation written as code.

```{r}
vasicek_default_rate <- function(pd, rho, confidence) {
  pnorm(
    (qnorm(pd) + sqrt(rho) * qnorm(confidence)) /
      sqrt(1 - rho)
  )
}

vasicek_credit_var <- function(ead, pd, recovery, rho, confidence) {
  ead * (1 - recovery) * vasicek_default_rate(pd, rho, confidence)
}
```

In that code, `vasicek_default_rate()` returns $V(\alpha,T)$, the tail default rate. The function `vasicek_credit_var()` then multiplies that rate by exposure and LGD to obtain the dollar loss percentile.

For the 99.9% confidence level, the numerical substitution is visible line by line.

```{r}
kable(
  credit_var_step_table,
  caption = "Numerical substitution for the 99.9% Credit VaR calculation.",
  row.names = FALSE,
  escape = FALSE
)
```

The table below evaluates the formula at several confidence levels. Expected loss uses the average default rate, while Credit VaR uses a tail default rate.

```{r}
kable(
  credit_var_summary_table,
  caption = "Gaussian Vasicek Credit VaR, expected loss, capital, and expected shortfall.",
  row.names = FALSE,
  escape = FALSE
)
```

At the 99.9% confidence level, the tail default rate is `r fmt_pct(credit_var_rate[3], 2)`. With $100 million of exposure and 40% LGD, the Credit VaR is `r fmt_musd(credit_var_loss[3])`. Expected loss is much smaller at `r fmt_musd(credit_expected_loss)`. The difference, `r fmt_musd(credit_var_loss[3] - credit_expected_loss)`, is an estimate of economic capital.

```{r}
kable(
  credit_var_999_table,
  caption = "The 99.9% Gaussian Credit VaR calculation.",
  row.names = FALSE,
  escape = FALSE
)
```

Expected shortfall adds one more tail-risk measure. Credit VaR identifies a high percentile of losses. Expected shortfall averages the simulated losses that exceed that percentile. The number is larger than VaR because it looks inside the tail after the VaR threshold has been crossed.

The simulation below uses the one-factor representation behind the approximation. A low value of the systematic factor represents a bad economy. Those bad common-factor states raise the conditional default rate and therefore raise portfolio loss.

```{r}
#| label: fig-vasicek-factor-loss
#| fig-cap: "Systematic factor and conditional portfolio loss."
vasicek_factor_plot_data <- data.frame(
  Systematic_factor = systematic_factor,
  Loss = vasicek_loss / 1e6
)

ggplot(vasicek_factor_plot_data, aes(x = Systematic_factor, y = Loss)) +
  geom_point(size = 2.15, alpha = 0.12, color = "#566573") +
  geom_hline(yintercept = credit_var_loss[3] / 1e6, color = "#C0392B", linetype = "dashed", linewidth = 1) +
  annotate(
    "label",
    x = quantile(systematic_factor, 0.96),
    y = credit_var_loss[3] / 1e6,
    label = paste0("99.9% Credit VaR ", fmt_musd(credit_var_loss[3])),
    hjust = 1,
    vjust = -0.25,
    size = 3.4,
    fill = "white",
    label.size = 0,
    linewidth = 0,
    color = "#7B241C"
  ) +
  labs(
    x = "Systematic credit factor",
    y = "Conditional portfolio loss ($ millions)"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

The t-copula comparison turns this into a model-risk question. The Gaussian Vasicek formula is the analytical benchmark. The Gaussian simulation checks the same logic with a granular portfolio of `r fmt_int(credit_model_n_names)` equal loans. The t-copula simulations keep the same one-year PD, total exposure, recovery, and correlation parameter, then change the tail shape of the credit-state scores.

```{r}
kable(
  credit_model_risk_table,
  caption = "Credit VaR model risk under Gaussian and t-copula assumptions.",
  row.names = FALSE,
  escape = FALSE
)
```

The expected loss rows stay near `r fmt_musd(credit_expected_loss)` because every model uses the same average one-year PD, EAD, and LGD. The capital rows respond to joint tail behavior. In this run, the Gaussian simulation produces 99.9% capital of `r fmt_musd(credit_gaussian_capital)`. The t-copula with 4 degrees of freedom produces `r fmt_musd(credit_t4_capital)`.

```{r}
#| label: fig-credit-var-copula-tail
#| fig-cap: "Loss-tail survival probabilities under Gaussian and t-copulas."
credit_tail_plot_data <- data.frame(
  Loss = rep(credit_tail_grid / 1e6, times = ncol(credit_tail_survival)),
  Survival = as.vector(credit_tail_survival),
  Model = factor(
    rep(colnames(credit_tail_survival), each = length(credit_tail_grid)),
    levels = colnames(credit_tail_survival)
  )
)

ggplot(
  subset(credit_tail_plot_data, !is.na(Survival) & Survival > 0),
  aes(x = Loss, y = Survival, color = Model)
) +
  geom_line(linewidth = 1.1, alpha = 0.9) +
  scale_color_manual(values = c("#2C7FB8", "#F39C12", "#C0392B")) +
  scale_y_log10(labels = function(x) fmt_pct(x, 2)) +
  labs(
    x = "Portfolio loss ($ millions)",
    y = "P(loss >= x)",
    color = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

The chapter's main gain is a portfolio view of credit risk. PDs describe average default risk name by name. The copula supplies the co-default mechanism. Exposures and recoveries translate default scenarios into dollar losses. Credit VaR, expected shortfall, and economic capital summarize the tail of the resulting loss distribution.

The final comparison turns the copula choice into a capital question. A Gaussian copula and a t-copula can use the same PD, exposure, recovery, and correlation inputs while producing different tail losses. The model choice is therefore a capital assumption as well as a dependence assumption. That is the practical bridge from Hull's copula example to credit portfolio management. The useful question is how large the portfolio loss can become when defaults arrive together.
