## From correlated defaults to portfolio losses

```{r gaussian-portfolio-losses-setup, include=FALSE}
source("R/gaussian-common.R")

loss_sims <- 100000
loss_n_firms <- 10
loss_total_ead <- 100e6
loss_ead <- rep(loss_total_ead / loss_n_firms, loss_n_firms)
loss_recovery <- rep(0.40, loss_n_firms)
loss_lgd <- 1 - loss_recovery
loss_threshold <- default_threshold[5]
loss_rhos <- c(0, 0.2, 0.5)
loss_expected_theoretical <- sum(loss_ead * loss_lgd * pd_cum[5])

simulate_loss_scenario <- function(rho, ead = loss_ead, recovery = loss_recovery) {
  latent <- simulate_latent_credit(
    n_sims = loss_sims,
    n_firms = length(ead),
    rho = rho,
    seed = 130575
  )
  defaults <- make_default_matrix(latent, loss_threshold)
  losses <- portfolio_losses(defaults, ead, recovery)
  list(defaults = defaults, losses = losses, summary = loss_risk_summary(losses))
}

loss_scenarios <- setNames(
  lapply(loss_rhos, simulate_loss_scenario),
  paste0("rho_", loss_rhos)
)

loss_summary_numeric <- do.call(rbind, lapply(loss_scenarios, function(x) x$summary))
loss_focus_rho_index <- which(loss_rhos == 0.2)
loss_focus_summary <- loss_summary_numeric[loss_focus_rho_index, ]
loss_summary_table <- data.frame(
  Copula_correlation = fmt_num(loss_rhos, 1),
  Expected_loss = fmt_musd(loss_summary_numeric[, "expected_loss"]),
  Median_loss = fmt_musd(loss_summary_numeric[, "median_loss"]),
  VaR_95 = fmt_musd(loss_summary_numeric[, "var_95"]),
  VaR_99 = fmt_musd(loss_summary_numeric[, "var_99"]),
  VaR_99_9 = fmt_musd(loss_summary_numeric[, "var_999"]),
  ES_99 = fmt_musd(loss_summary_numeric[, "es_99"]),
  Economic_capital_99_9 = fmt_musd(loss_summary_numeric[, "economic_capital_999"])
)
names(loss_summary_table) <- c(
  "rho",
  "EL",
  "Median",
  "VaR 95%",
  "VaR 99%",
  "VaR 99.9%",
  "ES 99%",
  "Capital 99.9%"
)

loss_distribution_matrix <- sapply(loss_scenarios, function(x) {
  as.numeric(tabulate(x$losses / (loss_ead[1] * loss_lgd[1]) + 1, nbins = loss_n_firms + 1)) / loss_sims
})
colnames(loss_distribution_matrix) <- paste0("rho = ", fmt_num(loss_rhos, 1))
```

Defaults become financially useful when we attach exposures and recoveries. In one simulated credit scenario $s$, portfolio loss is

$$
L_s=\sum_{i=1}^n EAD_i \times LGD_i \times Y_{i,s},
$$

where $Y_{i,s}$ equals 1 when firm $i$ defaults in scenario $s$ and 0 otherwise. The code applies the same calculation.

```{r}
ead <- rep(10e6, 10)
recovery <- rep(0.40, 10)
lgd <- 1 - recovery

losses <- defaults_5y %*% (ead * lgd)
```

The example below uses a homogeneous portfolio with 10 firms, total exposure of $100 million, $10 million exposure per firm, 40% recovery, and a 5-year cumulative PD of `r fmt_pct(pd_cum[5], 2)` for every firm. The recovery assumption implies an LGD of 60%.

Expected loss can be understood before running the portfolio scenarios.

$$
EL = \sum_{i=1}^{10} EAD_i \times LGD_i \times PD_i
= 10 \times 10{,}000{,}000 \times 0.60 \times 0.15
$$

The last expression is `r fmt_num(loss_expected_theoretical / 1e6, 2)` million dollars.

The simulation should produce an expected loss close to that value in every correlation scenario, because the threshold rule keeps each firm's marginal PD fixed. The tail of the loss distribution is the object that changes when shared shocks make default clusters more or less likely.

```{r}
kable(
  loss_summary_table,
  caption = "Portfolio loss summary for a homogeneous ten-firm portfolio.",
  escape = FALSE
)
```

The table separates three ideas. Expected loss is the average credit cost implied by PD, EAD, and LGD. VaR is a high percentile of the simulated loss distribution. Economic capital is the difference between a tail loss measure and expected loss.

For the Hull correlation case $\rho=`r fmt_num(loss_rhos[loss_focus_rho_index], 1)`$, the expected loss is `r fmt_musd(loss_focus_summary["expected_loss"])`, while the 99.9% VaR is `r fmt_musd(loss_focus_summary["var_999"])`. The economic-capital number, `r fmt_musd(loss_focus_summary["economic_capital_999"])`, is the extra loss buffer above expected loss at that confidence level. This is the practical reason the tail can dominate capital needs. Two portfolios can have similar expected loss and very different capital buffers.

The next figure shows the simulated loss distribution. Because the portfolio has equal exposures and equal recoveries, losses occur in steps of $6 million per default.

```{r}
#| label: fig-portfolio-loss-distribution
#| fig-cap: "Portfolio loss distribution under different copula correlations."
loss_steps <- 0:loss_n_firms
loss_distribution_colors <- c("#2C7FB8", "#F39C12", "#C0392B")
loss_distribution_plot_data <- data.frame(
  Loss = rep(loss_steps * loss_ead[1] * loss_lgd[1] / 1e6, times = ncol(loss_distribution_matrix)),
  Probability = as.vector(loss_distribution_matrix),
  Correlation = factor(
    rep(colnames(loss_distribution_matrix), each = length(loss_steps)),
    levels = colnames(loss_distribution_matrix)
  )
)

ggplot(
  loss_distribution_plot_data,
  aes(x = Loss, y = Probability, color = Correlation, group = Correlation)
) +
  geom_line(linewidth = 1.1, alpha = 0.85) +
  geom_point(size = 3.2, alpha = 0.82) +
  scale_color_manual(values = loss_distribution_colors) +
  scale_x_continuous(breaks = loss_steps * loss_ead[1] * loss_lgd[1] / 1e6) +
  scale_y_continuous(
    labels = function(x) fmt_pct(x, 1),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  labs(
    x = "Portfolio loss ($ millions)",
    y = "Simulated probability",
    color = "Copula correlation"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    plot.margin = margin(5.5, 12, 5.5, 5.5)
  )
```

This is where the copula becomes a credit-risk tool. The marginal PD tells us the average loss. The common-shock correlation tells us how much probability sits in the severe-loss states.
