```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r last-modified, echo=FALSE, results='asis'}
render_time <- Sys.time()
month_names <- c(
  "January", "February", "March", "April", "May", "June",
  "July", "August", "September", "October", "November", "December"
)
am_pm <- if (as.integer(format(render_time, "%H")) < 12) "am" else "pm"
formatted_modified <- paste0(
  month_names[as.integer(format(render_time, "%m"))],
  " ",
  as.integer(format(render_time, "%d")),
  format(render_time, ", %Y, %H:%M:%S "),
  am_pm,
  "."
)

cat(
  '<div class="home-last-modified">',
  '<span class="home-last-modified-label">Modified</span>',
  '<span class="date-modified">', formatted_modified, '</span>',
  '</div>\n',
  sep = ""
)
```

::: {.home-panel}
This tutorial introduces credit risk models with R, with emphasis on model logic, data analysis, graphical interpretation, and reproducible code.

The material relies on Hull's credit risk chapters, Lore Dirick's credit risk course materials, and the literate programming spirit of @knuth1984literate: the goal is to explain to human beings what we want a computer to do.

This is a work in progress and remains under revision.
:::

::: {.home-section-grid}
::: {.home-section}
**Credit scoring and data**

Start with loan-level default prediction, logistic regression, model evaluation, cutoffs, acceptance rates, bad rates, and ROC curves.

[Go to logistic credit scoring](01-logistic.qmd)
:::

::: {.home-section}
**Structural and portfolio models**

Continue with the Merton model, Gaussian copula dependence, and a compact Credit VaR example.

[Go to the Merton model](02-merton.qmd)
:::

::: {.home-section}
**References and support**

Use the reference list to trace the academic sources and technical materials behind the examples.

[Go to references](references.qmd)
:::
:::
