```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r book-edition, echo=FALSE, results='asis'}
source("R/book-edition.R")
write_book_edition_script()
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r publication-dates, echo=FALSE, results='asis'}
write_book_dates()
```

## Preface {.unnumbered}

::: {.home-panel}
This book is written for readers who want to close the gap between credit-risk models, the logic behind those models, and the code required to implement them. It is designed for business, finance, economics, and risk-management readers who want to understand quantitative credit risk through explicit calculations, reproducible R code, and careful financial interpretation.

A central feature of the book is its deliberately explicit style. It keeps intuition, mathematical structure, data work, graphics, tables, and R code in the same conversation. The computational details remain visible so the reader can see what the computer is being asked to do, why that operation is useful for the credit-risk problem, and how the numerical output should be interpreted.

The organizing principle follows the literate programming spirit of @knuth1984literate. Code appears as part of the explanation and stays close to the financial question it answers. A table is introduced because it answers a credit question. A figure is included because it makes a model mechanism easier to see. A formula is connected to the R object that implements it. This is the reason the book often walks slowly through steps that more advanced texts may compress.

This is an online book under active revision. The publication metadata above records the first online publication, the current publication number, and a content-based book edition. Those markers are included so readers can tell which version they are using and so revisions remain traceable over time. The current emphasis is clarity, reproducibility, practical interpretation, and a steady connection between credit-risk theory and applied financial decisions.
:::

## What's new in this edition {.unnumbered}

- **Credit-scoring workflow:** Chapter 1 now connects logistic PDs with cutoff rules, bad rates, calibration, ROC/AUC, Brier score, and a simple lending payoff exercise.
- **Tree-based benchmark:** Chapter 2 compares logistic regression, a single tree, and XGBoost using the same test set and the same credit-policy metrics.
- **Merton model rebuild:** Chapter 3 now gives a fuller risk-neutral valuation bridge, replicates Hull's equations, checks the numerical solution, and links equity to a call payoff on firm assets.
- **Credit-spread bridge:** Chapter 4 now connects default probabilities with corporate bond spreads, recovery assumptions, and CDS valuation.
- **Portfolio credit risk:** Chapter 5 now builds the Gaussian copula model around Hull Example 24.7, correlated defaults, portfolio losses, concentration risk, and capital.
- **Tail-risk extension:** The final Chapter 5 sections compare Gaussian and t-copula assumptions and connect the simulation logic to Credit VaR, expected shortfall, and economic capital.
- **Climate credit risk:** Chapter 6 now uses public NGFS/CLIMACRED scenario data to connect climate-adjusted PDs with expected loss, sector contributions, spread adjustments, CDS-style compensation, and simulated portfolio losses.
