```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r last-modified, echo=FALSE, results='asis'}
render_time <- Sys.time()
month_names <- c(
  "January", "February", "March", "April", "May", "June",
  "July", "August", "September", "October", "November", "December"
)
am_pm <- if (as.integer(format(render_time, "%H")) < 12) "am" else "pm"
formatted_modified <- paste0(
  month_names[as.integer(format(render_time, "%m"))],
  " ",
  as.integer(format(render_time, "%d")),
  format(render_time, ", %Y, %H:%M:%S "),
  am_pm,
  "."
)

cat(
  '<div class="home-last-modified">',
  '<span class="home-last-modified-label">Modified</span>',
  '<span class="date-modified">', formatted_modified, '</span>',
  '</div>\n',
  sep = ""
)
```

::: {.home-panel}
This tutorial introduces credit risk models with R, with emphasis on model logic, data analysis, graphical interpretation, and reproducible code.

The material relies on Hull's credit risk chapters, Lore Dirick's credit risk course materials, and the literate programming spirit of @knuth1984literate: the goal is to explain to human beings what we want a computer to do.

This is a work in progress and remains under revision.
:::
